### 2020-06-04 Outline.Rmd
Adding template for outline and list of packages

---
title: "Outline"
output: html_document
date: '`r format(Sys.time(), "%d %B, %Y")`'
---

<br/>

Journal: Current Landscape Ecology Reports
Type: Methodological Developments in Landscape Ecology

<br/>

< **TITLE** >

Maximillian H.K. Hesselbarth<sup>1</sup>, Jakub Nowosad<sup>2</sup>, ...

<br/>

<sup>1</sup>Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany

<sup>2</sup>Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland

<br/>

*Corresponding author: E-Mail address: maximilian.hesselbarth@uni-goettingen.de

<br/>

Keywords: keyword 1, keyword 2, keyword 3, keyword 4, keyword 5, …

### 1. Introduction
- First topic
    - First paragraph headline
    - Second paragraph headline
    - ...
- Second topic
    - First paragraph headline
    - Second paragraph headline
    - ...
- ...
- Last topic (often objectives and hypotheses)
    - ...

### 2. Methods
- First topic
    - First paragraph headline
    - Second paragraph headline
    - ...
- Second topic
    - First paragraph headline
    - Second paragraph headline
    - ...
- ...
- Last topic
    - ...

### 3. Results (same structure and order as Methods)
- First topic
    - First paragraph headline
    - Second paragraph headline
    - ...
- Second topic
    - First paragraph headline
    - Second paragraph headline
    - ...
- ...
- Last topic (often objectives and hypotheses)
    - ...

### 4. Discussion
-	Possible points to discuss
    - First point
    - Second point
- ...
- Conclusions
    - Conclusion 1 
    - Conclusion 2
    - ...

### 5. Acknowledgments
- ...

### Open questions:
- ...
- ...

### 2020-06-04 Manuscript/Manuscript.Rmd
Add Manuscript template

---
title: Open source landscape ecology tools
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Short form of title (if too long for head)
# authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: Maximilian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany 
  email: maximilian.hesselbarth@uni-goettingen.de

- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com

- name: Author 3
  address: Adress author 3
  email: E-Mail author 3

- name: ...
  address: ...
  email: ...

keywords:
- keyword 1
- keyword 2
- keyword 3
- keyword 4
- ...

#PACS: 
#- PAC1
#- superPAC

# MSC:
# - MSC code 1
# - MSC code 2

abstract: |
  The text of your abstract.

bibliography: bibliography.bib
bibstyle: spphys
# bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
---

# Introduction {#intro}

Your text comes here. Separate text sections with \cite{Mislevy06Cog}.

# Section title {#sec:1}

Text with citations by \cite{Galyardt14mmm}.

## Subsection title {#sec:2}

as required. Don't forget to give each section
and subsection a unique label (see Sect. \ref{sec:1}).

# Acknowledgments {#acknow}
Thanks to everyone.

#### Paragraph headings

Use paragraph headings as needed.

\begin{align}
a^2+b^2=c^2
\end{align}

### 2020-06-04 Manuscript/Manuscript.tex
Add Manuscript template

% !TeX program = pdfLaTeX
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

%% load any required packages here

\begin{document}

\title{Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page
should be placed here. General acknowledgments should be placed at the
end of the article.} }

\author{  Maximilian H.K. Hesselbarth \and  Jakub Nowosad \and  Author 3 \and  \ldots{} \and  }

\institute{
        Maximilian H.K. Hesselbarth \at
     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg
 4, 37077 Goettingen, Germany \\
     \email{\href{mailto:maximilian.hesselbarth@uni-goettingen.de}{\nolinkurl{maximilian.hesselbarth@uni-goettingen.de}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University,
 Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Author 3 \at
     Adress author 3 \\
     \email{E-Mail author 3}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        \ldots{} \at
     \ldots{} \\
     \email{\ldots{}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
The text of your abstract.
\\
\keywords{
        keyword 1 \and
        keyword 2 \and
        keyword 3 \and
        keyword 4 \and
        \ldots{} \and
    }

\end{abstract}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

Your text comes here. Separate text sections with \cite{Mislevy06Cog}.

\hypertarget{sec:1}{%
\section{Section title}\label{sec:1}}

Text with citations by \cite{Galyardt14mmm}.

\hypertarget{sec:2}{%
\subsection{Subsection title}\label{sec:2}}

as required. Don't forget to give each section and subsection a unique
label (see Sect. \ref{sec:1}).

\hypertarget{acknow}{%
\section{Acknowledgments}\label{acknow}}

Thanks to everyone.

\hypertarget{paragraph-headings}{%
\paragraph{Paragraph headings}\label{paragraph-headings}}

Use paragraph headings as needed.

\begin{align}
a^2+b^2=c^2
\end{align}

\bibliographystyle{spphys}
\bibliography{bibliography.bib}

\end{document}

### 2020-06-05 Manuscript/Manuscript.Rmd
Adding Manuscript .bib-file

@@
 # authorrunning: Short form of author list if too long for running head
+
 thanks: | 
@@
   email: maximilian.hesselbarth@uni-goettingen.de
-
 - name: Jakub Nowosad
@@
   email: nowosad.jakub@gmail.com
-
 - name: Author 3
@@
   email: E-Mail author 3
-
 - name: ...
@@
 
-#PACS: 
-#- PAC1
-#- superPAC
-
-# MSC:
-# - MSC code 1
-# - MSC code 2
-
 abstract: |
-  The text of your abstract.
+  (max.
+x words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
 
@@
 bibstyle: spphys
-# bibstyle options spbasic(default), spphys, spmpsci
+
 output: rticles::springer_article
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
-Your text comes here. Separate text sections with \cite{Mislevy06Cog}.
+Your text comes here.
 
@@
 
-Text with citations by \cite{Galyardt14mmm}.
+Text with citations by \cite{turner2005}.
 

### 2020-06-05 Manuscript/Manuscript.tex
Adding Manuscript .bib-file

@@
 
-\title{Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page
-should be placed here. General acknowledgments should be placed at the
-end of the article.} }
+\title{Open source landscape ecology tools \thanks{Grants or other notes
+about the article that should go on the front page should be placed
+here. General acknowledgments should be placed at the end of the
+article.} }
 
-\author{  Maximilian H.K. Hesselbarth \and  Jakub Nowosad \and  Author 3 \and  \ldots{} \and  }
+\author{  Maximilian H.K. Hesselbarth \and  Jakub Nowosad \and  Author
+3 \and  \ldots{} \and  }
 
@@
         Maximilian H.K. Hesselbarth \at
-     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg
- 4, 37077 Goettingen, Germany \\
+     Department of Ecosystem Modelling, University of Goettingen,
+Buesgenweg 4, 37077 Goettingen, Germany \\
      \email{\href{mailto:maximilian.hesselbarth@uni-goettingen.de}{\nolinkurl{maximilian.hesselbarth@uni-goettingen.de}}}  %  \\
@@
         Jakub Nowosad \at
-     Institute of Geoecology and Geoinformation, Adam Mickiewicz University,
- Krygowskiego 10, 61-680 Poznan, Poland \\
+     Institute of Geoecology and Geoinformation, Adam Mickiewicz
+University, Krygowskiego 10, 61-680 Poznan, Poland \\
      \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
@@
 \begin{abstract}
-The text of your abstract.
+(max. x words) Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract
 \\

### 2020-06-05 Outline.Rmd
Add working title to outline
Starting to work on possible outline

@@
 output: html_document
+theme: cosmo
 date: '`r format(Sys.time(), "%d %B, %Y")`'
+editor_options: 
+  chunk_output_type: console
 ---
@@
 Journal: Current Landscape Ecology Reports
+
 Type: Methodological Developments in Landscape Ecology
@@
 
-< **TITLE** >
+Title: **Open source landscape ecology tools**
 
@@
 
-<br/>
-
 *Corresponding author: E-Mail address: maximilian.hesselbarth@uni-goettingen.de
@@
 
-Keywords: keyword 1, keyword 2, keyword 3, keyword 4, keyword 5, …
+Keywords: keyword 1, keyword 2, keyword 3, keyword 4, ...
 
 ### 1. Introduction
-- First topic
-    - First paragraph headline
-    - Second paragraph headline
+- Intro to landscape ecology
+    - How do landscape characteristics influence ecological processes and vice verse
+    - Growing importance because often scales of policy making are adressed
+    - Need to preprocess, modify, model and analyze spatial data
+    - Short overview over common tasks/analysis in landscape ecology
     - ...
-- Second topic
-    - First paragraph headline
-    - Second paragraph headline
+- Intro to open-source
+    - What is open-source and why is it advantageous to closed source
+    - R environment
+    - ...
+- Landscape ecology in R
+    - One of the most used tools/powerful tool in ecology and spatial data analysis
+    - History of landscape ecology/spatial data analysis in R
+    - Growing body of specific textbooks and literature; e.g.:
+        - "Geocomputation with R"
+        - "Spatial Data Science"
+        - "Spatial Ecology and Conservation Modeling. Applications with R"
+        - "Learning Landscape Ecology. A Practical Guide to Concepts and Techniques"
+        - ...
+    - Growing body of specific R package for spatial data analysis
+    - ...
+- Objectives of review
+    - Existing/important R packages for landscape ecology
+    - Missing packages R packages for landscape ecology
     - ...
 - ...
-- Last topic (often objectives and hypotheses)
+
+### 2. Existing packages
+- Basic spatial data
+    - Basic data types (i.e. vector vs. raster data)
+        - raster, stars, terra(?)
+        - sp, sf
+        - rgeos, rgdal
         - ...
-
-### 2. Methods
-- First topic
-    - First paragraph headline
-    - Second paragraph headline
+    - Creating maps
+        - maptools, tmap, ggplot2
         - ...
-- Second topic
-    - First paragraph headline
-    - Second paragraph headline
+- Web-mapping frameworks
+    - mapview, leaflet
+    - ...
+- Spatial regression
+    - nlme, spdep, kriging
+    - ...
+- Ecological analysis
+    - Landscape metrics
+        - landscapmetrics
+        - ...
+    - Habitat selection and SDM
+        - ade4, sdm
+        - ...
+    - Various
+        - vegan
+        - NLMR
     - ...
 - ...
-- Last topic
-    - ...
-
-### 3. Results (same structure and order as Methods)
+### 3. Missing packages
 - First topic
@@
     - ...
+- ...
 - Conclusions
@@
     - Conclusion 2
+    - ...
 - ...
@@
 ### Open questions:
+- Remote sensing/GIS?
+- Landscape genetics?
+- Landscape Connectivity?
+- Some "results"/meta-analysis of increase of R in landscape ecology?
+- Missing packages
+    - surface metrics
+    - vector-based metrics
 - ...
-- ...

### 2020-06-07 paper/paper.Rmd
adds rtickles docs

---
title: Open source landscape ecology tools
subtitle: Do you have a subtitle? If so, write it here
titlerunning: Open source landscape ecology tools
authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: Maximillian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
  email: maximilian.hesselbarth@uni-goettingen.de

- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com

keywords:
- key
- dictionary
- word

abstract: |
  The text of your abstract.  150 -- 250 words.

bibliography: bibliography.bib
bibstyle: spphys
# bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
---

# Introduction {#intro}

Your text comes here. Separate text sections with \cite{Mislevy06Cog}.

# Section title {#sec:1}

Text with citations by \cite{Galyardt14mmm}.

## Subsection title {#sec:2}

as required. Don't forget to give each section
and subsection a unique label (see Sect. \ref{sec:1}).

#### Paragraph headings

Use paragraph headings as needed.

\begin{align}
a^2+b^2=c^2
\end{align}

### 2020-06-08 Outline.Rmd
Use paper-folder for manuscript

@@
 - Landscape Connectivity?
+  - ResistanceGA
+  - LQGraph
 - Some "results"/meta-analysis of increase of R in landscape ecology?
+- Open-source licenses
 - Missing packages

### 2020-06-08 paper/paper.Rmd
Use paper-folder for manuscript

@@
 title: Open source landscape ecology tools
-subtitle: Do you have a subtitle? If so, write it here
-titlerunning: Open source landscape ecology tools
-authorrunning: Short form of author list if too long for running head
+# subtitle: Do you have a subtitle? If so, write it here
+# titlerunning: Open source landscape ecology tools
+# authorrunning: Short form of author list if too long for running head
+
 thanks: | 
@@
   email: nowosad.jakub@gmail.com
+- name: Author 3
+  address: Adress author 3
+  email: E-Mail author 3
+- name: ...
+  address: ...
+  email: ...
 
 keywords:
-- key
-- dictionary
-- word
+- keyword 1
+- keyword 2
+- keyword 3
+- keyword 4
+- ...
 
 abstract: |
-  The text of your abstract.  150 -- 250 words.
+  max.
+(200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
 
 bibliography: bibliography.bib
-bibstyle: spphys
-# bibstyle options spbasic(default), spphys, spmpsci
+bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
 output: rticles::springer_article
@@
 
-Your text comes here. Separate text sections with \cite{Mislevy06Cog}.
+Landscape ecology does this and that \cite{turner2005}.
 
@@
 
-Text with citations by \cite{Galyardt14mmm}.
+Open-source is great \cite{sandve2013}.
 

### 2020-06-08 paper/paper.tex
adds rtickles docs
Use paper-folder for manuscript

--- 2020-06-04 Manuscript/Manuscript.tex
@@
 
-\title{Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page
-should be placed here. General acknowledgments should be placed at the
-end of the article.} }
+\title{Open source landscape ecology tools \thanks{Grants or other notes
+about the article that should go on the front page should be placed
+here. General acknowledgments should be placed at the end of the
+article.} }
 
-\author{  Maximilian H.K. Hesselbarth \and  Jakub Nowosad \and  Author 3 \and  \ldots{} \and  }
+\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Author
+3 \and  \ldots{} \and  }
 
 \institute{
-        Maximilian H.K. Hesselbarth \at
-     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg
- 4, 37077 Goettingen, Germany \\
+        Maximillian H.K. Hesselbarth \at
+     Department of Ecosystem Modelling, University of Goettingen,
+Buesgenweg 4, 37077 Goettingen, Germany \\
      \email{\href{mailto:maximilian.hesselbarth@uni-goettingen.de}{\nolinkurl{maximilian.hesselbarth@uni-goettingen.de}}}  %  \\
@@
 \begin{abstract}
-The text of your abstract.
+max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Abstract Abstract Abstract Abstract Abstract Abstract
 \\
@@
 label (see Sect. \ref{sec:1}).
-
-\hypertarget{acknow}{%
-\section{Acknowledgments}\label{acknow}}
-
-Thanks to everyone.
 

### 2020-06-16 Outline.Rmd
merges map creating
adds new idea
adds new set of ideas

@@
 ### 2. Existing packages
+
+- Base R
 - Basic spatial data
@@
         - raster, stars, terra(?)
-        - sp, sf
-        - rgeos, rgdal
+        - sp (+rgeos, rgdal), sf
         - ...
 - Creating maps
-        - maptools, tmap, ggplot2
-        - ...
-- Web-mapping frameworks
+    - tmap, ggplot2
     - mapview, leaflet
@@
 - Spatial regression
-    - nlme, spdep, kriging
+    - nlme, spdep, gstat
     - ...
+- Spatial statistics?? (point-patterns analysis)
+    - spatstat
 - Ecological analysis
@@
         - landscapmetrics
+        - belg
         - ...
     - Habitat selection and SDM
+        - lm(), glm(), gam, RandomForest (ranger)
         - ade4, sdm
+        - biomod2??
         - ...
@@
         - vegan
-        - NLMR
+        - NLMR, landscapetools
+        - ecospat?? (https://onlinelibrary.wiley.com/doi/full/10.1111/ecog.02671)
+        - PresenceAbsence?? (https://www.jstatsoft.org/article/view/v023i11)
+        - ENMeval?? (https://cran.r-project.org/web/packages/ENMeval/index.html)
+        - prioritizr?? (https://prioritizr.net/)
+        - data sources/API connections?
     - ...

### 2020-06-16 Outline.Rmd
adds new set of ideas

@@
         - data sources/API connections?
+        - ecodist?? (https://cran.r-project.org/web/packages/ecodist/index.html)
+        - LEA?? (http://membres-timc.imag.fr/Olivier.Francois/Frichot_etal_MEE_2015.pdf)
+        - remote sensing??
     - ...
@@
     - vector-based metrics
+    - Morphological Spatial Pattern Analysis (https://forest.jrc.ec.europa.eu/en/activities/lpa/mspa/)
 - ...

### 2020-06-17 Outline.Rmd
adds another sets of ideas

@@
 - Spatial regression
-    - nlme, spdep, gstat
+    - nlme
+    - spdep, gstat/geoR (spatial autocorrelation), + pgirmess??, (http://www.r-inla.org/download; INLA?? check license)
     - ...
 - Spatial statistics?? (point-patterns analysis)
-    - spatstat
+    - spatstat (e.g., https://link.springer.com/chapter/10.1007/978-3-030-01989-1_4)
 - Ecological analysis
@@
     - Habitat selection and SDM
-        - lm(), glm(), gam, RandomForest (ranger)
-        - ade4, sdm
+        - https://link.springer.com/chapter/10.1007/978-3-030-01989-1_7
+        - lm(), glm(), gam, rpart, RandomForest (ranger), lme4
+        - ade4, adehabitat, sdm
+        - some MAXENT package?
         - biomod2??
         - ...
+    - Connectivity? 
+        - https://link.springer.com/chapter/10.1007/978-3-030-01989-1_9
+        - gdistance??
+    - Spatial population dynamics
+        - https://link.springer.com/chapter/10.1007/978-3-030-01989-1_10
     - Various
         - vegan
-        - NLMR, landscapetools
+        - dismo
+        - NLMR (and RandomFields), landscapetools
         - ecospat?? (https://onlinelibrary.wiley.com/doi/full/10.1111/ecog.02671)
@@
         - remote sensing??
+        - agent-based modelling
     - ...
@@
 - Some "results"/meta-analysis of increase of R in landscape ecology?
-- Open-source licenses
+- Open-source licenses (!!)
 - Missing packages

### 2020-06-17 Outline.Rmd
Comments Outline

@@
     - ...
-- Spatial statistics?? (point-patterns analysis)
+- Spatial statistics?? (point-patterns analysis) // MH:I think that's way too much
     - spatstat (e.g., https://link.springer.com/chapter/10.1007/978-3-030-01989-1_4)
@@
         - remote sensing??
-        - agent-based modelling
+        - agent-based modelling // MH:I think that's way too much
     - ...

### 2020-07-08 paper/paper.Rmd
Starting to write on general introduction

@@
 
-# Introduction {#intro}
+# Introduction {#sec:intro}
 
-Landscape ecology does this and that \cite{turner2005}.
+## A short introduction to landscape ecology {#sec:landscape_ecology}
 
-# Section title {#sec:1}
+Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
+In this context, landscape ecology considers, besides others, spatial and temporal dynamics of heterogeneous landscapes, interactions, fluxes and exchange within these landscapes, how the landscapes influence ecological processes (and vice versa) and lastly how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
 
-Open-source is great \cite{sandve2013}.
+While human activities have altered their environment for millennials \cite{Ellis2011}, in the past centuries the effects of humans on the global environment have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}.
+Today, almost all ecosystems are directly or indirectly influenced by human activities \cite{Vitousek1997}.
+Thus, understanding the complex interactions between landscapes and ecological processes also becomes more important \cite{With2019}.
 
-## Subsection title {#sec:2}
+Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}.
+Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes spatial patterns \cite{Risser1984}.
+Consequently, the field of landscape ecology relies on tools to preprocess, modify, model, analyze and visualize spatial data.
 
-as required. Don't forget to give each section
-and subsection a unique label (see Sect. \ref{sec:1}).
+## Open-source software and R {#sec:open_source}
+
+Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
+Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
+
+One example of a successful open-source project is the R programming language and its *Comprehensive R Archive Network* (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
+Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
+
+## Landscape ecology in R {#sec:R}
 

### 2020-07-08 paper/paper.tex
Starting to write on general introduction

@@
 
-\hypertarget{intro}{%
-\section{Introduction}\label{intro}}
+\hypertarget{sec:intro}{%
+\section{Introduction}\label{sec:intro}}
 
-Landscape ecology does this and that \cite{turner2005}.
+\hypertarget{sec:landscape_ecology}{%
+\subsection{A short introduction to landscape
+ecology}\label{sec:landscape_ecology}}
 
-\hypertarget{sec:1}{%
-\section{Section title}\label{sec:1}}
+Landscape ecology focuses on how ecological processes are influenced and
+modified by the heterogeneous landscapes they occur in and
+simultaneously how the ecological processes themselves influence the
+landscapes \cite{Turner1989,Turner2005,With2019}. In this context,
+landscape ecology considers, besides others, spatial and temporal
+dynamics of heterogeneous landscapes, interactions, fluxes and exchange
+within these landscapes, how the landscapes influence ecological
+processes (and vice versa) and lastly how to manage these heterogeneous
+landscapes \cite{Risser1984,Turner1989}.
 
-Open-source is great \cite{sandve2013}.
+While human activities have altered their environment for millennials
+\cite{Ellis2011}, in the past centuries the effects of humans on the
+global environment have increased to an unknown high, known as the the
+Anthropocene \cite{Crutzen2002}. Today, almost all ecosystems are
+directly or indirectly influenced by human activities
+\cite{Vitousek1997}. Thus, understanding the complex interactions
+between landscapes and ecological processes also becomes more important
+\cite{With2019}.
 
-\hypertarget{sec:2}{%
-\subsection{Subsection title}\label{sec:2}}
+Because landscapes are defined as mosaics of different land covers,
+ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995},
+spatial context is important and ecological processes will vary
+spatially \cite{With2019}. Thus, in contrast to many other
+sub-disciplines of ecology, landscape ecology emphasizes spatial
+patterns \cite{Risser1984}. Consequently, the field of landscape ecology
+relies on tools to preprocess, modify, model, analyze and visualize
+spatial data.
 
-as required. Don't forget to give each section and subsection a unique
-label (see Sect. \ref{sec:1}).
+\hypertarget{sec:open_source}{%
+\subsection{Open-source software and R}\label{sec:open_source}}
+
+Open-source software includes all software which is released under a
+license that allows to freely use, modify and distribute the software
+\cite{St.Laurent2008}. Open-source software development has many
+advantages, such as fast innovation, transparency and reliability as
+well as longevity due to many diverse contributors
+\cite{vonKrogh2006,St.Laurent2008}.
+
+One example of a successful open-source project is the R programming
+language and its \emph{Comprehensive R Archive Network} (CRAN) for
+extensions (also called packages) \cite{RCoreTeam2019}. Firstly
+introduced in 1995 \cite{Smith2016}, today the programming language is
+among the most popular programming languages, especially in ecology
+\cite{Lai2019}.
+
+\hypertarget{sec:R}{%
+\subsection{Landscape ecology in R}\label{sec:R}}
 

### 2020-07-09 Outline.Rmd
Update Outline.Rmd
Adding landscape genetics to outline

@@
         - sp (+rgeos, rgdal), sf
+        - Access to spatial data (e.g. rnaturalearth) (?)
         - ...
@@
         - https://link.springer.com/chapter/10.1007/978-3-030-01989-1_10
+    - Landscape genetics
     - Various

### 2020-07-09 paper/paper.Rmd
Starting to work on manuscript

--- 2020-06-08 paper/paper.Rmd
@@
 
-# Introduction {#intro}
+# Introduction {#sec:intro}
 
-Landscape ecology does this and that \cite{turner2005}.
+## A short introduction to landscape ecology {#sec:landscape_ecology}
 
-# Section title {#sec:1}
+Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
+In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa) and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
 
-Open-source is great \cite{sandve2013}.
+While human activities have altered the landscapes for millennials \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans on landscapes have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}.
+Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
+Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important \cite{With2019}.
 
-## Subsection title {#sec:2}
+Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}.
+Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984}.
+Consequently, the field of landscape ecology relies on tools to preprocess, modify, model, analyze and visualize spatial data.
 
-as required. Don't forget to give each section
-and subsection a unique label (see Sect. \ref{sec:1}).
+<!-- MH: Do we need some more basic introduction to landscape ecology? -->
 
-#### Paragraph headings
+## Open-source software and R {#sec:open_source}
 
-Use paragraph headings as needed.
+Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
+Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
 
-\begin{align}
-a^2+b^2=c^2
-\end{align}
+<!-- MH: Is this enough about the advantages of open-source? I feel like this could be a bit more in contrast to proprietary software, but also don't want to repeat stuff -->
+
+One example of a successful open-source project is the *R* programming language and its *Comprehensive R Archive Network* (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
+Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
+Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
+This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} with currently almost 300 packages in total.
+The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
+
+Even though many other open-source tools for landscape ecology exist \cite{Baker1992,QGISDevelopmentTeam2016,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language.
+For more general overviews see e.g.
+\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Therefore, in the first part we are going to try to give a comprehensive overview over existing *R* packages for landscape ecology.
+In the second part, we will identify topics for which *R* packages are currently missing to the best of our knowledge.
+
+<!-- MH: Maybe there are better examples for open-source but not R than currently cited, e.g. GRASS -->
+
+# Existing packages {#sec:existing_packages}
+
+<!-- MH: I went for a structure where in the first paragraph I briefly introduce the topic and then list the most common packages. -->
+
+<!-- MH: Maybe we need to think about a table with the topics and the packages as a nice overview. Additionally, a figure where we show a typical workflow would be nice maybe -->
+
+## Basic spatial data
+
+## Creating maps
+
+## Spatial regression
+<!-- MH: Do we really want to include this? -->
+
+## Ecological analysis {#sec:ecological_analysis}
+
+#### Landscape metrics {#sec:landscape_metrics}
+
+One of the most fundamental steps of landscape ecology is to describe and quantify landscape characteristics \cite{Turner2005,Lausch2015}.
+For discrete land cover classes the composition (number and abundance) and configuration (spatial arrangement) of the landscape is often described using landscape metrics \cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}.
+These metrics allow to compare different landscapes, quantify temporal and spatial landscape changes and investigate interactions between landscape characteristics and ecological processes \cite{Uuemaa2009}.
+
+The use of landscape metrics was heavily facilitated by the introduction of the *FRAGSTATS* software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
+However, *FRAGSTATS* is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency and reproducibility.
+To provide a truly open-source solution to calculate landscape metrics, recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed.
+The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment.
+
+#### Species distribution modeling {#sec:SDM}
+
+Species distribution modeling (SDM) analysis how landscape patterns (e.g.
+habitat suitability, resources availability, etc.) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}.
+Originated in the 1970s, SDM has experienced numerous methodological advancement and a numerous body of literature exists today \cite{Zimmermann2010}.
+Additionally, textbooks introducing  basic concepts of SDM in R exists \cite{Guisan2017,Fletcher2019}.
+
+Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs.
+Popular approaches and packages include generalized linear models using e.g.
+the *stats* package; generalized additive models using e.g.the *mgcv* or *lme4* package; classification and regression trees (CART) using e.g.
+the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using e.g.
+the *ade4* package.
+Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}.
+However, currently the *maxent* package is not available on *CRAN* anymore.
+Of course, also packages specifically designed for SDM exists.
+This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.
+
+#### Landscape genetics {#sec:landscape_genetics}
+
+Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift and selection \cite{Manel2003}.
+Such insights improve our understanding of metapopulation dynamics, speciation, species’ distributions and conservation \cite{Storfer2007}.
+By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than e.g.
+metapopulation genetics \cite{Holderegger2006}.
+As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.
+
+<!-- MH: This is work in progress. However, there seems to be a lot of connection to basic connectivity. Maybe include here and/or combine the overall topic with connectivity. -->
+
+<!-- Resistance surface -->
+<!-- Commute distances -->
+<!-- Least cost path -->
+
+<!-- *graph4lg*  -->
+<!-- *gdistance* -->
+<!-- *lme4* -->
+<!-- *HierDpart* -->
+<!-- *PopGenReport* -->
+<!-- https://dyerlab.github.io/Landscape-Genetics-Data-Analysis/index.html -->

### 2020-07-09 paper/paper.tex
Starting to work on manuscript

--- 2020-06-08 paper/paper.tex
@@
 
-\hypertarget{intro}{%
-\section{Introduction}\label{intro}}
+\hypertarget{sec:intro}{%
+\section{Introduction}\label{sec:intro}}
 
-Landscape ecology does this and that \cite{turner2005}.
+\hypertarget{sec:landscape_ecology}{%
+\subsection{A short introduction to landscape
+ecology}\label{sec:landscape_ecology}}
 
-\hypertarget{sec:1}{%
-\section{Section title}\label{sec:1}}
+Landscape ecology focuses on how ecological processes are influenced and
+modified by the heterogeneous landscapes they occur in and
+simultaneously how the ecological processes themselves influence the
+landscapes \cite{Turner1989,Turner2005,With2019}. In this context,
+landscape ecology considers, besides others, i) spatial and temporal
+dynamics of heterogeneous landscapes, ii) interactions, fluxes and
+exchange within these landscapes, iii) how the landscapes influence
+ecological processes (and vice versa) and lastly iv) how to manage these
+heterogeneous landscapes \cite{Risser1984,Turner1989}.
 
-Open-source is great \cite{sandve2013}.
+While human activities have altered the landscapes for millennials
+\cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans
+on landscapes have increased to an unknown high, known as the the
+Anthropocene \cite{Crutzen2002}. Today, almost all landscapes are
+directly or indirectly influenced by human activities
+\cite{Vitousek1997}. Thus, understanding the complex interactions
+between landscapes and ecological processes becomes increasingly
+important \cite{With2019}.
 
-\hypertarget{sec:2}{%
-\subsection{Subsection title}\label{sec:2}}
+Because landscapes are defined as mosaics of different land covers,
+ecosystems, habitat types, or land uses
+\cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and
+ecological processes will vary spatially \cite{With2019}. Thus, in
+contrast to many other sub-disciplines of ecology, landscape ecology
+emphasizes spatial patterns to a high degree \cite{Risser1984}.
+Consequently, the field of landscape ecology relies on tools to
+preprocess, modify, model, analyze and visualize spatial data.
 
-as required. Don't forget to give each section and subsection a unique
-label (see Sect. \ref{sec:1}).
+\hypertarget{sec:open_source}{%
+\subsection{Open-source software and R}\label{sec:open_source}}
 
-\hypertarget{paragraph-headings}{%
-\paragraph{Paragraph headings}\label{paragraph-headings}}
+Open-source software includes all software which is released under a
+license that allows to freely use, modify and distribute the software
+\cite{St.Laurent2008}. Open-source software development has many
+advantages, such as fast innovation, transparency and reliability as
+well as longevity mainly due to many diverse contributors
+\cite{vonKrogh2006,St.Laurent2008}.
 
-Use paragraph headings as needed.
+One example of a successful open-source project is the \emph{R}
+programming language and its \emph{Comprehensive R Archive Network}
+(CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
+Firstly introduced in 1995 \cite{Smith2016}, today the programming
+language is among the most popular programming languages, especially in
+ecology \cite{Lai2019}. Originally introduced as a statistical
+programming language, a growing body of packages designed to analyze
+spatial data subsequently emerged for the \emph{R} programming language
+\cite{Bivand2006}. This is documented by the growing \emph{CRAN Task
+View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task
+View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020}
+with currently almost 300 packages in total. The growing popularity of
+the \emph{R} programming language for spatial data analysis and
+landscape ecology can also be seen by the increasing number of related
+textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
 
-\begin{align}
-a^2+b^2=c^2
-\end{align}
+Even though many other open-source tools for landscape ecology exist
+\cite{Baker1992,QGISDevelopmentTeam2016,Porta2017}, in this review we
+are going to focus on tools implemented in the \emph{R} programming
+language. For more general overviews see
+e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Therefore, in the first part we are going to try to give a comprehensive
+overview over existing \emph{R} packages for landscape ecology. In the
+second part, we will identify topics for which \emph{R} packages are
+currently missing to the best of our knowledge.
+
+\hypertarget{sec:existing_packages}{%
+\section{Existing packages}\label{sec:existing_packages}}
+
+\hypertarget{basic-spatial-data}{%
+\subsection{Basic spatial data}\label{basic-spatial-data}}
+
+\hypertarget{creating-maps}{%
+\subsection{Creating maps}\label{creating-maps}}
+
+\hypertarget{spatial-regression}{%
+\subsection{Spatial regression}\label{spatial-regression}}
+
+\hypertarget{sec:ecological_analysis}{%
+\subsection{Ecological analysis}\label{sec:ecological_analysis}}
+
+\hypertarget{sec:landscape_metrics}{%
+\paragraph{Landscape metrics}\label{sec:landscape_metrics}}
+
+One of the most fundamental steps of landscape ecology is to describe
+and quantify landscape characteristics \cite{Turner2005,Lausch2015}. For
+discrete land cover classes the composition (number and abundance) and
+configuration (spatial arrangement) of the landscape is often described
+using landscape metrics
+\cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}. These metrics
+allow to compare different landscapes, quantify temporal and spatial
+landscape changes and investigate interactions between landscape
+characteristics and ecological processes \cite{Uuemaa2009}.
+
+The use of landscape metrics was heavily facilitated by the introduction
+of the \emph{FRAGSTATS} software \cite{McGarigal2012}, firstly published
+in 1995 \cite{Kupfer2012,Gustafson2019}. However, \emph{FRAGSTATS} is
+not not open-source software (however it is free to download at the
+developers' homepage), which restricts transparency and reproducibility.
+To provide a truly open-source solution to calculate landscape metrics,
+recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was
+developed. The package allows to calculate the most widely used
+landscape metrics in a transparent and reproducible workflow within the
+\emph{R} environment.
+
+\hypertarget{sec:SDM}{%
+\paragraph{Species distribution modeling}\label{sec:SDM}}
+
+Species distribution modeling (SDM) analysis how landscape patterns
+(e.g.~habitat suitability, resources availability, etc.) influence and
+determine the patterns of species distribution, mainly to infer
+ecological processes and predict future species distributions
+\cite{Wiersma2011}. Originated in the 1970s, SDM has experienced
+numerous methodological advancement and a numerous body of literature
+exists today \cite{Zimmermann2010}. Additionally, textbooks introducing
+basic concepts of SDM in R exists \cite{Guisan2017,Fletcher2019}.
+
+Because the used modeling approaches are diverse
+\cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of
+\emph{R} packages used for SDMs. Popular approaches and packages include
+generalized linear models using e.g.~the \emph{stats} package;
+generalized additive models using e.g.the \emph{mgcv} or \emph{lme4}
+package; classification and regression trees (CART) using e.g.~the
+\emph{rpart}, \emph{randomForest} or \emph{ranger} package or
+multivariate data analysis using e.g.~the \emph{ade4} package. Another
+widely used SDM approach uses the concept of maximum entropy
+\cite{Phillips2006,Fletcher2019}. However, currently the \emph{maxent}
+package is not available on \emph{CRAN} anymore. Of course, also
+packages specifically designed for SDM exists. This includes the
+\emph{dismo}, \emph{sdm}, \emph{ecospat}, \emph{biomod2} and
+\emph{PresenceAbsence} packages.
+
+\hypertarget{sec:landscape_genetics}{%
+\paragraph{Landscape genetics}\label{sec:landscape_genetics}}
+
+Landscape genetics investigates how characteristics of landscapes
+interact with gene flow, genetic drift and selection \cite{Manel2003}.
+Such insights improve our understanding of metapopulation dynamics,
+speciation, species' distributions and conservation \cite{Storfer2007}.
+By explicitly including characteristics of landscape, landscape genetics
+is a way more realistic way of analysis than e.g.~metapopulation
+genetics \cite{Holderegger2006}. As a result of its interdisciplinary,
+landscape genetics combines methods from different disciplines including
+landscape ecology, spatial statistics, geography and population genetics
+\cite{Storfer2007}.
 

### 2020-07-10 paper/paper.Rmd
More work on existing packages

@@
 
+<!-- MH: Not super happy with the list i) - iv) -->
+
 While human activities have altered the landscapes for millennials \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans on landscapes have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}.
@@
 Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}.
-Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984}.
-Consequently, the field of landscape ecology relies on tools to preprocess, modify, model, analyze and visualize spatial data.
+Related to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}.
+Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}.
+Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze and visualize spatial data.
 
@@
 
+Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
 Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
 Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
-
-<!-- MH: Is this enough about the advantages of open-source? I feel like this could be a bit more in contrast to proprietary software, but also don't want to repeat stuff -->
+Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}.
+Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and customize existing software to their specific needs easily \cite{Steiniger2009}.
 
@@
 
-Even though many other open-source tools for landscape ecology exist \cite{Baker1992,QGISDevelopmentTeam2016,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language.
+Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language.
+Thus, in the first part we are going to try to give a comprehensive overview over existing *R* packages for landscape ecology.
+In the second part, we will identify topics for which *R* packages are currently missing to the best of our knowledge.
 For more general overviews see e.g.
 \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Therefore, in the first part we are going to try to give a comprehensive overview over existing *R* packages for landscape ecology.
-In the second part, we will identify topics for which *R* packages are currently missing to the best of our knowledge.
-
-<!-- MH: Maybe there are better examples for open-source but not R than currently cited, e.g. GRASS -->
 
@@
 
+<!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->
+
+<!-- MH: Add references for all packages -->
+
 ## Basic spatial data
+
+<!--MH: Short paragraph about scale; *grainchanger* https://doi.org/10.1111/2041-210X.13177 -->
 
@@
 ## Spatial regression
-<!-- MH: Do we really want to include this? -->
+<!-- MH: Do we really want to include this? Not really a landscape ecology topic but rather (geo) spatial stats I would argue -->
 
@@
 
+#### Connectivity {#sec:connectivity}
+Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
+Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
+While structural connectivity focuses only on landscape characteristics (e.g.
+movement corridors, barriers, etc.), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
+
+Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}.
+On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g.
+amount of suitable habitat) \cite{Kindlmann2008,With2019}.
+Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}).
+Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}.
+In graph theory  \cite{Laita2011} landscapes are described by nodes (i.e.
+habitat patches) connected by and functional connections called links \cite{Laita2011}.
+The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
+More general, the *igraph* packages \cite{Csardi2006} provides functionality related to graph theory.
+
+<!-- MH: Work-in-Progress -->
+
+<!-- - resistance surfaces (DOI: https://doi.org/10.1016/S0169-2046(02)00242-6): *gdistance* package DOI: 10.18637/jss.v076.i13-->
+
+<!-- - *samc*: an R package for connectivity modeling with spatial absorbing Markov chains; DOI: https://doi.org/10.1111/ecog.04891 -->
+
 #### Landscape genetics {#sec:landscape_genetics}
@@
 As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.
+
+<!-- MH: Work-in-Progress -->
+<!-- ISOLATION BY RESISTANCE; DOI: 10.1111/j.0014-3820.2006.tb00500.x -->
+
+#### Neutral landscape models {#sec:NLM}
 

### 2020-07-10 paper/paper.tex
More work on existing packages

@@
 \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and
-ecological processes will vary spatially \cite{With2019}. Thus, in
-contrast to many other sub-disciplines of ecology, landscape ecology
-emphasizes spatial patterns to a high degree \cite{Risser1984}.
-Consequently, the field of landscape ecology relies on tools to
-preprocess, modify, model, analyze and visualize spatial data.
+ecological processes will vary spatially \cite{With2019}. Related to
+this, the importance of scale was already raised decades ago
+\cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until
+today \cite{Simova2012,Estes2018}. Thus, in contrast to many other
+sub-disciplines of ecology, landscape ecology emphasizes especially
+spatial patterns to a high degree \cite{Risser1984}. Consequently, the
+field of landscape ecology relies on software to preprocess, modify,
+model, analyze and visualize spatial data.
 
@@
 
-Open-source software includes all software which is released under a
-license that allows to freely use, modify and distribute the software
-\cite{St.Laurent2008}. Open-source software development has many
-advantages, such as fast innovation, transparency and reliability as
-well as longevity mainly due to many diverse contributors
-\cite{vonKrogh2006,St.Laurent2008}.
+Software to manage and analyze data becomes increasingly important in
+modern scientific research \cite{Wilson2014} and many scientific studies
+would not have been possible without open-source software
+\cite{Prlic2012}. Open-source software includes all software which is
+released under a license that allows to freely use, modify and
+distribute the software \cite{St.Laurent2008}. Open-source software
+development has many advantages, such as fast innovation, transparency
+and reliability as well as longevity mainly due to many diverse
+contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use
+of open-source software facilitates (computational) reproducibility and
+can allow a better understanding of the methodology
+\cite{Prlic2012,Powers2019}. Lastly, open-source software allows other
+scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012}
+and customize existing software to their specific needs easily
+\cite{Steiniger2009}.
 
@@
 Even though many other open-source tools for landscape ecology exist
-\cite{Baker1992,QGISDevelopmentTeam2016,Porta2017}, in this review we
-are going to focus on tools implemented in the \emph{R} programming
-language. For more general overviews see
+\cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in
+this review we are going to focus on tools implemented in the \emph{R}
+programming language. Thus, in the first part we are going to try to
+give a comprehensive overview over existing \emph{R} packages for
+landscape ecology. In the second part, we will identify topics for which
+\emph{R} packages are currently missing to the best of our knowledge.
+For more general overviews see
 e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Therefore, in the first part we are going to try to give a comprehensive
-overview over existing \emph{R} packages for landscape ecology. In the
-second part, we will identify topics for which \emph{R} packages are
-currently missing to the best of our knowledge.
 
@@
 
+\hypertarget{sec:connectivity}{%
+\paragraph{Connectivity}\label{sec:connectivity}}
+
+Connectivity is one of the core elements of landscape structure
+\cite{Taylor1993} and thus one of the core concepts of landscape ecology
+\cite{With2019}. Landscape connectivity describes how landscape
+characteristics facilitate or hinder the movement of species
+\cite{Tischendorf2000} or other aspects of mobility, such as dispersal,
+gene or nutrient flow \cite{With2019}. While structural connectivity
+focuses only on landscape characteristics (e.g.~movement corridors,
+barriers, etc.), functional connectivity also includes behavior
+characteristics of the species \cite{Tischendorf2000,With2019}.
+
+Given its wide concept, many different measures of connectivity exists
+\cite{Kindlmann2008}. On patch level structural connectivity can be
+measured using nearest-neighbor distances or characterizations of the
+patch neighborhood (e.g.~amount of suitable habitat)
+\cite{Kindlmann2008,With2019}. Such measures are provided within the
+\emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
+Another way to describe connectivity is based on graph theory with the
+advantages that also functional connectivity can be included
+\cite{Kindlmann2008}. In graph theory \cite{Laita2011} landscapes are
+described by nodes (i.e.~habitat patches) connected by and functional
+connections called links \cite{Laita2011}. The \emph{grainscape} package
+\cite{Chubaty2020} provides a tool to model connectivity based on
+spatially explicit networks. More general, the \emph{igraph} packages
+\cite{Csardi2006} provides functionality related to graph theory.
+
 \hypertarget{sec:landscape_genetics}{%
@@
 
+\hypertarget{sec:NLM}{%
+\paragraph{Neutral landscape models}\label{sec:NLM}}
+
 \bibliographystyle{spphys}

### 2020-07-10 paper/paper.Rmd
Adding some comments to manuscript

@@
 
-<!--MH: Short paragraph about scale; *grainchanger* https://doi.org/10.1111/2041-210X.13177 -->
+<!--MH: Short paragraph about scale; *grainchanger*; DOI: 10.1111/2041-210X.13177 -->
+
+<!-- MH: landscapetools; DOI: 10.1111/2041-210X.13076 -->
 
@@
 <!-- MH: Work-in-Progress -->
+<!-- There seems to be a lot of connection to basic connectivity -->
 <!-- ISOLATION BY RESISTANCE; DOI: 10.1111/j.0014-3820.2006.tb00500.x -->
-
-#### Neutral landscape models {#sec:NLM}
-
-<!-- MH: This is work in progress. However, there seems to be a lot of connection to basic connectivity. Maybe include here and/or combine the overall topic with connectivity. -->
-
-<!-- Resistance surface -->
-<!-- Commute distances -->
-<!-- Least cost path -->
-
 <!-- *graph4lg*  -->
@@
 <!-- https://dyerlab.github.io/Landscape-Genetics-Data-Analysis/index.html -->
+
+#### Neutral landscape models {#sec:NLM}
+
+<!-- MH: NLMR; DOI:  https://doi.org/10.1111/2041-210X.13076 -->

### 2020-07-14 paper/paper.Rmd
Some more progress on existing packages

@@
 # authorrunning: Short form of author list if too long for running head
-
 thanks: | 
@@
     end of the article.
-
 authors: 
@@
   email: ...
-
 keywords:
@@
 - ...
-
 abstract: |
@@
 (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-
 bibliography: bibliography.bib
@@
 output: rticles::springer_article
+editor_options: 
+  chunk_output_type: console
 ---
@@
 Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
-Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
+Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
 Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}.
-Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and customize existing software to their specific needs easily \cite{Steiniger2009}.
+Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.
 
@@
 Originated in the 1970s, SDM has experienced numerous methodological advancement and a numerous body of literature exists today \cite{Zimmermann2010}.
-Additionally, textbooks introducing  basic concepts of SDM in R exists \cite{Guisan2017,Fletcher2019}.
+Additionally, textbooks introducing  basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2019}.
 
@@
 the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using e.g.
-the *ade4* package.
-Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}.
-However, currently the *maxent* package is not available on *CRAN* anymore.
+the *ade4* or *vegan* package.
 Of course, also packages specifically designed for SDM exists.
@@
 More general, the *igraph* packages \cite{Csardi2006} provides functionality related to graph theory.
-
-<!-- MH: Work-in-Progress -->
-
-<!-- - resistance surfaces (DOI: https://doi.org/10.1016/S0169-2046(02)00242-6): *gdistance* package DOI: 10.18637/jss.v076.i13-->
-
-<!-- - *samc*: an R package for connectivity modeling with spatial absorbing Markov chains; DOI: https://doi.org/10.1111/ecog.04891 -->
+Resistance surfaces and least-cost paths are another tool to model functional connectivity.
+The resistance surface describes effects of facilitating or hindering characteristics of the landscape for an organism moving within it \cite{Adriaensen2003}.
+Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}.
+Absorbing Markov chains that combines random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published *samc* package \cite{Marx2020}.
 
@@
 <!-- MH: Work-in-Progress -->
-<!-- There seems to be a lot of connection to basic connectivity -->
-<!-- ISOLATION BY RESISTANCE; DOI: 10.1111/j.0014-3820.2006.tb00500.x -->
-<!-- *graph4lg*  -->
-<!-- *gdistance* -->
-<!-- *lme4* -->
-<!-- *HierDpart* -->
-<!-- *PopGenReport* -->
-<!-- https://dyerlab.github.io/Landscape-Genetics-Data-Analysis/index.html -->
+<!-- MH: General population genetics -->
+<!-- MH: Isolation by resistance  -->
+<!-- MH: Bayesian Genetic Clustering -->
+<!-- MH: Multivariate Ordination Techniques -->
+
+Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also of importance for landscape genetics.
+Further functionality for landscape genetics can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, or *HierDpart* packages \cite{Qin2019}.
 
 #### Neutral landscape models {#sec:NLM}
+Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
+Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found (for a comprehensive list, see \cite{Sciaini2018}).
 
-<!-- MH: NLMR; DOI:  https://doi.org/10.1111/2041-210X.13076 -->
+Currently, to our knowledge, only one *R* packages exists that allows the use of neutral landscapes.
+Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters.
+
+<!-- # Missing packages -->
+
+<!-- Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}. However, currently the *maxent* package is not available on *CRAN* anymore. -->

### 2020-07-14 paper/paper.tex
Some more progress on existing packages

@@
 
-\title{Open source landscape ecology tools \thanks{Grants or other notes
-about the article that should go on the front page should be placed
-here. General acknowledgments should be placed at the end of the
-article.} }
+\title{Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page
+should be placed here. General acknowledgments should be placed at the
+end of the article.} }
 
-\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Author
-3 \and  \ldots{} \and  }
+\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Author 3 \and  \ldots{} \and  }
 
@@
         Maximillian H.K. Hesselbarth \at
-     Department of Ecosystem Modelling, University of Goettingen,
-Buesgenweg 4, 37077 Goettingen, Germany \\
+     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg
+ 4, 37077 Goettingen, Germany \\
      \email{\href{mailto:maximilian.hesselbarth@uni-goettingen.de}{\nolinkurl{maximilian.hesselbarth@uni-goettingen.de}}}  %  \\
@@
         Jakub Nowosad \at
-     Institute of Geoecology and Geoinformation, Adam Mickiewicz
-University, Krygowskiego 10, 61-680 Poznan, Poland \\
+     Institute of Geoecology and Geoinformation, Adam Mickiewicz University,
+ Krygowskiego 10, 61-680 Poznan, Poland \\
      \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
@@
 development has many advantages, such as fast innovation, transparency
-and reliability as well as longevity mainly due to many diverse
+and reliability as well as longevity, mainly due to many diverse
 contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use
@@
 scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012}
-and customize existing software to their specific needs easily
+and to customize existing software to their specific needs
 \cite{Steiniger2009}.
@@
 exists today \cite{Zimmermann2010}. Additionally, textbooks introducing
-basic concepts of SDM in R exists \cite{Guisan2017,Fletcher2019}.
+basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2019}.
 
@@
 \emph{rpart}, \emph{randomForest} or \emph{ranger} package or
-multivariate data analysis using e.g.~the \emph{ade4} package. Another
-widely used SDM approach uses the concept of maximum entropy
-\cite{Phillips2006,Fletcher2019}. However, currently the \emph{maxent}
-package is not available on \emph{CRAN} anymore. Of course, also
-packages specifically designed for SDM exists. This includes the
-\emph{dismo}, \emph{sdm}, \emph{ecospat}, \emph{biomod2} and
-\emph{PresenceAbsence} packages.
+multivariate data analysis using e.g.~the \emph{ade4} or \emph{vegan}
+package. Of course, also packages specifically designed for SDM exists.
+This includes the \emph{dismo}, \emph{sdm}, \emph{ecospat},
+\emph{biomod2} and \emph{PresenceAbsence} packages.
 
@@
 \cite{Csardi2006} provides functionality related to graph theory.
+Resistance surfaces and least-cost paths are another tool to model
+functional connectivity. The resistance surface describes effects of
+facilitating or hindering characteristics of the landscape for an
+organism moving within it \cite{Adriaensen2003}. Least-cost paths can be
+calculated using the \emph{gdistance} package \cite{vanEtten2017}.
+Absorbing Markov chains that combines random walk theory and mortality
+based on the landscape characteristics \cite{Fletcher2019a} and is
+provided by the recently published \emph{samc} package \cite{Marx2020}.
 
@@
 
+Since describing connectivity between two locations is one of the
+fundamental steps of landscape genetics, all packages useful for
+connectivity (see \ref{sec:landscape_genetics}) are also of importance
+for landscape genetics. Further functionality for landscape genetics can
+be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport}
+\cite{Adamack2014,Gruber2015}, or \emph{HierDpart} packages
+\cite{Qin2019}.
+
 \hypertarget{sec:NLM}{%
 \paragraph{Neutral landscape models}\label{sec:NLM}}
+
+Neutral landscape models are used to create structured landscapes in the
+absence of specific ecological and landscape processes as null models
+against which hypotheses including specific ecological and landscape
+processes can be tested statistically \cite{Gardner1987,With1997}.
+Because neutral landscape models are not based on ecological and
+landscape process, many different generic algorithms to create
+landscapes can be found (for a comprehensive list, see
+\cite{Sciaini2018}).
+
+Currently, to our knowledge, only one \emph{R} packages exists that
+allows the use of neutral landscapes. Using a wide collection of
+algorithms, the \emph{NLMR} package \cite{Sciaini2018} provides
+framework to simulate neutral landscape rasters.
 

### 2020-07-15 paper/paper.Rmd
Starting to add table

@@
 bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
+preamble: | 
+  \usepackage{caption}
+  \usepackage{ltablex}
+  \usepackage{lscape}
+  \usepackage{multirow}
 output: rticles::springer_article
@@
 The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment.
+Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{Nowosad2019a}.
 
@@
 
+\begin{landscape}
+
+\begin{center}
+  \begin{table}
+    \begin{tabularx}{0.9\linewidth}{XXXX}
+      \caption{Overview of commonly used \textit{R} packages for landscape ecology.
+Packages are sorted by their major application topic.
+Only packages especially for spatial data and landscape ecology are included.
+SDM = Species distribution modeling; NLM = Neutral landscape model} \\
+      \hline
+      Topic & Package & Description & Source \\
+      \hline
+      \multirow{5}{*}{Spatial data} & \textit{raster} & Methods for raster data & \cite{Hijmans2019} \\
+      & \textit{stars} & Methods for spatiotemporal arrays (raster and vector data cubes) & \cite{Pebesma2019} \\
+      & \textit{sp} & Methods for mainly vector data & \cite{Pebesma2005,Bivand2013} \\
+      & \textit{sf} & Methods for vector data as simple features & \cite{Pebesma2018} \\
+      & \textit{landscapetools} & Utility functions for landscape analysis & \cite{Sciaini2018} \\
+      \hline
+      \multirow{3}{*}{Creating maps} & \textit{tmap} & xxx & xxx \\
+      & \textit{mapview} & xxx & xxx \\
+      & \textit{leaflet} & xxx & xxx \\
+      \hline
+      \multirow{2}{*}{Landscape metrics} & \textit{landscapemetrics} & Calculates landscape metrics for categorical land cover maps & \cite{Hesselbarth2019a} \\
+      & \textit{belg} & Calculates the Boltzmann entropy of a landscape gradient & \cite{Nowosad2019a} \\
+      \hline
+      \multirow{5}{*}{SDM} & \textit{dismo} & xxx & xxx \\
+      & \textit{sdm} & xxx & xxx \\
+      & \textit{ecospat} & xxx & xxx \\
+      & \textit{biomod2} & xxx & xxx \\
+      & \textit{PresenceAbsence} & xxx & xxx \\
+      \hline
+      \multirow{3}{*}{Connectivity} & \textit{grainscape} & xxx & \cite{Chubaty2020} \\
+      & \textit{gdistance} & xxx & \cite{vanEtten2017} \\
+      & \textit{samc} & xxx & \cite{Marx2020} \\
+      \hline
+      \multirow{3}{*}{Landscape genetics} & \textit{graphs4lg} & xxx & \cite{Savary2020} \\
+      & \textit{PopGenReport} & xxx & \cite{Adamack2014,Gruber2015} \\
+      & \textit{HierDpart} & xxx & \cite{Qin2019} \\
+      \hline
+      NLM & \textit{NLMR} & xxx & \cite{Sciaini2018} \\
+      \hline
+    \end{tabularx}
+  \end{table}
+\end{center}
+
+\end{landscape}
+
 <!-- # Missing packages -->

### 2020-07-15 paper/paper.tex
Starting to add table

@@
 %% load any required packages here
+\usepackage{caption}
+\usepackage{ltablex}
+\usepackage{lscape}
+\usepackage{multirow}
 
@@
 
-\title{Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page
-should be placed here. General acknowledgments should be placed at the
-end of the article.} }
+\title{Open source landscape ecology tools \thanks{Grants or other notes
+about the article that should go on the front page should be placed
+here. General acknowledgments should be placed at the end of the
+article.} }
 
-\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Author 3 \and  \ldots{} \and  }
+\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Author
+3 \and  \ldots{} \and  }
 
@@
         Maximillian H.K. Hesselbarth \at
-     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg
- 4, 37077 Goettingen, Germany \\
+     Department of Ecosystem Modelling, University of Goettingen,
+Buesgenweg 4, 37077 Goettingen, Germany \\
      \email{\href{mailto:maximilian.hesselbarth@uni-goettingen.de}{\nolinkurl{maximilian.hesselbarth@uni-goettingen.de}}}  %  \\
@@
 landscape metrics in a transparent and reproducible workflow within the
-\emph{R} environment.
+\emph{R} environment. Additionally, the \emph{belg} packages allows to
+calculate the Boltzmann entropy of a landscape gradient
+\cite{Nowosad2019a}.
 
@@
 
+\begin{landscape}
+
+\begin{center}
+  \begin{table}
+    \begin{tabularx}{0.9\linewidth}{XXXX}
+      \caption{Overview of commonly used \textit{R} packages for landscape ecology.
+Packages are sorted by their major application topic.
+Only packages especially for spatial data and landscape ecology are included.
+SDM = Species distribution modeling; NLM = Neutral landscape model} \\
+      \hline
+      Topic & Package & Description & Source \\
+      \hline
+      \multirow{5}{*}{Spatial data} & \textit{raster} & Methods for raster data & \cite{Hijmans2019} \\
+      & \textit{stars} & Methods for spatiotemporal arrays (raster and vector data cubes) & \cite{Pebesma2019} \\
+      & \textit{sp} & Methods for mainly vector data & \cite{Pebesma2005,Bivand2013} \\
+      & \textit{sf} & Methods for vector data as simple features & \cite{Pebesma2018} \\
+      & \textit{landscapetools} & Utility functions for landscape analysis & \cite{Sciaini2018} \\
+      \hline
+      \multirow{3}{*}{Creating maps} & \textit{tmap} & xxx & xxx \\
+      & \textit{mapview} & xxx & xxx \\
+      & \textit{leaflet} & xxx & xxx \\
+      \hline
+      \multirow{2}{*}{Landscape metrics} & \textit{landscapemetrics} & Calculates landscape metrics for categorical land cover maps & \cite{Hesselbarth2019a} \\
+      & \textit{belg} & Calculates the Boltzmann entropy of a landscape gradient & \cite{Nowosad2019a} \\
+      \hline
+      \multirow{5}{*}{SDM} & \textit{dismo} & xxx & xxx \\
+      & \textit{sdm} & xxx & xxx \\
+      & \textit{ecospat} & xxx & xxx \\
+      & \textit{biomod2} & xxx & xxx \\
+      & \textit{PresenceAbsence} & xxx & xxx \\
+      \hline
+      \multirow{3}{*}{Connectivity} & \textit{grainscape} & xxx & \cite{Chubaty2020} \\
+      & \textit{gdistance} & xxx & \cite{vanEtten2017} \\
+      & \textit{samc} & xxx & \cite{Marx2020} \\
+      \hline
+      \multirow{3}{*}{Landscape genetics} & \textit{graphs4lg} & xxx & \cite{Savary2020} \\
+      & \textit{PopGenReport} & xxx & \cite{Adamack2014,Gruber2015} \\
+      & \textit{HierDpart} & xxx & \cite{Qin2019} \\
+      \hline
+      NLM & \textit{NLMR} & xxx & \cite{Sciaini2018} \\
+      \hline
+    \end{tabularx}
+  \end{table}
+\end{center}
+
+\end{landscape}
+
 \bibliographystyle{spphys}

### 2020-07-18 paper/paper.Rmd
splits text to one line per sentence and adds some new comments

@@
 
-While human activities have altered the landscapes for millennials \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans on landscapes have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}.
+While human activities have altered the landscapes for millenniums \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans on landscapes have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}.
 Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}. 
@@
 Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}.
-Related to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}.
+to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}. 
 Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}. 
@@
 <!-- MH: Do we need some more basic introduction to landscape ecology? -->
+
+<!-- JN:  We still can slightly update the landscape ecology introduction, but I think that the lenght of this text is good.
+Three paragraphs seems enough, as our goal is not to explain what is the landscape ecology, but what are the tools used in this field.
+-->
 
@@
 Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
+<!--probably we should also mention the cost here-->
 Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. 
@@
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
-This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} with currently almost 300 packages in total.
+This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total.
+<!-- we should add that the total number of spatial-related packages is, in fact, larger, as the task views, are manual lists, updated just by a few people -->
 The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
 
-Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language.
-Thus, in the first part we are going to try to give a comprehensive overview over existing *R* packages for landscape ecology.
-In the second part, we will identify topics for which *R* packages are currently missing to the best of our knowledge.
-For more general overviews see e.g.
-\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming.
+<!-- probably we should state why we are focusing on R only -->
+<!-- some answers: "vastness of spatial R packages", ... -->
+Thus, in the first part we are give a comprehensive overview over existing *R* packages for landscape ecology. 
+In the second part, we identify topics for which *R* packages are currently missing to the best of our knowledge. 
+For more general overviews see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
 
@@
 <!-- MH: I went for a structure where in the first paragraph I briefly introduce the topic and then list the most common packages. -->
+<!-- JN: I like this idea -->
 
 <!-- MH: Maybe we need to think about a table with the topics and the packages as a nice overview. Additionally, a figure where we show a typical workflow would be nice maybe -->
+<!-- JN: can you think of one/a few typical workflows? -->
 
 <!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->
+<!-- JN: if you think this package (or similar) are still useful, maybe we can mention them somewhere is the first part of the paper with annotation about their state, and then also write someting more about issues with archived packages in the second part (what's missing in R)...
+-->
 
@@
 The use of landscape metrics was heavily facilitated by the introduction of the *FRAGSTATS* software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
-However, *FRAGSTATS* is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency and reproducibility.
+However, *FRAGSTATS* is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
 To provide a truly open-source solution to calculate landscape metrics, recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed.
@@
 Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{Nowosad2019a}.
+<!-- JN: two notes: -->
+<!-- 1. I am working on a paper supporting the belg package - I plan to submit it in less than a month, so hopefully we could update the reference here -->
+<!-- 2.
+I also submitted a paper yesterday (2020-07-17) explaining a new package motif, which is aimed on calculations based on spatial signatures.
+I plan to add a paragraph about this package someplace in this paper.
+-->
 
@@
 Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. 
-Popular approaches and packages include generalized linear models using e.g.
-the *stats* package; generalized additive models using e.g.the *mgcv* or *lme4* package; classification and regression trees (CART) using e.g.
-the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using e.g.
-the *ade4* or *vegan* package.
+Popular approaches and packages include generalized linear models using, e.g., the *stats* package; generalized additive models using, e.g., the *mgcv* or *lme4* package; classification and regression trees (CART) using, e.g., the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using , e.g., the *ade4* or *vegan* package.
 Of course, also packages specifically designed for SDM exists. 
 This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.
+<!-- JN: I think all of the above packages should be cited -->
 
 #### Connectivity {#sec:connectivity}
+
 Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
 Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
-While structural connectivity focuses only on landscape characteristics (e.g.
-movement corridors, barriers, etc.), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
+While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
 
 Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}. 
-On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g.
-amount of suitable habitat) \cite{Kindlmann2008,With2019}.
+On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g. amount of suitable habitat) \cite{Kindlmann2008,With2019}. 
 Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}). 
 Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}. 
-In graph theory  \cite{Laita2011} landscapes are described by nodes (i.e.
-habitat patches) connected by and functional connections called links \cite{Laita2011}.
+In graph theory  \cite{Laita2011} landscapes are described by nodes (i.e. habitat patches) connected by and functional connections called links \cite{Laita2011}.
 The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
@@
 Such insights improve our understanding of metapopulation dynamics, speciation, species’ distributions and conservation \cite{Storfer2007}. 
-By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than e.g.
-metapopulation genetics \cite{Holderegger2006}.
+By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}. 
 As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.
@@
 #### Neutral landscape models {#sec:NLM}
+
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
@@
 Currently, to our knowledge, only one *R* packages exists that allows the use of neutral landscapes. 
+<!--JN: I am not sure if that's true as few NLMR functions are just wrappers around existing R functions from RandomFields.
+Also, it is possible to simulate landscapes (gradient landscapes) using geostatistical packages, e.g., gstat.
+-->
 Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters. 
+<!-- JN: I would also add landscapetools here as it allows to categorize outputs of the NLMR package -->
 

### 2020-07-18 paper/paper.tex
splits text to one line per sentence and adds some new comments

@@
 
-While human activities have altered the landscapes for millennials
+While human activities have altered the landscapes for millenniums
 \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans
@@
 \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and
-ecological processes will vary spatially \cite{With2019}. Related to
-this, the importance of scale was already raised decades ago
+ecological processes will vary spatially \cite{With2019}. to this, the
+importance of scale was already raised decades ago
 \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until
@@
 View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020}
-with currently almost 300 packages in total. The growing popularity of
-the \emph{R} programming language for spatial data analysis and
+that currently list almost 300 packages in total. The growing popularity
+of the \emph{R} programming language for spatial data analysis and
 landscape ecology can also be seen by the increasing number of related
@@
 this review we are going to focus on tools implemented in the \emph{R}
-programming language. Thus, in the first part we are going to try to
-give a comprehensive overview over existing \emph{R} packages for
-landscape ecology. In the second part, we will identify topics for which
-\emph{R} packages are currently missing to the best of our knowledge.
-For more general overviews see
+programming. Thus, in the first part we are give a comprehensive
+overview over existing \emph{R} packages for landscape ecology. In the
+second part, we identify topics for which \emph{R} packages are
+currently missing to the best of our knowledge. For more general
+overviews see
 e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 not not open-source software (however it is free to download at the
-developers' homepage), which restricts transparency and reproducibility.
-To provide a truly open-source solution to calculate landscape metrics,
-recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was
-developed. The package allows to calculate the most widely used
-landscape metrics in a transparent and reproducible workflow within the
-\emph{R} environment. Additionally, the \emph{belg} packages allows to
-calculate the Boltzmann entropy of a landscape gradient
-\cite{Nowosad2019a}.
+developers' homepage), which restricts transparency, reproducibility,
+and extensibility. To provide a truly open-source solution to calculate
+landscape metrics, recently the \emph{landscapemetrics} package
+\cite{Hesselbarth2019a} was developed. The package allows to calculate
+the most widely used landscape metrics in a transparent and reproducible
+workflow within the \emph{R} environment. Additionally, the \emph{belg}
+packages allows to calculate the Boltzmann entropy of a landscape
+gradient \cite{Nowosad2019a}.
 
@@
 \emph{R} packages used for SDMs. Popular approaches and packages include
-generalized linear models using e.g.~the \emph{stats} package;
-generalized additive models using e.g.the \emph{mgcv} or \emph{lme4}
-package; classification and regression trees (CART) using e.g.~the
+generalized linear models using, e.g., the \emph{stats} package;
+generalized additive models using, e.g., the \emph{mgcv} or \emph{lme4}
+package; classification and regression trees (CART) using, e.g., the
 \emph{rpart}, \emph{randomForest} or \emph{ranger} package or
-multivariate data analysis using e.g.~the \emph{ade4} or \emph{vegan}
+multivariate data analysis using , e.g., the \emph{ade4} or \emph{vegan}
 package. Of course, also packages specifically designed for SDM exists.
@@
 gene or nutrient flow \cite{With2019}. While structural connectivity
-focuses only on landscape characteristics (e.g.~movement corridors,
-barriers, etc.), functional connectivity also includes behavior
+focuses only on landscape characteristics (e.g., movement corridors,
+barriers), functional connectivity also includes behavior
 characteristics of the species \cite{Tischendorf2000,With2019}.
@@
 By explicitly including characteristics of landscape, landscape genetics
-is a way more realistic way of analysis than e.g.~metapopulation
+is a way more realistic way of analysis than, e.g., metapopulation
 genetics \cite{Holderegger2006}. As a result of its interdisciplinary,

### 2020-07-18 paper/paper.Rmd
I would suggest limiting abbriviations as much as possible

@@
 Packages are sorted by their major application topic.
-Only packages especially for spatial data and landscape ecology are included.
-SDM = Species distribution modeling; NLM = Neutral landscape model} \\
+Only packages especially for spatial data and landscape ecology are included.} \\
       \hline
@@
       \hline
-      \multirow{5}{*}{SDM} & \textit{dismo} & xxx & xxx \\
+      \multirow{5}{*}{}Species distribution modeling (SDM) & \textit{dismo} & xxx & xxx \\
       & \textit{sdm} & xxx & xxx \\
@@
       \hline
-      NLM & \textit{NLMR} & xxx & \cite{Sciaini2018} \\
+      Neutral Landscape Models (NLM) & \textit{NLMR} & xxx & \cite{Sciaini2018} \\
       \hline

### 2020-07-18 paper/paper.Rmd
adds a bunch of new ideas

@@
 <!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->
-<!-- JN: if you think this package (or similar) are still useful, maybe we can mention them somewhere is the first part of the paper with annotation about their state, and then also write someting more about issues with archived packages in the second part (what's missing in R)...
+<!-- JN: if you think this package (or similar) are still useful, maybe we can mention them somewhere is the first part of the paper with annotation about their state, and then also write something more about issues with archived packages in the second part (what's missing in R)...
 -->
@@
 
+<!-- JN: maybe we should also add an introduction to R packages here-->
+<!-- for example, CRAN vs GitHub vs other -->
+<!-- who creates R packages? -->
+<!-- who maintains them and what is an archived package -->
+<!-- resources on how to create your own packages - https://r-pkgs.org/ -->
+<!-- comment on never-ending changes in r world -->
+
+## Spatial data representations
+
+<!-- a paragraph or two about what are data representations, why do we need them, and what do they offer -->
+
+<!-- two major data types - raster and vector -->
+
+<!-- raster data representations -->
+<!-- raster (+ mention terra) -->
+<!-- stars -->
+
+<!-- vector data representation -->
+<!-- sp -->
+<!-- sf -->
+<!-- (conversion from text) -->
+<!-- terra new representations -->
+
+## Spatial data download
+
+<!-- access to spatial data -->
+
 ## Basic spatial data
+
+<!-- the role of data preparation - cropping, reprojecting, chaning resolution/simplifing, merging/joing -->
 
@@
 ## Creating maps
+
+<!-- role of making maps - from checking data, through the data exploration, to presenting the results  -->
+
+<!-- maps in the computer age -->
+<!-- static maps and interactive maps -->
+
+<!-- build-in plotting functions -->
+<!-- tmap -->
+<!-- leaflet/mapview -->
+<!-- mention ggplot2/ggspatial as well -->
 

### 2020-07-18 paper/paper.Rmd
adds new ideas

@@
 
+<!-- JN: maybe we can also add info about RMarkdown and similar, with R as an integrative tool for making scientific outputs -->
+
 # Existing packages {#sec:existing_packages}
@@
 
-<!-- the role of data preparation - cropping, reprojecting, chaning resolution/simplifing, merging/joing -->
+<!-- the role of data preparation - cropping, reprojecting, changing resolution/simplifying, merging/joining -->
 
@@
 <!-- MH: landscapetools; DOI: 10.1111/2041-210X.13076 -->
+
+<!-- JN: I am also thinking (unsure yet) about mentioning R as a bridge between other software (e.g. GRASSGIS) -->
 

### 2020-07-19 paper/paper.Rmd
adds some spatial data representations content

@@
 
-<!-- a paragraph or two about what are data representations, why do we need them, and what do they offer -->
+Most of the existing spatial data belong to one of two data models: spatial raster and spatial vectors.
+In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
+The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
+The vector data model consists of two main elements, geometries, such as points, lines, polygons, and attributes.
+Each geometry is connected to a row in an attribute table.
+<!-- ... -->
+A few hundred of different spatial data file formats were developed to store data in these formats.
 
-<!-- two major data types - raster and vector -->
+While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
+However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page ????).
+<!--history section-->
+Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.
+<!-- are pros/cons of that needed? -->
+<!-- info about GDAL/PROJ? -->
 
-<!-- raster data representations -->
-<!-- raster (+ mention terra) -->
+The most prominent package allowing for raster data representation is *raster*.
+<!-- maybe check number of reverse dependencies? -->
+A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed, however, it could take several years for this package to be adopted by users.
+Alternatively, *stars* package can be used to read and process raster data.
 <!-- stars -->
 
-<!-- vector data representation -->
-<!-- sp -->
-<!-- sf -->
-<!-- (conversion from text) -->
-<!-- terra new representations -->
+The *sp* package was a standard for vector data representation for more than 10 years.
+<!-- used by many packages -->
+However, it is now in a maintenance mode only with a recommended successor - the *sf* package.
+<!-- terra new vector representations -->
+<!-- maybe check number of reverse dependencies? -->
+
+Importantly, all of the above packages use the external libraries GDAL and PROJ, which allow for reading and writing of hundreds of spatial data formats and coordinates transformation.
+They also have dedicated methods or functions to convert between representations, for example between *sp* and *sf* objects or *raster* and *stars* objects.
+This can be useful when a given function expects different data representation than currently existing.
 
@@
 <!-- access to spatial data -->
+<!-- rgee -->
 
@@
 <!-- JN: I am also thinking (unsure yet) about mentioning R as a bridge between other software (e.g. GRASSGIS) -->
+<!-- rgee -->
 

### 2020-07-20 R/count-rev-deps.R
count rev deps

@@ -1,0 +1,14 @@
+library(crandep)
+library(dplyr)
+
+rev_sug_raster = get_dep_all("raster", type = c("Reverse_suggests"))
+rev_imp_raster = get_dep_all("raster", type = c("Reverse_imports"))
+rev_dep_raster = get_dep_all("raster", type = c("Reverse_depends"))
+
+rev_sug_sp = get_dep_all("sp", type = c("Reverse_suggests"))
+rev_imp_sp = get_dep_all("sp", type = c("Reverse_imports"))
+rev_dep_sp = get_dep_all("sp", type = c("Reverse_depends"))
+
+rev_sug_sf = get_dep_all("sf", type = c("Reverse_suggests"))
+rev_imp_sf = get_dep_all("sf", type = c("Reverse_imports"))
+rev_dep_sf = get_dep_all("sf", type = c("Reverse_depends"))

### 2020-07-20 paper/paper.Rmd
adds new code for rev deps

@@
 The most prominent package allowing for raster data representation is *raster*.
-<!-- maybe check number of reverse dependencies? -->
 A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed, however, it could take several years for this package to be adopted by users.
 Alternatively, *stars* package can be used to read and process raster data.
-<!-- stars -->
+It is interconnected with the *sf* package, and beside raster support, it provides tools for spatial data cubes.
 
 The *sp* package was a standard for vector data representation for more than 10 years.
-<!-- used by many packages -->
+As of 2020, more than 500 R packages directly depends or imports *sp*.
 However, it is now in a maintenance mode only with a recommended successor - the *sf* package.
 <!-- terra new vector representations -->
-<!-- maybe check number of reverse dependencies? -->
 

### 2020-07-20 paper/paper.Rmd
adds creating maps section

@@
   \usepackage{multirow}
-output: rticles::springer_article
+output:
+  bookdown::pdf_book:
+    base_format: rticles::springer_article
+    number_sections: yes
 editor_options: 
@@
 
-## Spatial data representations
+## Spatial data representations {#spatrep}
 
@@
 
+Creating maps is essential when working with spatial data.
+It has a role in checking the quality of data<!--(both coordinates and values)-->, data exploration, and finally in communicating the results.
 <!-- role of making maps - from checking data, through the data exploration, to presenting the results  -->
+<!-- maps in the computer age -->
+Two main types of maps, static maps with full creator control and interactive maps where more control is given to the user, can be created in R.
 
-<!-- maps in the computer age -->
-<!-- static maps and interactive maps -->
-
-<!-- build-in plotting functions -->
-<!-- tmap -->
-<!-- leaflet/mapview -->
-<!-- mention ggplot2/ggspatial as well -->
+All of the packages listed in Section \@ref(spatrep) have build-in methods for plotting spatial data using the `plot()` function.
+However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
+Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data.
+The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
+This package aims for creating publication-ready maps.
+Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library.<!--this sentence can be easily removed-->
+R's general-purpose plotting package *ggplot2* together with its extension *ggspatial* also offers some capabilities for plotting spatial data.
 

### 2020-07-20 paper/paper.tex
adds a bunch of new ideas
adds creating maps section

@@
 
-\title{Open source landscape ecology tools \thanks{Grants or other notes
-about the article that should go on the front page should be placed
-here. General acknowledgments should be placed at the end of the
-article.} }
+\title{Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front
+page should be placed here. General acknowledgments should be placed at the
+end of the article.} }
 
-\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Author
-3 \and  \ldots{} \and  }
+\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Author 3 \and  \ldots{} \and  }
 
@@
         Maximillian H.K. Hesselbarth \at
-     Department of Ecosystem Modelling, University of Goettingen,
-Buesgenweg 4, 37077 Goettingen, Germany \\
+     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany \\
      \email{\href{mailto:maximilian.hesselbarth@uni-goettingen.de}{\nolinkurl{maximilian.hesselbarth@uni-goettingen.de}}}  %  \\
@@
         Jakub Nowosad \at
-     Institute of Geoecology and Geoinformation, Adam Mickiewicz
-University, Krygowskiego 10, 61-680 Poznan, Poland \\
+     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
      \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
@@
 \begin{abstract}
-max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
-Abstract Abstract Abstract Abstract Abstract Abstract
+max.
+(200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
 \\
@@
 \hypertarget{sec:landscape_ecology}{%
-\subsection{A short introduction to landscape
-ecology}\label{sec:landscape_ecology}}
+\subsection{A short introduction to landscape ecology}\label{sec:landscape_ecology}}
 
-Landscape ecology focuses on how ecological processes are influenced and
-modified by the heterogeneous landscapes they occur in and
-simultaneously how the ecological processes themselves influence the
-landscapes \cite{Turner1989,Turner2005,With2019}. In this context,
-landscape ecology considers, besides others, i) spatial and temporal
-dynamics of heterogeneous landscapes, ii) interactions, fluxes and
-exchange within these landscapes, iii) how the landscapes influence
-ecological processes (and vice versa) and lastly iv) how to manage these
-heterogeneous landscapes \cite{Risser1984,Turner1989}.
+Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
+In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa) and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
 
-While human activities have altered the landscapes for millenniums
-\cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans
-on landscapes have increased to an unknown high, known as the the
-Anthropocene \cite{Crutzen2002}. Today, almost all landscapes are
-directly or indirectly influenced by human activities
-\cite{Vitousek1997}. Thus, understanding the complex interactions
-between landscapes and ecological processes becomes increasingly
-important \cite{With2019}.
+While human activities have altered the landscapes for millenniums \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans on landscapes have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}.
+Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
+Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important \cite{With2019}.
 
-Because landscapes are defined as mosaics of different land covers,
-ecosystems, habitat types, or land uses
-\cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and
-ecological processes will vary spatially \cite{With2019}. to this, the
-importance of scale was already raised decades ago
-\cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until
-today \cite{Simova2012,Estes2018}. Thus, in contrast to many other
-sub-disciplines of ecology, landscape ecology emphasizes especially
-spatial patterns to a high degree \cite{Risser1984}. Consequently, the
-field of landscape ecology relies on software to preprocess, modify,
-model, analyze and visualize spatial data.
+Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}.
+to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}.
+Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}.
+Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze and visualize spatial data.
 
@@
 
-Software to manage and analyze data becomes increasingly important in
-modern scientific research \cite{Wilson2014} and many scientific studies
-would not have been possible without open-source software
-\cite{Prlic2012}. Open-source software includes all software which is
-released under a license that allows to freely use, modify and
-distribute the software \cite{St.Laurent2008}. Open-source software
-development has many advantages, such as fast innovation, transparency
-and reliability as well as longevity, mainly due to many diverse
-contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use
-of open-source software facilitates (computational) reproducibility and
-can allow a better understanding of the methodology
-\cite{Prlic2012,Powers2019}. Lastly, open-source software allows other
-scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012}
-and to customize existing software to their specific needs
-\cite{Steiniger2009}.
+Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
+Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
+Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
+Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}.
+Lastly, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.
 
-One example of a successful open-source project is the \emph{R}
-programming language and its \emph{Comprehensive R Archive Network}
-(CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
-Firstly introduced in 1995 \cite{Smith2016}, today the programming
-language is among the most popular programming languages, especially in
-ecology \cite{Lai2019}. Originally introduced as a statistical
-programming language, a growing body of packages designed to analyze
-spatial data subsequently emerged for the \emph{R} programming language
-\cite{Bivand2006}. This is documented by the growing \emph{CRAN Task
-View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task
-View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020}
-that currently list almost 300 packages in total. The growing popularity
-of the \emph{R} programming language for spatial data analysis and
-landscape ecology can also be seen by the increasing number of related
-textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
+One example of a successful open-source project is the \emph{R} programming language and its \emph{Comprehensive R Archive Network} (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
+Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
+Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006}.
+This is documented by the growing \emph{CRAN Task View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} that currently list almost 300 packages in total.
+The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
 
-Even though many other open-source tools for landscape ecology exist
-\cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in
-this review we are going to focus on tools implemented in the \emph{R}
-programming. Thus, in the first part we are give a comprehensive
-overview over existing \emph{R} packages for landscape ecology. In the
-second part, we identify topics for which \emph{R} packages are
-currently missing to the best of our knowledge. For more general
-overviews see
-e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the \emph{R} programming.
+Thus, in the first part we are give a comprehensive overview over existing \emph{R} packages for landscape ecology.
+In the second part, we identify topics for which \emph{R} packages are currently missing to the best of our knowledge.
+For more general overviews see e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
 
@@
 \section{Existing packages}\label{sec:existing_packages}}
+
+\hypertarget{spatrep}{%
+\subsection{Spatial data representations}\label{spatrep}}
+
+Most of the existing spatial data belong to one of two data models: spatial raster and spatial vectors.
+In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
+The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
+The vector data model consists of two main elements, geometries, such as points, lines, polygons, and attributes.
+Each geometry is connected to a row in an attribute table.
+A few hundred of different spatial data file formats were developed to store data in these formats.
+
+While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
+However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page ????).
+Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.
+
+The most prominent package allowing for raster data representation is \emph{raster}.
+A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed, however, it could take several years for this package to be adopted by users.
+Alternatively, \emph{stars} package can be used to read and process raster data.
+It is interconnected with the \emph{sf} package, and beside raster support, it provides tools for spatial data cubes.
+
+The \emph{sp} package was a standard for vector data representation for more than 10 years.
+As of 2020, more than 500 R packages directly depends or imports \emph{sp}.
+However, it is now in a maintenance mode only with a recommended successor - the \emph{sf} package.
+
+Importantly, all of the above packages use the external libraries GDAL and PROJ, which allow for reading and writing of hundreds of spatial data formats and coordinates transformation.
+They also have dedicated methods or functions to convert between representations, for example between \emph{sp} and \emph{sf} objects or \emph{raster} and \emph{stars} objects.
+This can be useful when a given function expects different data representation than currently existing.
+
+\hypertarget{spatial-data-download}{%
+\subsection{Spatial data download}\label{spatial-data-download}}
 
@@
 \subsection{Creating maps}\label{creating-maps}}
+
+Creating maps is essential when working with spatial data.
+It has a role in checking the quality of data, data exploration, and finally in communicating the results.
+Two main types of maps, static maps with full creator control and interactive maps where more control is given to the user, can be created in R.
+
+All of the packages listed in Section \ref{spatrep} have build-in methods for plotting spatial data using the \texttt{plot()} function.
+However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
+Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data.
+The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
+This package aims for creating publication-ready maps.
+Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library.
+R's general-purpose plotting package \emph{ggplot2} together with its extension \emph{ggspatial} also offers some capabilities for plotting spatial data.
 
@@
 
-One of the most fundamental steps of landscape ecology is to describe
-and quantify landscape characteristics \cite{Turner2005,Lausch2015}. For
-discrete land cover classes the composition (number and abundance) and
-configuration (spatial arrangement) of the landscape is often described
-using landscape metrics
-\cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}. These metrics
-allow to compare different landscapes, quantify temporal and spatial
-landscape changes and investigate interactions between landscape
-characteristics and ecological processes \cite{Uuemaa2009}.
+One of the most fundamental steps of landscape ecology is to describe and quantify landscape characteristics \cite{Turner2005,Lausch2015}.
+For discrete land cover classes the composition (number and abundance) and configuration (spatial arrangement) of the landscape is often described using landscape metrics \cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}.
+These metrics allow to compare different landscapes, quantify temporal and spatial landscape changes and investigate interactions between landscape characteristics and ecological processes \cite{Uuemaa2009}.
 
-The use of landscape metrics was heavily facilitated by the introduction
-of the \emph{FRAGSTATS} software \cite{McGarigal2012}, firstly published
-in 1995 \cite{Kupfer2012,Gustafson2019}. However, \emph{FRAGSTATS} is
-not not open-source software (however it is free to download at the
-developers' homepage), which restricts transparency, reproducibility,
-and extensibility. To provide a truly open-source solution to calculate
-landscape metrics, recently the \emph{landscapemetrics} package
-\cite{Hesselbarth2019a} was developed. The package allows to calculate
-the most widely used landscape metrics in a transparent and reproducible
-workflow within the \emph{R} environment. Additionally, the \emph{belg}
-packages allows to calculate the Boltzmann entropy of a landscape
-gradient \cite{Nowosad2019a}.
+The use of landscape metrics was heavily facilitated by the introduction of the \emph{FRAGSTATS} software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
+However, \emph{FRAGSTATS} is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
+To provide a truly open-source solution to calculate landscape metrics, recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was developed.
+The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the \emph{R} environment.
+Additionally, the \emph{belg} packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{Nowosad2019a}.
 
@@
 
-Species distribution modeling (SDM) analysis how landscape patterns
-(e.g.~habitat suitability, resources availability, etc.) influence and
-determine the patterns of species distribution, mainly to infer
-ecological processes and predict future species distributions
-\cite{Wiersma2011}. Originated in the 1970s, SDM has experienced
-numerous methodological advancement and a numerous body of literature
-exists today \cite{Zimmermann2010}. Additionally, textbooks introducing
-basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2019}.
+Species distribution modeling (SDM) analysis how landscape patterns (e.g.~habitat suitability, resources availability, etc.) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}.
+Originated in the 1970s, SDM has experienced numerous methodological advancement and a numerous body of literature exists today \cite{Zimmermann2010}.
+Additionally, textbooks introducing basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2019}.
 
-Because the used modeling approaches are diverse
-\cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of
-\emph{R} packages used for SDMs. Popular approaches and packages include
-generalized linear models using, e.g., the \emph{stats} package;
-generalized additive models using, e.g., the \emph{mgcv} or \emph{lme4}
-package; classification and regression trees (CART) using, e.g., the
-\emph{rpart}, \emph{randomForest} or \emph{ranger} package or
-multivariate data analysis using , e.g., the \emph{ade4} or \emph{vegan}
-package. Of course, also packages specifically designed for SDM exists.
-This includes the \emph{dismo}, \emph{sdm}, \emph{ecospat},
-\emph{biomod2} and \emph{PresenceAbsence} packages.
+Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of \emph{R} packages used for SDMs.
+Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} package; generalized additive models using, e.g., the \emph{mgcv} or \emph{lme4} package; classification and regression trees (CART) using, e.g., the \emph{rpart}, \emph{randomForest} or \emph{ranger} package or multivariate data analysis using , e.g., the \emph{ade4} or \emph{vegan} package.
+Of course, also packages specifically designed for SDM exists.
+This includes the \emph{dismo}, \emph{sdm}, \emph{ecospat}, \emph{biomod2} and \emph{PresenceAbsence} packages.
 
@@
 
-Connectivity is one of the core elements of landscape structure
-\cite{Taylor1993} and thus one of the core concepts of landscape ecology
-\cite{With2019}. Landscape connectivity describes how landscape
-characteristics facilitate or hinder the movement of species
-\cite{Tischendorf2000} or other aspects of mobility, such as dispersal,
-gene or nutrient flow \cite{With2019}. While structural connectivity
-focuses only on landscape characteristics (e.g., movement corridors,
-barriers), functional connectivity also includes behavior
-characteristics of the species \cite{Tischendorf2000,With2019}.
+Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
+Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
+While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
 
-Given its wide concept, many different measures of connectivity exists
-\cite{Kindlmann2008}. On patch level structural connectivity can be
-measured using nearest-neighbor distances or characterizations of the
-patch neighborhood (e.g.~amount of suitable habitat)
-\cite{Kindlmann2008,With2019}. Such measures are provided within the
-\emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
-Another way to describe connectivity is based on graph theory with the
-advantages that also functional connectivity can be included
-\cite{Kindlmann2008}. In graph theory \cite{Laita2011} landscapes are
-described by nodes (i.e.~habitat patches) connected by and functional
-connections called links \cite{Laita2011}. The \emph{grainscape} package
-\cite{Chubaty2020} provides a tool to model connectivity based on
-spatially explicit networks. More general, the \emph{igraph} packages
-\cite{Csardi2006} provides functionality related to graph theory.
-Resistance surfaces and least-cost paths are another tool to model
-functional connectivity. The resistance surface describes effects of
-facilitating or hindering characteristics of the landscape for an
-organism moving within it \cite{Adriaensen2003}. Least-cost paths can be
-calculated using the \emph{gdistance} package \cite{vanEtten2017}.
-Absorbing Markov chains that combines random walk theory and mortality
-based on the landscape characteristics \cite{Fletcher2019a} and is
-provided by the recently published \emph{samc} package \cite{Marx2020}.
+Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}.
+On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g.~amount of suitable habitat) \cite{Kindlmann2008,With2019}.
+Such measures are provided within the \emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
+Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}.
+In graph theory \cite{Laita2011} landscapes are described by nodes (i.e.~habitat patches) connected by and functional connections called links \cite{Laita2011}.
+The \emph{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
+More general, the \emph{igraph} packages \cite{Csardi2006} provides functionality related to graph theory.
+Resistance surfaces and least-cost paths are another tool to model functional connectivity.
+The resistance surface describes effects of facilitating or hindering characteristics of the landscape for an organism moving within it \cite{Adriaensen2003}.
+Least-cost paths can be calculated using the \emph{gdistance} package \cite{vanEtten2017}.
+Absorbing Markov chains that combines random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published \emph{samc} package \cite{Marx2020}.
 
@@
 
-Landscape genetics investigates how characteristics of landscapes
-interact with gene flow, genetic drift and selection \cite{Manel2003}.
-Such insights improve our understanding of metapopulation dynamics,
-speciation, species' distributions and conservation \cite{Storfer2007}.
-By explicitly including characteristics of landscape, landscape genetics
-is a way more realistic way of analysis than, e.g., metapopulation
-genetics \cite{Holderegger2006}. As a result of its interdisciplinary,
-landscape genetics combines methods from different disciplines including
-landscape ecology, spatial statistics, geography and population genetics
-\cite{Storfer2007}.
+Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift and selection \cite{Manel2003}.
+Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions and conservation \cite{Storfer2007}.
+By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
+As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.
 
-Since describing connectivity between two locations is one of the
-fundamental steps of landscape genetics, all packages useful for
-connectivity (see \ref{sec:landscape_genetics}) are also of importance
-for landscape genetics. Further functionality for landscape genetics can
-be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport}
-\cite{Adamack2014,Gruber2015}, or \emph{HierDpart} packages
-\cite{Qin2019}.
+Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also of importance for landscape genetics.
+Further functionality for landscape genetics can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, or \emph{HierDpart} packages \cite{Qin2019}.
 
@@
 
-Neutral landscape models are used to create structured landscapes in the
-absence of specific ecological and landscape processes as null models
-against which hypotheses including specific ecological and landscape
-processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and
-landscape process, many different generic algorithms to create
-landscapes can be found (for a comprehensive list, see
-\cite{Sciaini2018}).
+Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
+Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found (for a comprehensive list, see \cite{Sciaini2018}).
 
-Currently, to our knowledge, only one \emph{R} packages exists that
-allows the use of neutral landscapes. Using a wide collection of
-algorithms, the \emph{NLMR} package \cite{Sciaini2018} provides
-framework to simulate neutral landscape rasters.
+Currently, to our knowledge, only one \emph{R} packages exists that allows the use of neutral landscapes.
+Using a wide collection of algorithms, the \emph{NLMR} package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters.
 

### 2020-07-20 paper/paper.Rmd
Replying to some comments (either by comments or changes in the text)

@@
   address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
-  email: maximilian.hesselbarth@uni-goettingen.de
+  email: mhk.hesselbarthh@gmail.com
 - name: Jakub Nowosad
@@
 
+<!-- MH: Not super happy with the list i) - iv) -->
+
 Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
 In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa) and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
-
-<!-- MH: Not super happy with the list i) - iv) -->
 
@@
 
-<!-- MH: Do we need some more basic introduction to landscape ecology? -->
-
-<!-- JN:  We still can slightly update the landscape ecology introduction, but I think that the lenght of this text is good.
-Three paragraphs seems enough, as our goal is not to explain what is the landscape ecology, but what are the tools used in this field.
--->
-
 ## Open-source software and R {#sec:open_source}
@@
 Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
-<!--probably we should also mention the cost here-->
 Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. 
-Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.
+Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}. 
+Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to many quite expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
 
@@
 This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total.
-<!-- we should add that the total number of spatial-related packages is, in fact, larger, as the task views, are manual lists, updated just by a few people -->
+Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even much higher.
 The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
 
-Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming.
-<!-- probably we should state why we are focusing on R only -->
-<!-- some answers: "vastness of spatial R packages", ... -->
+Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language.
+The growing body of *R* packages related to spatial data processing and analyzing results in a high capability of the *R* programming language for landscape ecology.
+Because additionally to the handling of spatial data, also further task such as statistical modeling, the preparation of publication-ready figures and even the preparation of complete reports using *RMarkdown* \cite{Xie2018} can be done in the same working environment, the *R* programming language can be a very powerful tool.
 Thus, in the first part we are give a comprehensive overview over existing *R* packages for landscape ecology. 
@@
 
-<!-- JN: maybe we can also add info about RMarkdown and similar, with R as an integrative tool for making scientific outputs -->
-
 # Existing packages {#sec:existing_packages}
-
-<!-- MH: I went for a structure where in the first paragraph I briefly introduce the topic and then list the most common packages. -->
-<!-- JN: I like this idea -->
 
@@
 <!-- comment on never-ending changes in r world -->
+<!-- MH: I like this idea. I will add a short paragraph -->
 
@@
 -->
+<!-- MH: Sounds great -->
 
@@
 <!-- JN: I think all of the above packages should be cited -->
+<!-- MH: Yeah, will do. I was just to lazy  -->
 
@@
 Currently, to our knowledge, only one *R* packages exists that allows the use of neutral landscapes. 
-<!--JN: I am not sure if that's true as few NLMR functions are just wrappers around existing R functions from RandomFields.
+<!-- JN: I am not sure if that's true as few NLMR functions are just wrappers around existing R functions from RandomFields.
 Also, it is possible to simulate landscapes (gradient landscapes) using geostatistical packages, e.g., gstat.
 -->
+<!-- MH: Well, but RandomFields and geostatistcal are not directly designed for NLM. Maybe need to relax the statement a little bit -->
 Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters. 
 <!-- JN: I would also add landscapetools here as it allows to categorize outputs of the NLMR package -->
+<!-- MH: I would rahter add a short sentence about landscapetools in the general spatial data section since the package can be used for all different kinds of tasks not only for NLMR -->
 

### 2020-07-20 paper/paper.tex
Replying to some comments (either by comments or changes in the text)

@@
      Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany \\
-     \email{\href{mailto:maximilian.hesselbarth@uni-goettingen.de}{\nolinkurl{maximilian.hesselbarth@uni-goettingen.de}}}  %  \\
+     \email{\href{mailto:mhk.hesselbarthh@gmail.com}{\nolinkurl{mhk.hesselbarthh@gmail.com}}}  %  \\
 %             \emph{Present address:} of F. Author  %  if needed
@@
 Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}.
-Lastly, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.
+Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.
+Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to many quite expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
 
@@
 This is documented by the growing \emph{CRAN Task View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} that currently list almost 300 packages in total.
+Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely even much higher.
 The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
 
-Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the \emph{R} programming.
+Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the \emph{R} programming language.
+The growing body of \emph{R} packages related to spatial data processing and analyzing results in a high capability of the \emph{R} programming language for landscape ecology.
+Because additionally to the handling of spatial data, also further task such as statistical modeling, the preparation of publication-ready figures and even the preparation of complete reports using \emph{RMarkdown} \cite{Xie2018} can be done in the same working environment, the \emph{R} programming language can be a very powerful tool.
 Thus, in the first part we are give a comprehensive overview over existing \emph{R} packages for landscape ecology.

### 2020-07-20 paper/paper.Rmd
Fixing typo in email

@@
   address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
-  email: mhk.hesselbarthh@gmail.com
+  email: mhk.hesselbarth@gmail.com
 - name: Jakub Nowosad

### 2020-07-21 paper/paper.Rmd
Adding references and section about CRAN/GitHub

@@
 
-One example of a successful open-source project is the *R* programming language and its *Comprehensive R Archive Network* (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}. 
+One example of a successful open-source project is the *R* programming language and its extensions (also called packages) \cite{RCoreTeam2019}. 
 Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}. 
@@
 
+There are two major platforms to host *R* packages and make them easily accessible to potential users. Firstly, hosting the package on *GitHub* allows users to install it with only one line of code. 
+Additionally, hosting the package on *GitHub* provides many useful tools to communicate between developers and users, such as bug reports \cite{Wickham2015}.
+Secondly, the *Comprehensive R Archive Network* (*CRAN*) provides large visibility to the community, ease of installation and a certain quality standard \cite{Wickham2015}.
+However, at the same time, releasing a *R* package on *CRAN* requires more initial work for the developers compared to *GitHub*.
+Since most *R* packages are developed and maintained by the community, this is a good example of how open-source software development can facilitate innovation, reproducibility and reuse of code, especially since now many *R* packages depend on each other (i.e.
+package *x* requires and uses code from package *y*) \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+For more information about *R* package development, see \cite{Wickham2015}.
+
 <!-- MH: Maybe we need to think about a table with the topics and the packages as a nice overview. Additionally, a figure where we show a typical workflow would be nice maybe -->
@@
 
-<!-- MH: Add references for all packages -->
-
-<!-- JN: maybe we should also add an introduction to R packages here-->
-<!-- for example, CRAN vs GitHub vs other -->
-<!-- who creates R packages? -->
-<!-- who maintains them and what is an archived package -->
-<!-- resources on how to create your own packages - https://r-pkgs.org/ -->
-<!-- comment on never-ending changes in r world -->
-<!-- MH: I like this idea. I will add a short paragraph -->
-
-## Spatial data representations {#spatrep}
+## Spatial data representations {#sec:spatrep}
 
@@
 <!-- rgee -->
+<!-- MH: Maybe one short sentence that R allows to connect to other open-source software. We already introduce the topic previously anyhow -->
 
@@
 
-All of the packages listed in Section \@ref(spatrep) have build-in methods for plotting spatial data using the `plot()` function.
+All of the packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the `plot()` function.
 However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
@@
 Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. 
-Popular approaches and packages include generalized linear models using, e.g., the *stats* package; generalized additive models using, e.g., the *mgcv* or *lme4* package; classification and regression trees (CART) using, e.g., the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using , e.g., the *ade4* or *vegan* package.
+Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package.
 Of course, also packages specifically designed for SDM exists. 
-This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.
-<!-- JN: I think all of the above packages should be cited -->
-<!-- MH: Yeah, will do. I was just to lazy  -->
+This includes the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020} and *PresenceAbsence* \cite{Freeman2008} packages.
 

### 2020-07-21 paper/paper.tex
Adding references and section about CRAN/GitHub

@@
      Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany \\
-     \email{\href{mailto:mhk.hesselbarthh@gmail.com}{\nolinkurl{mhk.hesselbarthh@gmail.com}}}  %  \\
+     \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
 %             \emph{Present address:} of F. Author  %  if needed
@@
 
-One example of a successful open-source project is the \emph{R} programming language and its \emph{Comprehensive R Archive Network} (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
+One example of a successful open-source project is the \emph{R} programming language and its extensions (also called packages) \cite{RCoreTeam2019}.
 Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
@@
 
-\hypertarget{spatrep}{%
-\subsection{Spatial data representations}\label{spatrep}}
+There are two major platforms to host \emph{R} packages and make them easily accessible to potential users.
+Firstly, hosting the package on \emph{GitHub} allows users to install it with only one line of code.
+Additionally, hosting the package on \emph{GitHub} provides many useful tools to communicate between developers and users, such as bug reports \cite{Wickham2015}.
+Secondly, the \emph{Comprehensive R Archive Network} (\emph{CRAN}) provides large visibility to the community, ease of installation and a certain quality standard \cite{Wickham2015}.
+However, at the same time, releasing a \emph{R} package on \emph{CRAN} requires more initial work for the developers compared to \emph{GitHub}.
+Since most \emph{R} packages are developed and maintained by the community, this is a good example of how open-source software development can facilitate innovation, reproducibility and reuse of code, especially since now many \emph{R} packages depend on each other (i.e.~package \emph{x} requires and uses code from package \emph{y}) \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+For more information about \emph{R} package development, see \cite{Wickham2015}.
+
+\hypertarget{sec:spatrep}{%
+\subsection{Spatial data representations}\label{sec:spatrep}}
 
@@
 
-All of the packages listed in Section \ref{spatrep} have build-in methods for plotting spatial data using the \texttt{plot()} function.
+All of the packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the \texttt{plot()} function.
 However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
@@
 Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of \emph{R} packages used for SDMs.
-Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} package; generalized additive models using, e.g., the \emph{mgcv} or \emph{lme4} package; classification and regression trees (CART) using, e.g., the \emph{rpart}, \emph{randomForest} or \emph{ranger} package or multivariate data analysis using , e.g., the \emph{ade4} or \emph{vegan} package.
+Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
 Of course, also packages specifically designed for SDM exists.
-This includes the \emph{dismo}, \emph{sdm}, \emph{ecospat}, \emph{biomod2} and \emph{PresenceAbsence} packages.
+This includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020} and \emph{PresenceAbsence} \cite{Freeman2008} packages.
 

### 2020-07-21 paper/paper.Rmd
adds page number

@@
 While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
-However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page ????).
+However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page 10).
 <!--history section-->
@@
 
-## Basic spatial data
+## Spatial data processing
 

### 2020-07-21 paper/paper.Rmd
adds some new ideas

@@
 
+<!-- mention different data types/formats/projections/extents/resolutions -->
 <!-- access to spatial data -->
@@
 
-<!-- the role of data preparation - cropping, reprojecting, changing resolution/simplifying, merging/joining -->
+As we mentioned in the previous sections, obtained data can be represented by different data types and stored in various file formats and read into R using one of the packages from Section \@ref(sec:spatrep).
+However, several potential obstacles to applying this data for ecological analysis still exist.
+
+<!-- geographic/projected CRS -->
+<!-- reprojecting -->
+<!-- Different data can also used one of many coordinate reference systems (CRS) -->
+
+<!-- different extents -->
+<!-- cropping -->
+
+<!-- the role of data preparation - changing resolution/simplifying, merging/joining -->
 

### 2020-07-21 paper/paper.Rmd
adds spatial data processing section

--- 2020-07-21 paper/paper.Rmd
@@
 
-<!-- the role of data preparation - cropping, reprojecting, changing resolution/simplifying, merging/joining -->
+As we mentioned in the previous sections, gathered or obtained data can be represented by different data types and stored in various file formats.
+<!-- and read into R using one of the packages from Section \@ref(sec:spatrep). -->
+However, several potential obstacles to applying this data for ecological analysis still exist.
+
+The main issue for many landscape ecology studies is related to projections of the used data.
+It is not only required to have all of the used datasets in the same projection, but also selection of a proper projection is crucial.
+Coordinates in spatial data can represent one of many coordinate reference systems.
+Two main groups of coordinate reference systems (CRS) exist, namely geographical and projected, each having many members.
+In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
+<!---->
+However, in most landscape ecology studies, projected CRSs, which use some measurement units (e.g., meters), should be utilized.
+<!-- Projected CRS  -->
+Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
+<!-- Where to find advice on how to select projections-->
+Gladly, all of the packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations.
+
+Second possible issue with spatial data is when the obtained data extends over a larger area than the study region.
+Then, some spatial preprecessing, such as vector clipping or raster cropping, is needed.
+Packages from Section \@ref(sec:spatrep) also make these operations possible.
+Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications.
 
 <!--MH: Short paragraph about scale; *grainchanger*; DOI: 10.1111/2041-210X.13177 -->
-
 <!-- MH: landscapetools; DOI: 10.1111/2041-210X.13076 -->
 
-<!-- JN: I am also thinking (unsure yet) about mentioning R as a bridge between other software (e.g. GRASSGIS) -->
-<!-- rgee -->
-<!-- MH: Maybe one short sentence that R allows to connect to other open-source software. We already introduce the topic previously anyhow -->
+Finally, there is a number of tools for landscape ecology implemented in some GIS software, such as ... for GRASS GIS or ... for ... .
+Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* for ..., ... for ..., and *rgee* for Google Earth Engine.
 

### 2020-07-21 paper/paper.Rmd
stars with spatial data download

@@
 
-<!-- mention different data types/formats/projections/extents/resolutions -->
-<!-- access to spatial data -->
+Nowadays, spatial data for many spatial scales is available from abundance of sources.
+Some of them require registration or having appropriate credentials and allows downloading only from the web interface.
+However, many of the sources made their data publicly available either as a direct download or through an API connection.
+This is used by some R packages allowing for downloading various environmental spatial dataset.
+
+<!-- elevatr + ?? - elevation -->
+<!-- land cover -->
+<!-- rgbif + ?? -->
 <!-- rgee -->
@@
 <!-- MH: Do we really want to include this? Not really a landscape ecology topic but rather (geo) spatial stats I would argue -->
+<!-- JN: we could wait with this section to after receiving the results of the questionnaire -->
 

### 2020-07-26 paper/paper.Rmd
adds new ideas
splits lines
improves spatial data processing section

@@
 
-There are two major platforms to host *R* packages and make them easily accessible to potential users. Firstly, hosting the package on *GitHub* allows users to install it with only one line of code. 
+There are two major platforms to host *R* packages and make them easily accessible to potential users.
+Firstly, hosting the package on *GitHub* allows users to install it with only one line of code. 
 Additionally, hosting the package on *GitHub* provides many useful tools to communicate between developers and users, such as bug reports \cite{Wickham2015}.
@@
 However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page 10).
-<!--history section-->
 Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.
@@
 <!-- rgbif + ?? -->
+<!-- climate data? -->
+<!-- https://github.com/rspatial/geodata/blob/master/R/worldclim.R -->
 <!-- rgee -->
+<!-- look for rOpenSci packages -->
 
@@
 
-Finally, there is a number of tools for landscape ecology implemented in some GIS software, such as ... for GRASS GIS or ... for ... .
-Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* for ..., ... for ..., and *rgee* for Google Earth Engine.
+Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass}<!--ADD R.LI CITATION-->, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R-->
+Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
+
+<!--JN: should we mention ArcGIS here and some of its modules?-->
+<!--JN: in the same vain as above: should we mention Google Earth Engine?-->
 
@@
 <!-- Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}. However, currently the *maxent* package is not available on *CRAN* anymore. -->
+
+<!-- missing packages: data sources, such as global land cover, worldclim 2.0, etc. -->

### 2020-07-26 paper/paper.Rmd
adds references to data representation section

@@
 
-The most prominent package allowing for raster data representation is *raster*.
-A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed, however, it could take several years for this package to be adopted by users.
-Alternatively, *stars* package can be used to read and process raster data.
+The most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
+A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
+Alternatively, *stars* package can be used to read and process raster data \cite{Pebesma2019}.
 It is interconnected with the *sf* package, and beside raster support, it provides tools for spatial data cubes.
 
-The *sp* package was a standard for vector data representation for more than 10 years.
+The *sp* package was a standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
 As of 2020, more than 500 R packages directly depends or imports *sp*.

### 2020-07-26 paper/paper.Rmd
adds maps creation references

@@
 However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
-Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data.
+Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
 The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
 This package aims for creating publication-ready maps.
-Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library.<!--this sentence can be easily removed-->
-R's general-purpose plotting package *ggplot2* together with its extension *ggspatial* also offers some capabilities for plotting spatial data.
+Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.<!--this sentence can be easily removed-->
+R's general-purpose plotting package *ggplot2* together with its extension *ggspatial* also offers some capabilities for plotting spatial data \cite{Wickham2016a,R-ggspatial}.
 

### 2020-07-26 paper/paper.tex
adds updates files versions

@@
 While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
-However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page ????).
+However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page 10).
 Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.
 
-The most prominent package allowing for raster data representation is \emph{raster}.
-A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed, however, it could take several years for this package to be adopted by users.
-Alternatively, \emph{stars} package can be used to read and process raster data.
+The most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
+A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
+Alternatively, \emph{stars} package can be used to read and process raster data \cite{Pebesma2019}.
 It is interconnected with the \emph{sf} package, and beside raster support, it provides tools for spatial data cubes.
 
-The \emph{sp} package was a standard for vector data representation for more than 10 years.
+The \emph{sp} package was a standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
 As of 2020, more than 500 R packages directly depends or imports \emph{sp}.
@@
 
-\hypertarget{basic-spatial-data}{%
-\subsection{Basic spatial data}\label{basic-spatial-data}}
+Nowadays, spatial data for many spatial scales is available from abundance of sources.
+Some of them require registration or having appropriate credentials and allows downloading only from the web interface.
+However, many of the sources made their data publicly available either as a direct download or through an API connection.
+This is used by some R packages allowing for downloading various environmental spatial dataset.
+
+A plethora of land cover products on global, continental, and country scales has been developed, however, most of them cannot be directly downloaded from R.
+The exception is the National Land Cover Database (NLCD) data for the USA, which can be accessed through the \emph{FedData} package \cite{R-FedData}.
+The \emph{elevatr} package provides a synthesis raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}.
+It uses different elevation products for a given area and selected level of details (zoom/resolution).
+The \emph{rgbif} gives access to data from the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}.
+The \texttt{get\_data()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim version 1.4, while the new in-development package \emph{geodata} provides an access to WorldClim version 2 data.
+Additionally, the \emph{rgee package} \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
+
+\hypertarget{spatial-data-processing}{%
+\subsection{Spatial data processing}\label{spatial-data-processing}}
+
+As we mentioned in the previous sections, gathered or obtained data can be represented by different data types and stored in various file formats.
+However, several potential obstacles to applying this data for ecological analysis still exist.
+
+The main issue for many landscape ecology studies is related to projections of the used data.
+It is not only required to have all of the used datasets in the same projection, but also selection of a proper projection is crucial.
+Coordinates in spatial data can represent one of many coordinate reference systems.
+Two main groups of coordinate reference systems (CRS) exist, namely geographical and projected, each having many members.
+In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
+However, in most landscape ecology studies, projected CRSs, which use some measurement units (e.g., meters), should be utilized.
+Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
+Gladly, all of the packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations.
+
+Second possible issue with spatial data is when the obtained data extends over a larger area than the study region.
+Then, some spatial preprecessing, such as vector clipping or raster cropping, is needed.
+Packages from Section \ref{sec:spatrep} also make these operations possible.
+Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications.
+
+Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
+Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.
 
@@
 However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
-Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data.
+Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
 The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
 This package aims for creating publication-ready maps.
-Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library.
-R's general-purpose plotting package \emph{ggplot2} together with its extension \emph{ggspatial} also offers some capabilities for plotting spatial data.
+Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
+R's general-purpose plotting package \emph{ggplot2} together with its extension \emph{ggspatial} also offers some capabilities for plotting spatial data \cite{Wickham2016a,R-ggspatial}.
 

### 2020-07-27 paper/paper.Rmd

@@
 
-<!-- elevatr + ?? - elevation -->
-<!-- land cover -->
-<!-- rgbif + ?? -->
-<!-- climate data? -->
-<!-- https://github.com/rspatial/geodata/blob/master/R/worldclim.R -->
-<!-- rgee -->
-<!-- look for rOpenSci packages -->
+A plethora of land cover products on global, continental, and country scales has been developed, however, most of them cannot be directly downloaded from R. 
+The exception is the National Land Cover Database (NLCD) data for the USA, which can be accessed through the *FedData* package \cite{R-FedData}.
+The *elevatr* package provides a synthesis raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}.
+It uses different elevation products for a given area and selected level of details (zoom/resolution).
+<!-- how to cite (or maybe we should just add a link to?) - https://github.com/tilezen/joerd/blob/master/docs/data-sources.md -->
+The *rgbif* gives access to data from the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}.
+The `get_data()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim version 1.4, while the new in-development package *geodata*<!-- https://github.com/rspatial/geodata/blob/master/R/worldclim.R --> provides an access to WorldClim version 2 data.
+Additionally, the *rgee package* \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
+<!-- it is worth to search for more as there are a lot of changes going on -->
+
+<!--JN: Max, do you want to add something to this list?-->
 

### 2020-09-04 survey_stats.Rmd
adds initial resutls of the survey

---
title: "Survey's results"
output: 
  html_document:
    code_folding: hide
    toc: true
theme: cosmo
date: '`r format(Sys.time(), "%Y-%m-%d")`'
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, fig.align = "center")
options(scipen = 99*99)
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(googlesheets4)
library(forcats)
library(stringr)
```

```{r}
# read and prepare --------------------------------------------------------
x = read_sheet("https://docs.google.com/spreadsheets/d/1mV_wZWb4desjdzUWHALIB-uFRCauSTOGSvENsnkiAhI/edit#gid=1189808333")

old_names = names(x)

new_names = c("time", "position", "how_often", "expertise",
              "major_topic", "important_tasks", "data_model",
              "r_packages", "own_package", "own_package_location",
              "missing_methods", "usefulness", "comments")

names(x) = new_names
```

# `r old_names[2]`

```{r}
# position
ggplot(x, aes(y = fct_infreq(position))) +
    geom_bar() +
    labs(x = "Count", y = NULL, title = old_names[2])
```

# `r old_names[3]`

```{r}
# how_often
ggplot(x, aes(y = fct_infreq(how_often))) +
    geom_bar() +
    labs(x = "Count", y = NULL, title = old_names[3])
```

# `r old_names[4]`

```{r}
# expertise
ggplot(x, aes(y = fct_infreq(expertise))) +
    geom_bar() +
    labs(x = "Count", y = NULL, title = old_names[4])
```

# `r old_names[5]`

```{r}
# JN: more manual cleaning could be useful here
major_topics = unlist(str_split(x$major_topic, pattern = ","))
major_topics = str_trim(major_topics)
major_topics = str_to_lower(major_topics)
major_topics_df = data.frame(major_topic = major_topics)
# major_topic
ggplot(major_topics_df, aes(y = fct_infreq(major_topic))) +
    geom_bar() +
    labs(x = "Count", y = NULL, title = old_names[5])
```

# `r old_names[6]`

```{r}
# JN: more manual cleaning could be useful here
important_tasks = unlist(str_split(x$important_tasks, pattern = ","))
important_tasks = str_trim(important_tasks)
important_tasks = str_to_lower(important_tasks)
important_tasks_df = data.frame(important_tasks = important_tasks)

# important_tasks
ggplot(important_tasks_df, aes(y = fct_infreq(important_tasks))) +
    geom_bar() +
    labs(x = "Count", y = NULL, title = old_names[6])
```

# `r old_names[7]`

```{r}
# data_model
ggplot(x, aes(y = fct_infreq(data_model))) +
    geom_bar() +
    labs(x = "Count", y = NULL, title = old_names[7])
```

# `r old_names[8]`

```{r}
# JN: more manual cleaning could be useful here
r_packages = unlist(str_split(x$r_packages, pattern = ","))
r_packages = unlist(str_split(r_packages, pattern = ";"))
r_packages = unlist(str_split(r_packages, pattern = "\\*"))
r_packages = str_trim(r_packages)
# r_packages = str_to_lower(r_packages)
r_packages_df = data.frame(r_package = r_packages)

# r_packages
ggplot(r_packages_df, aes(y = fct_infreq(r_package))) +
    geom_bar() +
    labs(x = "Count", y = NULL, title = old_names[8])
```

# `r old_names[9]`

```{r}
# own_package
ggplot(x, aes(y = fct_infreq(own_package))) +
    geom_bar() +
    labs(x = "Count", y = NULL, title = old_names[9])
```

# `r old_names[10]`

```{r}
# own_package_location
ggplot(drop_na(x, own_package_location),
       aes(y = fct_infreq(own_package_location))) +
    geom_bar() +
    labs(x = "Count", y = NULL, title = old_names[10])
```

# `r old_names[11]`

```{r}
missing_methods = na.omit(x$missing_methods)
missing_methods_df = data.frame(id = seq_along(missing_methods),
                         missing_method = missing_methods)
knitr::kable(missing_methods_df)
```

# `r old_names[12]`

```{r}
# usefulness
x$usefulness = factor(x$usefulness, levels = as.character(c(1:5)))
x$usefulness = fct_rev(fct_infreq(x$usefulness))
ggplot(x, aes(x = usefulness)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE) +
  labs(x = "Count", y = NULL, title = old_names[10])
```

# `r old_names[13]`

```{r}
comments = na.omit(x$comments)
comments_df = data.frame(id = seq_along(comments),
                         comment = comments)
knitr::kable(comments_df)
```

### 2020-10-10 R/survey_stats.Rmd
Starting to work on results of survey

--- 2020-09-04 survey_stats.Rmd
@@
 
-new_names = c("time", "position", "how_often", "expertise",
-              "major_topic", "important_tasks", "data_model",
-              "r_packages", "own_package", "own_package_location",
-              "missing_methods", "usefulness", "comments")
+new_names = c("time", "position", "usage_freq", "expertise", "major_topic",
+              "important_tasks", "data_model", "r_packages", "own_package",
+              "own_package_location", "missing_methods", "usefulness", "comments")
 
@@
 ```
+
+Number of answers: `r nrow(x)`.
 
@@
 # how_often
-ggplot(x, aes(y = fct_infreq(how_often))) +
+ggplot(x, aes(y = fct_infreq(usage_freq))) +
     geom_bar() +

### 2020-10-10 paper/paper.Rmd
Starting to work on results of survey

@@
 ---
+
+```{r setup R, include = FALSE}
+# load packages
+library(dplyr)
+library(forcats)
+library(ggplot2)
+library(googlesheets4)
+library(stringr)
+library(tidyr)
+
+answers <- read_sheet("https://docs.google.com/spreadsheets/d/1mV_wZWb4desjdzUWHALIB-uFRCauSTOGSvENsnkiAhI/edit#gid=1189808333")
+
+old_names <- names(answers)
+
+new_names <- c("time", "position", "usage_freq", "expertise", "major_topic",
+               "important_tasks", "data_model", "r_packages", "own_package",
+               "own_package_location", "missing_methods", "usefulness", "comments")
+
+names(answers) <-  new_names
+```
 
@@
 
-## Spatial regression
-<!-- MH: Do we really want to include this? Not really a landscape ecology topic but rather (geo) spatial stats I would argue -->
-<!-- JN: we could wait with this section to after receiving the results of the questionnaire -->
-
 ## Ecological analysis {#sec:ecological_analysis}
@@
 
+# Survey of R usage by landscape ecology community
+
+```{r participation, include = FALSE}
+digits <- 1
+
+# number of people that took part
+n_people <- nrow(answers)
+
+# current position
+current_position <- dplyr::group_by(answers, position) %>% 
+  dplyr::summarise(n_rel = n() / n_people * 100) %>% 
+  dplyr::mutate(n_rel = round(n_rel, digits = digits)) %>% 
+  dplyr::arrange(-n_rel)
+
+# how often do you use R?
+usage_freq <- dplyr::group_by(answers, usage_freq) %>% 
+  dplyr::summarise(n_rel = n() / n_people * 100) %>% 
+  dplyr::mutate(n_rel = round(n_rel, digits = digits)) %>% 
+  dplyr::arrange(-n_rel)
+
+# expertise
+expertise <- dplyr::group_by(answers, expertise) %>% 
+  dplyr::summarise(n_rel = n() / n_people * 100) %>% 
+  dplyr::mutate(n_rel = round(n_rel, digits = digits)) %>% 
+  dplyr::arrange(-n_rel)
+```
+
+To better understand how *R* is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the landscape ecology community. 
+In total, the survey was answered by `r n_people` people, of which the the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
+Other less frequent answers were "data scientists", "government employees" or "bachelor's and master's degree students".
+Most people used R either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%) and and about half of all participants described themselves as "advanced" user (`r expertise[[1, 2]]`%), while still `r expertise[[2, 2]]`% described themselves as "intermediate" users.
+
+```{r data_model_packages, include = FALSE}
+# data model
+data_model <- dplyr::group_by(answers, data_model) %>% 
+  dplyr::summarise(n_rel = n() / n_people * 100) %>% 
+  dplyr::mutate(n_rel = round(n_rel, digits = digits)) %>% 
+  dplyr::arrange(-n_rel)
+
+# split answers r packages
+r_packages <- unlist(str_split(answers$r_packages, pattern = ","))
+r_packages <- unlist(str_split(r_packages, pattern = ";"))
+r_packages <- unlist(str_split(r_packages, pattern = "\\*"))
+r_packages <- str_trim(r_packages)
+
+# clean some special cases
+r_packages[str_detect(r_packages, pattern = "landscape metrics")] <- "landscapemetrics"
+r_packages[str_detect(r_packages, pattern = "spectre*")] <- "spectre"
+r_packages[str_detect(r_packages, pattern = "map tools")] <- "maptools"
+r_packages[str_detect(r_packages, pattern = "gdal and many more")] <- "rgdal"
+r_packages[str_detect(r_packages, pattern = "I would like to know")] <- "NA"
+r_packages[str_detect(r_packages, pattern = "no")] <- "NA"
+r_packages[str_detect(r_packages, pattern = "currently none")] <- "NA"
+r_packages[!str_detect(r_packages, pattern = "")] <- "NA"
+
+# split last packages
+r_packages <- unlist(str_split(r_packages, pattern = " "))
+
+# remove NA answers
+r_packages <- r_packages[!str_detect(r_packages, pattern = "NA")]
+
+# make sure all are lower case
+r_packages <- str_to_lower(r_packages)
+
+# create data.frame
+r_packages_df <- tibble(r_package = r_packages) %>%
+  dplyr::group_by(r_package) %>% 
+  dplyr::summarise(n = dplyr::n()) %>% 
+  dplyr::mutate(n_rel = n / sum(n) * 100, 
+                r_package_reclass = dplyr::case_when(n >= 5 ~ r_package, 
+                                                     n < 5 ~ "others")) %>%
+  dplyr::group_by(r_package_reclass) %>% 
+  dplyr::summarise(n_rel = sum(n_rel)) %>% 
+  dplyr::mutate(label = paste0(round(n_rel, digits = 1), "%"), 
+                col = dplyr::case_when(r_package_reclass == "others" ~ 0, 
+                                       r_package_reclass != "others" ~ 1), 
+                col = factor(col)) %>% 
+  dplyr::arrange(-n_rel)
+
+# change ordering
+new_levels <- c("others", rev(r_packages_df$r_package_reclass[-1]))
+
+# reorder levels
+r_packages_df$r_package_reclass <- factor(r_packages_df$r_package_reclass, 
+                                          levels = new_levels)
+```
+Interestingly, most people used the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
+
+```{r used_package, echo = FALSE, fig.align = "center"}
+ggplot(data = r_packages_df) +
+  geom_bar(aes(y = r_package_reclass, x = n_rel, fill = col), stat = "identity") + 
+  geom_text(aes(y = r_package_reclass, x = n_rel, label = label, col = col), 
+            nudge_x = 2.5) + 
+  theme_classic() + 
+  scale_x_continuous(breaks = seq(0, 45, 5), limits = c(0, 45)) + 
+  scale_fill_manual(values = c("grey50", "black")) +
+  scale_colour_manual(values = c("grey50", "black")) +
+  guides(fill = FALSE, colour = FALSE) +
+  labs(x = "Relative value [%]", y = expression(paste(italic(R), " package")))
+```
+
 <!-- # Missing packages -->

### 2020-10-10 paper/paper.tex
Starting to work on results of survey

@@
 
-\hypertarget{spatial-regression}{%
-\subsection{Spatial regression}\label{spatial-regression}}
-
 \hypertarget{sec:ecological_analysis}{%
@@
 
+\hypertarget{survey-of-r-usage-by-landscape-ecology-community}{%
+\section{Survey of R usage by landscape ecology community}\label{survey-of-r-usage-by-landscape-ecology-community}}
+
+To better understand how \emph{R} is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the landscape ecology community.
+In total, the survey was answered by 94 people, of which the the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (12.8\%).
+Other less frequent answers were ``data scientists'', ``government employees'' or ``bachelor's and master's degree students''.
+Most people used R either ``daily'' (58.5\%) or a ``few times a week'' (34\%) and and about half of all participants described themselves as ``advanced'' user (48.9\%), while still 39.4\% described themselves as ``intermediate'' users.
+
+Interestingly, most people used the raster data model (70.2\%) in comparison to the vector data model (29.8\%).
+
+\begin{center}\includegraphics{paper_files/figure-latex/used_package-1} \end{center}
+
 \bibliographystyle{spphys}

### 2020-10-11 paper/paper.Rmd
Add further results of survery

@@
   \usepackage{ltablex}
-  \usepackage{lscape}
+  \usepackage{pdflscape}
   \usepackage{multirow}
+  \usepackage{placeins}
 output:
@@
     number_sections: yes
+    citation_package: default
 editor_options: 
@@
 library(googlesheets4)
+library(patchwork)
 library(stringr)
+library(tibble)
 library(tidyr)
 
-answers <- read_sheet("https://docs.google.com/spreadsheets/d/1mV_wZWb4desjdzUWHALIB-uFRCauSTOGSvENsnkiAhI/edit#gid=1189808333")
+# MAKE SURE TO RUN R/get_survey_data.R TO GET NEWST SURVERY RESULTS
 
-old_names <- names(answers)
+results <- readRDS("results.rds")
 
-new_names <- c("time", "position", "usage_freq", "expertise", "major_topic",
-               "important_tasks", "data_model", "r_packages", "own_package",
-               "own_package_location", "missing_methods", "usefulness", "comments")
+answers <- results$answers
 
-names(answers) <-  new_names
+old_names <- results$old_names
+
+new_names <- results$new_names
 ```
@@
 
-\begin{center}
   \begin{table}
@@
   \end{table}
-\end{center}
 
@@
   dplyr::arrange(-n_rel)
+
+# own package
+own_package <- dplyr::group_by(answers, own_package) %>% 
+  dplyr::summarise(n_rel = n() / n_people * 100) %>% 
+  dplyr::mutate(n_rel = round(n_rel, digits = digits)) %>% 
+  dplyr::arrange(-n_rel)
 ```
@@
 To better understand how *R* is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the landscape ecology community. 
-In total, the survey was answered by `r n_people` people, of which the the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
+In total, the survey was answered by `r n_people` participants, of which the the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
 Other less frequent answers were "data scientists", "government employees" or "bachelor's and master's degree students".
 Most people used R either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%) and and about half of all participants described themselves as "advanced" user (`r expertise[[1, 2]]`%), while still `r expertise[[2, 2]]`% described themselves as "intermediate" users.
+Related to this, about half of the participants either implemented their own package (`r own_package[[3, 2]]`%), or plan to do so in the future (`r own_package[[2, 2]]`%) and most of these packages are hosted on *GitHub* and/or *CRAN*.
 
-```{r data_model_packages, include = FALSE}
+```{r data-model-packages, include = FALSE}
 # data model
@@
                                           levels = new_levels)
-```
-Interestingly, most people used the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
 
-```{r used_package, echo = FALSE, fig.align = "center"}
-ggplot(data = r_packages_df) +
+ggplot_packages <- ggplot(data = r_packages_df) +
   geom_bar(aes(y = r_package_reclass, x = n_rel, fill = col), stat = "identity") + 
-  geom_text(aes(y = r_package_reclass, x = n_rel, label = label, col = col), 
-            nudge_x = 2.5) + 
-  theme_classic() + 
-  scale_x_continuous(breaks = seq(0, 45, 5), limits = c(0, 45)) + 
+  # geom_text(aes(y = r_package_reclass, x = n_rel, label = label, col = col), 
+  #           nudge_x = 2, size = 2.5) + 
+  theme_classic(base_size = 7.5) + 
+  scale_x_continuous(breaks = seq(0, 35, 5), limits = c(0, 35)) + 
   scale_fill_manual(values = c("grey50", "black")) +
@@
   guides(fill = FALSE, colour = FALSE) +
-  labs(x = "Relative value [%]", y = expression(paste(italic(R), " package")))
+  labs(x = "Number of package mentionings [%]", y = "List the three R packages related to landscape ecology you use the most!")
 ```
+
+Interestingly, most people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%). 
+This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-packages-tasks) A))
+When asked for the three most used packages, participants of the survey listed `r length(unique(r_packages))` packages in total.
+The *raster* packages was listed the most, followed by the *sf* package.
+Because both the *raster* and the *sf* are designed for basic and advanced data handling and processing of raster and vector data, respectively, it is not surprising that "(pre-)processing of spatial data" was the most important workflow task for the majority of the survey participants (Fig.
+\@ref(fig:fig-packages-tasks) B)).
+
+```{r tasks, include = FALSE}
+# get data 
+important_tasks <- unlist(str_split(answers$important_tasks, pattern = ","))
+
+# remove empty space at beginning
+important_tasks <- str_trim(important_tasks)
+
+# make sure all lower case
+important_tasks <- str_to_lower(important_tasks)
+
+# shorten labels
+important_tasks[important_tasks == "(pre-)processing of spatial data"] <- "(pre-)processing data"
+
+important_tasks[important_tasks == "spatial point pattern analysis"] <- "point pattern analysis"
+
+important_tasks[important_tasks == "programming individual based models"] <- "individual based modelling"
+
+important_tasks[important_tasks == "creating neutral landscape models"] <- "neutral landscape models"
+
+important_tasks[important_tasks == "quantifying landscape characteristics (e.g. landscape metrics)"] <- "quantifying landscape characteristics"
+
+# add line breaks
+important_tasks <- str_replace_all(important_tasks, pattern = " ", replacement = "\n")
+
+# create data.frame
+tasks_df <- tibble(tasks = important_tasks) %>%
+  dplyr::group_by(tasks) %>% 
+  dplyr::summarise(n = dplyr::n()) %>% 
+  dplyr::mutate(n_rel = n / sum(n) * 100, 
+                tasks_reclass = dplyr::case_when(n >= 5 ~ tasks, 
+                                                 n < 5 ~ "others")) %>%
+  dplyr::group_by(tasks_reclass) %>% 
+  dplyr::summarise(n_rel = sum(n_rel)) %>% 
+  dplyr::mutate(label = paste0(round(n_rel, digits = 1), "%"), 
+                col = dplyr::case_when(tasks_reclass == "others" ~ 0, 
+                                       tasks_reclass != "others" ~ 1), 
+                col = factor(col)) %>% 
+  dplyr::arrange(-n_rel)
+
+# change ordering
+new_levels <- c("others", rev(tasks_df$tasks_reclass[!tasks_df$tasks_reclass == "others"]))
+
+# reorder levels
+tasks_df$tasks_reclass <- factor(tasks_df$tasks_reclass, levels = new_levels)
+
+ggplot_tasks <- ggplot(data = tasks_df) +
+  geom_bar(aes(y = tasks_reclass, x = n_rel, fill = col), stat = "identity") + 
+  # geom_text(aes(y = tasks_reclass, x = n_rel, label = label, col = col), 
+  #           nudge_x = 1.25, size = 2.5) + 
+  theme_classic(base_size = 7.5) + 
+  scale_x_continuous(breaks = seq(0, 20, 5), limits = c(0, 20)) + 
+  scale_fill_manual(values = c("grey50", "black")) +
+  scale_colour_manual(values = c("grey50", "black")) +
+  guides(fill = FALSE, colour = FALSE) +
+  labs(x = "Number of task mentionings [%]", y = "Which are the most important tasks to your workflow?")
+```
+
+```{r fig-packages-tasks, echo = FALSE, fig.align="center", fig.cap = "A) Most used R packages and B) most important workflow tasks as listed by the participants of the survery.
+'others' includes all packages/tasks with less than five total mentions."}
+
+ggplot_packages + ggplot_tasks + plot_layout(ncol = 2, nrow = 1) + plot_annotation(tag_levels = "A", tag_suffix = ")")
+
+```
+
+```{r usefulness, include = FALSE}
+# create vector with classses
+
+classes <- c("very useful", "useful", "intermediate", 
+             "not useful", "not useful at all")
+
+# create data.frame
+usefulness_df <- dplyr::group_by(answers, usefulness) %>% 
+  dplyr::summarise(n = dplyr::n()) %>% 
+  dplyr::mutate(n_rel = n / sum(n) * 100) %>%
+  tibble::add_row(usefulness = 1, n = 0, n_rel = 0) %>% 
+  dplyr::arrange(-usefulness) %>% 
+  dplyr::mutate(label = paste0(round(n_rel, digits = 1), "%"), 
+                usefulness_class = classes,
+                usefulness_class = factor(usefulness_class,
+                                          levels = rev(classes))) %>% 
+  dplyr::arrange(-n_rel)
+
+good_eval <- dplyr::filter(usefulness_df, usefulness %in% c(5, 4)) %>% 
+  dplyr::pull(n_rel) %>% 
+  sum() %>% 
+  round(digits = 2)
+
+bad_eval <- dplyr::filter(usefulness_df, usefulness %in% c(3, 2, 1)) %>% 
+  dplyr::pull(n_rel) %>% 
+  sum() %>% 
+  round(digits = 2)
+```
+
+When asked "`r old_names[12]`", the vast majority of participants answered with either "very useful" or "useful" (`r good_eval`%) and only only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (`r bad_eval`%; Fig.
+\@ref(fig:fig-usefulness)).
+
+```{r fig-usefulness, echo = FALSE, fig.align="center", fig.cap = "Evaluation of the usefullnes o fR for landscape ecology by the particpants of the survey. Answers were possible on a 5-point scale."}
+ggplot(data = usefulness_df) +
+  geom_bar(aes(x = usefulness_class, y = n_rel), 
+           stat = "identity") + 
+  theme_classic(base_size = 7.5) + 
+  scale_y_continuous(breaks = seq(0, 60, 5), limits = c(0, 60)) + 
+  guides(fill = FALSE, colour = FALSE) +
+  labs(x = "How useful do you think is R currently for landscape ecology?", 
+       y = "Number of participants [%]")
+```
+
+\FloatBarrier
 

### 2020-10-11 paper/paper.tex
Add further results of survery

@@
 \usepackage{ltablex}
-\usepackage{lscape}
+\usepackage{pdflscape}
 \usepackage{multirow}
+\usepackage{placeins}
 
@@
 To better understand how \emph{R} is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the landscape ecology community.
-In total, the survey was answered by 94 people, of which the the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (12.8\%).
+In total, the survey was answered by 94 participants, of which the the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (12.8\%).
 Other less frequent answers were ``data scientists'', ``government employees'' or ``bachelor's and master's degree students''.
 Most people used R either ``daily'' (58.5\%) or a ``few times a week'' (34\%) and and about half of all participants described themselves as ``advanced'' user (48.9\%), while still 39.4\% described themselves as ``intermediate'' users.
+Related to this, about half of the participants either implemented their own package (20.2\%), or plan to do so in the future (22.3\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
 
-Interestingly, most people used the raster data model (70.2\%) in comparison to the vector data model (29.8\%).
+Interestingly, most people use the raster data model (70.2\%) in comparison to the vector data model (29.8\%).
+This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-packages-tasks} A))
+When asked for the three most used packages, participants of the survey listed 75 packages in total. The \emph{raster} packages was listed the most, followed by the \emph{sf} package.
+Because both the \emph{raster} and the \emph{sf} are designed for basic and advanced data handling and processing of raster and vector data, respectively, it is not surprising that ``(pre-)processing of spatial data'' was the most important workflow task for the majority of the survey participants (Fig.
+\ref{fig:fig-packages-tasks} B)).
 
-\begin{center}\includegraphics{paper_files/figure-latex/used_package-1} \end{center}
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/fig-packages-tasks-1}
+
+}
+
+\caption{A) Most used R packages and B) most important workflow tasks as listed by the participants of the survery.
+'others' includes all packages/tasks with less than five total mentions.}\label{fig:fig-packages-tasks}
+\end{figure}
+
+When asked ``How useful do you think is R currently for landscape ecology?'', the vast majority of participants answered with either ``very useful'' or ``useful'' (91.49\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (8.51\%; Fig.
+\ref{fig:fig-usefulness}).
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/fig-usefulness-1}
+
+}
+
+\caption{Evaluation of the usefullnes o fR for landscape ecology by the particpants of the survey. Answers were possible on a 5-point scale.}\label{fig:fig-usefulness}
+\end{figure}
+
+\FloatBarrier
 

### 2020-10-11 R/get_survery_data.R

@@ -1,0 +1,17 @@
+# Make sure survery data is up-to date
+
+library(googlesheets4)
+
+answers <- read_sheet("https://docs.google.com/spreadsheets/d/1mV_wZWb4desjdzUWHALIB-uFRCauSTOGSvENsnkiAhI/edit#gid=1189808333")
+
+old_names <- names(answers)
+
+new_names <- c("time", "position", "usage_freq", "expertise", "major_topic",
+               "important_tasks", "data_model", "r_packages", "own_package",
+               "own_package_location", "missing_methods", "usefulness", "comments")
+
+names(answers) <-  new_names
+
+results <- list(answers = answers, old_names = old_names, new_names = new_names)
+
+saveRDS(object = results, file = "paper/results.rds")

### 2020-10-15 paper/paper.Rmd
improves plots

@@
   theme_classic(base_size = 7.5) +
+  theme(plot.title.position = "plot") +
   scale_x_continuous(breaks = seq(0, 35, 5), limits = c(0, 35)) + 
@@
   guides(fill = FALSE, colour = FALSE) +
-  labs(x = "Number of package mentionings [%]", y = "List the three R packages related to landscape ecology you use the most!")
+  labs(x = "Number of package mentioned [%]", y = NULL,
+       title = "List the three R packages related to landscape\necology you use the most")
 ```
@@
   theme_classic(base_size = 7.5) +
+  theme(plot.title.position = "plot") +
   scale_x_continuous(breaks = seq(0, 20, 5), limits = c(0, 20)) + 
@@
   guides(fill = FALSE, colour = FALSE) +
-  labs(x = "Number of task mentionings [%]", y = "Which are the most important tasks to your workflow?")
+  labs(x = "Number of task mentioned [%]", y = NULL,
+       title = "Which are the most important tasks to your\nworkflow?")
 ```
@@
 
-```{r fig-usefulness, echo = FALSE, fig.align="center", fig.cap = "Evaluation of the usefullnes o fR for landscape ecology by the particpants of the survey. Answers were possible on a 5-point scale."}
+```{r fig-usefulness, echo = FALSE, fig.align="center", fig.cap = "Evaluation of the usefullnes of R for landscape ecology by the particpants of the survey. Answers were possible on a 5-point scale."}
 ggplot(data = usefulness_df) +
@@
            stat = "identity") + 
-  theme_classic(base_size = 7.5) + 
+  theme_classic(base_size = 10) +
+  theme(plot.title.position = "plot") +
   scale_y_continuous(breaks = seq(0, 60, 5), limits = c(0, 60)) + 
   guides(fill = FALSE, colour = FALSE) +
-  labs(x = "How useful do you think is R currently for landscape ecology?", 
+  labs(x = NULL,
+       title = "How useful do you think is R currently for landscape ecology?", 
        y = "Number of participants [%]")

### 2020-10-15 paper/paper.Rmd
cleans text

@@
 In total, the survey was answered by `r n_people` participants, of which the the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
-Other less frequent answers were "data scientists", "government employees" or "bachelor's and master's degree students".
-Most people used R either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%) and and about half of all participants described themselves as "advanced" user (`r expertise[[1, 2]]`%), while still `r expertise[[2, 2]]`% described themselves as "intermediate" users.
-Related to this, about half of the participants either implemented their own package (`r own_package[[3, 2]]`%), or plan to do so in the future (`r own_package[[2, 2]]`%) and most of these packages are hosted on *GitHub* and/or *CRAN*.
+Other, less frequent answers, were "data scientists", "government employees" or "bachelor's and master's degree students".
+Most people used R either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%) and about half of all participants described themselves as "advanced" user (`r expertise[[1, 2]]`%), while `r expertise[[2, 2]]`% described themselves as "intermediate" users.
+Related to this, about half of the participants either implemented their own package (`r own_package[[3, 2]]`%) or plan to do so in the future (`r own_package[[2, 2]]`%) and most of these packages are hosted on *GitHub* and/or *CRAN*.
 
@@
 Interestingly, most people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%). 
-This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-packages-tasks) A))
+This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-packages-tasks) A).
 When asked for the three most used packages, participants of the survey listed `r length(unique(r_packages))` packages in total.
-The *raster* packages was listed the most, followed by the *sf* package.
+The *raster* package was mentioned the most, followed by the *sf* package. 
 Because both the *raster* and the *sf* are designed for basic and advanced data handling and processing of raster and vector data, respectively, it is not surprising that "(pre-)processing of spatial data" was the most important workflow task for the majority of the survey participants (Fig.
-\@ref(fig:fig-packages-tasks) B)).
+\@ref(fig:fig-packages-tasks) B).
 

### 2020-10-15 paper/paper.tex
improves plots
cleans text

@@
 In total, the survey was answered by 94 participants, of which the the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (12.8\%).
-Other less frequent answers were ``data scientists'', ``government employees'' or ``bachelor's and master's degree students''.
-Most people used R either ``daily'' (58.5\%) or a ``few times a week'' (34\%) and and about half of all participants described themselves as ``advanced'' user (48.9\%), while still 39.4\% described themselves as ``intermediate'' users.
-Related to this, about half of the participants either implemented their own package (20.2\%), or plan to do so in the future (22.3\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
+Other, less frequent answers, were ``data scientists'', ``government employees'' or ``bachelor's and master's degree students''.
+Most people used R either ``daily'' (58.5\%) or a ``few times a week'' (34\%) and about half of all participants described themselves as ``advanced'' user (48.9\%), while 39.4\% described themselves as ``intermediate'' users.
+Related to this, about half of the participants either implemented their own package (20.2\%) or plan to do so in the future (22.3\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
 
 Interestingly, most people use the raster data model (70.2\%) in comparison to the vector data model (29.8\%).
-This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-packages-tasks} A))
-When asked for the three most used packages, participants of the survey listed 75 packages in total. The \emph{raster} packages was listed the most, followed by the \emph{sf} package.
+This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-packages-tasks} A).
+When asked for the three most used packages, participants of the survey listed 75 packages in total.
+The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
 Because both the \emph{raster} and the \emph{sf} are designed for basic and advanced data handling and processing of raster and vector data, respectively, it is not surprising that ``(pre-)processing of spatial data'' was the most important workflow task for the majority of the survey participants (Fig.
-\ref{fig:fig-packages-tasks} B)).
+\ref{fig:fig-packages-tasks} B).
 
@@
 
-\caption{Evaluation of the usefullnes o fR for landscape ecology by the particpants of the survey. Answers were possible on a 5-point scale.}\label{fig:fig-usefulness}
+\caption{Evaluation of the usefullnes of R for landscape ecology by the particpants of the survey. Answers were possible on a 5-point scale.}\label{fig:fig-usefulness}
 \end{figure}

### 2020-10-15 paper/paper.Rmd
updates ref

@@
 The use of landscape metrics was heavily facilitated by the introduction of the *FRAGSTATS* software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
-However, *FRAGSTATS* is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
+However, *FRAGSTATS* is not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
 To provide a truly open-source solution to calculate landscape metrics, recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed.
 The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment. 
-Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{Nowosad2019a}.
+Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
 <!-- JN: two notes: -->
@@
       \multirow{2}{*}{Landscape metrics} & \textit{landscapemetrics} & Calculates landscape metrics for categorical land cover maps & \cite{Hesselbarth2019a} \\
-      & \textit{belg} & Calculates the Boltzmann entropy of a landscape gradient & \cite{Nowosad2019a} \\
+      & \textit{belg} & Calculates the Boltzmann entropy of a landscape gradient & \cite{R-belgpaper} \\
       \hline

### 2020-10-15 paper/paper.tex
updates ref

@@
 The use of landscape metrics was heavily facilitated by the introduction of the \emph{FRAGSTATS} software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
-However, \emph{FRAGSTATS} is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
+However, \emph{FRAGSTATS} is not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
 To provide a truly open-source solution to calculate landscape metrics, recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was developed.
 The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the \emph{R} environment.
-Additionally, the \emph{belg} packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{Nowosad2019a}.
+Additionally, the \emph{belg} packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
 

### 2020-10-19 paper/paper.Rmd
Update results survey

@@
 new_names <- results$new_names
+
+base_size <-  7.5
 ```
@@
 
-To better understand how *R* is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the landscape ecology community. 
+To better understand how *R* is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the community. 
 In total, the survey was answered by `r n_people` participants, of which the the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
-Other, less frequent answers, were "data scientists", "government employees" or "bachelor's and master's degree students".
-Most people used R either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%) and about half of all participants described themselves as "advanced" user (`r expertise[[1, 2]]`%), while `r expertise[[2, 2]]`% described themselves as "intermediate" users.
+Other, less frequent answers, were "data scientists", "none of the above", "government employees" "Master's degree student" and "Bachelor's degree student" (in decreasing order).
+
+Most people use *R* either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%) and about half of all participants described themselves as "advanced" user (`r expertise[[1, 2]]`%), while `r expertise[[2, 2]]`% described themselves as "intermediate" users.
 Related to this, about half of the participants either implemented their own package (`r own_package[[3, 2]]`%) or plan to do so in the future (`r own_package[[2, 2]]`%) and most of these packages are hosted on *GitHub* and/or *CRAN*.
 
-```{r data-model-packages, include = FALSE}
+```{r research-topics, include = FALSE}
+
+major_topics <- unlist(str_split(answers$major_topic, pattern = ","))
+
+major_topics <- str_trim(major_topics)
+
+major_topics <- str_to_lower(major_topics)
+
+major_topics[major_topics == "-"] <- NA
+
+topics_df <- tibble(topics = major_topics) %>%
+  dplyr::group_by(topics) %>%
+  dplyr::summarise(n = dplyr::n()) %>%
+  dplyr::mutate(n_rel = n / sum(n) * 100,
+                topics_reclass = dplyr::case_when(n >= 5 ~ topics,
+                                                  n < 5 ~ "others")) %>%
+  dplyr::group_by(topics_reclass) %>%
+  dplyr::summarise(n_rel = sum(n_rel)) %>% 
+  dplyr::mutate(label = paste0(round(n_rel, digits = 1), "%"), 
+                col = dplyr::case_when(topics_reclass == "others" ~ 0, 
+                                       topics_reclass != "others" ~ 1), 
+                col = factor(col)) %>% 
+  dplyr::arrange(-n_rel)
+
+# change ordering
+new_levels <- c("others", rev(topics_df$topics_reclass[!topics_df$topics_reclass == "others"]))
+
+# reorder levels
+topics_df$topics_reclass <- factor(topics_df$topics_reclass, levels = new_levels)
+
+# major_topic
+ggplot_topics <-  ggplot(data = topics_df) +
+  geom_bar(aes(y = topics_reclass, x = n_rel, fill = col), stat = "identity") + 
+  # geom_text(aes(y = topics_reclass, x = n_rel, label = label, col = col),
+  #           nudge_x = 1.25, size = 2.5) +
+  theme_classic(base_size = base_size) +
+  theme(plot.title.position = "plot") +
+  scale_x_continuous(breaks = seq(0, 20, 5), limits = c(0, 20)) + 
+  scale_fill_manual(values = c("grey50", "black")) +
+  scale_colour_manual(values = c("grey50", "black")) +
+  guides(fill = FALSE, colour = FALSE) +
+  labs(x = "Number of task mentioned [%]", y = NULL,
+       title = "Which terms describe your\nmajor research topics the best?")
+
+```
+
+We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10% of participants included "biodiversity", followed by "land use management", "landscape connectivity" and "nature conservation" (Fig.
+\@ref(fig:fig-survey) A)).
+
+```{r tasks, include = FALSE}
+# get data 
+important_tasks <- unlist(str_split(answers$important_tasks, pattern = ","))
+
+# remove empty space at beginning
+important_tasks <- str_trim(important_tasks)
+
+# make sure all lower case
+important_tasks <- str_to_lower(important_tasks)
+
+# shorten labels
+important_tasks[important_tasks == "(pre-)processing of spatial data"] <- "(pre-)processing data"
+
+important_tasks[important_tasks == "spatial point pattern analysis"] <- "point pattern analysis"
+
+important_tasks[important_tasks == "programming individual based models"] <- "individual based modelling"
+
+important_tasks[important_tasks == "creating neutral landscape models"] <- "neutral landscape models"
+
+important_tasks[important_tasks == "quantifying landscape characteristics (e.g. landscape metrics)"] <- "quantifying landscape characteristics"
+
+# # add line breaks
+# important_tasks <- str_replace_all(important_tasks, pattern = " ", replacement = "\n")
+
+# create data.frame
+tasks_df <- tibble(tasks = important_tasks) %>%
+  dplyr::group_by(tasks) %>% 
+  dplyr::summarise(n = dplyr::n()) %>% 
+  dplyr::mutate(n_rel = n / sum(n) * 100, 
+                tasks_reclass = dplyr::case_when(n >= 5 ~ tasks, 
+                                                 n < 5 ~ "others")) %>%
+  dplyr::group_by(tasks_reclass) %>% 
+  dplyr::summarise(n_rel = sum(n_rel)) %>% 
+  dplyr::mutate(label = paste0(round(n_rel, digits = 1), "%"), 
+                col = dplyr::case_when(tasks_reclass == "others" ~ 0, 
+                                       tasks_reclass != "others" ~ 1), 
+                col = factor(col)) %>% 
+  dplyr::arrange(-n_rel)
+
+# change ordering
+new_levels <- c("others", rev(tasks_df$tasks_reclass[!tasks_df$tasks_reclass == "others"]))
+
+# reorder levels
+tasks_df$tasks_reclass <- factor(tasks_df$tasks_reclass, levels = new_levels)
+
+ggplot_tasks <- ggplot(data = tasks_df) +
+  geom_bar(aes(y = tasks_reclass, x = n_rel, fill = col), stat = "identity") + 
+  # geom_text(aes(y = tasks_reclass, x = n_rel, label = label, col = col),
+  #           nudge_x = 1.25, size = 2.5) +
+  theme_classic(base_size = base_size) +
+  theme(plot.title.position = "plot") +
+  scale_x_continuous(breaks = seq(0, 20, 5), limits = c(0, 20)) + 
+  scale_fill_manual(values = c("grey50", "black")) +
+  scale_colour_manual(values = c("grey50", "black")) +
+  guides(fill = FALSE, colour = FALSE) +
+  labs(x = "Number of task mentioned [%]", y = NULL,
+       title = "Which are the most important\ntasks to your workflow?")
+```
+
+Next, we were interested in the most important tasks to the workflow of the participants.
+Not surprisingly, "(pre-)processing of data", "spatial statistics" and "creating maps" were the most selected options (Fig.
+\@ref(fig:fig-survey) B)).
+Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
+
+```{r data-model, include = FALSE}
 # data model
@@
   dplyr::arrange(-n_rel)
+
+```
+
+````{r packages, include =FALSE}
 
@@
   #           nudge_x = 2, size = 2.5) + 
-  theme_classic(base_size = 7.5) +
+  theme_classic(base_size = base_size) +
   theme(plot.title.position = "plot") +
@@
   labs(x = "Number of package mentioned [%]", y = NULL,
-       title = "List the three R packages related to landscape\necology you use the most")
+       title = "List the three R packages related\nto landscape ecology you use the most")
 ```
@@
 Interestingly, most people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
-This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-packages-tasks) A).
+This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-survey) C)).
 When asked for the three most used packages, participants of the survey listed `r length(unique(r_packages))` packages in total.
 The *raster* package was mentioned the most, followed by the *sf* package. 
-Because both the *raster* and the *sf* are designed for basic and advanced data handling and processing of raster and vector data, respectively, it is not surprising that "(pre-)processing of spatial data" was the most important workflow task for the majority of the survey participants (Fig.
-\@ref(fig:fig-packages-tasks) B).
-
-```{r tasks, include = FALSE}
-# get data 
-important_tasks <- unlist(str_split(answers$important_tasks, pattern = ","))
-
-# remove empty space at beginning
-important_tasks <- str_trim(important_tasks)
-
-# make sure all lower case
-important_tasks <- str_to_lower(important_tasks)
-
-# shorten labels
-important_tasks[important_tasks == "(pre-)processing of spatial data"] <- "(pre-)processing data"
-
-important_tasks[important_tasks == "spatial point pattern analysis"] <- "point pattern analysis"
-
-important_tasks[important_tasks == "programming individual based models"] <- "individual based modelling"
-
-important_tasks[important_tasks == "creating neutral landscape models"] <- "neutral landscape models"
-
-important_tasks[important_tasks == "quantifying landscape characteristics (e.g. landscape metrics)"] <- "quantifying landscape characteristics"
-
-# add line breaks
-important_tasks <- str_replace_all(important_tasks, pattern = " ", replacement = "\n")
-
-# create data.frame
-tasks_df <- tibble(tasks = important_tasks) %>%
-  dplyr::group_by(tasks) %>% 
-  dplyr::summarise(n = dplyr::n()) %>% 
-  dplyr::mutate(n_rel = n / sum(n) * 100, 
-                tasks_reclass = dplyr::case_when(n >= 5 ~ tasks, 
-                                                 n < 5 ~ "others")) %>%
-  dplyr::group_by(tasks_reclass) %>% 
-  dplyr::summarise(n_rel = sum(n_rel)) %>% 
-  dplyr::mutate(label = paste0(round(n_rel, digits = 1), "%"), 
-                col = dplyr::case_when(tasks_reclass == "others" ~ 0, 
-                                       tasks_reclass != "others" ~ 1), 
-                col = factor(col)) %>% 
-  dplyr::arrange(-n_rel)
-
-# change ordering
-new_levels <- c("others", rev(tasks_df$tasks_reclass[!tasks_df$tasks_reclass == "others"]))
-
-# reorder levels
-tasks_df$tasks_reclass <- factor(tasks_df$tasks_reclass, levels = new_levels)
-
-ggplot_tasks <- ggplot(data = tasks_df) +
-  geom_bar(aes(y = tasks_reclass, x = n_rel, fill = col), stat = "identity") + 
-  # geom_text(aes(y = tasks_reclass, x = n_rel, label = label, col = col), 
-  #           nudge_x = 1.25, size = 2.5) + 
-  theme_classic(base_size = 7.5) +
-  theme(plot.title.position = "plot") +
-  scale_x_continuous(breaks = seq(0, 20, 5), limits = c(0, 20)) + 
-  scale_fill_manual(values = c("grey50", "black")) +
-  scale_colour_manual(values = c("grey50", "black")) +
-  guides(fill = FALSE, colour = FALSE) +
-  labs(x = "Number of task mentioned [%]", y = NULL,
-       title = "Which are the most important tasks to your\nworkflow?")
-```
-
-```{r fig-packages-tasks, echo = FALSE, fig.align="center", fig.cap = "A) Most used R packages and B) most important workflow tasks as listed by the participants of the survery.
-'others' includes all packages/tasks with less than five total mentions."}
-
-ggplot_packages + ggplot_tasks + plot_layout(ncol = 2, nrow = 1) + plot_annotation(tag_levels = "A", tag_suffix = ")")
-
-```
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey) C).
+Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; 32.6%).
 
@@
   round(digits = 2)
-```
 
-When asked "`r old_names[12]`", the vast majority of participants answered with either "very useful" or "useful" (`r good_eval`%) and only only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (`r bad_eval`%; Fig.
-\@ref(fig:fig-usefulness)).
-
-```{r fig-usefulness, echo = FALSE, fig.align="center", fig.cap = "Evaluation of the usefullnes of R for landscape ecology by the particpants of the survey. Answers were possible on a 5-point scale."}
-ggplot(data = usefulness_df) +
+ggplot_usefulness <- ggplot(data = usefulness_df) +
   geom_bar(aes(x = usefulness_class, y = n_rel), 
@@
 
+Lastly, when asked "`r old_names[12]`", the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
+\@ref(fig:fig-survey) D)).
+
+```{r fig-survey, echo = FALSE, fig.align="center", fig.cap = "Results of the online survey about open-source software tools in R for landscape ecology.
+Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology.
+'others' includes all answers with less than five total mentions."}
+
+(ggplot_topics | ggplot_tasks | ggplot_packages ) / ggplot_usefulness + 
+  plot_layout(heights = c(2, 3)) +
+  plot_annotation(tag_levels = "A", tag_suffix = ")")
+
+```
+
 \FloatBarrier
 
-<!-- # Missing packages -->
+# Missing packages
+
+```{r comments, include = FALSE}
+
+missing_methods <- answers$missing_methods[!is.na(answers$missing_methods)]
+
+missing_methods <- str_to_lower(missing_methods)
+
+# no answer, no missing packages
+id_na <- c("none", "na", "no", "not sure", "not sure right now", "unsure",
+           "i have no additional needs for my current workflow", 
+           "none that i can think of", "i would like to know",
+           "i don't know, sorry!", "r havent fail me yet", 
+           "i'm not missing anything at present.",
+           "have not yet identified any.", "no package missing",
+           "i am currently missing no methods",  
+           "i still have to fully understand what is there", 
+           "não posso responder", "not enough overview to answer this",
+           ";", ".", "?")
+
+length(id_na) / nrow(answers) * 100
+
+missing_methods <- missing_methods[!missing_methods %in% id_na]
+
+# classify answers as good as possible
+
+methods_performance <- missing_methods[c(2, 4, 8, 9, 10, 21, 23, 27, 33, 41, 45, 
+                                         48, 56)]
+perc_performance <- round(length(methods_performance) / nrow(answers) * 100, 1)
+
+methods_viz <- missing_methods[c(3, 7, 13, 22, 29, 33, 41, 57, 58)]
+perc_viz <- round(length(methods_viz) / nrow(answers) * 100, 1)
+
+methods_connectivity <- missing_methods[c(12, 17, 26, 31, 35, 36, 40, 51)]
+perc_connectivity <- round(length(methods_connectivity) / nrow(answers) * 100, 1)
+
+methods_characterisation <- missing_methods[c(7, 20, 24, 39, 42, 47, 59)]
+perc_characterisation <- round(length(methods_characterisation) / nrow(answers) * 100, 1)
+
+methods_tutorial <- missing_methods[c(1, 14, 44, 52, 53)]
+
+methods_datahandling <- missing_methods[c(5, 11, 22,  60, 63)]
+
+methods_gis <- missing_methods[c(43, 50, 55)]
+
+methods_various <- missing_methods[c(6, 11, 15, 16, 18, 19, 22, 25, 28, 30, 32, 
+                                     34, 37, 38, 46, 49, 54, 61, 62, 63, 64)]
+
+```
+
+The survey also included a section in which participants could list methods and tools that are currently missing in *R* and answers to this question were very diverse.
+
+Overall, `r round(length(id_na) / nrow(answers) * 100, 1)`% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
+There were two most common topics across the answers of the participants . 
+Firstly, many participants (`r perc_performance`%) wished for a better computational performance of *R* in terms of speed and required RAM for larger data.
+Secondly, many participants (`r perc_viz`%) are currently missing advanced and easy-to-apply methods for creating high-quality maps or other visualization-related functionality.
+Fewer participants were missing tools for various connectivity-related issues (`r perc_viz`%) and characterization of landscape characteristics (`r perc_characterisation`%).
 

### 2020-10-19 paper/paper.tex
Update results survey

@@
 
-To better understand how \emph{R} is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the landscape ecology community.
-In total, the survey was answered by 94 participants, of which the the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (12.8\%).
-Other, less frequent answers, were ``data scientists'', ``government employees'' or ``bachelor's and master's degree students''.
-Most people used R either ``daily'' (58.5\%) or a ``few times a week'' (34\%) and about half of all participants described themselves as ``advanced'' user (48.9\%), while 39.4\% described themselves as ``intermediate'' users.
-Related to this, about half of the participants either implemented their own package (20.2\%) or plan to do so in the future (22.3\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
+To better understand how \emph{R} is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the community.
+In total, the survey was answered by 96 participants, of which the the majority were either ``PhD students'' (33.3\%), followed by ``Post-Docs'' (27.1\%) and ``Professors'' (12.5\%).
+Other, less frequent answers, were ``data scientists'', ``none of the above'', ``government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
-Interestingly, most people use the raster data model (70.2\%) in comparison to the vector data model (29.8\%).
-This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-packages-tasks} A).
+Most people use \emph{R} either ``daily'' (57.3\%) or a ``few times a week'' (35.4\%) and about half of all participants described themselves as ``advanced'' user (47.9\%), while 40.6\% described themselves as ``intermediate'' users.
+Related to this, about half of the participants either implemented their own package (20.8\%) or plan to do so in the future (21.9\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
+
+We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'' and ``nature conservation'' (Fig.
+\ref{fig:fig-survey} A)).
+
+Next, we were interested in the most important tasks to the workflow of the participants.
+Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'' and ``creating maps'' were the most selected options (Fig.
+\ref{fig:fig-survey} B)).
+Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
+
+Interestingly, most people use the raster data model (70.8\%) in comparison to the vector data model (29.2\%).
+This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey} C)).
 When asked for the three most used packages, participants of the survey listed 75 packages in total.
 The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
-Because both the \emph{raster} and the \emph{sf} are designed for basic and advanced data handling and processing of raster and vector data, respectively, it is not surprising that ``(pre-)processing of spatial data'' was the most important workflow task for the majority of the survey participants (Fig.
-\ref{fig:fig-packages-tasks} B).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey} C).
+Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 32.6\%).
+
+Lastly, when asked ``How useful do you think is R currently for landscape ecology?'', the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.67\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.33\%; Fig.
+\ref{fig:fig-survey} D)).
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-packages-tasks-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-survey-1}
 
@@
 
-\caption{A) Most used R packages and B) most important workflow tasks as listed by the participants of the survery.
-'others' includes all packages/tasks with less than five total mentions.}\label{fig:fig-packages-tasks}
-\end{figure}
-
-When asked ``How useful do you think is R currently for landscape ecology?'', the vast majority of participants answered with either ``very useful'' or ``useful'' (91.49\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (8.51\%; Fig.
-\ref{fig:fig-usefulness}).
-
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/fig-usefulness-1}
-
-}
-
-\caption{Evaluation of the usefullnes of R for landscape ecology by the particpants of the survey. Answers were possible on a 5-point scale.}\label{fig:fig-usefulness}
+\caption{Results of the online survey about open-source software tools in R for landscape ecology.
+Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology.
+'others' includes all answers with less than five total mentions.}\label{fig:fig-survey}
 \end{figure}
@@
 \FloatBarrier
+
+\hypertarget{missing-packages}{%
+\section{Missing packages}\label{missing-packages}}
+
+The survey also included a section in which participants could list methods and tools that are currently missing in \emph{R} and answers to this question were very diverse.
+Overall, 21.9\% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
+There were two most common topics across the answers of the participants .
+Firstly, many participants (13.5\%) wished for a better computational performance of \emph{R} in terms of speed and required RAM for larger data.
+Secondly, many participants (9.4\%) are currently missing advanced and easy-to-apply methods for creating high-quality maps or other visualization-related functionality.
+Fewer participants were missing tools for various connectivity-related issues (9.4\%) and characterization of landscape characteristics (7.3\%).
 

### 2020-10-20 paper/paper.Rmd
adds motif bib and updates the packages table

@@
 Packages are sorted by their major application topic.
-Only packages especially for spatial data and landscape ecology are included.} \\
+Only packages focussed on spatial data and landscape ecology are included.} \\
       \hline
@@
       & \textit{sp} & Methods for mainly vector data & \cite{Pebesma2005,Bivand2013} \\
-      & \textit{sf} & Methods for vector data as simple features & \cite{Pebesma2018} \\
+      & \textit{sf} & Successor of the sp package. Methods for vector data as simple features & \cite{Pebesma2018} \\
       & \textit{landscapetools} & Utility functions for landscape analysis & \cite{Sciaini2018} \\
       \hline
-      \multirow{3}{*}{Creating maps} & \textit{tmap} & xxx & xxx \\
-      & \textit{mapview} & xxx & xxx \\
-      & \textit{leaflet} & xxx & xxx \\
+      \multirow{3}{*}{Creating maps} & \textit{tmap} & Plotting system for static and interactive maps & \cite{Tennekes2018} \\
+      & \textit{mapview} & Quick, interactive viewing of spatial data & \cite{R-mapview} \\
+      & \textit{leaflet} & Creating interactive web maps & \cite{R-leaflet} \\
       \hline
-      \multirow{2}{*}{Landscape metrics} & \textit{landscapemetrics} & Calculates landscape metrics for categorical land cover maps & \cite{Hesselbarth2019a} \\
+      \multirow{3}{*}{}Quantifying landscape characteristics & \textit{landscapemetrics} & Calculates landscape metrics for categorical land cover maps & \cite{Hesselbarth2019a} \\
       & \textit{belg} & Calculates the Boltzmann entropy of a landscape gradient & \cite{R-belgpaper} \\
+      & \textit{motif} & Enables several type of spatial analysis such as search, change detection, and clustering to be performed on categorical maps  & \cite{R-motif} \\
       \hline

### 2020-10-20 paper/paper.Rmd
adds motif paragraph

@@
 Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
-<!-- JN: two notes: -->
-<!-- 1. I am working on a paper supporting the belg package - I plan to submit it in less than a month, so hopefully we could update the reference here -->
-<!-- 2.
-I also submitted a paper yesterday (2020-07-17) explaining a new package motif, which is aimed on calculations based on spatial signatures.
-I plan to add a paragraph about this package someplace in this paper.
--->
-<!-- MH: Sounds great -->
+
+Most of landscape metrics are a single number depicting specific characteristics of a local landscape.
+Another possibility is to derive spatial signatures -  a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
+Spatial signatures calculated for many landscapes can be compared using one of the existing distance measures.
+This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
+Methods for pattern-based spatial analysis are implemented in the *motif* package \cite{R-motif}.
 

### 2020-10-20 paper/paper.Rmd
adds new comments

@@
 Overall, `r round(length(id_na) / nrow(answers) * 100, 1)`% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
-There were two most common topics across the answers of the participants . 
+There were two most common topics across the answers of the participants. 
 Firstly, many participants (`r perc_performance`%) wished for a better computational performance of *R* in terms of speed and required RAM for larger data.
@@
 <!-- missing packages: data sources, such as global land cover, worldclim 2.0, etc. -->
+
+<!--JN: I think we should answer to some of the comments in the missing packages section: point people to some packages (if possible), mention which of the missing tools are not going to be implemented, and also forward people to methods and tools on how to improve some problems (e.g.
+speed and required RAM for larger data -> use of Rcpp, use external servers, or read data in chunks, etc.)-->

### 2020-10-20 paper/paper.tex
adds new comments

@@
 
+Most of landscape metrics are a single number depicting specific characteristics of a local landscape.
+Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
+Spatial signatures calculated for many landscapes can be compared using one of the existing distance measures.
+This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
+Methods for pattern-based spatial analysis are implemented in the \emph{motif} package \cite{R-motif}.
+
 \hypertarget{sec:SDM}{%
@@
 The survey also included a section in which participants could list methods and tools that are currently missing in \emph{R} and answers to this question were very diverse.
+
 Overall, 21.9\% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.

### 2020-12-12 paper/paper.Rmd
:pencil: Work on introduction and overview of R packages

@@
   email: nowosad.jakub@gmail.com
-- name: Author 3
-  address: Adress author 3
-  email: E-Mail author 3
 - name: ...
@@
   chunk_output_type: console
+  markdown: 
+    wrap: sentence
 ---
@@
 library(googlesheets4)
+library(knitr)
 library(patchwork)
@@
 
-results <- readRDS("results.rds")
+results <- readRDS("data/results.rds")
 
@@
 
+<!-- MH: We really need to make sure, we introduce all technical terms well enough throughout the paper -->
+
 ## A short introduction to landscape ecology {#sec:landscape_ecology}
 
-<!-- MH: Not super happy with the list i) - iv) -->
-
 Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
-In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa) and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
+In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
 
@@
 Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}.
-to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}. 
+To this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}.
 Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}.
-Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze and visualize spatial data.
+Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze, and visualize spatial data.
 
-## Open-source software and R {#sec:open_source}
+## Open-source software and *R* {#sec:open_source}
 
 Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
-Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
-Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
-Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. 
-Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}. 
-Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to many quite expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+Open-source software includes all software which is released under licenses that allow to freely use, modify and distribute the software \cite{St.Laurent2008}.
+Open-source software development has many advantages, such as fast innovation, transparency, and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
+Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
+Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
+Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
 
-One example of a successful open-source project is the *R* programming language and its extensions (also called packages) \cite{RCoreTeam2019}. 
+One successful example of an open-source project is the *R* programming language and its extensions (also called packages) \cite{RCoreTeam2019}.
 Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
-This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total.
-Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even much higher.
+This is documented by the growing *CRAN Task Views: Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total.
+Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even higher.
 The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
 
-Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language.
 The growing body of *R* packages related to spatial data processing and analyzing results in a high capability of the *R* programming language for landscape ecology.
-Because additionally to the handling of spatial data, also further task such as statistical modeling, the preparation of publication-ready figures and even the preparation of complete reports using *RMarkdown* \cite{Xie2018} can be done in the same working environment, the *R* programming language can be a very powerful tool.
-Thus, in the first part we are give a comprehensive overview over existing *R* packages for landscape ecology. 
-In the second part, we identify topics for which *R* packages are currently missing to the best of our knowledge. 
-For more general overviews see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on software implemented in the *R* programming language.
+For more general overviews of open-source software for landscape ecology see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Because additionally to handling spatial data, also further tasks such as statistical modeling, preparation of publication-ready figures and even preparation of complete reports can be done in the same working environment, the *R* programming language can be a very powerful tool.
+Thus, in the first part we present an overview over existing *R* packages for landscape ecology.
+In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are currently missing for landscape ecology.
 
@@
 
-There are two major platforms to host *R* packages and make them easily accessible to potential users.
-Firstly, hosting the package on *GitHub* allows users to install it with only one line of code. 
-Additionally, hosting the package on *GitHub* provides many useful tools to communicate between developers and users, such as bug reports \cite{Wickham2015}.
-Secondly, the *Comprehensive R Archive Network* (*CRAN*) provides large visibility to the community, ease of installation and a certain quality standard \cite{Wickham2015}.
-However, at the same time, releasing a *R* package on *CRAN* requires more initial work for the developers compared to *GitHub*.
-Since most *R* packages are developed and maintained by the community, this is a good example of how open-source software development can facilitate innovation, reproducibility and reuse of code, especially since now many *R* packages depend on each other (i.e.
-package *x* requires and uses code from package *y*) \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+Since most *R* packages are developed and maintained by the community, this shows how open-source software development can facilitate innovation, reproducibility and reuse of code.
+There are two major online platforms to host *R* packages and make them accessible to potential users.
+Firstly, *GitHub* allows to host source code under version control and users can install packages with one line using the *remotes* \cite{Hester2020} package.
+Additionally, hosting the package on *GitHub* provides many useful features to collaborate and communicate between developers as well as users \cite{Wickham2015}, or integrated unit testing.
+Secondly, the *Comprehensive R Archive Network* (*CRAN*) provides a large visibility to the community, ease of installation and a technical quality standard, including checks for common problems on all major operating systems.
+\cite{Wickham2015}.
+For developers, the guaranteed technical quality standard on *CRAN* requires more initial work for compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
+Additionally, the technical quality standard on *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e.
+package *x* requires and uses code from package *y* \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
 For more information about *R* package development, see \cite{Wickham2015}.
 
-<!-- MH: Maybe we need to think about a table with the topics and the packages as a nice overview. Additionally, a figure where we show a typical workflow would be nice maybe -->
-<!-- JN: can you think of one/a few typical workflows? -->
+```{r fig-workflow, echo = FALSE, out.width = "100%", out.height = "65%", fig.align="center", fig.cap = "Exemplary worklow of spatial data analyses for landscape ecology using the R programming language.
+For all major tasks, a few example *R* packages are listed.
+Of course, many more packages for the specific tasks exist."}
+include_graphics("data/Workflow.pdf")
+```
 
 <!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->
+
 <!-- JN: if you think this package (or similar) are still useful, maybe we can mention them somewhere is the first part of the paper with annotation about their state, and then also write something more about issues with archived packages in the second part (what's missing in R)...
@@
 
+<!-- MH: I would argue to only include packages that are actively maintained.
+It's already such a huge amount and I guess there is a reason why the packages were archived.
+Doesnt speak for their popularity -->
+
 ## Spatial data representations {#sec:spatrep}
 
-Most of the existing spatial data belong to one of two data models: spatial raster and spatial vectors.
+<!-- MH: Make sure all libraries etc. have reference -->
+
+While *R* has several build-in data structures, including vectors, matrices or lists, the base *R* programming language has no internal support for reading, processing, or visualizing spatial data.
+However, because there is a notable interest in spatial data analysis, support for spatial data is provided by many *R* packages (\cite{Lovelace2019}, page 10).
+Most spatial data belong to one of two contrasting data models, namely spatial raster and spatial vectors, and both data models have several implementations in the *R* language.
+Importantly, basically all *R* packages for spatial data use the external *GDAL* and *PROJ* libraries, which allow for e.g.
+reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Additionally, *R*, *GDAL* and *PROJ* allow to convert between data models and specific implementations, which can be useful if given methods only exist for a specific data model or implementation.
+
 In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
 The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
-The vector data model consists of two main elements, geometries, such as points, lines, polygons, and attributes.
-Each geometry is connected to a row in an attribute table.
-<!-- ... -->
-A few hundred of different spatial data file formats were developed to store data in these formats.
+The probably most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
+A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
+Alternatively, the *stars* package can be used to read and process raster data with a focus on spatial-temporal data cubes \cite{Pebesma2019}.
 
-While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
-However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page 10).
-Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.
-<!-- are pros/cons of that needed? -->
-<!-- info about GDAL/PROJ? -->
+The vector data model consists of two main elements i) geometries, such as points, lines, polygons, and ii) attributes and each geometry is connected to a value in an attribute table.
+In many cases, this data model allows a more natural representation of landscape features, however, with the cost of higher computational demands <!--(add reference)-->.
+The *sp* package was the standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
+As of 2020, more than 500 *R* packages directly depend or imports *sp*.
+However, more recently *sp* will only be maintained (and not actively developed any more), and the recommended successor is the improved *sf* package \cite{Pebesma2018}.
 
-The most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
-A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
-Alternatively, *stars* package can be used to read and process raster data \cite{Pebesma2019}.
-It is interconnected with the *sf* package, and beside raster support, it provides tools for spatial data cubes.
+<!-- JN: terra new vector representations -->
 
-The *sp* package was a standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
-As of 2020, more than 500 R packages directly depends or imports *sp*.
-However, it is now in a maintenance mode only with a recommended successor - the *sf* package.
-<!-- terra new vector representations -->
-
-Importantly, all of the above packages use the external libraries GDAL and PROJ, which allow for reading and writing of hundreds of spatial data formats and coordinates transformation.
-They also have dedicated methods or functions to convert between representations, for example between *sp* and *sf* objects or *raster* and *stars* objects.
-This can be useful when a given function expects different data representation than currently existing.
+<!-- MH: I probably would not add terra here.
+As far as I see, the vector support in terra is basically just support to combine vector with raster data.
+The main focuse of the package is raster data though -->
 
@@
 
-Nowadays, spatial data for many spatial scales is available from abundance of sources.
-Some of them require registration or having appropriate credentials and allows downloading only from the web interface.
-However, many of the sources made their data publicly available either as a direct download or through an API connection.
-This is used by some R packages allowing for downloading various environmental spatial dataset.
+Nowadays, spatial data for many spatial scales is available from an abundance of online accessible sources.
+A lot of this data is publicly available, either as a direct download or through an API connection, and this can be used by several packages to download the spatial data directly into a *R* session.
+Since public available data is becoming more and more prominent, so are *R* packages to accesses them.
+Packages include the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA, the *elevatr* package to access raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}, the *rgbif* to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, *rnaturalearth* to access the Natural Earth database \cite{South2017}, or *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}.
+Furthermore, the `get_data()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim v1.4, while the new in-development package *geodata* <!--(add reference)--> provides an access to data from WorldClim v2.
+Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
 
-A plethora of land cover products on global, continental, and country scales has been developed, however, most of them cannot be directly downloaded from R. 
-The exception is the National Land Cover Database (NLCD) data for the USA, which can be accessed through the *FedData* package \cite{R-FedData}.
-The *elevatr* package provides a synthesis raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}.
-It uses different elevation products for a given area and selected level of details (zoom/resolution).
-<!-- how to cite (or maybe we should just add a link to?) - https://github.com/tilezen/joerd/blob/master/docs/data-sources.md -->
-The *rgbif* gives access to data from the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}.
-The `get_data()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim version 1.4, while the new in-development package *geodata*<!-- https://github.com/rspatial/geodata/blob/master/R/worldclim.R --> provides an access to WorldClim version 2 data.
-Additionally, the *rgee package* \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
-<!-- it is worth to search for more as there are a lot of changes going on -->
+<!-- JN: how to cite (or maybe we should just add a link to?) - https://github.com/tilezen/joerd/blob/master/docs/data-sources.md -->
 
-<!--JN: Max, do you want to add something to this list?-->
+<!-- MH: I dont really get it. For which package do we need this link? -->
 
@@
 
-As we mentioned in the previous sections, gathered or obtained data can be represented by different data types and stored in various file formats.
-<!-- and read into R using one of the packages from Section \@ref(sec:spatrep). -->
-However, several potential obstacles to applying this data for ecological analysis still exist.
+Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps <!-- (add reference) -->.
+This is also referred to as the projection of the data and is often a first barrier in spatial data analysis.
+It is not only required to have all of the used data in the same projection, but also selection of a proper projection is crucial.
+This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion and different CRS are optimized for different regions of the world and scales <!-- (add reference) -->.
+Coordinates in spatial data can represent one of many coordinate reference systems.
+Two main groups of CRS exist, namely geographical and projected, each having many members.
+In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
+However, in most landscape ecology studies, projected CRSs, which use some measurement units (e.g., meters), should be utilized.
+Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
+Tools to find an appropriate CRS for a certain region can be found at <https://spatialreference.org>, <https://epsg.org>, or <http://epsg.io>.
+In used data have different CRS, all packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations.
 
-The main issue for many landscape ecology studies is related to projections of the used data.
-It is not only required to have all of the used datasets in the same projection, but also selection of a proper projection is crucial.
-Coordinates in spatial data can represent one of many coordinate reference systems.
-Two main groups of coordinate reference systems (CRS) exist, namely geographical and projected, each having many members.
-In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
-<!---->
-However, in most landscape ecology studies, projected CRSs, which use some measurement units (e.g., meters), should be utilized.
-<!-- Projected CRS  -->
-Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
-<!-- Where to find advice on how to select projections-->
-Gladly, all of the packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations.
+A second common spatial data processing task is required when the available data extends over a larger area than the study region.
+In this case, the pre-processing of spatial data can include vector clipping or raster cropping. 
+Related to that, also masking of certain areas of the study region can be required by using e.g. certain spatial filters such as water bodies.
+Packages from Section \@ref(sec:spatrep) also allow for these operations.
+Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications.
+A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the *grainchanger* package \cite{Graham2019}.
+Furthermore, *landscapetools* is a collection of various utility functions for the raster data model.
 
-Second possible issue with spatial data is when the obtained data extends over a larger area than the study region.
-Then, some spatial preprecessing, such as vector clipping or raster cropping, is needed.
-Packages from Section \@ref(sec:spatrep) also make these operations possible.
-Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications.
-
-<!--MH: Short paragraph about scale; *grainchanger*; DOI: 10.1111/2041-210X.13177 -->
-<!-- MH: landscapetools; DOI: 10.1111/2041-210X.13076 -->
-
-Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass}<!--ADD R.LI CITATION-->, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R-->
-Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
+Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
 
 <!--JN: should we mention ArcGIS here and some of its modules?-->
+<!-- MH: I don't think this is really important -->
+
 <!--JN: in the same vain as above: should we mention Google Earth Engine?-->
+<!-- MH: Again, I dont think thats really of relevance  -->
 
@@
 Creating maps is essential when working with spatial data.
-It has a role in checking the quality of data<!--(both coordinates and values)-->, data exploration, and finally in communicating the results.
-<!-- role of making maps - from checking data, through the data exploration, to presenting the results  -->
-<!-- maps in the computer age -->
-Two main types of maps, static maps with full creator control and interactive maps where more control is given to the user, can be created in R.
-
-All of the packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the `plot()` function.
-However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
+It has a role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
+*R* allows to create two major types of maps.
+Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.
+changing the displayed values.
+All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function.
+However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
 Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
-The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
-This package aims for creating publication-ready maps.
-Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.<!--this sentence can be easily removed-->
-R's general-purpose plotting package *ggplot2* together with its extension *ggspatial* also offers some capabilities for plotting spatial data \cite{Wickham2016a,R-ggspatial}.
+The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
+Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}. 
+Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
 
@@
 While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
-
 Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}.
@@
 Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift and selection \cite{Manel2003}.
-Such insights improve our understanding of metapopulation dynamics, speciation, species’ distributions and conservation \cite{Storfer2007}. 
+Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions and conservation \cite{Storfer2007}.
 By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
@@
 <!-- MH: Work-in-Progress -->
+
 <!-- MH: General population genetics -->
+
 <!-- MH: Isolation by resistance  -->
+
 <!-- MH: Bayesian Genetic Clustering -->
+
 <!-- MH: Multivariate Ordination Techniques -->
@@
 Also, it is possible to simulate landscapes (gradient landscapes) using geostatistical packages, e.g., gstat.
--->
-<!-- MH: Well, but RandomFields and geostatistcal are not directly designed for NLM. Maybe need to relax the statement a little bit -->
-Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters. 
-<!-- JN: I would also add landscapetools here as it allows to categorize outputs of the NLMR package -->
-<!-- MH: I would rahter add a short sentence about landscapetools in the general spatial data section since the package can be used for all different kinds of tasks not only for NLMR -->
-
-\begin{landscape}
+--> <!-- MH: Well, but RandomFields and geostatistcal are not directly designed for NLM.
+Maybe need to relax the statement a little bit --> Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters.
 
 \begin{table}
-    \begin{tabularx}{0.9\linewidth}{XXXX}
+
 \caption{Overview of commonly used \textit{R} packages for landscape ecology.
-Packages are sorted by their major application topic.
-Only packages focussed on spatial data and landscape ecology are included.} \\
+Packages are sorted by their major application task.
+Only packages focused on spatial data and landscape ecology are included.}
+
+\begin{tabular}{l | r r }
+
 \hline
-      Topic & Package & Description & Source \\
+Task & R package & Reference \\
 \hline
-      \multirow{5}{*}{Spatial data} & \textit{raster} & Methods for raster data & \cite{Hijmans2019} \\
-      & \textit{stars} & Methods for spatiotemporal arrays (raster and vector data cubes) & \cite{Pebesma2019} \\
-      & \textit{sp} & Methods for mainly vector data & \cite{Pebesma2005,Bivand2013} \\
-      & \textit{sf} & Successor of the sp package. Methods for vector data as simple features & \cite{Pebesma2018} \\
-      & \textit{landscapetools} & Utility functions for landscape analysis & \cite{Sciaini2018} \\
+\multirow{5}{*}{Spatial data} & raster & \cite{Hijmans2019} \\
+& terra & \cite{R-terra} \\
+& stars & \cite{Pebesma2019} \\
+& sp & \cite{Pebesma2005,Bivand2013} \\
+& sf & \cite{Pebesma2018} \\
 \hline
-      \multirow{3}{*}{Creating maps} & \textit{tmap} & Plotting system for static and interactive maps & \cite{Tennekes2018} \\
-      & \textit{mapview} & Quick, interactive viewing of spatial data & \cite{R-mapview} \\
-      & \textit{leaflet} & Creating interactive web maps & \cite{R-leaflet} \\
+\multirow{6}{*}{Spatial data download} & FedData & \cite{R-FedData} \\
+& elevatr & \cite{R-elevatr} \\
+& rgbif & \cite{R-rgbif} \\
+& rnaturalearth & \cite{South2017} \\
+& marmap & \cite{Pante2013} \\
+& rgee & \cite{R-rgee} \\
 \hline
-      \multirow{3}{*}{}Quantifying landscape characteristics & \textit{landscapemetrics} & Calculates landscape metrics for categorical land cover maps & \cite{Hesselbarth2019a} \\
-      & \textit{belg} & Calculates the Boltzmann entropy of a landscape gradient & \cite{R-belgpaper} \\
-      & \textit{motif} & Enables several type of spatial analysis such as search, change detection, and clustering to be performed on categorical maps  & \cite{R-motif} \\
+\multirow{4}{*}{Creating maps} & tmap & \cite{Tennekes2018} \\
+& mapview & \cite{R-mapview} \\
+& leaflet & \cite{R-leaflet} \\
+& ggspatial & \cite{Dunnington2020} \\
 \hline
-      \multirow{5}{*}{}Species distribution modeling (SDM) & \textit{dismo} & xxx & xxx \\
-      & \textit{sdm} & xxx & xxx \\
-      & \textit{ecospat} & xxx & xxx \\
-      & \textit{biomod2} & xxx & xxx \\
-      & \textit{PresenceAbsence} & xxx & xxx \\
+\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & \cite{Hesselbarth2019a} \\
+& belg &  \cite{R-belgpaper} \\
+& motif & \cite{R-motif} \\
+& VLSM & xxx \\
 \hline
-      \multirow{3}{*}{Connectivity} & \textit{grainscape} & xxx & \cite{Chubaty2020} \\
-      & \textit{gdistance} & xxx & \cite{vanEtten2017} \\
-      & \textit{samc} & xxx & \cite{Marx2020} \\
+\multirow{5}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & \cite{Hijmans2017} \\
+& sdm & \cite{Naimi2016} \\
+& ecospat & \cite{Broennimann2020} \\
+& biomod2 & \cite{Thuiller2020} \\
+& PresenceAbsence & \cite{Freeman2008} \\
 \hline
-      \multirow{3}{*}{Landscape genetics} & \textit{graphs4lg} & xxx & \cite{Savary2020} \\
-      & \textit{PopGenReport} & xxx & \cite{Adamack2014,Gruber2015} \\
-      & \textit{HierDpart} & xxx & \cite{Qin2019} \\
+\multirow{3}{*}{Connectivity} & grainscape & \cite{Chubaty2020} \\
+& gdistance & \cite{vanEtten2017} \\
+& samc & \cite{Marx2020} \\
 \hline
-      Neutral Landscape Models (NLM) & \textit{NLMR} & xxx & \cite{Sciaini2018} \\
+\multirow{3}{*}{Landscape genetics} & raphs4lg & \cite{Savary2020} \\
+& PopGenReport & \cite{Adamack2014,Gruber2015} \\
+& HierDpart & \cite{Qin2019} \\
 \hline
-    \end{tabularx}
+\multirow{4}{*}{various} & NLMR & \cite{Sciaini2018} \\
+& RandomFields & \cite{Schlather2015} \\
+& landscapetools & \cite{Sciaini2018}  \\
+& grainchanger & \cite{Graham2019} \\
+\hline
+
+\end{tabular}
 \end{table}
-
-\end{landscape}
 
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, "(pre-)processing of data", "spatial statistics" and "creating maps" were the most selected options (Fig.
-\@ref(fig:fig-survey) B)).
+Not surprisingly, "(pre-)processing of data", "spatial statistics" and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey) B)).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
@@
 
-````{r packages, include =FALSE}
+```{r packages, include =FALSE}
 
@@
 The *raster* package was mentioned the most, followed by the *sf* package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey) C).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig.
+\@ref(fig:fig-survey) C).
 Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; 32.6%).
@@
 
-Overall, `r round(length(id_na) / nrow(answers) * 100, 1)`% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
+Overall, `r round(length(id_na) / nrow(answers) * 100, digits)`% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
 There were two most common topics across the answers of the participants.

### 2020-12-12 paper/paper.tex
:pencil: Work on introduction and overview of R packages

@@
 
+% Pandoc citation processing
+
 \begin{document}
@@
 
-\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Author 3 \and  \ldots{} \and  }
+\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  \ldots{} \and  }
 
@@
      \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-    \and
-        Author 3 \at
-     Adress author 3 \\
-     \email{E-Mail author 3}  %  \\
 %             \emph{Present address:} of F. Author  %  if needed
@@
 Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
-In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa) and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
+In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
 
@@
 Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}.
-to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}.
+To this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}.
 Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}.
-Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze and visualize spatial data.
+Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze, and visualize spatial data.
 
 \hypertarget{sec:open_source}{%
-\subsection{Open-source software and R}\label{sec:open_source}}
+\subsection{\texorpdfstring{Open-source software and \emph{R}}{Open-source software and R}}\label{sec:open_source}}
 
 Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
-Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
-Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
-Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}.
-Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.
-Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to many quite expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+Open-source software includes all software which is released under licenses that allow to freely use, modify and distribute the software \cite{St.Laurent2008}.
+Open-source software development has many advantages, such as fast innovation, transparency, and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
+Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
+Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
+Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
 
-One example of a successful open-source project is the \emph{R} programming language and its extensions (also called packages) \cite{RCoreTeam2019}.
+One successful example of an open-source project is the \emph{R} programming language and its extensions (also called packages) \cite{RCoreTeam2019}.
 Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006}.
-This is documented by the growing \emph{CRAN Task View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} that currently list almost 300 packages in total.
-Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely even much higher.
+This is documented by the growing \emph{CRAN Task Views: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} that currently list almost 300 packages in total.
+Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely even higher.
 The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
 
-Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the \emph{R} programming language.
 The growing body of \emph{R} packages related to spatial data processing and analyzing results in a high capability of the \emph{R} programming language for landscape ecology.
-Because additionally to the handling of spatial data, also further task such as statistical modeling, the preparation of publication-ready figures and even the preparation of complete reports using \emph{RMarkdown} \cite{Xie2018} can be done in the same working environment, the \emph{R} programming language can be a very powerful tool.
-Thus, in the first part we are give a comprehensive overview over existing \emph{R} packages for landscape ecology.
-In the second part, we identify topics for which \emph{R} packages are currently missing to the best of our knowledge.
-For more general overviews see e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on software implemented in the \emph{R} programming language.
+For more general overviews of open-source software for landscape ecology see e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Because additionally to handling spatial data, also further tasks such as statistical modeling, preparation of publication-ready figures and even preparation of complete reports can be done in the same working environment, the \emph{R} programming language can be a very powerful tool.
+Thus, in the first part we present an overview over existing \emph{R} packages for landscape ecology.
+In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are currently missing for landscape ecology.
 
@@
 
-There are two major platforms to host \emph{R} packages and make them easily accessible to potential users.
-Firstly, hosting the package on \emph{GitHub} allows users to install it with only one line of code.
-Additionally, hosting the package on \emph{GitHub} provides many useful tools to communicate between developers and users, such as bug reports \cite{Wickham2015}.
-Secondly, the \emph{Comprehensive R Archive Network} (\emph{CRAN}) provides large visibility to the community, ease of installation and a certain quality standard \cite{Wickham2015}.
-However, at the same time, releasing a \emph{R} package on \emph{CRAN} requires more initial work for the developers compared to \emph{GitHub}.
-Since most \emph{R} packages are developed and maintained by the community, this is a good example of how open-source software development can facilitate innovation, reproducibility and reuse of code, especially since now many \emph{R} packages depend on each other (i.e.~package \emph{x} requires and uses code from package \emph{y}) \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+Since most \emph{R} packages are developed and maintained by the community, this shows how open-source software development can facilitate innovation, reproducibility and reuse of code.
+There are two major online platforms to host \emph{R} packages and make them accessible to potential users.
+Firstly, \emph{GitHub} allows to host source code under version control and users can install packages with one line using the \emph{remotes} \cite{Hester2020} package.
+Additionally, hosting the package on \emph{GitHub} provides many useful features to collaborate and communicate between developers as well as users \cite{Wickham2015}, or integrated unit testing.
+Secondly, the \emph{Comprehensive R Archive Network} (\emph{CRAN}) provides a large visibility to the community, ease of installation and a technical quality standard, including checks for common problems on all major operating systems.
+\cite{Wickham2015}.
+For developers, the guaranteed technical quality standard on \emph{CRAN} requires more initial work for compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
+Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e.~package \emph{x} requires and uses code from package \emph{y} \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
 For more information about \emph{R} package development, see \cite{Wickham2015}.
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth,height=0.65\textheight]{data/Workflow}
+
+}
+
+\caption{Exemplary worklow of spatial data analyses for landscape ecology using the R programming language.
+For all major tasks, a few example *R* packages are listed.
+Of course, many more packages for the specific tasks exist.}\label{fig:fig-workflow}
+\end{figure}
 
@@
 
-Most of the existing spatial data belong to one of two data models: spatial raster and spatial vectors.
+While \emph{R} has several build-in data structures, including vectors, matrices or lists, the base \emph{R} programming language has no internal support for reading, processing, or visualizing spatial data.
+However, because there is a notable interest in spatial data analysis, support for spatial data is provided by many \emph{R} packages (\cite{Lovelace2019}, page 10).
+Most spatial data belong to one of two contrasting data models, namely spatial raster and spatial vectors, and both data models have several implementations in the \emph{R} language.
+Importantly, basically all \emph{R} packages for spatial data use the external \emph{GDAL} and \emph{PROJ} libraries, which allow for e.g.~reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Additionally, \emph{R}, \emph{GDAL} and \emph{PROJ} allow to convert between data models and specific implementations, which can be useful if given methods only exist for a specific data model or implementation.
+
 In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
 The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
-The vector data model consists of two main elements, geometries, such as points, lines, polygons, and attributes.
-Each geometry is connected to a row in an attribute table.
-A few hundred of different spatial data file formats were developed to store data in these formats.
+The probably most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
+A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
+Alternatively, the \emph{stars} package can be used to read and process raster data with a focus on spatial-temporal data cubes \cite{Pebesma2019}.
 
-While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
-However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page 10).
-Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.
-
-The most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
-A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
-Alternatively, \emph{stars} package can be used to read and process raster data \cite{Pebesma2019}.
-It is interconnected with the \emph{sf} package, and beside raster support, it provides tools for spatial data cubes.
-
-The \emph{sp} package was a standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
-As of 2020, more than 500 R packages directly depends or imports \emph{sp}.
-However, it is now in a maintenance mode only with a recommended successor - the \emph{sf} package.
-
-Importantly, all of the above packages use the external libraries GDAL and PROJ, which allow for reading and writing of hundreds of spatial data formats and coordinates transformation.
-They also have dedicated methods or functions to convert between representations, for example between \emph{sp} and \emph{sf} objects or \emph{raster} and \emph{stars} objects.
-This can be useful when a given function expects different data representation than currently existing.
+The vector data model consists of two main elements i) geometries, such as points, lines, polygons, and ii) attributes and each geometry is connected to a value in an attribute table.
+In many cases, this data model allows a more natural representation of landscape features, however, with the cost of higher computational demands .
+The \emph{sp} package was the standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
+As of 2020, more than 500 \emph{R} packages directly depend or imports \emph{sp}.
+However, more recently \emph{sp} will only be maintained (and not actively developed any more), and the recommended successor is the improved \emph{sf} package \cite{Pebesma2018}.
 
@@
 
-Nowadays, spatial data for many spatial scales is available from abundance of sources.
-Some of them require registration or having appropriate credentials and allows downloading only from the web interface.
-However, many of the sources made their data publicly available either as a direct download or through an API connection.
-This is used by some R packages allowing for downloading various environmental spatial dataset.
-
-A plethora of land cover products on global, continental, and country scales has been developed, however, most of them cannot be directly downloaded from R.
-The exception is the National Land Cover Database (NLCD) data for the USA, which can be accessed through the \emph{FedData} package \cite{R-FedData}.
-The \emph{elevatr} package provides a synthesis raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}.
-It uses different elevation products for a given area and selected level of details (zoom/resolution).
-The \emph{rgbif} gives access to data from the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}.
-The \texttt{get\_data()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim version 1.4, while the new in-development package \emph{geodata} provides an access to WorldClim version 2 data.
-Additionally, the \emph{rgee package} \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
+Nowadays, spatial data for many spatial scales is available from an abundance of online accessible sources.
+A lot of this data is publicly available, either as a direct download or through an API connection, and this can be used by several packages to download the spatial data directly into a \emph{R} session.
+Since public available data is becoming more and more prominent, so are \emph{R} packages to accesses them.
+Packages include the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA, the \emph{elevatr} package to access raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}, the \emph{rgbif} to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, \emph{rnaturalearth} to access the Natural Earth database \cite{South2017}, or \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}.
+Furthermore, the \texttt{get\_data()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim v1.4, while the new in-development package \emph{geodata} provides an access to data from WorldClim v2.
+Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
 
@@
 
-As we mentioned in the previous sections, gathered or obtained data can be represented by different data types and stored in various file formats.
-However, several potential obstacles to applying this data for ecological analysis still exist.
-
-The main issue for many landscape ecology studies is related to projections of the used data.
-It is not only required to have all of the used datasets in the same projection, but also selection of a proper projection is crucial.
+Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps .
+This is also referred to as the projection of the data and is often a first barrier in spatial data analysis.
+It is not only required to have all of the used data in the same projection, but also selection of a proper projection is crucial.
+This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion and different CRS are optimized for different regions of the world and scales .
 Coordinates in spatial data can represent one of many coordinate reference systems.
-Two main groups of coordinate reference systems (CRS) exist, namely geographical and projected, each having many members.
+Two main groups of CRS exist, namely geographical and projected, each having many members.
 In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
@@
 Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
-Gladly, all of the packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations.
+Tools to find an appropriate CRS for a certain region can be found at \url{https://spatialreference.org}, \url{https://epsg.org}, or \url{http://epsg.io}.
+In used data have different CRS, all packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations.
 
-Second possible issue with spatial data is when the obtained data extends over a larger area than the study region.
-Then, some spatial preprecessing, such as vector clipping or raster cropping, is needed.
-Packages from Section \ref{sec:spatrep} also make these operations possible.
+A second common spatial data processing task is required when the available data extends over a larger area than the study region.
+In this case, the pre-processing of spatial data can include vector clipping or raster cropping.
+Related to that, also masking of certain areas of the study region can be required by using e.g.~certain spatial filters such as water bodies.
+Packages from Section \ref{sec:spatrep} also allow for these operations.
 Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications.
+A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the \emph{grainchanger} package \cite{Graham2019}.
+Furthermore, \emph{landscapetools} is a collection of various utility functions for the raster data model.
 
-Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
+Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
 Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.
@@
 Creating maps is essential when working with spatial data.
-It has a role in checking the quality of data, data exploration, and finally in communicating the results.
-Two main types of maps, static maps with full creator control and interactive maps where more control is given to the user, can be created in R.
-
-All of the packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the \texttt{plot()} function.
-However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
+It has a role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
+\emph{R} allows to create two major types of maps.
+Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.~changing the displayed values.
+All packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the generic \texttt{plot()} function.
+However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
 Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
-The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
-This package aims for creating publication-ready maps.
+The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
 Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
-R's general-purpose plotting package \emph{ggplot2} together with its extension \emph{ggspatial} also offers some capabilities for plotting spatial data \cite{Wickham2016a,R-ggspatial}.
+Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
 
@@
 
-\begin{landscape}
+\begin{table}
 
-  \begin{table}
-    \begin{tabularx}{0.9\linewidth}{XXXX}
 \caption{Overview of commonly used \textit{R} packages for landscape ecology.
-Packages are sorted by their major application topic.
-Only packages focussed on spatial data and landscape ecology are included.} \\
+Packages are sorted by their major application task.
+Only packages focused on spatial data and landscape ecology are included.}
+
+\begin{tabular}{l | r r }
+
 \hline
-      Topic & Package & Description & Source \\
+Task & R package & Reference \\
 \hline
-      \multirow{5}{*}{Spatial data} & \textit{raster} & Methods for raster data & \cite{Hijmans2019} \\
-      & \textit{stars} & Methods for spatiotemporal arrays (raster and vector data cubes) & \cite{Pebesma2019} \\
-      & \textit{sp} & Methods for mainly vector data & \cite{Pebesma2005,Bivand2013} \\
-      & \textit{sf} & Successor of the sp package. Methods for vector data as simple features & \cite{Pebesma2018} \\
-      & \textit{landscapetools} & Utility functions for landscape analysis & \cite{Sciaini2018} \\
+\multirow{5}{*}{Spatial data} & raster & \cite{Hijmans2019} \\
+& terra & \cite{R-terra} \\
+& stars & \cite{Pebesma2019} \\
+& sp & \cite{Pebesma2005,Bivand2013} \\
+& sf & \cite{Pebesma2018} \\
 \hline
-      \multirow{3}{*}{Creating maps} & \textit{tmap} & Plotting system for static and interactive maps & \cite{Tennekes2018} \\
-      & \textit{mapview} & Quick, interactive viewing of spatial data & \cite{R-mapview} \\
-      & \textit{leaflet} & Creating interactive web maps & \cite{R-leaflet} \\
+\multirow{6}{*}{Spatial data download} & FedData & \cite{R-FedData} \\
+& elevatr & \cite{R-elevatr} \\
+& rgbif & \cite{R-rgbif} \\
+& rnaturalearth & \cite{South2017} \\
+& marmap & \cite{Pante2013} \\
+& rgee & \cite{R-rgee} \\
 \hline
-      \multirow{3}{*}{}Quantifying landscape characteristics & \textit{landscapemetrics} & Calculates landscape metrics for categorical land cover maps & \cite{Hesselbarth2019a} \\
-      & \textit{belg} & Calculates the Boltzmann entropy of a landscape gradient & \cite{R-belgpaper} \\
-      & \textit{motif} & Enables several type of spatial analysis such as search, change detection, and clustering to be performed on categorical maps  & \cite{R-motif} \\
+\multirow{4}{*}{Creating maps} & tmap & \cite{Tennekes2018} \\
+& mapview & \cite{R-mapview} \\
+& leaflet & \cite{R-leaflet} \\
+& ggspatial & \cite{Dunnington2020} \\
 \hline
-      \multirow{5}{*}{}Species distribution modeling (SDM) & \textit{dismo} & xxx & xxx \\
-      & \textit{sdm} & xxx & xxx \\
-      & \textit{ecospat} & xxx & xxx \\
-      & \textit{biomod2} & xxx & xxx \\
-      & \textit{PresenceAbsence} & xxx & xxx \\
+\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & \cite{Hesselbarth2019a} \\
+& belg &  \cite{R-belgpaper} \\
+& motif & \cite{R-motif} \\
+& VLSM & xxx \\
 \hline
-      \multirow{3}{*}{Connectivity} & \textit{grainscape} & xxx & \cite{Chubaty2020} \\
-      & \textit{gdistance} & xxx & \cite{vanEtten2017} \\
-      & \textit{samc} & xxx & \cite{Marx2020} \\
+\multirow{5}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & \cite{Hijmans2017} \\
+& sdm & \cite{Naimi2016} \\
+& ecospat & \cite{Broennimann2020} \\
+& biomod2 & \cite{Thuiller2020} \\
+& PresenceAbsence & \cite{Freeman2008} \\
 \hline
-      \multirow{3}{*}{Landscape genetics} & \textit{graphs4lg} & xxx & \cite{Savary2020} \\
-      & \textit{PopGenReport} & xxx & \cite{Adamack2014,Gruber2015} \\
-      & \textit{HierDpart} & xxx & \cite{Qin2019} \\
+\multirow{3}{*}{Connectivity} & grainscape & \cite{Chubaty2020} \\
+& gdistance & \cite{vanEtten2017} \\
+& samc & \cite{Marx2020} \\
 \hline
-      Neutral Landscape Models (NLM) & \textit{NLMR} & xxx & \cite{Sciaini2018} \\
+\multirow{3}{*}{Landscape genetics} & raphs4lg & \cite{Savary2020} \\
+& PopGenReport & \cite{Adamack2014,Gruber2015} \\
+& HierDpart & \cite{Qin2019} \\
 \hline
-    \end{tabularx}
+\multirow{4}{*}{various} & NLMR & \cite{Sciaini2018} \\
+& RandomFields & \cite{Schlather2015} \\
+& landscapetools & \cite{Sciaini2018}  \\
+& grainchanger & \cite{Graham2019} \\
+\hline
+
+\end{tabular}
 \end{table}
-
-\end{landscape}
 
@@
 To better understand how \emph{R} is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the community.
-In total, the survey was answered by 96 participants, of which the the majority were either ``PhD students'' (33.3\%), followed by ``Post-Docs'' (27.1\%) and ``Professors'' (12.5\%).
-Other, less frequent answers, were ``data scientists'', ``none of the above'', ``government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
+Other, less frequent answers, were ``data scientists,'' ``none of the above,'' ``government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
-Most people use \emph{R} either ``daily'' (57.3\%) or a ``few times a week'' (35.4\%) and about half of all participants described themselves as ``advanced'' user (47.9\%), while 40.6\% described themselves as ``intermediate'' users.
-Related to this, about half of the participants either implemented their own package (20.8\%) or plan to do so in the future (21.9\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
+Most people use \emph{R} either ``daily'' (54.4\%) or a ``few times a week'' (36.9\%) and about half of all participants described themselves as ``advanced'' user (46.6\%), while 40.8\% described themselves as ``intermediate'' users.
+Related to this, about half of the participants either implemented their own package (21.4\%) or plan to do so in the future (23.3\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
 
-We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'' and ``nature conservation'' (Fig.
+We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity'' and ``nature conservation'' (Fig.
 \ref{fig:fig-survey} A)).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'' and ``creating maps'' were the most selected options (Fig.
-\ref{fig:fig-survey} B)).
+Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
 
-Interestingly, most people use the raster data model (70.8\%) in comparison to the vector data model (29.2\%).
+Interestingly, most people use the raster data model (72.8\%) in comparison to the vector data model (27.2\%).
 This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey} C)).
-When asked for the three most used packages, participants of the survey listed 75 packages in total.
+When asked for the three most used packages, participants of the survey listed 83 packages in total.
 The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey} C).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig.
+\ref{fig:fig-survey} C).
 Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 32.6\%).
 
-Lastly, when asked ``How useful do you think is R currently for landscape ecology?'', the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.67\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.33\%; Fig.
+Lastly, when asked ``How useful do you think is R currently for landscape ecology?'' the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
 \ref{fig:fig-survey} D)).
+
+\begin{verbatim}
+## Warning: Removed 1 rows containing missing values (position_stack).
+\end{verbatim}
 
@@
 
-Overall, 21.9\% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
-There were two most common topics across the answers of the participants .
-Firstly, many participants (13.5\%) wished for a better computational performance of \emph{R} in terms of speed and required RAM for larger data.
-Secondly, many participants (9.4\%) are currently missing advanced and easy-to-apply methods for creating high-quality maps or other visualization-related functionality.
-Fewer participants were missing tools for various connectivity-related issues (9.4\%) and characterization of landscape characteristics (7.3\%).
+Overall, 20.4\% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
+There were two most common topics across the answers of the participants.
+Firstly, many participants (12.6\%) wished for a better computational performance of \emph{R} in terms of speed and required RAM for larger data.
+Secondly, many participants (8.7\%) are currently missing advanced and easy-to-apply methods for creating high-quality maps or other visualization-related functionality.
+Fewer participants were missing tools for various connectivity-related issues (8.7\%) and characterization of landscape characteristics (6.8\%).
 

### 2020-12-12 R/get_survery_data.R

@@ -14,4 +14,4 @@ names(answers) <-  new_names
 
 results <- list(answers = answers, old_names = old_names, new_names = new_names)
 
-saveRDS(object = results, file = "paper/results.rds")
+saveRDS(object = results, file = "paper/data/results.rds")

### 2020-12-14 paper/paper.Rmd
Update figures and text

@@
   \usepackage{caption}
-  \usepackage{ltablex}
-  \usepackage{pdflscape}
+  \usepackage{tabularx}
   \usepackage{multirow}
@@
 For more general overviews of open-source software for landscape ecology see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Because additionally to handling spatial data, also further tasks such as statistical modeling, preparation of publication-ready figures and even preparation of complete reports can be done in the same working environment, the *R* programming language can be a very powerful tool.
-Thus, in the first part we present an overview over existing *R* packages for landscape ecology.
+Because additionally to handling spatial data, also further tasks such as statistical modeling, preparation of publication-ready figures and even preparation of complete reports can be done in the same working environment (Fig.
+\@ref(fig:fig-workflow)), the *R* programming language can be a very powerful tool.
+Thus, in the first part we present an overview over existing *R* packages for landscape ecology (Table \@ref(tab:packages)).
 In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are currently missing for landscape ecology.
@@
 
-```{r fig-workflow, echo = FALSE, out.width = "100%", out.height = "65%", fig.align="center", fig.cap = "Exemplary worklow of spatial data analyses for landscape ecology using the R programming language.
-For all major tasks, a few example *R* packages are listed.
+```{r fig-workflow, echo = FALSE, out.width="100%", out.height="30%", fig.cap = "Exemplary worklow of spatial data analyses for landscape ecology using the R programming language.
+For all major tasks, a few example R packages are listed.
 Of course, many more packages for the specific tasks exist."}
-include_graphics("data/Workflow.pdf")
+include_graphics("data/Workflow.png")
 ```
@@
 
+<!-- Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}. However, currently the *maxent* package is not available on *CRAN* anymore. -->
+
 #### Connectivity {#sec:connectivity}
@@
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found (for a comprehensive list, see \cite{Sciaini2018}).
-
-Currently, to our knowledge, only one *R* packages exists that allows the use of neutral landscapes.
-<!-- JN: I am not sure if that's true as few NLMR functions are just wrappers around existing R functions from RandomFields.
-Also, it is possible to simulate landscapes (gradient landscapes) using geostatistical packages, e.g., gstat.
---> <!-- MH: Well, but RandomFields and geostatistcal are not directly designed for NLM.
-Maybe need to relax the statement a little bit --> Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters.
+Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found across many different *R* packages. 
+A comprehensive collections of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the *NLMR* package \cite{Sciaini2018}.
+Furthermore, the *RandomFields* package \cite{Schlather2015} allows to simulate Gaussian fields which could be used as neutral landscape models.
 
@@
 
-\caption{Overview of commonly used \textit{R} packages for landscape ecology.
-Packages are sorted by their major application task.
-Only packages focused on spatial data and landscape ecology are included.}
+\caption{Overview of commonly used R packages for landscape ecology. Packages are sorted by their major application task. Only packages focused on spatial data and landscape ecology are included.}
 
-\begin{tabular}{l | r r }
+\begin{tabularx}{\textwidth}{X r r }
 
@@
 
-\end{tabular}
+\end{tabularx}
+
+\label{tab:packages}
+
 \end{table}
@@
 
-To better understand how *R* is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the community.
+To better understand how *R* is used by the landscape ecology community, we conducted a short survey and used mailing lists and social media to reach the community.
 In total, the survey was answered by `r n_people` participants, of which the the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
-Other, less frequent answers, were "data scientists", "none of the above", "government employees" "Master's degree student" and "Bachelor's degree student" (in decreasing order).
+Other, less frequent answers, were "Data scientists", "None of the above", "Government employees" "Master's degree student" and "Bachelor's degree student" (in decreasing order).
 
-Most people use *R* either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%) and about half of all participants described themselves as "advanced" user (`r expertise[[1, 2]]`%), while `r expertise[[2, 2]]`% described themselves as "intermediate" users.
+Most people use *R* either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%). 
+About half of all participants described themselves as "advanced" user (`r expertise[[1, 2]]`%), while `r expertise[[2, 2]]`% described themselves as "intermediate" users.
 Related to this, about half of the participants either implemented their own package (`r own_package[[3, 2]]`%) or plan to do so in the future (`r own_package[[2, 2]]`%) and most of these packages are hosted on *GitHub* and/or *CRAN*.
@@
 
-Interestingly, most people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
+Interestingly, more people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
 This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-survey) C)).
@@
 The *raster* package was mentioned the most, followed by the *sf* package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig.
-\@ref(fig:fig-survey) C).
-Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; 32.6%).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey) C).
+Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r r_packages_df[[1, 2]]`%).
 
@@
   theme(plot.title.position = "plot") +
-  scale_y_continuous(breaks = seq(0, 60, 5), limits = c(0, 60)) + 
+  scale_y_continuous(breaks = seq(0, 60, 10), limits = c(0, 60)) + 
   guides(fill = FALSE, colour = FALSE) +
@@
 
-Lastly, when asked "`r old_names[12]`", the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
+Lastly, when asked how use *R* us currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
 \@ref(fig:fig-survey) D)).
@@
 # no answer, no missing packages
-id_na <- c("none", "na", "no", "not sure", "not sure right now", "unsure",
-           "i have no additional needs for my current workflow", 
-           "none that i can think of", "i would like to know",
-           "i don't know, sorry!", "r havent fail me yet", 
-           "i'm not missing anything at present.",
-           "have not yet identified any.", "no package missing",
-           "i am currently missing no methods",  
-           "i still have to fully understand what is there", 
-           "não posso responder", "not enough overview to answer this",
-           ";", ".", "?")
+id_na <- c("not sure", "not sure right now",  
+           "i have no additional needs for my current workflow", "none that i can think of", 
+           "i would like to know", "none", "i still have to fully understand what is there", 
+           "i don't know, sorry!", "r havent fail me yet", "na", 
+           "i'm not missing anything at present.", "não posso responder", 
+           "no", "unsure", "have not yet identified any.", "no package missing", 
+           "not enough overview to answer this", "?", "i am currently missing no methods",
+           ";", "nothing yet", "i'm thinking", "nil")
 
@@
 # classify answers as good as possible
+methods_performance <- missing_methods[c(2, 4, 8, 9, 10, 21, 23, 27, 30, 33, 41, 45, 
+                                         48, 56)]
 
-methods_performance <- missing_methods[c(2, 4, 8, 9, 10, 21, 23, 27, 33, 41, 45, 
-                                         48, 56)]
 perc_performance <- round(length(methods_performance) / nrow(answers) * 100, 1)
@@
 methods_viz <- missing_methods[c(3, 7, 13, 22, 29, 33, 41, 57, 58)]
+
 perc_viz <- round(length(methods_viz) / nrow(answers) * 100, 1)
 
-methods_connectivity <- missing_methods[c(12, 17, 26, 31, 35, 36, 40, 51)]
-perc_connectivity <- round(length(methods_connectivity) / nrow(answers) * 100, 1)
+methods_characterisation <- missing_methods[c(7, 14, 20, 24, 39, 42, 47, 51, 59, 65)]
 
-methods_characterisation <- missing_methods[c(7, 20, 24, 39, 42, 47, 59)]
 perc_characterisation <- round(length(methods_characterisation) / nrow(answers) * 100, 1)
 
-methods_tutorial <- missing_methods[c(1, 14, 44, 52, 53)]
-
-methods_datahandling <- missing_methods[c(5, 11, 22,  60, 63)]
-
-methods_gis <- missing_methods[c(43, 50, 55)]
-
-methods_various <- missing_methods[c(6, 11, 15, 16, 18, 19, 22, 25, 28, 30, 32, 
-                                     34, 37, 38, 46, 49, 54, 61, 62, 63, 64)]
+# methods_connectivity <- missing_methods[c(12, 17, 26, 31, 35, 36, 40, 51)]
+# 
+# perc_connectivity <- round(length(methods_connectivity) / nrow(answers) * 100, 1)
 
@@
 Overall, `r round(length(id_na) / nrow(answers) * 100, digits)`% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
-There were two most common topics across the answers of the participants.
-Firstly, many participants (`r perc_performance`%) wished for a better computational performance of *R* in terms of speed and required RAM for larger data.
-Secondly, many participants (`r perc_viz`%) are currently missing advanced and easy-to-apply methods for creating high-quality maps or other visualization-related functionality.
-Fewer participants were missing tools for various connectivity-related issues (`r perc_viz`%) and characterization of landscape characteristics (`r perc_characterisation`%).
-
-<!-- Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}. However, currently the *maxent* package is not available on *CRAN* anymore. -->
-
-<!-- missing packages: data sources, such as global land cover, worldclim 2.0, etc. -->
+There were three most common topics across the answers of the participants.
+Firstly, many participants (`r perc_performance`%) wished for a better computational performance of *R* in terms of speed and required RAM, especially for larger data.
+Secondly, participant are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to  quantify landscape characteristics (`r perc_characterisation`%).
+Thirdly, many participants (`r perc_viz`%) are currently missing advanced and easy-to-apply methods to create high-quality maps or other visualization-related functionality.
 
@@
 speed and required RAM for larger data -> use of Rcpp, use external servers, or read data in chunks, etc.)-->
+
+<!-- MH: I am not sure if we should do this explicitly here  That would somehow repeat the section we already have about the different packages.
+We should, however, make sure that in case we know a pacakge that answers one of the points here, is included in the first section of the papaer -->
+
+# Conclusions
+
+Since its first introduction in 1995, *R* has come a long way from a statistical programming language to a powerful tool for landscape ecology. 
+Today, many *R* packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed for spatial data handling and analysis.
+The highly dynamic development of *R* packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
+However, since the scientific field of landscape ecology constantly changes and improves, also the *R* programming language and its packages need to change and adapt these changes.

### 2020-12-14 paper/paper.tex
Update figures and text

@@
 \usepackage{caption}
-\usepackage{ltablex}
-\usepackage{pdflscape}
+\usepackage{tabularx}
 \usepackage{multirow}
@@
 For more general overviews of open-source software for landscape ecology see e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Because additionally to handling spatial data, also further tasks such as statistical modeling, preparation of publication-ready figures and even preparation of complete reports can be done in the same working environment, the \emph{R} programming language can be a very powerful tool.
-Thus, in the first part we present an overview over existing \emph{R} packages for landscape ecology.
+Because additionally to handling spatial data, also further tasks such as statistical modeling, preparation of publication-ready figures and even preparation of complete reports can be done in the same working environment (Fig.
+\ref{fig:fig-workflow}), the \emph{R} programming language can be a very powerful tool.
+Thus, in the first part we present an overview over existing \emph{R} packages for landscape ecology (Table \ref{tab:packages}).
 In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are currently missing for landscape ecology.
@@
 \begin{figure}
-
-{\centering \includegraphics[width=1\linewidth,height=0.65\textheight]{data/Workflow}
-
-}
-
-\caption{Exemplary worklow of spatial data analyses for landscape ecology using the R programming language.
-For all major tasks, a few example *R* packages are listed.
+\includegraphics[width=1\linewidth,height=0.3\textheight]{data/Workflow} \caption{Exemplary worklow of spatial data analyses for landscape ecology using the R programming language.
+For all major tasks, a few example R packages are listed.
 Of course, many more packages for the specific tasks exist.}\label{fig:fig-workflow}
@@
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found (for a comprehensive list, see \cite{Sciaini2018}).
-
-Currently, to our knowledge, only one \emph{R} packages exists that allows the use of neutral landscapes.
-Using a wide collection of algorithms, the \emph{NLMR} package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters.
+Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found across many different \emph{R} packages.
+A comprehensive collections of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \emph{NLMR} package \cite{Sciaini2018}.
+Furthermore, the \emph{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields which could be used as neutral landscape models.
 
@@
 
-To better understand how \emph{R} is used by the landscape ecology community, we designed a short survey and used mailing lists and social media to reach the community.
+To better understand how \emph{R} is used by the landscape ecology community, we conducted a short survey and used mailing lists and social media to reach the community.
 In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
-Other, less frequent answers, were ``data scientists,'' ``none of the above,'' ``government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers, were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
-Most people use \emph{R} either ``daily'' (54.4\%) or a ``few times a week'' (36.9\%) and about half of all participants described themselves as ``advanced'' user (46.6\%), while 40.8\% described themselves as ``intermediate'' users.
+Most people use \emph{R} either ``daily'' (54.4\%) or a ``few times a week'' (36.9\%).
+About half of all participants described themselves as ``advanced'' user (46.6\%), while 40.8\% described themselves as ``intermediate'' users.
 Related to this, about half of the participants either implemented their own package (21.4\%) or plan to do so in the future (23.3\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
 
-We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity'' and ``nature conservation'' (Fig.
+We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'' and ``nature conservation'' (Fig.
 \ref{fig:fig-survey} A)).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
+Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
 
-Interestingly, most people use the raster data model (72.8\%) in comparison to the vector data model (27.2\%).
+Interestingly, more people use the raster data model (72.8\%) in comparison to the vector data model (27.2\%).
 This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey} C)).
@@
 The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig.
-\ref{fig:fig-survey} C).
-Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 32.6\%).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey} C).
+Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 33.4415584\%).
 
-Lastly, when asked ``How useful do you think is R currently for landscape ecology?'' the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
+Lastly, when asked how use \emph{R} us currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
 \ref{fig:fig-survey} D)).
@@
 
-Overall, 20.4\% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
-There were two most common topics across the answers of the participants.
-Firstly, many participants (12.6\%) wished for a better computational performance of \emph{R} in terms of speed and required RAM for larger data.
-Secondly, many participants (8.7\%) are currently missing advanced and easy-to-apply methods for creating high-quality maps or other visualization-related functionality.
-Fewer participants were missing tools for various connectivity-related issues (8.7\%) and characterization of landscape characteristics (6.8\%).
+Overall, 22.3\% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
+There were three most common topics across the answers of the participants.
+Firstly, many participants (13.6\%) wished for a better computational performance of \emph{R} in terms of speed and required RAM, especially for larger data.
+Secondly, participant are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (9.7\%).
+Thirdly, many participants (8.7\%) are currently missing advanced and easy-to-apply methods to create high-quality maps or other visualization-related functionality.
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+Since its first introduction in 1995, \emph{R} has come a long way from a statistical programming language to a powerful tool for landscape ecology.
+Today, many \emph{R} packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed for spatial data handling and analysis.
+The highly dynamic development of \emph{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
+However, since the scientific field of landscape ecology constantly changes and improves, also the \emph{R} programming language and its packages need to change and adapt these changes.
 

### 2020-12-15 paper/paper.Rmd
:pencil: Adding  packages addressing some comments section of survey

@@
 package *x* requires and uses code from package *y* \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+Thus, most packages can be found on both platforms and many developers use *GitHub* for daily developing and *CRAN* to publish stable releases of the packages.
 For more information about *R* package development, see \cite{Wickham2015}.
@@
 
-<!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->
-
-<!-- JN: if you think this package (or similar) are still useful, maybe we can mention them somewhere is the first part of the paper with annotation about their state, and then also write something more about issues with archived packages in the second part (what's missing in R)...
--->
-
-<!-- MH: I would argue to only include packages that are actively maintained.
-It's already such a huge amount and I guess there is a reason why the packages were archived.
-Doesnt speak for their popularity -->
-
 ## Spatial data representations {#sec:spatrep}
-
-<!-- MH: Make sure all libraries etc. have reference -->
 
@@
 Alternatively, the *stars* package can be used to read and process raster data with a focus on spatial-temporal data cubes \cite{Pebesma2019}.
+Additionally, there are packages that improve some basics raster operations in terms of computational performance or compatibility between raster and vector operations, such as *fasterize* \cite{Ross2020}, *rasterDT* \cite{OBrien2020}, or *exactextractr* \cite{Baston2020}.
 
@@
 <!-- MH: I probably would not add terra here.
-As far as I see, the vector support in terra is basically just support to combine vector with raster data.
-The main focuse of the package is raster data though -->
+As far as I see it, the vector support in terra is basically just support to combine vector with raster data.
+The main focus of the package is raster data though -->
 
@@
 Since public available data is becoming more and more prominent, so are *R* packages to accesses them.
-Packages include the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA, the *elevatr* package to access raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}, the *rgbif* to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, *rnaturalearth* to access the Natural Earth database \cite{South2017}, or *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}.
-Furthermore, the `get_data()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim v1.4, while the new in-development package *geodata* <!--(add reference)--> provides an access to data from WorldClim v2.
+Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows to download data from the Landsat 8, while the *MODIS* package \cite{Mattiuzzi2020} allows to download MODIS products.
+Also, the `get_data()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the  in-development package *geodata* <!--(add reference)-->).
 Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
@@
 <!--JN: in the same vain as above: should we mention Google Earth Engine?-->
-<!-- MH: Again, I dont think thats really of relevance  -->
+<!-- MH: Again, I dont think thats really of relevance. I think as it is now its ok  -->
 
@@
 However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
+Advanced plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
 Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
@@
 Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
+Thematic maps, including besides others proportional symbols, chroropleth, or typology, can be created with the *cartography* package \cite{Giraud2016}.
 Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
+
+A slightly different approach of visualizing spatial data is the *rayshader* package \cite{Morgen-Wall2020} which allows to create topographical 2D and 3D maps.
 
@@
 
-#### Landscape metrics {#sec:landscape_metrics}
+#### Quantify landscape characteristics {#sec:landscape_metrics}
 
@@
 Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
+
+More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
+The *geodiv* package \cite{Smith2020} allows to gradient surface metrics to characterize continuous analysis of landscape features.
 
@@
 Of course, also packages specifically designed for SDM exists.
-This includes the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020} and *PresenceAbsence* \cite{Freeman2008} packages.
+This includes the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008} or *zoon* \cite{Golding2018} packages.
+Related to SDM, there are several packages to estimate home ranges and habitat selection, including *adehabitatHR* and *adehabitatHS* \cite{Calenge2006} or *amt \cite{Signer2019}*
 
@@
 Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}).
+Furthermore, the *lconnect* package \cite{Mestre2019} and *Makurhini* package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
 Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}.
@@
 
-<!-- MH: Work-in-Progress -->
-
-<!-- MH: General population genetics -->
-
-<!-- MH: Isolation by resistance  -->
-
-<!-- MH: Bayesian Genetic Clustering -->
-
-<!-- MH: Multivariate Ordination Techniques -->
-
 Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also of importance for landscape genetics.
-Further functionality for landscape genetics can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, or *HierDpart* packages \cite{Qin2019}.
+Further functionality for landscape genetics can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.
 
@@
 
+#### Various
+
+<!-- MH: What about spatial autocorrelation? -->
+
+<!-- MH: SpatialKDE, geoR, gstat, spdep, siland -->
+
+```{=tex}
 \begin{table}
@@
 \hline
-\multirow{5}{*}{Spatial data} & raster & \cite{Hijmans2019} \\
+\multirow{8}{*}{Spatial data} & raster & \cite{Hijmans2019} \\
 & terra & \cite{R-terra} \\
 & stars & \cite{Pebesma2019} \\
+& fasterize & \cite{Ross2020} \\
+& rasterDT & \cite{OBrien2020} \\
+& exactextractr & \cite{Baston2020} \\
 & sp & \cite{Pebesma2005,Bivand2013} \\
@@
 \hline
-\multirow{6}{*}{Spatial data download} & FedData & \cite{R-FedData} \\
+\multirow{6}{*}{Spatial data download} & rnaturalearth & \cite{South2017} \\
 & elevatr & \cite{R-elevatr} \\
 & rgbif & \cite{R-rgbif} \\
-& rnaturalearth & \cite{South2017} \\
+& BIEN & \cite{Maitner2020} \\
 & marmap & \cite{Pante2013} \\
-& rgee & \cite{R-rgee} \\
+& FedData & \cite{R-FedData} \\
+& getlandsat & \cite{Chamberlain2018} \\
+& MODIS & \cite{Mattiuzzi2020} \\
 \hline
-\multirow{4}{*}{Creating maps} & tmap & \cite{Tennekes2018} \\
+\multirow{7}{*}{Creating maps} & rasterVis & \cite{Lamigueiro2020} \\
 & mapview & \cite{R-mapview} \\
+& tmap & \cite{Tennekes2018} \\
 & leaflet & \cite{R-leaflet} \\
+& cartography & \cite{Giraud2016} \\
 & ggspatial & \cite{Dunnington2020} \\
+& rayshader & \cite{Morgen-Wall2020} \\
 \hline
@@
 & motif & \cite{R-motif} \\
-& VLSM & xxx \\
+& geodiv & \cite{Smith2020} \\
 \hline
-\multirow{5}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & \cite{Hijmans2017} \\
+\multirow{8}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & \cite{Hijmans2017} \\
 & sdm & \cite{Naimi2016} \\
@@
 & PresenceAbsence & \cite{Freeman2008} \\
+& zoon & \cite{Golding2018} \\
+& adehabitatHR, adehabitatHS & \cite{Calenge2006} \\
+& amt & \cite{Signer2019} \\
 \hline
-\multirow{3}{*}{Connectivity} & grainscape & \cite{Chubaty2020} \\
+\multirow{5}{*}{Connectivity} & lconnect & \cite{Mestre2019} \\
+& Makurhini & \cite{Godinez-Gomez2020} \\
+& grainscape & \cite{Chubaty2020} \\
 & gdistance & \cite{vanEtten2017} \\
@@
 \hline
-\multirow{3}{*}{Landscape genetics} & raphs4lg & \cite{Savary2020} \\
+\multirow{4}{*}{Landscape genetics} & raphs4lg & \cite{Savary2020} \\
 & PopGenReport & \cite{Adamack2014,Gruber2015} \\
 & HierDpart & \cite{Qin2019} \\
+& GeNetIt & \cite{Murphy2010} \\
 \hline
@@
 \end{table}
-
+```
 # Survey of R usage by landscape ecology community
@@
 The *raster* package was mentioned the most, followed by the *sf* package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey) C).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig.
+\@ref(fig:fig-survey) C).
 Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r r_packages_df[[1, 2]]`%).
@@
 
-\FloatBarrier
-
-# Missing packages
+The survey also included a section in which participants could list methods and tools that are currently missing in *R* and answers to this question were very diverse.
 
@@
 
-The survey also included a section in which participants could list methods and tools that are currently missing in *R* and answers to this question were very diverse.
-
 Overall, `r round(length(id_na) / nrow(answers) * 100, digits)`% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
@@
 
-Since its first introduction in 1995, *R* has come a long way from a statistical programming language to a powerful tool for landscape ecology. 
+Since its first introduction in 1995, *R* has come a long way from an exclusively statistical programming language to a powerful tool for landscape ecology.
 Today, many *R* packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed for spatial data handling and analysis.
 The highly dynamic development of *R* packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since the scientific field of landscape ecology constantly changes and improves, also the *R* programming language and its packages need to change and adapt these changes.
+However, since the scientific field of landscape ecology constantly develops and improves, also the *R* programming language and its packages need to change and adapt these changes.
+
+## Acknowledgments
+
+#### Conflict of interest
+
+MHKH and JN are authors of the *landscapemetrics* and *landscapetools* package.
+JN is author of *belg* and *motif* package.
+
+#### Funding
+
+MHKH was supported by the German Research Association (DFG) Research Training Group 1644 'Scaling Problems in Statistics', grant number 152112243.
+<!-- MH: Add UMich Funding? -->
+
+#### Author contributions
+
+MHKH and JN designed the survey form and analyzed the responses of the participants.
+MHKH and JN drafted the manuscript and all authors contributed critically to the manuscript and gave final approval for publication.
+We used the 'sequence--determines--credit' approach (SDC) for the sequence of authors.
+
+#### Human and Animal Rights and Informed Consent
+
+This article does not contain any studies with human or animal subjects performed by any of the authors.

### 2020-12-15 paper/paper.tex
:pencil: Adding  packages addressing some comments section of survey

@@
 Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e.~package \emph{x} requires and uses code from package \emph{y} \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+Thus, most packages can be found on both platforms and many developers use \emph{GitHub} for daily developing and \emph{CRAN} to publish stable releases of the packages.
 For more information about \emph{R} package development, see \cite{Wickham2015}.
@@
 Alternatively, the \emph{stars} package can be used to read and process raster data with a focus on spatial-temporal data cubes \cite{Pebesma2019}.
+Additionally, there are packages that improve some basics raster operations in terms of computational performance or compatibility between raster and vector operations, such as \emph{fasterize} \cite{Ross2020}, \emph{rasterDT} \cite{OBrien2020}, or \emph{exactextractr} \cite{Baston2020}.
 
@@
 Since public available data is becoming more and more prominent, so are \emph{R} packages to accesses them.
-Packages include the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA, the \emph{elevatr} package to access raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}, the \emph{rgbif} to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, \emph{rnaturalearth} to access the Natural Earth database \cite{South2017}, or \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}.
-Furthermore, the \texttt{get\_data()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim v1.4, while the new in-development package \emph{geodata} provides an access to data from WorldClim v2.
+Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows to download data from the Landsat 8, while the \emph{MODIS} package \cite{Mattiuzzi2020} allows to download MODIS products.
+Also, the \texttt{get\_data()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
 Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
@@
 However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
+Advanced plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
 Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
@@
 Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
+Thematic maps, including besides others proportional symbols, chroropleth, or typology, can be created with the \emph{cartography} package \cite{Giraud2016}.
 Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
+
+A slightly different approach of visualizing spatial data is the \emph{rayshader} package \cite{Morgen-Wall2020} which allows to create topographical 2D and 3D maps.
 
@@
 \hypertarget{sec:landscape_metrics}{%
-\paragraph{Landscape metrics}\label{sec:landscape_metrics}}
+\paragraph{Quantify landscape characteristics}\label{sec:landscape_metrics}}
 
@@
 Additionally, the \emph{belg} packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
+
+More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
+The \emph{geodiv} package \cite{Smith2020} allows to gradient surface metrics to characterize continuous analysis of landscape features.
 
@@
 Of course, also packages specifically designed for SDM exists.
-This includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020} and \emph{PresenceAbsence} \cite{Freeman2008} packages.
+This includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008} or \emph{zoon} \cite{Golding2018} packages.
+Related to SDM, there are several packages to estimate home ranges and habitat selection, including \emph{adehabitatHR} and \emph{adehabitatHS} \cite{Calenge2006} or \emph{amt \cite{Signer2019}}
 
@@
 Such measures are provided within the \emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
+Furthermore, the \emph{lconnect} package \cite{Mestre2019} and \emph{Makurhini} package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
 Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}.
@@
 Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also of importance for landscape genetics.
-Further functionality for landscape genetics can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, or \emph{HierDpart} packages \cite{Qin2019}.
+Further functionality for landscape genetics can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.
 
@@
 Furthermore, the \emph{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields which could be used as neutral landscape models.
+
+\hypertarget{various}{%
+\paragraph{Various}\label{various}}
 
@@
 In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
-Other, less frequent answers, were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers, were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
@@
 
-We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'' and ``nature conservation'' (Fig.
+We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity'' and ``nature conservation'' (Fig.
 \ref{fig:fig-survey} A)).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
+Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey} C).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig.
+\ref{fig:fig-survey} C).
 Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 33.4415584\%).
 
-Lastly, when asked how use \emph{R} us currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
+Lastly, when asked how use \emph{R} us currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
 \ref{fig:fig-survey} D)).
@@
 
-\FloatBarrier
-
-\hypertarget{missing-packages}{%
-\section{Missing packages}\label{missing-packages}}
-
 The survey also included a section in which participants could list methods and tools that are currently missing in \emph{R} and answers to this question were very diverse.
@@
 
-Since its first introduction in 1995, \emph{R} has come a long way from a statistical programming language to a powerful tool for landscape ecology.
+Since its first introduction in 1995, \emph{R} has come a long way from an exclusively statistical programming language to a powerful tool for landscape ecology.
 Today, many \emph{R} packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed for spatial data handling and analysis.
 The highly dynamic development of \emph{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since the scientific field of landscape ecology constantly changes and improves, also the \emph{R} programming language and its packages need to change and adapt these changes.
+However, since the scientific field of landscape ecology constantly develops and improves, also the \emph{R} programming language and its packages need to change and adapt these changes.
+
+\hypertarget{acknowledgments}{%
+\subsection{Acknowledgments}\label{acknowledgments}}
+
+\hypertarget{conflict-of-interest}{%
+\paragraph{Conflict of interest}\label{conflict-of-interest}}
+
+MHKH and JN are authors of the \emph{landscapemetrics} and \emph{landscapetools} package.
+JN is author of \emph{belg} and \emph{motif} package.
+
+\hypertarget{funding}{%
+\paragraph{Funding}\label{funding}}
+
+MHKH was supported by the German Research Association (DFG) Research Training Group 1644 `Scaling Problems in Statistics,' grant number 152112243.
+
+\hypertarget{author-contributions}{%
+\paragraph{Author contributions}\label{author-contributions}}
+
+MHKH and JN designed the survey form and analyzed the responses of the participants.
+MHKH and JN drafted the manuscript and all authors contributed critically to the manuscript and gave final approval for publication.
+We used the `sequence--determines--credit' approach (SDC) for the sequence of authors.
+
+\hypertarget{human-and-animal-rights-and-informed-consent}{%
+\paragraph{Human and Animal Rights and Informed Consent}\label{human-and-animal-rights-and-informed-consent}}
+
+This article does not contain any studies with human or animal subjects performed by any of the authors.
 

### 2020-12-15 R/cran_logs.R

@@ -1,0 +1,30 @@
+library(cranlogs)
+library(devtools)
+library(tidyverse)
+
+# raster #
+revrep_raster <- devtools::revdep(pkg = "raster")
+
+download_ras <- cranlogs::cran_downloads(packages = revrep_raster,
+                                         from = "2012-10-01", to = Sys.Date()) %>%
+    dplyr::group_by(package) %>%
+    dplyr::summarise(count = sum(count)) %>%
+    dplyr::arrange(-count)
+
+# sf #
+revrep_sf <- devtools::revdep(pkg = "sf")
+
+download_sf <- cranlogs::cran_downloads(packages = revrep_sf,
+                                        from = "2012-10-01", to = Sys.Date()) %>%
+    dplyr::group_by(package) %>%
+    dplyr::summarise(count = sum(count)) %>%
+    dplyr::arrange(-count)
+
+# sp #
+revrep_sp <- devtools::revdep(pkg = "sp")
+
+download_sp <- cranlogs::cran_downloads(packages = revrep_sp,
+                                        from = "2012-10-01", to = Sys.Date()) %>%
+    dplyr::group_by(package) %>%
+    dplyr::summarise(count = sum(count)) %>%
+    dplyr::arrange(-count)

### 2020-12-16 paper/paper.tex
Add Bivand 2020 Paper

@@
 The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
+A recent overview over the progress of \emph{R} to handle spatial data in general can be found e.g.~here \cite{Bivand2020}.
 
@@
 In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
-Other, less frequent answers, were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers, were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
@@
 
-We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity'' and ``nature conservation'' (Fig.
+We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'' and ``nature conservation'' (Fig.
 \ref{fig:fig-survey} A)).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
+Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 
-Lastly, when asked how use \emph{R} us currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
+Lastly, when asked how use \emph{R} us currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
 \ref{fig:fig-survey} D)).
@@
 
-MHKH was supported by the German Research Association (DFG) Research Training Group 1644 `Scaling Problems in Statistics,' grant number 152112243.
+MHKH was supported by the German Research Association (DFG) Research Training Group 1644 `Scaling Problems in Statistics', grant number 152112243.
 

### 2020-12-20 paper/paper.Rmd
Add Bivand 2020 Paper
updates title
improves 1st section

@@
 ---
-title: Open source landscape ecology tools
+title: Open-source tools in R for landscape ecology
 # subtitle: Do you have a subtitle? If so, write it here
@@
 
-Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
-In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
+<!-- JN: my first thought was - "do we even need this kind of introduction in the 'Current Landscape Ecology Reports' journal"; however, after a while, I like it a lot.
+It could be very helpful for people from other fields.
+-->
 
-While human activities have altered the landscapes for millenniums \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans on landscapes have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}.
+Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
+In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
+
+While human activities have altered the landscapes for millenniums \cite{Ellis2011,Ellis2015}, in the past centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
 Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
@@
 
-Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}.
+Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes vary spatially \cite{With2019}.
 To this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}.
-Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}.
+Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984}.
 Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze, and visualize spatial data.
@@
 Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
-Open-source software includes all software which is released under licenses that allow to freely use, modify and distribute the software \cite{St.Laurent2008}.
-Open-source software development has many advantages, such as fast innovation, transparency, and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
+Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
+Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
 Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
+<!--JN: I like the second argument above. We can also add that the use of open-source software democratize scientific research (as opposed to the closed+expensive software) -->
 Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
-Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
 
-One successful example of an open-source project is the *R* programming language and its extensions (also called packages) \cite{RCoreTeam2019}.
-Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
-Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
-This is documented by the growing *CRAN Task Views: Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total.
+One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}.
+Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
+<!--language is older (1991?), but in 1995 it was officially released for the first time-->
+Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
+The expanding CRAN Task Views document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
 Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even higher.
-The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
+The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
+<!--JN: should we cite Pebesma2019a as it is a work in progress? -->
+<!--JN: also - maybe we should cite http://book.ecosens.org/RSEbook/?-->
+A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.
 
-The growing body of *R* packages related to spatial data processing and analyzing results in a high capability of the *R* programming language for landscape ecology.
-Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on software implemented in the *R* programming language.
-For more general overviews of open-source software for landscape ecology see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Because additionally to handling spatial data, also further tasks such as statistical modeling, preparation of publication-ready figures and even preparation of complete reports can be done in the same working environment (Fig.
+The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
+Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review, we focus on software implemented in the *R* programming language.
+For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Because in addition to handling spatial data, also tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
 \@ref(fig:fig-workflow)), the *R* programming language can be a very powerful tool.
-Thus, in the first part we present an overview over existing *R* packages for landscape ecology (Table \@ref(tab:packages)).
-In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are currently missing for landscape ecology.
+Thus, in the first part, we present an overview of existing *R* packages for landscape ecology (Table \@ref(tab:packages)).
+In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are presently missing for landscape ecology.
 

### 2020-12-20 paper/paper.Rmd
improves Existing packages

@@
 
-Since most *R* packages are developed and maintained by the community, this shows how open-source software development can facilitate innovation, reproducibility and reuse of code.
-There are two major online platforms to host *R* packages and make them accessible to potential users.
-Firstly, *GitHub* allows to host source code under version control and users can install packages with one line using the *remotes* \cite{Hester2020} package.
-Additionally, hosting the package on *GitHub* provides many useful features to collaborate and communicate between developers as well as users \cite{Wickham2015}, or integrated unit testing.
-Secondly, the *Comprehensive R Archive Network* (*CRAN*) provides a large visibility to the community, ease of installation and a technical quality standard, including checks for common problems on all major operating systems.
-\cite{Wickham2015}.
-For developers, the guaranteed technical quality standard on *CRAN* requires more initial work for compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
-Additionally, the technical quality standard on *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e.
-package *x* requires and uses code from package *y* \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
-Thus, most packages can be found on both platforms and many developers use *GitHub* for daily developing and *CRAN* to publish stable releases of the packages.
+Most *R* packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
+There are three major online platforms to host *R* packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
+The last one focuses on tools for the analysis of genomic data; therefore, we focus on only the former two in this review.
+CRAN (the *Comprehensive R Archive Network*) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
+*GitHub* hosts source code under version control, and allows users to install packages with one line of code using the *remotes* \cite{Hester2020} package.
+Additionally, hosting the package on *GitHub* provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing.
+
+The guaranteed technical quality standard on *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
+Additionally, the technical quality standard on *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y* \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
 For more information about *R* package development, see \cite{Wickham2015}.
 
-```{r fig-workflow, echo = FALSE, out.width="100%", out.height="30%", fig.cap = "Exemplary worklow of spatial data analyses for landscape ecology using the R programming language.
+```{r fig-workflow, echo = FALSE, out.width="100%", out.height="30%", fig.cap = "Exemplary workflow of spatial data analyses for landscape ecology using the R programming language.
 For all major tasks, a few example R packages are listed.
-Of course, many more packages for the specific tasks exist."}
+."}
 include_graphics("data/Workflow.png")
 ```
+
+<!--JN: the above figure is nice; we should update it when everything else is done-->
 

### 2020-12-20 paper/paper.Rmd
improves spatial data representation

@@
 
-While *R* has several build-in data structures, including vectors, matrices or lists, the base *R* programming language has no internal support for reading, processing, or visualizing spatial data.
-However, because there is a notable interest in spatial data analysis, support for spatial data is provided by many *R* packages (\cite{Lovelace2019}, page 10).
-Most spatial data belong to one of two contrasting data models, namely spatial raster and spatial vectors, and both data models have several implementations in the *R* language.
-Importantly, basically all *R* packages for spatial data use the external *GDAL* and *PROJ* libraries, which allow for e.g.
-reading and writing of hundreds of spatial data formats, and coordinates transformation.
-Additionally, *R*, *GDAL* and *PROJ* allow to convert between data models and specific implementations, which can be useful if given methods only exist for a specific data model or implementation.
+While *R* has several build-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10).
+Most spatial data belong to two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
+Importantly, main *R* packages for spatial data use the external *GDAL* and *PROJ* libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
+<!--JN: references??-->
+Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
 
@@
 The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
-The probably most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
-A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
-Alternatively, the *stars* package can be used to read and process raster data with a focus on spatial-temporal data cubes \cite{Pebesma2019}.
+Currently, the most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
+A *raster* successor, *terra*,  aimed at the simpler interface and improved performance is now being developed \cite{R-terra}; however, it could take several years for this package to be adopted by other developers and users.
+Alternatively, the *stars* package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
 Additionally, there are packages that improve some basics raster operations in terms of computational performance or compatibility between raster and vector operations, such as *fasterize* \cite{Ross2020}, *rasterDT* \cite{OBrien2020}, or *exactextractr* \cite{Baston2020}.
 
-The vector data model consists of two main elements i) geometries, such as points, lines, polygons, and ii) attributes and each geometry is connected to a value in an attribute table.
-In many cases, this data model allows a more natural representation of landscape features, however, with the cost of higher computational demands <!--(add reference)-->.
-The *sp* package was the standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
+The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
+In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands <!--(add reference)-->.
+The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
 As of 2020, more than 500 *R* packages directly depend or imports *sp*.
-However, more recently *sp* will only be maintained (and not actively developed any more), and the recommended successor is the improved *sf* package \cite{Pebesma2018}.
+However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
 

### 2020-12-20 paper/paper.Rmd
adds new icte + update data donwload

@@
 
-Nowadays, spatial data for many spatial scales is available from an abundance of online accessible sources.
-A lot of this data is publicly available, either as a direct download or through an API connection, and this can be used by several packages to download the spatial data directly into a *R* session.
-Since public available data is becoming more and more prominent, so are *R* packages to accesses them.
+Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
+A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
+Since publicly available data is becoming more prominent, so are *R* packages to access them.
 Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows to download data from the Landsat 8, while the *MODIS* package \cite{Mattiuzzi2020} allows to download MODIS products.
-Also, the `get_data()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the  in-development package *geodata* <!--(add reference)-->).
+Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows to download data from the Landsat 8, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download  Sentinel-2 optical images.
+Also, the `getData()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package *geodata* <!--(add reference) https://github.com/rspatial/geodata-->).
 Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
-
-<!-- JN: how to cite (or maybe we should just add a link to?) - https://github.com/tilezen/joerd/blob/master/docs/data-sources.md -->
-
-<!-- MH: I dont really get it. For which package do we need this link? -->
 

### 2020-12-20 paper/paper.Rmd
improves spataial data processing part 1

@@
 Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps <!-- (add reference) -->.
-This is also referred to as the projection of the data and is often a first barrier in spatial data analysis.
-It is not only required to have all of the used data in the same projection, but also selection of a proper projection is crucial.
-This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion and different CRS are optimized for different regions of the world and scales <!-- (add reference) -->.
-Coordinates in spatial data can represent one of many coordinate reference systems.
-Two main groups of CRS exist, namely geographical and projected, each having many members.
+This is also referred to as the spatial projection and is often the first barrier in spatial data analysis.
+It is not only required to have all of the used data in the same projection, but also to select a proper CRS.
+This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales <!-- (add reference) -->.
+Coordinates in spatial data represent one of many coordinate reference systems.
+Two main groups of CRS, namely geographical and projected, exist, with each having many members.
 In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
-However, in most landscape ecology studies, projected CRSs, which use some measurement units (e.g., meters), should be utilized.
-Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
+However, most landscape ecology studies should utilize projected CRSs, which use some measurement units (e.g., meters).
+The selection of projected CRS should be based on the property of spatial data that we want to keep intact (e.g., does not distort areas, shapes, distances, or angles) and be appropriate for a given study area.
 Tools to find an appropriate CRS for a certain region can be found at <https://spatialreference.org>, <https://epsg.org>, or <http://epsg.io>.
-In used data have different CRS, all packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations.
+All packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations, allowing to unify spatial projections when the used data have different CRS.
 
@@
 Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
-
-<!--JN: should we mention ArcGIS here and some of its modules?-->
-<!-- MH: I don't think this is really important -->
-
-<!--JN: in the same vain as above: should we mention Google Earth Engine?-->
-<!-- MH: Again, I dont think thats really of relevance. I think as it is now its ok  -->
 

### 2020-12-20 paper/paper.Rmd
improved data provessing

@@
 A second common spatial data processing task is required when the available data extends over a larger area than the study region.
-In this case, the pre-processing of spatial data can include vector clipping or raster cropping.
-Related to that, also masking of certain areas of the study region can be required by using e.g. certain spatial filters such as water bodies.
+In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
+Related to that, masking certain areas of the study region can be required by using, e.g., certain spatial filters such as water bodies.
 Packages from Section \@ref(sec:spatrep) also allow for these operations.
-Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications.
+Additionally, they offer many other operations, such as merging or joining spatial data, extracting values from one dataset into another, raster resolution changes, or vector data simplifications.
 A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the *grainchanger* package \cite{Graham2019}.
-Furthermore, *landscapetools* is a collection of various utility functions for the raster data model.
+Furthermore, *landscapetools* is a collection of various utility functions for the raster data model \cite{Sciaini2018}.
 

### 2020-12-20 paper/paper.Rmd
improves spatial viz

@@
 However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
-Advanced plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
-Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
+
 The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
+Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
+Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
+<!-- Advanced plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}. -->
+<!-- JN: I do not know if we should add rasterVis here... -->
+Quick interactive visualization of spatial data can be done with the *mapview* package\cite{R-mapview}.
 Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
-Thematic maps, including besides others proportional symbols, chroropleth, or typology, can be created with the *cartography* package \cite{Giraud2016}.
-Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
 
-A slightly different approach of visualizing spatial data is the *rayshader* package \cite{Morgen-Wall2020} which allows to create topographical 2D and 3D maps.
+A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
+
+<!--JN: idea - a figure comparing viz of several packages-->
 

### 2020-12-20 paper/paper.Rmd
improves Quantify landscape characteristics

@@
 
-One of the most fundamental steps of landscape ecology is to describe and quantify landscape characteristics \cite{Turner2005,Lausch2015}.
-For discrete land cover classes the composition (number and abundance) and configuration (spatial arrangement) of the landscape is often described using landscape metrics \cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}.
+One of the most fundamental steps of landscape ecology analyses is to describe and quantify landscape characteristics \cite{Turner2005,Lausch2015}.
+For discrete land cover classes, the composition (number and abundance) and configuration (spatial arrangement) of the landscape are often described using landscape metrics \cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}.
 These metrics allow to compare different landscapes, quantify temporal and spatial landscape changes and investigate interactions between landscape characteristics and ecological processes \cite{Uuemaa2009}.
 
-The use of landscape metrics was heavily facilitated by the introduction of the *FRAGSTATS* software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
-However, *FRAGSTATS* is not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
-To provide a truly open-source solution to calculate landscape metrics, recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed.
-The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment.
-Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
+The introduction of the *FRAGSTATS* in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019}.
+However, while *FRAGSTATS* is free to download at the developers' homepage, it is not open-source software. 
+This restricts its transparency, reproducibility, and extensibility.
+Recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed to provide a truly open-source solution to calculate landscape metrics.
+The package allows calculating the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment.
 
 More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
-The *geodiv* package \cite{Smith2020} allows to gradient surface metrics to characterize continuous analysis of landscape features.
+The *geodiv* package \cite{Smith2020} allows to calculate gradient surface metrics to characterize continuous analysis of landscape features. 
+Additionally, the *belg* package allows calculating the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
 
-Most of landscape metrics are a single number depicting specific characteristics of a local landscape.
+Most landscape metrics are a single number depicting specific characteristics of a local landscape.
 Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
@@
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
-Methods for pattern-based spatial analysis are implemented in the *motif* package \cite{R-motif}.
+All of the methods mentioned above are implemented in the *motif* package \cite{R-motif}.
 

### 2020-12-20 paper/paper.Rmd
improves Species distribution modeling

@@
 
-Species distribution modeling (SDM) analysis how landscape patterns (e.g.
-habitat suitability, resources availability, etc.) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}.
-Originated in the 1970s, SDM has experienced numerous methodological advancement and a numerous body of literature exists today \cite{Zimmermann2010}.
+Species distribution modeling (SDM) examine how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}.
+Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010}.
 Additionally, textbooks introducing basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2019}.
@@
 Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package.
-Of course, also packages specifically designed for SDM exists.
-This includes the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008} or *zoon* \cite{Golding2018} packages.
+Also, packages specifically designed for SDM exists, including includes the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages.
 Related to SDM, there are several packages to estimate home ranges and habitat selection, including *adehabitatHR* and *adehabitatHS* \cite{Calenge2006} or *amt \cite{Signer2019}*

### 2020-12-20 paper/paper.Rmd
improves Connectivity

@@
 While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
-Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}.
-On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g. amount of suitable habitat) \cite{Kindlmann2008,With2019}.
+Given its broad concept, many different measures of connectivity exist \cite{Kindlmann2008}.
+On patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
 Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}).
 Furthermore, the *lconnect* package \cite{Mestre2019} and *Makurhini* package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
-Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}.
-In graph theory \cite{Laita2011} landscapes are described by nodes (i.e. habitat patches) connected by and functional connections called links \cite{Laita2011}.
+Another way to describe connectivity is based on graph theory with the advantage that functional connectivity can also be included \cite{Kindlmann2008}.
+In graph theory \cite{Laita2011}, landscapes are described by nodes (i.e., habitat patches) connected by and functional connections called links (or edges) \cite{Laita2011}.
 The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
-More general, the *igraph* packages \cite{Csardi2006} provides functionality related to graph theory.
-Resistance surfaces and least-cost paths are another tool to model functional connectivity.
-The resistance surface describes effects of facilitating or hindering characteristics of the landscape for an organism moving within it \cite{Adriaensen2003}.
+More general, the *igraph* package \cite{Csardi2006} provides functionality related to graph theory.
+Resistance surfaces and least-cost paths are other tools to model functional connectivity.
+The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
 Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}.
-Absorbing Markov chains that combines random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published *samc* package \cite{Marx2020}.
+Absorbing Markov chains that combine random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published *samc* package \cite{Marx2020}.
 

### 2020-12-20 paper/paper.Rmd
adds a bunch of improv

@@
 
-Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift and selection \cite{Manel2003}.
-Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions and conservation \cite{Storfer2007}.
-By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
-As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.
+Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift, and selection \cite{Manel2003}.
+Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions, and conservation \cite{Storfer2007}.
+By explicitly including landscape characteristics, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
+As a result of its interdisciplinary, landscape genetics combines methods from different disciplines, including landscape ecology, spatial statistics, geography, and population genetics \cite{Storfer2007}.
 
-Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also of importance for landscape genetics.
+Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also important for landscape genetics.
 Further functionality for landscape genetics can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.
@@
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found across many different *R* packages.
-A comprehensive collections of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the *NLMR* package \cite{Sciaini2018}.
-Furthermore, the *RandomFields* package \cite{Schlather2015} allows to simulate Gaussian fields which could be used as neutral landscape models.
+Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across various *R* packages.
+A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the *NLMR* package \cite{Sciaini2018}.
+Furthermore, the *RandomFields* package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.
 
@@
 <!-- MH: What about spatial autocorrelation? -->
-
 <!-- MH: SpatialKDE, geoR, gstat, spdep, siland -->
@@
 & MODIS & \cite{Mattiuzzi2020} \\
+& sen2r & \cite{R-sen2r} \\
 \hline
-\multirow{7}{*}{Creating maps} & rasterVis & \cite{Lamigueiro2020} \\
-& mapview & \cite{R-mapview} \\
+\multirow{7}{*}{Creating maps} & mapview & \cite{R-mapview} \\
 & tmap & \cite{Tennekes2018} \\
@@
 
-To better understand how *R* is used by the landscape ecology community, we conducted a short survey and used mailing lists and social media to reach the community.
+To better understand how the landscape ecology community uses* R*, we conducted a short survey and used mailing lists and social media to reach the community.
 In total, the survey was answered by `r n_people` participants, of which the the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
-Other, less frequent answers, were "Data scientists", "None of the above", "Government employees" "Master's degree student" and "Bachelor's degree student" (in decreasing order).
+Other, less frequent answers were "Data scientists", "None of the above", "Government employees" "Master's degree student" and "Bachelor's degree student" (in decreasing order).
 
 Most people use *R* either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%).
-About half of all participants described themselves as "advanced" user (`r expertise[[1, 2]]`%), while `r expertise[[2, 2]]`% described themselves as "intermediate" users.
+Almost half of all participants described themselves as "advanced" users (`r expertise[[1, 2]]`%), while `r expertise[[2, 2]]`% described themselves as "intermediate" users.
 Related to this, about half of the participants either implemented their own package (`r own_package[[3, 2]]`%) or plan to do so in the future (`r own_package[[2, 2]]`%) and most of these packages are hosted on *GitHub* and/or *CRAN*.

### 2020-12-20 paper/paper.Rmd
adds fixes to survey and ideas for conclusions

@@
 
-We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10% of participants included "biodiversity", followed by "land use management", "landscape connectivity" and "nature conservation" (Fig.
+We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10% of participants included "biodiversity", followed by "land use management", "landscape connectivity", and "nature conservation" (Fig.
 \@ref(fig:fig-survey) A)).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, "(pre-)processing of data", "spatial statistics" and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey) B)).
+Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey) B)).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
@@
 The *raster* package was mentioned the most, followed by the *sf* package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig.
-\@ref(fig:fig-survey) C).
-Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r r_packages_df[[1, 2]]`%).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey) C).
+Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], 2)`%).
 
@@
 
-Lastly, when asked how use *R* us currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
+Lastly, when asked how useful *R* is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
 \@ref(fig:fig-survey) D)).
@@
 Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology.
-'others' includes all answers with less than five total mentions."}
+'others' includes all answers with less than five total mentions.", warning=FALSE}
 
@@
 
-The survey also included a section in which participants could list methods and tools that are currently missing in *R* and answers to this question were very diverse.
+The survey also included a section in which participants could list methods and tools currently missing in *R* and answers to this question were very diverse.
 
@@
 Firstly, many participants (`r perc_performance`%) wished for a better computational performance of *R* in terms of speed and required RAM, especially for larger data.
-Secondly, participant are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (`r perc_characterisation`%).
+Secondly, participants are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (`r perc_characterisation`%).
 Thirdly, many participants (`r perc_viz`%) are currently missing advanced and easy-to-apply methods to create high-quality maps or other visualization-related functionality.
@@
 
-Since its first introduction in 1995, *R* has come a long way from an exclusively statistical programming language to a powerful tool for landscape ecology.
-Today, many *R* packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed for spatial data handling and analysis.
+<!--intro-->
+Since its first introduction in 1995, *R* has come a long way from an exclusively statistical programming language to a powerful landscape ecology tool.
+Today, many *R* packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed at spatial data handling and analysis.
 The highly dynamic development of *R* packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since the scientific field of landscape ecology constantly develops and improves, also the *R* programming language and its packages need to change and adapt these changes.
+However, since landscape ecology constantly develops and improves, the *R* programming language and its packages need to change and adapt to these changes.
+
+<!--a paragraph summarizing possibilities-->
+
+<!--a paragraph summarizing the survey-->
+
+<!--a paragraph about missing features/future work-->
 

### 2020-12-20 paper/paper.Rmd
improves figures

@@
            stat = "identity") + 
-  theme_classic(base_size = 10) +
+  theme_classic(base_size = base_size) +
   theme(plot.title.position = "plot") +
-  scale_y_continuous(breaks = seq(0, 60, 10), limits = c(0, 60)) + 
+  # scale_x_continuous(breaks = seq(0, 60, 10), limits = c(0, 60)) +
   guides(fill = FALSE, colour = FALSE) +
   labs(x = NULL,
-       title = "How useful do you think is R currently for landscape ecology?", 
-       y = "Number of participants [%]")
+       title = "How useful do you think is R currently\nfor landscape ecology?", 
+       y = "Number of participants [%]") +
+  coord_flip()
 ```
@@
 Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology.
-'others' includes all answers with less than five total mentions.", warning=FALSE}
+The 'others' category includes all answers with less than five total mentions.", warning=FALSE, out.width="100%"}
 
-(ggplot_topics | ggplot_tasks | ggplot_packages ) / ggplot_usefulness + 
-  plot_layout(heights = c(2, 3)) +
+(ggplot_topics | ggplot_tasks) / (ggplot_packages | ggplot_usefulness) + 
+  # plot_layout(heights = c(2, 3)) +
   plot_annotation(tag_levels = "A", tag_suffix = ")")

### 2020-12-20 paper/paper.tex
improves figures

--- 2020-12-15 paper/paper.tex
@@
 
-\title{Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front
+\title{Open-source tools in R for landscape ecology \thanks{Grants or other notes about the article that should go on the front
 page should be placed here. General acknowledgments should be placed at the
@@
 
-Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
-In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
+Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
+In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
 
-While human activities have altered the landscapes for millenniums \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans on landscapes have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}.
+While human activities have altered the landscapes for millenniums \cite{Ellis2011,Ellis2015}, in the past centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
 Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
@@
 Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
-Open-source software includes all software which is released under licenses that allow to freely use, modify and distribute the software \cite{St.Laurent2008}.
-Open-source software development has many advantages, such as fast innovation, transparency, and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
+Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
+Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
 Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
 Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
-Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
 
-One successful example of an open-source project is the \emph{R} programming language and its extensions (also called packages) \cite{RCoreTeam2019}.
-Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
-Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006}.
-This is documented by the growing \emph{CRAN Task Views: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} that currently list almost 300 packages in total.
+One successful example of an open-source project is the \emph{R} programming language, and its extensions called packages \cite{RCoreTeam2019}.
+Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
+Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
+The expanding CRAN Task Views document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
 Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely even higher.
-The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
+The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
+A recent overview over the progress of \emph{R} to handle spatial data in general can be found in \cite{Bivand2020}.
 
-The growing body of \emph{R} packages related to spatial data processing and analyzing results in a high capability of the \emph{R} programming language for landscape ecology.
-Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on software implemented in the \emph{R} programming language.
-For more general overviews of open-source software for landscape ecology see e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Because additionally to handling spatial data, also further tasks such as statistical modeling, preparation of publication-ready figures and even preparation of complete reports can be done in the same working environment (Fig.
+The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
+Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review, we focus on software implemented in the \emph{R} programming language.
+For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+Because in addition to handling spatial data, also tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
 \ref{fig:fig-workflow}), the \emph{R} programming language can be a very powerful tool.
-Thus, in the first part we present an overview over existing \emph{R} packages for landscape ecology (Table \ref{tab:packages}).
-In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are currently missing for landscape ecology.
+Thus, in the first part, we present an overview of existing \emph{R} packages for landscape ecology (Table \ref{tab:packages}).
+In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are presently missing for landscape ecology.
 
@@
 
-Since most \emph{R} packages are developed and maintained by the community, this shows how open-source software development can facilitate innovation, reproducibility and reuse of code.
-There are two major online platforms to host \emph{R} packages and make them accessible to potential users.
-Firstly, \emph{GitHub} allows to host source code under version control and users can install packages with one line using the \emph{remotes} \cite{Hester2020} package.
-Additionally, hosting the package on \emph{GitHub} provides many useful features to collaborate and communicate between developers as well as users \cite{Wickham2015}, or integrated unit testing.
-Secondly, the \emph{Comprehensive R Archive Network} (\emph{CRAN}) provides a large visibility to the community, ease of installation and a technical quality standard, including checks for common problems on all major operating systems.
-\cite{Wickham2015}.
-For developers, the guaranteed technical quality standard on \emph{CRAN} requires more initial work for compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
-Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e.~package \emph{x} requires and uses code from package \emph{y} \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
-Thus, most packages can be found on both platforms and many developers use \emph{GitHub} for daily developing and \emph{CRAN} to publish stable releases of the packages.
+Most \emph{R} packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
+There are three major online platforms to host \emph{R} packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
+The last one focuses on tools for the analysis of genomic data; therefore, we focus on only the former two in this review.
+CRAN (the \emph{Comprehensive R Archive Network}) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
+\emph{GitHub} hosts source code under version control, and allows users to install packages with one line of code using the \emph{remotes} \cite{Hester2020} package.
+Additionally, hosting the package on \emph{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing.
+
+The guaranteed technical quality standard on \emph{CRAN} requires more initial work for developers compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
+Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y} \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+Thus, most packages can be found on both platforms, and many developers use \emph{GitHub} for regular development and \emph{CRAN} to publish stable releases of the packages.
 For more information about \emph{R} package development, see \cite{Wickham2015}.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth,height=0.3\textheight]{data/Workflow} \caption{Exemplary worklow of spatial data analyses for landscape ecology using the R programming language.
+\includegraphics[width=1\linewidth,height=0.3\textheight]{data/Workflow} \caption{Exemplary workflow of spatial data analyses for landscape ecology using the R programming language.
 For all major tasks, a few example R packages are listed.
-Of course, many more packages for the specific tasks exist.}\label{fig:fig-workflow}
+.}\label{fig:fig-workflow}
 \end{figure}
@@
 
-While \emph{R} has several build-in data structures, including vectors, matrices or lists, the base \emph{R} programming language has no internal support for reading, processing, or visualizing spatial data.
-However, because there is a notable interest in spatial data analysis, support for spatial data is provided by many \emph{R} packages (\cite{Lovelace2019}, page 10).
-Most spatial data belong to one of two contrasting data models, namely spatial raster and spatial vectors, and both data models have several implementations in the \emph{R} language.
-Importantly, basically all \emph{R} packages for spatial data use the external \emph{GDAL} and \emph{PROJ} libraries, which allow for e.g.~reading and writing of hundreds of spatial data formats, and coordinates transformation.
-Additionally, \emph{R}, \emph{GDAL} and \emph{PROJ} allow to convert between data models and specific implementations, which can be useful if given methods only exist for a specific data model or implementation.
+While \emph{R} has several build-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10).
+Most spatial data belong to two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
+Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} and \emph{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Additionally, \emph{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
 
@@
 The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
-The probably most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
-A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
-Alternatively, the \emph{stars} package can be used to read and process raster data with a focus on spatial-temporal data cubes \cite{Pebesma2019}.
+Currently, the most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
+A \emph{raster} successor, \emph{terra}, aimed at the simpler interface and improved performance is now being developed \cite{R-terra}; however, it could take several years for this package to be adopted by other developers and users.
+Alternatively, the \emph{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
 Additionally, there are packages that improve some basics raster operations in terms of computational performance or compatibility between raster and vector operations, such as \emph{fasterize} \cite{Ross2020}, \emph{rasterDT} \cite{OBrien2020}, or \emph{exactextractr} \cite{Baston2020}.
 
-The vector data model consists of two main elements i) geometries, such as points, lines, polygons, and ii) attributes and each geometry is connected to a value in an attribute table.
-In many cases, this data model allows a more natural representation of landscape features, however, with the cost of higher computational demands .
-The \emph{sp} package was the standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
+The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
+In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands .
+The \emph{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
 As of 2020, more than 500 \emph{R} packages directly depend or imports \emph{sp}.
-However, more recently \emph{sp} will only be maintained (and not actively developed any more), and the recommended successor is the improved \emph{sf} package \cite{Pebesma2018}.
+However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
 
@@
 
-Nowadays, spatial data for many spatial scales is available from an abundance of online accessible sources.
-A lot of this data is publicly available, either as a direct download or through an API connection, and this can be used by several packages to download the spatial data directly into a \emph{R} session.
-Since public available data is becoming more and more prominent, so are \emph{R} packages to accesses them.
+Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
+A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
+Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
 Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows to download data from the Landsat 8, while the \emph{MODIS} package \cite{Mattiuzzi2020} allows to download MODIS products.
-Also, the \texttt{get\_data()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
+Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows to download data from the Landsat 8, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
+Also, the \texttt{getData()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
 Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
@@
 Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps .
-This is also referred to as the projection of the data and is often a first barrier in spatial data analysis.
-It is not only required to have all of the used data in the same projection, but also selection of a proper projection is crucial.
-This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion and different CRS are optimized for different regions of the world and scales .
-Coordinates in spatial data can represent one of many coordinate reference systems.
-Two main groups of CRS exist, namely geographical and projected, each having many members.
+This is also referred to as the spatial projection and is often the first barrier in spatial data analysis.
+It is not only required to have all of the used data in the same projection, but also to select a proper CRS.
+This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales .
+Coordinates in spatial data represent one of many coordinate reference systems.
+Two main groups of CRS, namely geographical and projected, exist, with each having many members.
 In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
-However, in most landscape ecology studies, projected CRSs, which use some measurement units (e.g., meters), should be utilized.
-Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
+However, most landscape ecology studies should utilize projected CRSs, which use some measurement units (e.g., meters).
+The selection of projected CRS should be based on the property of spatial data that we want to keep intact (e.g., does not distort areas, shapes, distances, or angles) and be appropriate for a given study area.
 Tools to find an appropriate CRS for a certain region can be found at \url{https://spatialreference.org}, \url{https://epsg.org}, or \url{http://epsg.io}.
-In used data have different CRS, all packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations.
+All packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations, allowing to unify spatial projections when the used data have different CRS.
 
 A second common spatial data processing task is required when the available data extends over a larger area than the study region.
-In this case, the pre-processing of spatial data can include vector clipping or raster cropping.
-Related to that, also masking of certain areas of the study region can be required by using e.g.~certain spatial filters such as water bodies.
+In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
+Related to that, masking certain areas of the study region can be required by using, e.g., certain spatial filters such as water bodies.
 Packages from Section \ref{sec:spatrep} also allow for these operations.
-Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications.
+Additionally, they offer many other operations, such as merging or joining spatial data, extracting values from one dataset into another, raster resolution changes, or vector data simplifications.
 A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the \emph{grainchanger} package \cite{Graham2019}.
-Furthermore, \emph{landscapetools} is a collection of various utility functions for the raster data model.
+Furthermore, \emph{landscapetools} is a collection of various utility functions for the raster data model \cite{Sciaini2018}.
 
@@
 However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
-Advanced plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
-Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
+
 The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
+Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
+Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
+Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{R-mapview}.
 Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
-Thematic maps, including besides others proportional symbols, chroropleth, or typology, can be created with the \emph{cartography} package \cite{Giraud2016}.
-Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
 
-A slightly different approach of visualizing spatial data is the \emph{rayshader} package \cite{Morgen-Wall2020} which allows to create topographical 2D and 3D maps.
+A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
 
@@
 
-One of the most fundamental steps of landscape ecology is to describe and quantify landscape characteristics \cite{Turner2005,Lausch2015}.
-For discrete land cover classes the composition (number and abundance) and configuration (spatial arrangement) of the landscape is often described using landscape metrics \cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}.
+One of the most fundamental steps of landscape ecology analyses is to describe and quantify landscape characteristics \cite{Turner2005,Lausch2015}.
+For discrete land cover classes, the composition (number and abundance) and configuration (spatial arrangement) of the landscape are often described using landscape metrics \cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}.
 These metrics allow to compare different landscapes, quantify temporal and spatial landscape changes and investigate interactions between landscape characteristics and ecological processes \cite{Uuemaa2009}.
 
-The use of landscape metrics was heavily facilitated by the introduction of the \emph{FRAGSTATS} software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
-However, \emph{FRAGSTATS} is not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
-To provide a truly open-source solution to calculate landscape metrics, recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was developed.
-The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the \emph{R} environment.
-Additionally, the \emph{belg} packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
+The introduction of the \emph{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019}.
+However, while \emph{FRAGSTATS} is free to download at the developers' homepage, it is not open-source software.
+This restricts its transparency, reproducibility, and extensibility.
+Recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was developed to provide a truly open-source solution to calculate landscape metrics.
+The package allows calculating the most widely used landscape metrics in a transparent and reproducible workflow within the \emph{R} environment.
 
 More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
-The \emph{geodiv} package \cite{Smith2020} allows to gradient surface metrics to characterize continuous analysis of landscape features.
+The \emph{geodiv} package \cite{Smith2020} allows to calculate gradient surface metrics to characterize continuous analysis of landscape features.
+Additionally, the \emph{belg} package allows calculating the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
 
-Most of landscape metrics are a single number depicting specific characteristics of a local landscape.
+Most landscape metrics are a single number depicting specific characteristics of a local landscape.
 Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
@@
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
-Methods for pattern-based spatial analysis are implemented in the \emph{motif} package \cite{R-motif}.
+All of the methods mentioned above are implemented in the \emph{motif} package \cite{R-motif}.
 
@@
 
-Species distribution modeling (SDM) analysis how landscape patterns (e.g.~habitat suitability, resources availability, etc.) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}.
-Originated in the 1970s, SDM has experienced numerous methodological advancement and a numerous body of literature exists today \cite{Zimmermann2010}.
+Species distribution modeling (SDM) examine how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}.
+Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010}.
 Additionally, textbooks introducing basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2019}.
@@
 Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
-Of course, also packages specifically designed for SDM exists.
-This includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008} or \emph{zoon} \cite{Golding2018} packages.
+Also, packages specifically designed for SDM exists, including includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008}, or \emph{zoon} \cite{Golding2018} packages.
 Related to SDM, there are several packages to estimate home ranges and habitat selection, including \emph{adehabitatHR} and \emph{adehabitatHS} \cite{Calenge2006} or \emph{amt \cite{Signer2019}}
@@
 While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
-Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}.
-On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g.~amount of suitable habitat) \cite{Kindlmann2008,With2019}.
+Given its broad concept, many different measures of connectivity exist \cite{Kindlmann2008}.
+On patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
 Such measures are provided within the \emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
 Furthermore, the \emph{lconnect} package \cite{Mestre2019} and \emph{Makurhini} package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
-Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}.
-In graph theory \cite{Laita2011} landscapes are described by nodes (i.e.~habitat patches) connected by and functional connections called links \cite{Laita2011}.
+Another way to describe connectivity is based on graph theory with the advantage that functional connectivity can also be included \cite{Kindlmann2008}.
+In graph theory \cite{Laita2011}, landscapes are described by nodes (i.e., habitat patches) connected by and functional connections called links (or edges) \cite{Laita2011}.
 The \emph{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
-More general, the \emph{igraph} packages \cite{Csardi2006} provides functionality related to graph theory.
-Resistance surfaces and least-cost paths are another tool to model functional connectivity.
-The resistance surface describes effects of facilitating or hindering characteristics of the landscape for an organism moving within it \cite{Adriaensen2003}.
+More general, the \emph{igraph} package \cite{Csardi2006} provides functionality related to graph theory.
+Resistance surfaces and least-cost paths are other tools to model functional connectivity.
+The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
 Least-cost paths can be calculated using the \emph{gdistance} package \cite{vanEtten2017}.
-Absorbing Markov chains that combines random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published \emph{samc} package \cite{Marx2020}.
+Absorbing Markov chains that combine random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published \emph{samc} package \cite{Marx2020}.
 
@@
 
-Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift and selection \cite{Manel2003}.
-Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions and conservation \cite{Storfer2007}.
-By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
-As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.
+Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift, and selection \cite{Manel2003}.
+Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions, and conservation \cite{Storfer2007}.
+By explicitly including landscape characteristics, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
+As a result of its interdisciplinary, landscape genetics combines methods from different disciplines, including landscape ecology, spatial statistics, geography, and population genetics \cite{Storfer2007}.
 
-Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also of importance for landscape genetics.
+Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also important for landscape genetics.
 Further functionality for landscape genetics can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.
@@
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found across many different \emph{R} packages.
-A comprehensive collections of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \emph{NLMR} package \cite{Sciaini2018}.
-Furthermore, the \emph{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields which could be used as neutral landscape models.
+Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across various \emph{R} packages.
+A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \emph{NLMR} package \cite{Sciaini2018}.
+Furthermore, the \emph{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.
 
@@
 
-To better understand how \emph{R} is used by the landscape ecology community, we conducted a short survey and used mailing lists and social media to reach the community.
+To better understand how the landscape ecology community uses* R*, we conducted a short survey and used mailing lists and social media to reach the community.
 In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
-Other, less frequent answers, were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
 Most people use \emph{R} either ``daily'' (54.4\%) or a ``few times a week'' (36.9\%).
-About half of all participants described themselves as ``advanced'' user (46.6\%), while 40.8\% described themselves as ``intermediate'' users.
+Almost half of all participants described themselves as ``advanced'' users (46.6\%), while 40.8\% described themselves as ``intermediate'' users.
 Related to this, about half of the participants either implemented their own package (21.4\%) or plan to do so in the future (23.3\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
 
-We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity'' and ``nature conservation'' (Fig.
+We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig.
 \ref{fig:fig-survey} A)).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
+Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig.
-\ref{fig:fig-survey} C).
-Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 33.4415584\%).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey} C).
+Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 33.44\%).
 
-Lastly, when asked how use \emph{R} us currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
+Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
 \ref{fig:fig-survey} D)).
-
-\begin{verbatim}
-## Warning: Removed 1 rows containing missing values (position_stack).
-\end{verbatim}
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-survey-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-survey-1}
 
@@
 
-The survey also included a section in which participants could list methods and tools that are currently missing in \emph{R} and answers to this question were very diverse.
+The survey also included a section in which participants could list methods and tools currently missing in \emph{R} and answers to this question were very diverse.
 
@@
 Firstly, many participants (13.6\%) wished for a better computational performance of \emph{R} in terms of speed and required RAM, especially for larger data.
-Secondly, participant are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (9.7\%).
+Secondly, participants are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (9.7\%).
 Thirdly, many participants (8.7\%) are currently missing advanced and easy-to-apply methods to create high-quality maps or other visualization-related functionality.
@@
 
-Since its first introduction in 1995, \emph{R} has come a long way from an exclusively statistical programming language to a powerful tool for landscape ecology.
-Today, many \emph{R} packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed for spatial data handling and analysis.
+Since its first introduction in 1995, \emph{R} has come a long way from an exclusively statistical programming language to a powerful landscape ecology tool.
+Today, many \emph{R} packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed at spatial data handling and analysis.
 The highly dynamic development of \emph{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since the scientific field of landscape ecology constantly develops and improves, also the \emph{R} programming language and its packages need to change and adapt these changes.
+However, since landscape ecology constantly develops and improves, the \emph{R} programming language and its packages need to change and adapt to these changes.
 

### 2020-12-21 paper/paper.Rmd
:pencil: Address some comments

@@
 -->
+<!-- MH: Yeah, I kind of like it.
+It sets the stage I think.
+Also, I hope the paper will be read by people that are not really familiar with landscape ecology as a field and just came across some spatial data -->
 
@@
 Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
-<!--JN: I like the second argument above. We can also add that the use of open-source software democratize scientific research (as opposed to the closed+expensive software) -->
 Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
 Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+This democratize scientific research as free-of-cost software removes one gatekeeper for researches without access to proprietary software.
 
@@
 Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
-<!--language is older (1991?), but in 1995 it was officially released for the first time-->
+<!--JN: language is older (1991?), but in 1995 it was officially released for the first time-->
+<!-- MH: Since we write "released" I think it's ok to use 1995? Especially since this was most likely the time a broader audience used it? -->
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
@@
 <!--JN: should we cite Pebesma2019a as it is a work in progress? -->
+<!-- MH: I think this does not speak against citing it, but also happy to remove it -->
 <!--JN: also - maybe we should cite http://book.ecosens.org/RSEbook/?-->
+<!-- MH: I dont know it, but sure if it fits -->
 A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.
@@
 <!-- JN: terra new vector representations -->
-
 <!-- MH: I probably would not add terra here.
@@
 <!--JN: idea - a figure comparing viz of several packages-->
+<!-- MH: I like the idea: Maybe one base plot, one fancy map (tmap) and a screenshot of an interactive map? -->
 

### 2020-12-21 paper/paper.tex
:pencil: Address some comments

@@
 Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+This democratize scientific research as free-of-cost software removes one gatekeeper for researches without access to proprietary software.
 
@@
 In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
-Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
@@
 
-We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig.
+We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'', and ``nature conservation'' (Fig.
 \ref{fig:fig-survey} A)).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
+Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 
-Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
+Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
 \ref{fig:fig-survey} D)).
@@
 Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology.
-'others' includes all answers with less than five total mentions.}\label{fig:fig-survey}
+The 'others' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
 \end{figure}

### 2021-01-04 paper/paper.Rmd
Some small changes in the ms.

@@
 *GitHub* hosts source code under version control, and allows users to install packages with one line of code using the *remotes* \cite{Hester2020} package.
-Additionally, hosting the package on *GitHub* provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing.
+Additionally, hosting a package on *GitHub* provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing.
 
@@
 Additionally, the technical quality standard on *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y* \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability of code.
 Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
@@
 However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10).
-Most spatial data belong to two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
+Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
 Importantly, main *R* packages for spatial data use the external *GDAL* and *PROJ* libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.

### 2021-01-04 paper/paper.Rmd
Adding authors, keywords and ROpenSci

@@
 - name: Maximillian H.K. Hesselbarth
-  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
+  address: 
+    Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
+    Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
   email: mhk.hesselbarth@gmail.com
@@
   email: nowosad.jakub@gmail.com
-- name: ...
-  address: ...
-  email: ...
+- name: Johannes M. Signer
+  address: Department of Wildlife Sciences, University of Göttingen, Büsgenweg 3, 37077 Göttingen, Germany
+  email: j.m.signer@gmail.com
+- name: Laura J. Graham
+  address: Geography and Environment, University of Southampton, Highfield Rd, Southampton SO17 1BJ, United Kingdom
+  email: l.graham@bham.ac.uk
 keywords:
-- keyword 1
-- keyword 2
-- keyword 3
-- keyword 4
+- spatial data
+- statistical programming language
+- R packages
+- reproducibility
 - ...
@@
 Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
+Furthermore, communities like *rOpenSci* also provide a peer review process for code quality.
 For more information about *R* package development, see \cite{Wickham2015}.

### 2021-01-04 paper/paper.tex
Adding authors, keywords and ROpenSci

@@
 
-\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  \ldots{} \and  }
+\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Johannes M. Signer \and  Laura J. Graham \and  }
 
@@
         Maximillian H.K. Hesselbarth \at
-     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany \\
+     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA \\
      \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
@@
     \and
-        \ldots{} \at
-     \ldots{} \\
-     \email{\ldots{}}  %  \\
+        Johannes M. Signer \at
+     Department of Wildlife Sciences, University of Göttingen, Büsgenweg 3, 37077 Göttingen, Germany \\
+     \email{\href{mailto:j.m.signer@gmail.com}{\nolinkurl{j.m.signer@gmail.com}}}  %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+        Laura J. Graham \at
+     Geography and Environment, University of Southampton, Highfield Rd, Southampton SO17 1BJ, United Kingdom \\
+     \email{\href{mailto:l.graham@bham.ac.uk}{\nolinkurl{l.graham@bham.ac.uk}}}  %  \\
 %             \emph{Present address:} of F. Author  %  if needed
@@
 \keywords{
-        keyword 1 \and
-        keyword 2 \and
-        keyword 3 \and
-        keyword 4 \and
+        spatial data \and
+        statistical programming language \and
+        R packages \and
+        reproducibility \and
         \ldots{} \and
@@
 \emph{GitHub} hosts source code under version control, and allows users to install packages with one line of code using the \emph{remotes} \cite{Hester2020} package.
-Additionally, hosting the package on \emph{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing.
+Additionally, hosting a package on \emph{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing.
 
@@
 Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y} \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability of code.
 Thus, most packages can be found on both platforms, and many developers use \emph{GitHub} for regular development and \emph{CRAN} to publish stable releases of the packages.
+Furthermore, communities like \emph{rOpenSci} also provide a peer review process for code quality.
 For more information about \emph{R} package development, see \cite{Wickham2015}.
@@
 However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10).
-Most spatial data belong to two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
+Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
 Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} and \emph{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.

### 2021-01-04 paper/paper.Rmd
Add linenumbers (Modulo 5)

@@
 # authorrunning: Short form of author list if too long for running head
-thanks: | 
-    Grants or other notes about the article that should go on the front 
-    page should be placed here. General acknowledgments should be placed at the
-    end of the article.
+# thanks: |
+#    Grants or other notes about the article that should go on the front
+#    page should be placed here. General acknowledgments should be placed at the
+#    end of the article.
 authors: 
@@
   \usepackage{placeins}
+  \usepackage[modulo]{lineno}
 output:
@@
 ```
+
+\linenumbers
 

### 2021-01-05 paper/paper.Rmd
ljg edits

--- 2021-01-04 paper/paper.Rmd
@@
 - name: Laura J. Graham
-  address: Geography and Environment, University of Southampton, Highfield Rd, Southampton SO17 1BJ, United Kingdom
+  address: Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
   email: l.graham@bham.ac.uk
@@
 
-While human activities have altered the landscapes for millenniums \cite{Ellis2011,Ellis2015}, in the past centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
+While human activities have altered the landscapes for millennia \cite{Ellis2011,Ellis2015}, in the past centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
 Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
@@
 Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
-Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
-This democratize scientific research as free-of-cost software removes one gatekeeper for researches without access to proprietary software.
+Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows use of the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.
 
@@
 Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review, we focus on software implemented in the *R* programming language.
+<!-- LJG: Do we need something about python capabilities for landscape ecology too? -->
 For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Because in addition to handling spatial data, also tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
+Because in addition to handling spatial data, tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
 \@ref(fig:fig-workflow)), the *R* programming language can be a very powerful tool.
@@
 
+<!-- LJG: I think that we might need to add a caveat about R packages around the fact that it's continual update by the community can lead to breaking changes (see http://datacolada.org/95) -->
+
 ```{r fig-workflow, echo = FALSE, out.width="100%", out.height="30%", fig.cap = "Exemplary workflow of spatial data analyses for landscape ecology using the R programming language.
@@
 
-While *R* has several build-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+While *R* has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
 However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10).
@@
 Alternatively, the *stars* package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
-Additionally, there are packages that improve some basics raster operations in terms of computational performance or compatibility between raster and vector operations, such as *fasterize* \cite{Ross2020}, *rasterDT* \cite{OBrien2020}, or *exactextractr* \cite{Baston2020}.
+Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as *fasterize* \cite{Ross2020}, *rasterDT* \cite{OBrien2020}, or *exactextractr* \cite{Baston2020}.
 
@@
 Since publicly available data is becoming more prominent, so are *R* packages to access them.
-Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows to download data from the Landsat 8, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download  Sentinel-2 optical images.
-Also, the `getData()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package *geodata* <!--(add reference) https://github.com/rspatial/geodata-->).
+Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download Sentinel-2 optical images.
+Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package *geodata* <!--(add reference) https://github.com/rspatial/geodata-->).
 Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
@@
 Tools to find an appropriate CRS for a certain region can be found at <https://spatialreference.org>, <https://epsg.org>, or <http://epsg.io>.
-All packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations, allowing to unify spatial projections when the used data have different CRS.
+All packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations, allowing unification of spatial projections when the used data have different CRS.
 
@@
 In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
-Related to that, masking certain areas of the study region can be required by using, e.g., certain spatial filters such as water bodies.
+Related to that, masking certain areas of the study region using spatial filters (e.g. water bodies, urban areas) can be required.
 Packages from Section \@ref(sec:spatrep) also allow for these operations.
@@
 
-Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
+Finally, there are a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
 
@@
 <!-- JN: I do not know if we should add rasterVis here... -->
+<!-- LJG: out of interest why not? -->
+
 Quick interactive visualization of spatial data can be done with the *mapview* package\cite{R-mapview}.
@@
 <!-- MH: I like the idea: Maybe one base plot, one fancy map (tmap) and a screenshot of an interactive map? -->
+<!-- LJG: great idea - happy to be assigned some maps to make! -->
 
@@
 For discrete land cover classes, the composition (number and abundance) and configuration (spatial arrangement) of the landscape are often described using landscape metrics \cite{Gustafson1998,Uuemaa2009,Uuemaa2013,Gustafson2019}.
-These metrics allow to compare different landscapes, quantify temporal and spatial landscape changes and investigate interactions between landscape characteristics and ecological processes \cite{Uuemaa2009}.
+These metrics allow the comparison of different landscapes, quantification of temporal and spatial landscape changes and investigation of interactions between landscape characteristics and ecological processes \cite{Uuemaa2009}.
 
@@
 Recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed to provide a truly open-source solution to calculate landscape metrics.
-The package allows calculating the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment.
+The package allows calculation of the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment.
 
 More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
-The *geodiv* package \cite{Smith2020} allows to calculate gradient surface metrics to characterize continuous analysis of landscape features. 
-Additionally, the *belg* package allows calculating the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
+The *geodiv* package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
+Additionally, the *belg* package allows calculation of the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
 
-Most landscape metrics are a single number depicting specific characteristics of a local landscape.
+Most landscape metrics are a represented by a single number depicting specific characteristics of a local landscape.
 Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
-Spatial signatures calculated for many landscapes can be compared using one of the existing distance measures.
+Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
+<!-- LJG added some examples, but not 100% if they're the distances you're thinking of -->
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
@@
 
-Species distribution modeling (SDM) examine how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}.
+Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
 Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010}.
+<!-- LJG: I'd use a more up to date reference. How about Norberg et al. 2019? https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecm.1370 -->
 Additionally, textbooks introducing basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2019}.
@@
 Also, packages specifically designed for SDM exists, including includes the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages.
+<!-- These packages [check!] allow the user to create ensemble models of SDMs based on different underlying model structures such as Maxent, regression trees, and generalized linear or additive models.
+This is considered the gold standard in contemporary SDM \cite(Araujo2007).-->
 Related to SDM, there are several packages to estimate home ranges and habitat selection, including *adehabitatHR* and *adehabitatHS* \cite{Calenge2006} or *amt \cite{Signer2019}*
@@
 <!-- Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}. However, currently the *maxent* package is not available on *CRAN* anymore. -->
+<!-- LJG: maxent can be applied through biomod2 - I think it's worth mentioning because it's pretty much the most widely used SDM (whether this is a good thing or not is another question)-->
+<!-- LJG: Another thought - it's generally accepted within SDM that ensemble models are superior to using single models.
+I know that biomod and zoon facilitate this.
+Can have a look at the other ones to see if they do.
+Have added a potential sentence as a comment -->
 
@@
 Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
-While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
+While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species such as habitat associations and dispersal distances \cite{Tischendorf2000,With2019}.
 Given its broad concept, many different measures of connectivity exist \cite{Kindlmann2008}.
-On patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
+At the patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
 Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}).
@@
 The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
-More general, the *igraph* package \cite{Csardi2006} provides functionality related to graph theory.
-Resistance surfaces and least-cost paths are other tools to model functional connectivity.
+More generally, the *igraph* package \cite{Csardi2006} provides functionality for graph theoretic analyses.
+Resistance surfaces and least-cost paths are other tools to model functional connectivity which include attributes of the matrix.
 The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
 Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}.
-Absorbing Markov chains that combine random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published *samc* package \cite{Marx2020}.
+Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019a}, and is provided by the recently published *samc* package \cite{Marx2020}.
 
@@
 Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions, and conservation \cite{Storfer2007}.
-By explicitly including landscape characteristics, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
-As a result of its interdisciplinary, landscape genetics combines methods from different disciplines, including landscape ecology, spatial statistics, geography, and population genetics \cite{Storfer2007}.
+By explicitly including landscape characteristics, landscape genetics provides a more detailed analysis than more abstract concepts (e.g., metapopulation genetics) \cite{Holderegger2006}.
+As a result of its interdisciplinarity, landscape genetics draws together methods from multiple fields, including landscape ecology, spatial statistics, geography, and population genetics \cite{Storfer2007}.
 
 Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also important for landscape genetics.
-Further functionality for landscape genetics can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.
+<!-- LJG: this might be me forgetting how refs in markdown work, but should the ref above be to the connectivity section? -->
+Further functionality for landscape genetics such as ...
+<!-- LJG: I think a little more info needed here - happy to dig through the packages to add in, but thought I'd leave for now incase anyone knows off top of their heads -->can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.
 

### 2021-01-05 paper/paper.tex
Add linenumbers (Modulo 5)
ljg edits

@@
         Laura J. Graham \at
-     Geography and Environment, University of Southampton, Highfield Rd, Southampton SO17 1BJ, United Kingdom \\
+     Geography, Earth \& Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom \\
      \email{\href{mailto:l.graham@bham.ac.uk}{\nolinkurl{l.graham@bham.ac.uk}}}  %  \\
@@
 Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
-Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
-This democratize scientific research as free-of-cost software removes one gatekeeper for researches without access to proprietary software.
+Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows use of the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.
 
@@
 For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Because in addition to handling spatial data, also tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
+Because in addition to handling spatial data, tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
 \ref{fig:fig-workflow}), the \emph{R} programming language can be a very powerful tool.
@@
 Alternatively, the \emph{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
-Additionally, there are packages that improve some basics raster operations in terms of computational performance or compatibility between raster and vector operations, such as \emph{fasterize} \cite{Ross2020}, \emph{rasterDT} \cite{OBrien2020}, or \emph{exactextractr} \cite{Baston2020}.
+Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as \emph{fasterize} \cite{Ross2020}, \emph{rasterDT} \cite{OBrien2020}, or \emph{exactextractr} \cite{Baston2020}.
 
@@
 Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
-Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows to download data from the Landsat 8, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
-Also, the \texttt{getData()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
+Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
+Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
 Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
@@
 Tools to find an appropriate CRS for a certain region can be found at \url{https://spatialreference.org}, \url{https://epsg.org}, or \url{http://epsg.io}.
-All packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations, allowing to unify spatial projections when the used data have different CRS.
+All packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations, allowing unification of spatial projections when the used data have different CRS.
 
@@
 In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
-Related to that, masking certain areas of the study region can be required by using, e.g., certain spatial filters such as water bodies.
+Related to that, masking certain areas of the study region using spatial filters (e.g.~water bodies, urban areas) can be required.
 Packages from Section \ref{sec:spatrep} also allow for these operations.
@@
 
-Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
+Finally, there are a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
 Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.
@@
 Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
+
 Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{R-mapview}.
@@
 Recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was developed to provide a truly open-source solution to calculate landscape metrics.
-The package allows calculating the most widely used landscape metrics in a transparent and reproducible workflow within the \emph{R} environment.
+The package allows calculation of the most widely used landscape metrics in a transparent and reproducible workflow within the \emph{R} environment.
 
 More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
-The \emph{geodiv} package \cite{Smith2020} allows to calculate gradient surface metrics to characterize continuous analysis of landscape features.
-Additionally, the \emph{belg} package allows calculating the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
+The \emph{geodiv} package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
+Additionally, the \emph{belg} package allows calculation of the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
 
-Most landscape metrics are a single number depicting specific characteristics of a local landscape.
+Most landscape metrics are a represented by a single number depicting specific characteristics of a local landscape.
 Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
-Spatial signatures calculated for many landscapes can be compared using one of the existing distance measures.
+Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g.~Euclidian, Manhattan, Mahlanobis).
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
@@
 
-Species distribution modeling (SDM) examine how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}.
+Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
 Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010}.
@@
 Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
-While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
+While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species such as habitat associations and dispersal distances \cite{Tischendorf2000,With2019}.
 Given its broad concept, many different measures of connectivity exist \cite{Kindlmann2008}.
-On patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
+At the patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
 Such measures are provided within the \emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
@@
 The \emph{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
-More general, the \emph{igraph} package \cite{Csardi2006} provides functionality related to graph theory.
-Resistance surfaces and least-cost paths are other tools to model functional connectivity.
+More generally, the \emph{igraph} package \cite{Csardi2006} provides functionality for graph theoretic analyses.
+Resistance surfaces and least-cost paths are other tools to model functional connectivity which include attributes of the matrix.
 The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
 Least-cost paths can be calculated using the \emph{gdistance} package \cite{vanEtten2017}.
-Absorbing Markov chains that combine random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published \emph{samc} package \cite{Marx2020}.
+Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019a}, and is provided by the recently published \emph{samc} package \cite{Marx2020}.
 
@@
 Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions, and conservation \cite{Storfer2007}.
-By explicitly including landscape characteristics, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
-As a result of its interdisciplinary, landscape genetics combines methods from different disciplines, including landscape ecology, spatial statistics, geography, and population genetics \cite{Storfer2007}.
+By explicitly including landscape characteristics, landscape genetics provides a more detailed analysis than more abstract concepts (e.g., metapopulation genetics) \cite{Holderegger2006}.
+As a result of its interdisciplinarity, landscape genetics draws together methods from multiple fields, including landscape ecology, spatial statistics, geography, and population genetics \cite{Storfer2007}.
 
 Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also important for landscape genetics.
-Further functionality for landscape genetics can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.
+Further functionality for landscape genetics such as \ldots{} can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.
 

### 2021-01-05 paper/paper.Rmd

@@
 # authorrunning: Short form of author list if too long for running head
-thanks: |
-    Grants or other notes about the article that should go on the front
-    page should be placed here. General acknowledgments should be placed at the
-    end of the article.
+# thanks: |
+#    Grants or other notes about the article that should go on the front
+#    page should be placed here. General acknowledgments should be placed at the
+#    end of the article.
 authors: 

### 2021-01-05 paper/paper.Rmd
:pencil: Address some comments within MS

@@
 
-<!-- JN: my first thought was - "do we even need this kind of introduction in the 'Current Landscape Ecology Reports' journal"; however, after a while, I like it a lot.
-It could be very helpful for people from other fields.
--->
-<!-- MH: Yeah, I kind of like it.
-It sets the stage I think.
-Also, I hope the paper will be read by people that are not really familiar with landscape ecology as a field and just came across some spatial data -->
-
 Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
@@
 Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
-<!--JN: language is older (1991?), but in 1995 it was officially released for the first time-->
-<!-- MH: Since we write "released" I think it's ok to use 1995? Especially since this was most likely the time a broader audience used it? -->
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
@@
 Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even higher.
-The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
-<!--JN: should we cite Pebesma2019a as it is a work in progress? -->
-<!-- MH: I think this does not speak against citing it, but also happy to remove it -->
-<!--JN: also - maybe we should cite http://book.ecosens.org/RSEbook/?-->
-<!-- MH: I dont know it, but sure if it fits -->
+The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
 A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.
@@
 The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
-Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review, we focus on software implemented in the *R* programming language.
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable and common programming languages (e.g.
+Python) for landscape ecology exist, in this review, we focus on software implemented in the *R* programming language.
 <!-- LJG: Do we need something about python capabilities for landscape ecology too? -->
+<!-- MH: I tried to incorporate it. Not really happy with the sentence though. If somebody has a better idea/formulation, please feel free to change it :) -->
 For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 Additionally, the technical quality standard on *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y* \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+<!-- MH: Not sure if these references belong here? -->
 CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability of code.
@@
 Furthermore, communities like *rOpenSci* also provide a peer review process for code quality.
-For more information about *R* package development, see \cite{Wickham2015}.
+However, while the package environment has many advantages, its highly dynamic characterstic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
+Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
+For more information about *R* package development in general, see \cite{Wickham2015}.
 
 <!-- LJG: I think that we might need to add a caveat about R packages around the fact that it's continual update by the community can lead to breaking changes (see http://datacolada.org/95) -->
+<!-- MH: Good idea, tried to come up with a sentence. Again, feel free to edit if it's not in the direction you thought of. -->
 
@@
 <!-- LJG: out of interest why not? -->
+<!-- MH: I would also be interesed. I sometimes like to use it for specific raster :)  -->
 
@@
 <!-- LJG: great idea - happy to be assigned some maps to make! -->
+<!-- MH: I already wrote that somewhere on GitHub I think.
+It would be super cool if we could download some data using one of the packages of the download section and then create the maps with it.
+That would allow to add the code to the appendix and give an example of two sections at once.
+If sombody wants to start with that, just assign yourself here https://github.com/r-spatialecology/Paper_open_source_tools/issues/4 -->
 
@@
 Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
-Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010}.
-<!-- LJG: I'd use a more up to date reference. How about Norberg et al. 2019? https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecm.1370 -->
+Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
 Additionally, textbooks introducing basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2019}.
@@
 Have added a potential sentence as a comment -->
+<!-- MH: I basically have no idea about SDM at all! So please feel free to change this section as you want. Pretty sure Johannes has some inputs about this as well. -->
 
@@
 
-Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also important for landscape genetics.
-<!-- LJG: this might be me forgetting how refs in markdown work, but should the ref above be to the connectivity section? -->
+Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
 Further functionality for landscape genetics such as ...
 <!-- LJG: I think a little more info needed here - happy to dig through the packages to add in, but thought I'd leave for now incase anyone knows off top of their heads -->can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.
+
+<!-- MH: Johannes, do you have more insights into the packages here? See comment LJG in L307 -->
 
@@
 
-#### Various
+#### Geostatistics
 
@@
 <!-- MH: I am not sure if we should do this explicitly here  That would somehow repeat the section we already have about the different packages.
-We should, however, make sure that in case we know a pacakge that answers one of the points here, is included in the first section of the papaer -->
+We should, however, make sure that in case we know a package that answers one of the points here, is included in the first section of the paper -->
 
@@
 
+<!-- MH: Make sure to update all sections with information of all authors. Out-of-date currently -->
+
 #### Conflict of interest
@@
 JN is author of *belg* and *motif* package.
+<!-- MH: I wasn't sure if we need to list this here, but it kind of felt like a not completely stupid idea. What do you think? -->
 

### 2021-01-05 paper/paper.tex
:pencil: Address some comments within MS

@@
 \usepackage{placeins}
+\usepackage[modulo]{lineno}
 
@@
 
-\title{Open-source tools in R for landscape ecology \thanks{Grants or other notes about the article that should go on the front
-page should be placed here. General acknowledgments should be placed at the
-end of the article.} }
+\title{Open-source tools in R for landscape ecology }
 
@@
 
+\linenumbers
+
 \hypertarget{sec:intro}{%
@@
 Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely even higher.
-The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
+The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
 A recent overview over the progress of \emph{R} to handle spatial data in general can be found in \cite{Bivand2020}.
@@
 The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
-Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review, we focus on software implemented in the \emph{R} programming language.
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable and common programming languages (e.g.~Python) for landscape ecology exist, in this review, we focus on software implemented in the \emph{R} programming language.
 For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 Furthermore, communities like \emph{rOpenSci} also provide a peer review process for code quality.
-For more information about \emph{R} package development, see \cite{Wickham2015}.
+However, while the package environment has many advantages, its highly dynamic characterstic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
+Packages that deal with such issue include \emph{groundhog} \cite{Simonsohn2021}, \emph{packrat} \cite{Ushey2018}, or \emph{renv} \cite{Ushey2020}.
+For more information about \emph{R} package development in general, see \cite{Wickham2015}.
 
@@
 
-While \emph{R} has several build-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+While \emph{R} has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
 However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10).
@@
 Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
-Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010}.
+Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
 Additionally, textbooks introducing basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2019}.
@@
 
-Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also important for landscape genetics.
+Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
 Further functionality for landscape genetics such as \ldots{} can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.
@@
 
-\hypertarget{various}{%
-\paragraph{Various}\label{various}}
+\hypertarget{geostatistics}{%
+\paragraph{Geostatistics}\label{geostatistics}}
 

### 2021-01-06 paper/paper.Rmd
some edits

@@
   email: nowosad.jakub@gmail.com
-- name: Johannes M. Signer
-  address: Department of Wildlife Sciences, University of Göttingen, Büsgenweg 3, 37077 Göttingen, Germany
-  email: j.m.signer@gmail.com
+- name: Johannes Signer
+  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany
+  email: jsigner@uni-goettingen.de
 - name: Laura J. Graham
@@
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
-Importantly, main *R* packages for spatial data use the external *GDAL* and *PROJ* libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
-<!--JN: references??-->
+Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL} and *PROJ* \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
 Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
@@
 
-Finally, there are a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
+Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
 
@@
 Creating maps is essential when working with spatial data.
-It has a role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
+Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
 *R* allows to create two major types of maps.
@@
 Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
+<!--- JS: I wonder if we should also cite js library? However, doing a quick google, I was not able to find a correct citation -->
 
@@
 MHKH and JN are authors of the *landscapemetrics* and *landscapetools* package.
-JN is author of *belg* and *motif* package.
+JN is author of *belg* and *motif* package. JS is author of the *amt* package. 
 <!-- MH: I wasn't sure if we need to list this here, but it kind of felt like a not completely stupid idea. What do you think? -->
+<!-- JS: I don't think its critical, but surely it doesn't do any arm listing this here -->
 

### 2021-01-06 paper/paper.tex
some edits

@@
 
-\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Johannes M. Signer \and  Laura J. Graham \and  }
+\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Johannes Signer \and  Laura J. Graham \and  }
 
@@
     \and
-        Johannes M. Signer \at
-     Department of Wildlife Sciences, University of Göttingen, Büsgenweg 3, 37077 Göttingen, Germany \\
-     \email{\href{mailto:j.m.signer@gmail.com}{\nolinkurl{j.m.signer@gmail.com}}}  %  \\
+        Johannes Signer \at
+     Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany \\
+     \email{\href{mailto:jsigner@uni-goettingen.de}{\nolinkurl{jsigner@uni-goettingen.de}}}  %  \\
 %             \emph{Present address:} of F. Author  %  if needed
@@
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
-Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} and \emph{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} \cite{GDAL} and \emph{PROJ} \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
 Additionally, \emph{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
@@
 
-Finally, there are a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
+Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
 Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.
@@
 Creating maps is essential when working with spatial data.
-It has a role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
+Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
 \emph{R} allows to create two major types of maps.
@@
 MHKH and JN are authors of the \emph{landscapemetrics} and \emph{landscapetools} package.
-JN is author of \emph{belg} and \emph{motif} package.
+JN is author of \emph{belg} and \emph{motif} package. JS is author of the \emph{amt} package.
 

### 2021-01-08 paper/paper.Rmd
Fix some minor typos

@@
   address:
-    Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
-    Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
+    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
+    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
   email: mhk.hesselbarth@gmail.com
@@
 
-To better understand how the landscape ecology community uses* R*, we conducted a short survey and used mailing lists and social media to reach the community.
+To better understand how the landscape ecology community uses *R*, we conducted a short survey and used mailing lists and social media to reach the community.
 In total, the survey was answered by `r n_people` participants, of which the the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
@@
 MHKH and JN are authors of the *landscapemetrics* and *landscapetools* package.
-JN is author of *belg* and *motif* package. JS is author of the *amt* package. 
+JN is author of *belg* and *motif* package. JS is author of the *amt* package. LJG is author of the *grainchanger* package.
 <!-- MH: I wasn't sure if we need to list this here, but it kind of felt like a not completely stupid idea. What do you think? -->
@@
 MHKH and JN designed the survey form and analyzed the responses of the participants.
-MHKH and JN drafted the manuscript and all authors contributed critically to the manuscript and gave final approval for publication.
+MHKH and JN drafted the manuscript with contributions of JS and LJG and all authors contributed critically to the manuscript and gave final approval for publication.
 We used the 'sequence--determines--credit' approach (SDC) for the sequence of authors.

### 2021-01-08 paper/paper.tex
Fix some minor typos

@@
         Maximillian H.K. Hesselbarth \at
-     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA \\
+     a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA \\
      \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
@@
 
-To better understand how the landscape ecology community uses* R*, we conducted a short survey and used mailing lists and social media to reach the community.
+To better understand how the landscape ecology community uses \emph{R}, we conducted a short survey and used mailing lists and social media to reach the community.
 In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
@@
 MHKH and JN are authors of the \emph{landscapemetrics} and \emph{landscapetools} package.
-JN is author of \emph{belg} and \emph{motif} package. JS is author of the \emph{amt} package.
+JN is author of \emph{belg} and \emph{motif} package. JS is author of the \emph{amt} package. LJG is author of the \emph{grainchanger} package.
 
@@
 MHKH and JN designed the survey form and analyzed the responses of the participants.
-MHKH and JN drafted the manuscript and all authors contributed critically to the manuscript and gave final approval for publication.
+MHKH and JN drafted the manuscript with contributions of JS and LJG and all authors contributed critically to the manuscript and gave final approval for publication.
 We used the `sequence--determines--credit' approach (SDC) for the sequence of authors.

### 2021-01-12 R/code_example.R
:art: Start to work on example code/plotting

@@ -1,0 +1,37 @@
+library(ggplot2)
+library(cowplot)
+library(sf)
+library(raster)
+library(rnaturalearth)
+library(tmap)
+
+# get precipitation data
+precipitation <- raster::getData("worldclim", lon = 3, lat = 38,
+                                 res = 0.5, var = "prec", path = "R/")
+
+# get admin boundaries of switzerland
+switzerland <- rnaturalearth::ne_countries(country = "switzerland", returnclass = "sf")
+
+# crop and mask
+precipitation_ch <- raster:::crop(precipitation, switzerland)
+
+precipitation_ch_mean <- raster::calc(x = precipitation_ch, fun = mean)
+
+plot(precipitation_ch_mean)
+
+plot(switzerland$geometry, add = TRUE)
+
+plot_base <- recordPlot()
+
+precipitation_ch_class <- raster::cut(precipitation_ch_mean, breaks = 5)
+
+plot_gg <- ggplot(raster::as.data.frame(precipitation_ch_class, xy = TRUE)) +
+    geom_raster(aes(x = x, y = y, fill = factor(layer))) +
+    geom_sf(data = switzerland, fill = NA, col = "white") +
+    scale_fill_viridis_d(name = "Precipitation\nclassified") +
+    coord_sf() +
+    labs(x = "", y = "") +
+    theme_classic()
+
+cowplot::plot_grid(plot_base, plot_gg,
+                   rel_widths = c(1, 1))

### 2021-01-12 R/get_survery_data.R

@@ -1,4 +1,4 @@
-# Make sure survery data is up-to date
+# Make sure survey data is up-to date
 
 library(googlesheets4)
 

### 2021-01-12 R/code_example.R
:art: Update example maps

@@ -1,5 +1,4 @@
 library(ggplot2)
-library(cowplot)
 library(sf)
 library(raster)
 library(rnaturalearth)
@@ -9,23 +8,38 @@ library(tmap)
 precipitation <- raster::getData("worldclim", lon = 3, lat = 38,
                                  res = 0.5, var = "prec", path = "R/")
 
-# get admin boundaries of switzerland
-switzerland <- rnaturalearth::ne_countries(country = "switzerland", returnclass = "sf")
+# calculate total precipitation
+precipitation_sum <- raster::calc(x = precipitation, fun = sum)
 
-# crop and mask
-precipitation_ch <- raster:::crop(precipitation, switzerland)
+# reproject to geographical CRS
+precipitation_sum <- raster::projectRaster(precipitation_sum, crs = "+init=epsg:21781")
 
-precipitation_ch_mean <- raster::calc(x = precipitation_ch, fun = mean)
+# get admin boundaries of Switzerland
+switzerland <- rnaturalearth::ne_countries(country = "switzerland",
+                                           returnclass = "sf")
 
-plot(precipitation_ch_mean)
+# reproject to geographical CRS
+switzerland <- sf::st_transform(x = switzerland, crs = "epsg:21781")
+
+# crop and mask with buffer
+precipitation_sum_ch <- raster:::crop(x = precipitation_sum,
+                                      y = sf::st_buffer(x = switzerland, dist = 5000))
+
+# create base plot
+# pdf(file = "R/Figures/base_plot.pdf")
+
+plot(precipitation_sum_ch)
 
 plot(switzerland$geometry, add = TRUE)
 
-plot_base <- recordPlot()
+# dev.off()
 
-precipitation_ch_class <- raster::cut(precipitation_ch_mean, breaks = 5)
+# reclassify into 5 classes
+precipitation_class_ch <- raster::cut(precipitation_sum_ch, breaks = 5)
 
-plot_gg <- ggplot(raster::as.data.frame(precipitation_ch_class, xy = TRUE)) +
+# create ggplot
+# MH: Why is this on long/lat? Coordinates seem to be okay?
+plot_gg <- ggplot(raster::as.data.frame(precipitation_class_ch, xy = TRUE)) +
     geom_raster(aes(x = x, y = y, fill = factor(layer))) +
     geom_sf(data = switzerland, fill = NA, col = "white") +
[…]

### 2021-01-12 R/code_example.R
added mask

@@ -13,7 +13,10 @@ precipitation <- raster::getData("worldclim", lon = 3, lat = 38,
 switzerland <- rnaturalearth::ne_countries(country = "switzerland", returnclass = "sf")
 
 # crop and mask
-precipitation_ch <- raster:::crop(precipitation, switzerland)
+precipitation_ch <- raster::crop(precipitation, switzerland)
+precipitation_ch <- raster::mask(precipitation_ch, switzerland)
+plot(precipitation_ch[[1]])
+
 
 precipitation_ch_mean <- raster::calc(x = precipitation_ch, fun = mean)
 

### 2021-01-13 paper/paper.Rmd
updated to include geostats section

@@
 
+#### Geostatistics
+
+Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyse patterns in continuous data (e.g.
+normalised difference vegetation index, disturbance intensity, topography).
+In landscape ecology, geostatistics has three key uses: (1) detecting and correcting for spatial autocorrelation; (2) quantifying and comparing landscape pattern; (3) interpolating data.
+
+Point pattern analysis is one such method. 
+This uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
+The *spatstat* packge contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modelling fitting, inference and diagnostics.
+
+Distance-based methods allow us to detect and correct for spatial autocorrelation in data.
+It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modelling. 
+The *spdep* package has methods for quantifying multiple metrics of spatial autocorrelation, and to correct for these in a spatial autoregressive model.
+The *rinla* and *inlabru* packages also provide functionality for modelling of spatially structured data.
+
+Finally, with variograms and correlograms, the spatial structure of continuous landscapes can be quantified and compared. 
+R packages *geoR* and *gstat* provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+
 #### Species distribution modeling {#sec:SDM}
@@
 Furthermore, the *RandomFields* package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.
-
-#### Geostatistics
-
-<!-- MH: What about spatial autocorrelation? -->
-<!-- MH: SpatialKDE, geoR, gstat, spdep, siland -->
 

### 2021-01-13 paper/paper.tex
updated to include geostats section

@@
 
+\hypertarget{geostatistics}{%
+\paragraph{Geostatistics}\label{geostatistics}}
+
+Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyse patterns in continuous data (e.g.~normalised difference vegetation index, disturbance intensity, topography).
+In landscape ecology, geostatistics has three key uses: (1) detecting and correcting for spatial autocorrelation; (2) quantifying and comparing landscape pattern; (3) interpolating data.
+
+Point pattern analysis is one such method.
+This uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
+The \emph{spatstat} packge contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modelling fitting, inference and diagnostics.
+
+Distance-based methods allow us to detect and correct for spatial autocorrelation in data.
+It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modelling.
+The \emph{spdep} package has methods for quantifying multiple metrics of spatial autocorrelation, and to correct for these in a spatial autoregressive model.
+The \emph{rinla} and \emph{inlabru} packages also provide functionality for modelling of spatially structured data.
+
+Finally, with variograms and correlograms, the spatial structure of continuous landscapes can be quantified and compared.
+R packages \emph{geoR} and \emph{gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+
 \hypertarget{sec:SDM}{%
@@
 Furthermore, the \emph{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.
-
-\hypertarget{geostatistics}{%
-\paragraph{Geostatistics}\label{geostatistics}}
 

### 2021-01-14 paper/paper.Rmd
Added to the movement ecology paragraph

@@
 This is considered the gold standard in contemporary SDM \cite(Araujo2007).-->
-Related to SDM, there are several packages to estimate home ranges and habitat selection, including *adehabitatHR* and *adehabitatHS* \cite{Calenge2006} or *amt \cite{Signer2019}*
+
+Related to SDMs, there are growing number of R packages to analyze data from tracked animals, to study their movement characteristics, space use and interaction with other animals and the environment.
+These analyses often use results of landscape ecological analyses as explanatory variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
+Widely used R packages include *ctmm* \cite{calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
+A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.
 

### 2021-01-14 paper/paper.tex
Added to the movement ecology paragraph

@@
 Also, packages specifically designed for SDM exists, including includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008}, or \emph{zoon} \cite{Golding2018} packages.
-Related to SDM, there are several packages to estimate home ranges and habitat selection, including \emph{adehabitatHR} and \emph{adehabitatHS} \cite{Calenge2006} or \emph{amt \cite{Signer2019}}
+
+Related to SDMs, there are growing number of R packages to analyze data from tracked animals, to study their movement characteristics, space use and interaction with other animals and the environment.
+These analyses often use results of landscape ecological analyses as explanatory variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
+Widely used R packages include \emph{ctmm} \cite{calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
+A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.
 

### 2021-01-14 paper/paper.Rmd
:pencil: Add glossary and describtion to table 1. Also some wording and typos

@@
 
+Glossary
+
+\begin{tabularx}{\textwidth}{XX}
+
+\hline
+
+CRAN & Comprehensive R Archive Network \\
+CRS & Coordinate references systems \\
+GIS & Geographical information system \\
+NLM & Neutral landscape model \\
+raster & Cell grid based data model \\
+R packages & Software extension to R programming language \\
+SDM & Species distribution modelling \\
+vector & Points and lines based data model \\
+
+\hline
+
+\end{tabularx}
+
 <!-- MH: We really need to make sure, we introduce all technical terms well enough throughout the paper -->
@@
 
-Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
-In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
-
-While human activities have altered the landscapes for millennia \cite{Ellis2011,Ellis2015}, in the past centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
+While human activities have altered the landscapes for millennia \cite{Ellis2011,Ellis2015}, in the recent centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
 Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
@@
 
+Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
+In this context, landscape ecology predominantly focuses on i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchanges within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) and can guide how to to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
+
 Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes vary spatially \cite{With2019}.
-To this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}.
-Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984}.
-Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze, and visualize spatial data.
+The importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance today \cite{Simova2012,Estes2018}.
+Thus, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize spatial data.
 
@@
 Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
-Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
+Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
 Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
-Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows use of the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
 This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.
@@
 The expanding CRAN Task Views document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
-Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even higher.
+Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
 The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
@@
 Furthermore, communities like *rOpenSci* also provide a peer review process for code quality.
-However, while the package environment has many advantages, its highly dynamic characterstic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
+However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
 Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
@@
 While *R* has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10).
+However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
@@
 Two main groups of CRS, namely geographical and projected, exist, with each having many members.
-In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
+Using geographical CRS, positions are specified by latitude and longitude coordinates in degrees.
 However, most landscape ecology studies should utilize projected CRSs, which use some measurement units (e.g., meters).
@@
 
-A second common spatial data processing task is required when the available data extends over a larger area than the study region.
+Another common spatial data processing task is required when the available data extends over a larger area than the study region.
 In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
@@
 Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
-*R* allows to create two major types of maps.
+*R* allows two major types of maps.
 Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.
@@
 
-Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyse patterns in continuous data (e.g.
-normalised difference vegetation index, disturbance intensity, topography).
+Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g.
+normalized difference vegetation index, disturbance intensity, topography).
 In landscape ecology, geostatistics has three key uses: (1) detecting and correcting for spatial autocorrelation; (2) quantifying and comparing landscape pattern; (3) interpolating data.
@@
 This uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
-The *spatstat* packge contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modelling fitting, inference and diagnostics.
+The *spatstat* package contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modelling fitting, inference and diagnostics. 
+A comprehensive text book covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
 
@@
 
-Related to SDMs, there are growing number of R packages to analyze data from tracked animals, to study their movement characteristics, space use and interaction with other animals and the environment.
+Related to SDMs, there are growing number of *R* packages to analyze data from tracked animals, to study their movement characteristics, space use and interaction with other animals and the environment.
 These analyses often use results of landscape ecological analyses as explanatory variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
@@
 
-\begin{tabularx}{\textwidth}{X r r }
+\begin{tabularx}{450pt}{lrrr}
 
 \hline
-Task & R package & Reference \\
+Task & R package & Describtion & Reference \\
 \hline
-\multirow{8}{*}{Spatial data} & raster & \cite{Hijmans2019} \\
-& terra & \cite{R-terra} \\
-& stars & \cite{Pebesma2019} \\
-& fasterize & \cite{Ross2020} \\
-& rasterDT & \cite{OBrien2020} \\
-& exactextractr & \cite{Baston2020} \\
-& sp & \cite{Pebesma2005,Bivand2013} \\
-& sf & \cite{Pebesma2018} \\
+\multirow{8}{*}{Spatial data} & raster & Raster data handling and analysis & \cite{Hijmans2019} \\
+& terra & Basic data handling and analysis & \cite{R-terra} \\
+& stars & Spatiotemporal raster data & \cite{Pebesma2019} \\
+& fasterize & Polygon to raster conversion & \cite{Ross2020} \\
+& rasterDT & Faster alternatives for some raster functions & \cite{OBrien2020} \\
+& exactextractr & Summarize raster values over polygonal areas & \cite{Baston2020} \\
+& sp & Vector data handling and analysis & \cite{Pebesma2005,Bivand2013} \\
+& sf & Vector data handling and analysis & \cite{Pebesma2018} \\
 \hline
-\multirow{6}{*}{Spatial data download} & rnaturalearth & \cite{South2017} \\
-& elevatr & \cite{R-elevatr} \\
-& rgbif & \cite{R-rgbif} \\
-& BIEN & \cite{Maitner2020} \\
-& marmap & \cite{Pante2013} \\
-& FedData & \cite{R-FedData} \\
-& getlandsat & \cite{Chamberlain2018} \\
-& MODIS & \cite{Mattiuzzi2020} \\
-& sen2r & \cite{R-sen2r} \\
+\multirow{6}{*}{Spatial data download} & rnaturalearth & Download region and country data & \cite{South2017} \\
+& elevatr & Download elevation data & \cite{R-elevatr} \\
+& rgbif & Download biodiversity data & \cite{R-rgbif} \\
+& BIEN & Download plant diversity, function and distribution data & \cite{Maitner2020} \\
+& marmap & Download and manipulate bathymetric data & \cite{Pante2013} \\
+& FedData & Download US federal data & \cite{R-FedData} \\
+& getlandsat & Download satellite data from Landsat 8 & \cite{Chamberlain2018} \\
+& MODIS & Download satellite data from MODIS & \cite{Mattiuzzi2020} \\
+& sen2r & Download satellite data from Sentinel & \cite{R-sen2r} \\
 \hline
-\multirow{7}{*}{Creating maps} & mapview & \cite{R-mapview} \\
-& tmap & \cite{Tennekes2018} \\
-& leaflet & \cite{R-leaflet} \\
-& cartography & \cite{Giraud2016} \\
-& ggspatial & \cite{Dunnington2020} \\
-& rayshader & \cite{Morgen-Wall2020} \\
+\multirow{7}{*}{Creating maps} & mapview & Interactive viewing of spatial data & \cite{R-mapview} \\
+& tmap & Create thematic maps & \cite{Tennekes2018} \\
+& leaflet & Create interactive web maps & \cite{R-leaflet} \\
+& cartography & Create cartographic maps & \cite{Giraud2016} \\
+& ggspatial & Spatial extension for ggplot2 & \cite{Dunnington2020} \\
+& rayshader & Create 2D and 3D data visualizations & \cite{Morgen-Wall2020} \\
 \hline
-\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & \cite{Hesselbarth2019a} \\
-& belg &  \cite{R-belgpaper} \\
-& motif & \cite{R-motif} \\
-& geodiv & \cite{Smith2020} \\
+\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & Quantify categorical landscape patterns & \cite{Hesselbarth2019a} \\
+& belg & Calculates the Boltzmann entropy & \cite{R-belgpaper} \\
+& motif & Not on CRAN? & \cite{R-motif} \\
+& geodiv & Quantify continuous landscape patterns & \cite{Smith2020} \\
 \hline
-\multirow{8}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & \cite{Hijmans2017} \\
-& sdm & \cite{Naimi2016} \\
-& ecospat & \cite{Broennimann2020} \\
-& biomod2 & \cite{Thuiller2020} \\
-& PresenceAbsence & \cite{Freeman2008} \\
-& zoon & \cite{Golding2018} \\
-& adehabitatHR, adehabitatHS & \cite{Calenge2006} \\
-& amt & \cite{Signer2019} \\
+\multirow{8}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & Add describtion & \cite{Hijmans2017} \\
+& sdm & Add describtion & \cite{Naimi2016} \\
+& ecospat & Add describtion & \cite{Broennimann2020} \\
+& biomod2 & Add describtion & \cite{Thuiller2020} \\
+& PresenceAbsence & Add describtion & \cite{Freeman2008} \\
+& zoon & Add describtion & \cite{Golding2018} \\
+& adehabitatHR, adehabitatHS & Add describtion & \cite{Calenge2006} \\
+& amt &  Add describtion &\cite{Signer2019} \\
 \hline
-\multirow{5}{*}{Connectivity} & lconnect & \cite{Mestre2019} \\
-& Makurhini & \cite{Godinez-Gomez2020} \\
-& grainscape & \cite{Chubaty2020} \\
-& gdistance & \cite{vanEtten2017} \\
-& samc & \cite{Marx2020} \\
+\multirow{5}{*}{Connectivity} & lconnect & Add describtion & \cite{Mestre2019} \\
+& Makurhini & Add describtion & \cite{Godinez-Gomez2020} \\
+& grainscape & Add describtion & \cite{Chubaty2020} \\
+& gdistance & Add describtion & \cite{vanEtten2017} \\
+& samc & Add describtion & \cite{Marx2020} \\
 \hline
-\multirow{4}{*}{Landscape genetics} & raphs4lg & \cite{Savary2020} \\
-& PopGenReport & \cite{Adamack2014,Gruber2015} \\
-& HierDpart & \cite{Qin2019} \\
-& GeNetIt & \cite{Murphy2010} \\
+\multirow{4}{*}{Landscape genetics} & raphs4lg & Add describtion & \cite{Savary2020} \\
+& PopGenReport & Add describtion & \cite{Adamack2014,Gruber2015} \\
+& HierDpart & Add describtion & \cite{Qin2019} \\
+& GeNetIt & Add describtion & \cite{Murphy2010} \\
 \hline
-\multirow{4}{*}{various} & NLMR & \cite{Sciaini2018} \\
-& RandomFields & \cite{Schlather2015} \\
-& landscapetools & \cite{Sciaini2018}  \\
-& grainchanger & \cite{Graham2019} \\
+\multirow{4}{*}{various} & NLMR & Simulate neutral landscape models & \cite{Sciaini2018} \\
+& RandomFields & Add describtion & \cite{Schlather2015} \\
+& landscapetools & Utility functions for raster data & \cite{Sciaini2018}  \\
+& grainchanger & Data aggregation methods for raster data & \cite{Graham2019} \\
 \hline
@@
 
-To better understand how the landscape ecology community uses *R*, we conducted a short survey and used mailing lists and social media to reach the community.
-In total, the survey was answered by `r n_people` participants, of which the the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
+To better understand how the landscape ecology community uses *R*, we conducted a short survey using mailing lists and social media to reach the community.
+In total, the survey was answered by `r n_people` participants, of which the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
 Other, less frequent answers were "Data scientists", "None of the above", "Government employees" "Master's degree student" and "Bachelor's degree student" (in decreasing order).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey) B)).
+Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey) B).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
@@
   geom_bar(aes(x = usefulness_class, y = n_rel),
-           stat = "identity") +
+           stat = "identity", fill = "black") +
   theme_classic(base_size = base_size) +
@@
 
-Lastly, when asked how useful *R* is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
+Lastly, when asked how useful *R* is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
 \@ref(fig:fig-survey) D)).
@@
 
+We are thankful to all members of the 'Coastal Ecology and Conservation Lab' at University of Michigan for comments on an early draft of the manuscript.
+
 <!-- MH: Make sure to update all sections with information of all authors. Out-of-date currently -->

### 2021-01-14 paper/paper.tex
:pencil: Add glossary and describtion to table 1. Also some wording and typos

@@
 
+Glossary
+
+\begin{tabularx}{\textwidth}{XX}
+
+\hline
+
+CRAN & Comprehensive R Archive Network \\
+CRS & Coordinate references systems \\
+GIS & Geographical information system \\
+NLM & Neutral landscape model \\
+raster & Cell grid based data model \\
+R packages & Software extension to R programming language \\
+SDM & Species distribution modelling \\
+vector & Points and lines based data model \\
+
+\hline
+
+\end{tabularx}
+
 \hypertarget{sec:landscape_ecology}{%
@@
 Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
-Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
+Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
 Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
-Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows use of the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
+Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
 This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.
@@
 The expanding CRAN Task Views document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
-Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely even higher.
+Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely higher.
 The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
@@
 Furthermore, communities like \emph{rOpenSci} also provide a peer review process for code quality.
-However, while the package environment has many advantages, its highly dynamic characterstic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
+However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
 Packages that deal with such issue include \emph{groundhog} \cite{Simonsohn2021}, \emph{packrat} \cite{Ushey2018}, or \emph{renv} \cite{Ushey2020}.
@@
 While \emph{R} has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10).
+However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
@@
 Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
-\emph{R} allows to create two major types of maps.
+\emph{R} allows two major types of maps.
 Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.~changing the displayed values.
@@
 
-Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyse patterns in continuous data (e.g.~normalised difference vegetation index, disturbance intensity, topography).
+Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g.~normalized difference vegetation index, disturbance intensity, topography).
 In landscape ecology, geostatistics has three key uses: (1) detecting and correcting for spatial autocorrelation; (2) quantifying and comparing landscape pattern; (3) interpolating data.
@@
 This uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
-The \emph{spatstat} packge contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modelling fitting, inference and diagnostics.
+The \emph{spatstat} package contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modelling fitting, inference and diagnostics.
+A comprehensive text book covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
 
@@
 
-Related to SDMs, there are growing number of R packages to analyze data from tracked animals, to study their movement characteristics, space use and interaction with other animals and the environment.
+Related to SDMs, there are growing number of \emph{R} packages to analyze data from tracked animals, to study their movement characteristics, space use and interaction with other animals and the environment.
 These analyses often use results of landscape ecological analyses as explanatory variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
@@
 
-To better understand how the landscape ecology community uses \emph{R}, we conducted a short survey and used mailing lists and social media to reach the community.
-In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
+To better understand how the landscape ecology community uses \emph{R}, we conducted a short survey using mailing lists and social media to reach the community.
+In total, the survey was answered by 103 participants, of which the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
 Other, less frequent answers were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
+Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 
-Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
+Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
 \ref{fig:fig-survey} D)).
@@
 
+We are thankful to all members of the `Coastal Ecology and Conservation Lab' at University of Michigan for comments on an early draft of the manuscript.
+
 \hypertarget{conflict-of-interest}{%

### 2021-01-15 paper/paper.Rmd
:pencil: Add missing describtions to table and glossary

@@
 
-Glossary
-
-\begin{tabularx}{\textwidth}{XX}
+\begin{tabularx}{450pt}{lX}
 
@@
 
+\multicolumn{2}{c}{Glossary} \\
+
+\hline
+
+API & Application programming interface \\
+cell & Smallest, rectangular unit of raster data model \\
 CRAN & Comprehensive R Archive Network \\
@@
 GIS & Geographical information system \\
+landscapes & Mosaics of land covers, ecosystems, habitat types, land uses \\
 NLM & Neutral landscape model \\
+open-source software & Software released under licenses that allow to use, modify and distribute it \\
+patch & Neighboring cells with same value \\
 raster & Cell grid based data model \\
-R packages & Software extension to R programming language \\
+resolution & Size in meter or degree of one cell \\
+R packages & User-created software extension to R programming language \\
 SDM & Species distribution modelling \\
@@
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
-The expanding CRAN Task Views document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
+The expanding CRAN Task Views (lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
 Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
@@
 Since publicly available data is becoming more prominent, so are *R* packages to access them.
-Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
+Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to downlad region and country data, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
 Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download Sentinel-2 optical images.
@@
 normalized difference vegetation index, disturbance intensity, topography).
-In landscape ecology, geostatistics has three key uses: (1) detecting and correcting for spatial autocorrelation; (2) quantifying and comparing landscape pattern; (3) interpolating data.
+In landscape ecology, geostatistics has three key uses: i) detecting and correcting for spatial autocorrelation; ii) quantifying and comparing landscape pattern; iii) interpolating data.
 
@@
 
-\begin{tabularx}{450pt}{lrrr}
+\begin{tabularx}{450pt}{lrXr}
 
@@
 \multirow{8}{*}{Spatial data} & raster & Raster data handling and analysis & \cite{Hijmans2019} \\
-& terra & Basic data handling and analysis & \cite{R-terra} \\
-& stars & Spatiotemporal raster data & \cite{Pebesma2019} \\
+& terra & Raster (and some vector) data handling and analysis & \cite{R-terra} \\
+& stars & Spatiotemporal raster data handling and analysis & \cite{Pebesma2019} \\
 & fasterize & Polygon to raster conversion & \cite{Ross2020} \\
@@
 & marmap & Download and manipulate bathymetric data & \cite{Pante2013} \\
-& FedData & Download US federal data & \cite{R-FedData} \\
+& FedData & Download geospatial data from US federal sources & \cite{R-FedData} \\
 & getlandsat & Download satellite data from Landsat 8 & \cite{Chamberlain2018} \\
@@
 \multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & Quantify categorical landscape patterns & \cite{Hesselbarth2019a} \\
-& belg & Calculates the Boltzmann entropy & \cite{R-belgpaper} \\
-& motif & Not on CRAN? & \cite{R-motif} \\
+& belg & Calculate the Boltzmann entropy & \cite{R-belgpaper} \\
+& motif & Pattern-based spatial analysis & \cite{R-motif} \\
 & geodiv & Quantify continuous landscape patterns & \cite{Smith2020} \\
 \hline
-\multirow{8}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & Add describtion & \cite{Hijmans2017} \\
-& sdm & Add describtion & \cite{Naimi2016} \\
-& ecospat & Add describtion & \cite{Broennimann2020} \\
-& biomod2 & Add describtion & \cite{Thuiller2020} \\
-& PresenceAbsence & Add describtion & \cite{Freeman2008} \\
-& zoon & Add describtion & \cite{Golding2018} \\
-& adehabitatHR, adehabitatHS & Add describtion & \cite{Calenge2006} \\
-& amt &  Add describtion &\cite{Signer2019} \\
+\multirow{8}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & Methods for species distribution modeling & \cite{Hijmans2017} \\
+& sdm & Species distribution models using individual and community-based approaches & \cite{Naimi2016} \\
+& ecospat & Species distribution, niche quantification and community assembly & \cite{Broennimann2020} \\
+& biomod2 & Species distribution modeling, ensemble of models and ensemble forecasting & \cite{Thuiller2020} \\
+& PresenceAbsence & Evaluating of presence-absence models & \cite{Freeman2008} \\
+& zoon & Reproducible and remixable species distribution modelling & \cite{Golding2018} \\
+& adehabitatHR, adehabitatHS & Home range and habitat selection modelling & \cite{Calenge2006} \\
+& amt & Manage and analyze animal movement data &\cite{Signer2019} \\
 \hline
-\multirow{5}{*}{Connectivity} & lconnect & Add describtion & \cite{Mestre2019} \\
-& Makurhini & Add describtion & \cite{Godinez-Gomez2020} \\
-& grainscape & Add describtion & \cite{Chubaty2020} \\
-& gdistance & Add describtion & \cite{vanEtten2017} \\
-& samc & Add describtion & \cite{Marx2020} \\
+\multirow{5}{*}{Connectivity} & lconnect & Calculate landscape connectivity metrics & \cite{Mestre2019} \\
+& Makurhini & Calculate fragmentation and landscape connectivity indices & \cite{Godinez-Gomez2020} \\
+& grainscape & Calculate minimum planar graph and grains of connectivity models & \cite{Chubaty2020} \\
+& gdistance & Distances and routes on geographical grids & \cite{vanEtten2017} \\
+& samc & Functions for working with absorbing Markov chains & \cite{Marx2020} \\
 \hline
-\multirow{4}{*}{Landscape genetics} & raphs4lg & Add describtion & \cite{Savary2020} \\
-& PopGenReport & Add describtion & \cite{Adamack2014,Gruber2015} \\
-& HierDpart & Add describtion & \cite{Qin2019} \\
-& GeNetIt & Add describtion & \cite{Murphy2010} \\
+\multirow{4}{*}{Landscape genetics} & graph4lg & Build graphs for landscape genetics analysis & \cite{Savary2020} \\
+& PopGenReport & Framework to analyse population genetic data & \cite{Adamack2014,Gruber2015} \\
+& HierDpart & Calculating and decomposing hierarchical diversity metrics & \cite{Qin2019} \\
+& GeNetIt & Spatial graph-theoretic genetic gravity models & \cite{Murphy2010} \\
 \hline
 \multirow{4}{*}{various} & NLMR & Simulate neutral landscape models & \cite{Sciaini2018} \\
-& RandomFields & Add describtion & \cite{Schlather2015} \\
+& RandomFields & Simulation and analysis of random fields & \cite{Schlather2015} \\
 & landscapetools & Utility functions for raster data & \cite{Sciaini2018}  \\

### 2021-01-15 paper/paper.tex
:pencil: Add missing describtions to table and glossary

@@
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
-The expanding CRAN Task Views document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
+The expanding CRAN Task Views (lists of packages related to a certain topic) document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
 Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely higher.
@@
 Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
-Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
+Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database to downlad region and country data, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
 Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
@@
 Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g.~normalized difference vegetation index, disturbance intensity, topography).
-In landscape ecology, geostatistics has three key uses: (1) detecting and correcting for spatial autocorrelation; (2) quantifying and comparing landscape pattern; (3) interpolating data.
+In landscape ecology, geostatistics has three key uses: i) detecting and correcting for spatial autocorrelation; ii) quantifying and comparing landscape pattern; iii) interpolating data.
 

### 2021-01-20 paper/paper.Rmd
improves glossary

@@
 GIS & Geographical information system \\
-landscapes & Mosaics of land covers, ecosystems, habitat types, land uses \\
+landscape & A mosaic of land covers, ecosystems, habitat types, land uses \\
 NLM & Neutral landscape model \\
@@
 resolution & Size in meter or degree of one cell \\
-R packages & User-created software extension to R programming language \\
-SDM & Species distribution modelling \\
-vector & Points and lines based data model \\
+R package & User-created software extension to R programming language \\
+SDM & Species distribution modeling \\
+vector & Points, lines, and polygons-based data model \\
 

### 2021-01-20 paper/paper.Rmd
adds a keyword

@@
 - reproducibility
+- scientific software
 - ...

### 2021-01-20 paper/paper.Rmd
improves spatial stat section

@@
 
-#### Geostatistics
+#### Spatial statistics
 
-Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g.
-normalized difference vegetation index, disturbance intensity, topography).
-In landscape ecology, geostatistics has three key uses: i) detecting and correcting for spatial autocorrelation; ii) quantifying and comparing landscape pattern; iii) interpolating data.
+<!--JN: I would rather say that geostatistics is a subbranch of spatial statistics (e.g. https://www.geos.ed.ac.uk/~gisteac/gis_book_abridged/files/ch16.pdf). 
+However, we can revert my changes if you want-->
+
+Spatial statistics are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g., normalized difference vegetation index, disturbance intensity, topography). 
+In landscape ecology, spatial statistics has three key uses: i) detecting and correcting for spatial autocorrelation; ii) quantifying and comparing landscape patterns; iii) interpolating data.
 
@@
 This uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
-The *spatstat* package contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modelling fitting, inference and diagnostics. 
-A comprehensive text book covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
+The *spatstat* package \cite{R-spatstat} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
+A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
 
 Distance-based methods allow us to detect and correct for spatial autocorrelation in data.
-It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modelling. 
-The *spdep* package has methods for quantifying multiple metrics of spatial autocorrelation, and to correct for these in a spatial autoregressive model.
-The *rinla* and *inlabru* packages also provide functionality for modelling of spatially structured data.
+It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling. 
+The *spdep* package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
+The *rinla* \cite{rue2009approximate} (\url{https://www.r-inla.org}) and *inlabru* \cite{R-inlabru} packages also provide functionality for modeling of spatially structured data.
 
-Finally, with variograms and correlograms, the spatial structure of continuous landscapes can be quantified and compared. 
-R packages *geoR* and *gstat* provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms. 
+R packages *geoR* \cite{R-geoR} and *gstat* \cite{R-gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+Geostatistics also allows for spatial data simulations.
 

### 2021-01-20 R/code_example.R
improves code example

@@ -11,16 +11,12 @@ precipitation <- raster::getData("worldclim", lon = 3, lat = 38,
 # calculate total precipitation
 precipitation_sum <- raster::calc(x = precipitation, fun = sum)
 
-# crop and mask
-precipitation_ch <- raster::crop(precipitation, switzerland)
-precipitation_ch <- raster::mask(precipitation_ch, switzerland)
-plot(precipitation_ch[[1]])
-
 # reproject to geographical CRS
 precipitation_sum <- raster::projectRaster(precipitation_sum, crs = "+init=epsg:21781")
 
 # get admin boundaries of Switzerland
-switzerland <- rnaturalearth::ne_countries(country = "switzerland",
+switzerland <- rnaturalearth::ne_countries(scale = 50,
+                                           country = "switzerland",
                                            returnclass = "sf")
 
 # reproject to geographical CRS
@@ -55,9 +51,19 @@ plot_gg <- ggplot(raster::as.data.frame(precipitation_class_ch, xy = TRUE)) +
 ggplot2::ggsave(filename = "R/Figures/ggplot2.pdf")
 
 # create interactive tmap (take screenshot for paper)
+# tmap_mode("plot")
 tmap_mode("view")
 
-tm_shape(precipitation_class_ch) +
-    tm_raster() +
+plot_tm = tm_shape(precipitation_class_ch) +
+    tm_graticules() +
+    tm_raster(title = "Precipitation\nclassified", style = "cont",
+              palette = "viridis") +
     tm_shape(switzerland) +
-    tm_borders()
+    tm_borders(lwd = 2, col = "white") +
+    tm_scale_bar(breaks = c(0, 25, 50),
+                 bg.color = "white") +
+    tm_layout(legend.outside = TRUE)
+
+tmap::tmap_save(plot_tm, "R/Figures/tmap.pdf")
+plot_tml = tmap_leaflet(plot_tm)
+mapview::mapshot(plot_tml, file = "R/Figures/tmap2.pdf")

### 2021-01-20 R/code_example.R
combine plots

@@ -30,8 +30,7 @@ precipitation_sum_ch <- raster:::crop(x = precipitation_sum,
 # pdf(file = "R/Figures/base_plot.pdf")
 
 plot(precipitation_sum_ch)
-
-plot(switzerland$geometry, add = TRUE)
+plot(st_geometry(switzerland), add = TRUE)
 
 # dev.off()
 
@@ -67,3 +66,24 @@ plot_tm = tm_shape(precipitation_class_ch) +
 tmap::tmap_save(plot_tm, "R/Figures/tmap.pdf")
 plot_tml = tmap_leaflet(plot_tm)
 mapview::mapshot(plot_tml, file = "R/Figures/tmap2.pdf")
+
+
+
+# combine plots -----------------------------------------------------------
+# https://wilkelab.org/cowplot/articles/mixing_plot_frameworks.html
+library(cowplot)
+tmap_mode("plot")
+
+# prep
+tm_grob = tmap_grob(plot_tm)
+
+p1 = function() {
+    plot(precipitation_class_ch)
+    plot(st_geometry(switzerland), add = TRUE)
+}
+
+plot_all = plot_grid(p1, plot_gg, tm_grob,
+          nrow = 1, labels = "auto")
+
+ggsave("R/Figures/plot_all.pdf",
+       width = 12, height = 4)

### 2021-01-21 paper/paper.tex
added movement packages

@@
         reproducibility \and
+        scientific software \and
         \ldots{} \and
@@
 
-\hypertarget{geostatistics}{%
-\paragraph{Geostatistics}\label{geostatistics}}
+\hypertarget{spatial-statistics}{%
+\paragraph{Spatial statistics}\label{spatial-statistics}}
 
-Geostatistics, or spatial statistics, are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g.~normalized difference vegetation index, disturbance intensity, topography).
-In landscape ecology, geostatistics has three key uses: i) detecting and correcting for spatial autocorrelation; ii) quantifying and comparing landscape pattern; iii) interpolating data.
+Spatial statistics are complimentary to landscape metrics, and can be used to analyze patterns in continuous data (e.g., normalized difference vegetation index, disturbance intensity, topography).
+In landscape ecology, spatial statistics has three key uses: i) detecting and correcting for spatial autocorrelation; ii) quantifying and comparing landscape patterns; iii) interpolating data.
 
@@
 This uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
-The \emph{spatstat} package contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modelling fitting, inference and diagnostics.
-A comprehensive text book covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
+The \emph{spatstat} package \cite{R-spatstat} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
+A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
 
 Distance-based methods allow us to detect and correct for spatial autocorrelation in data.
-It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modelling.
-The \emph{spdep} package has methods for quantifying multiple metrics of spatial autocorrelation, and to correct for these in a spatial autoregressive model.
-The \emph{rinla} and \emph{inlabru} packages also provide functionality for modelling of spatially structured data.
+It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling.
+The \emph{spdep} package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
+The \emph{rinla} \cite{rue2009approximate} (\url{https://www.r-inla.org}) and \emph{inlabru} \cite{R-inlabru} packages also provide functionality for modeling of spatially structured data.
 
-Finally, with variograms and correlograms, the spatial structure of continuous landscapes can be quantified and compared.
-R packages \emph{geoR} and \emph{gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms.
+R packages \emph{geoR} \cite{R-geoR} and \emph{gstat} \cite{R-gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+Geostatistics also allows for spatial data simulations.
 
@@
 & amt & Manage and analyze animal movement data &\cite{Signer2019} \\
+& ctmm & Fit continuous time movement models &\cite{calabrese2016} \\
+& moveHMM & Fit hidden Markov models to movement data &\cite{michelot2016} \\
 \hline

### 2021-01-22 paper/paper.Rmd
added movement packages
add in spatial stats section to the table

@@
 \hline
+\multirow{6}{*}{\shortstack[l]{Spatial statistics}} & spatstat & Point pattern analysis & \cite{R-spatstat} \\
+& spdep & Quantify and correct for spatial autocorrelation & \cite{Bivand2013} \\
+& rinla & Fitting Bayesian spatio-temporal models using INLA & \cite{rue2009approximate} \\
+& inlabru & Fitting Bayesian spatio-temporal models using INLA & \cite{R-inlabru} \\
+& geoR & Variograms, correlograms and (co-)kriging & \cite{R-geoR} \\
+& gstat & Variograms, correlograms and (co-)kriging & \cite{R-gstat} \\
+\hline
 \multirow{8}{*}{\shortstack[l]{Species distribution\\modeling}} & dismo & Methods for species distribution modeling & \cite{Hijmans2017} \\

### 2021-01-23 paper/paper.Rmd
Update Conclusions

@@
 # thanks: |
-#    Grants or other notes about the article that should go on the front
-#    page should be placed here. General acknowledgments should be placed at the
-#    end of the article.
 authors: 
@@
 GIS & Geographical information system \\
-landscape & A mosaic of land covers, ecosystems, habitat types, land uses \\
+landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
 NLM & Neutral landscape model \\
@@
 
-<!-- MH: We really need to make sure, we introduce all technical terms well enough throughout the paper -->
-
 ## A short introduction to landscape ecology {#sec:landscape_ecology}
@@
 Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
-Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important \cite{With2019}.
+Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important for science, conservation, or management \cite{With2019}.
 
@@
 
-Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes vary spatially \cite{With2019}.
+Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
 The importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance today \cite{Simova2012,Estes2018}.
@@
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
-The expanding CRAN Task Views (lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
+The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
 Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
@@
 The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
-Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable and common programming languages (e.g.
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.
 Python) for landscape ecology exist, in this review, we focus on software implemented in the *R* programming language.
-<!-- LJG: Do we need something about python capabilities for landscape ecology too? -->
-<!-- MH: I tried to incorporate it. Not really happy with the sentence though. If somebody has a better idea/formulation, please feel free to change it :) -->
 For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 *GitHub* hosts source code under version control, and allows users to install packages with one line of code using the *remotes* \cite{Hester2020} package.
-Additionally, hosting a package on *GitHub* provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing.
+Additionally, hosting a package on *GitHub* provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e.
+testing if functions return an expected value).
 
@@
 Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
-Furthermore, communities like *rOpenSci* also provide a peer review process for code quality.
+Furthermore, online communities like *rOpenSci* also provide a peer-review process for code quality.
 However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
@@
 
-<!-- LJG: I think that we might need to add a caveat about R packages around the fact that it's continual update by the community can lead to breaking changes (see http://datacolada.org/95) -->
-<!-- MH: Good idea, tried to come up with a sentence. Again, feel free to edit if it's not in the direction you thought of. -->
-
 ```{r fig-workflow, echo = FALSE, out.width="100%", out.height="30%", fig.cap = "Exemplary workflow of spatial data analyses for landscape ecology using the R programming language.
-For all major tasks, a few example R packages are listed.
-."}
+For all major tasks, some example R packages are listed."}
 include_graphics("data/Workflow.png")
@@
 Currently, the most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
-A *raster* successor, *terra*,  aimed at the simpler interface and improved performance is now being developed \cite{R-terra}; however, it could take several years for this package to be adopted by other developers and users.
+A *raster* successor, *terra*, aimed at the simpler interface and improved performance is being developed \cite{R-terra}, however, it could take several years for this package to be adopted by other developers and users.
 Alternatively, the *stars* package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
@@
 However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
-
-<!-- JN: terra new vector representations -->
-<!-- MH: I probably would not add terra here.
-As far as I see it, the vector support in terra is basically just support to combine vector with raster data.
-The main focus of the package is raster data though -->
+Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019a}.
 
@@
 Since publicly available data is becoming more prominent, so are *R* packages to access them.
-Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to downlad region and country data, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download Sentinel-2 optical images.
-Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package *geodata* <!--(add reference) https://github.com/rspatial/geodata-->).
+Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to download region and country data, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download Sentinel-2 optical images.
+Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4.
 Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
@@
 However, most landscape ecology studies should utilize projected CRSs, which use some measurement units (e.g., meters).
-The selection of projected CRS should be based on the property of spatial data that we want to keep intact (e.g., does not distort areas, shapes, distances, or angles) and be appropriate for a given study area.
+The selection of projected CRS should be based on the property of spatial data that needs to be kept intact (e.g., no distortion of areas, shapes, distances, or angles) and be appropriate for a given study area.
 Tools to find an appropriate CRS for a certain region can be found at <https://spatialreference.org>, <https://epsg.org>, or <http://epsg.io>.
@@
 
-<!--JN: idea - a figure comparing viz of several packages-->
-<!-- MH: I like the idea: Maybe one base plot, one fancy map (tmap) and a screenshot of an interactive map? -->
-<!-- LJG: great idea - happy to be assigned some maps to make! -->
-<!-- MH: I already wrote that somewhere on GitHub I think.
-It would be super cool if we could download some data using one of the packages of the download section and then create the maps with it.
-That would allow to add the code to the appendix and give an example of two sections at once.
-If sombody wants to start with that, just assign yourself here https://github.com/r-spatialecology/Paper_open_source_tools/issues/4 -->
+```{r fig-map, echo = FALSE, out.width="100%", out.height="50%", fig.cap = "Comparison of different options to create maps using a) base plot, b) ggplot2, and c) tmap.
+All maps show the total annual precipitation of Switzerland.
+Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders  package.
+The code to create the maps can be found in the appendix."}
+include_graphics("data/plot_all.pdf")
+```
 
@@
 
-The introduction of the *FRAGSTATS* in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019}.
-However, while *FRAGSTATS* is free to download at the developers' homepage, it is not open-source software.
-This restricts its transparency, reproducibility, and extensibility.
-Recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed to provide a truly open-source solution to calculate landscape metrics.
-The package allows calculation of the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment.
+The introduction of the *FRAGSTATS* in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the *landscapemetrics* package \cite{Hesselbarth2019a} allows to calculate the most widely used landscape metrics in within the *R* environment.
 
@@
 <!-- LJG added some examples, but not 100% if they're the distances you're thinking of -->
+<!-- MH: Do we need a reference for L266f.? -->
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
-All of the methods mentioned above are implemented in the *motif* package \cite{R-motif}.
+All of the spatial signatures methods mentioned above are implemented in the *motif* package \cite{R-motif}.
 
@@
 
-Point pattern analysis is one such method. 
-This uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
+Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
 The *spatstat* package \cite{R-spatstat} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
@@
 
-Distance-based methods allow us to detect and correct for spatial autocorrelation in data.
+Distance-based methods allow to detect and correct for spatial autocorrelation in data.
 It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling. 
@@
 Geostatistics also allows for spatial data simulations.
+<!-- MH: I am not sure if I understand the last sentence, but is it worth to reference the NLM section here? -->
 
@@
 These analyses often use results of landscape ecological analyses as explanatory variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
-Widely used R packages include *ctmm* \cite{calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
+Widely used *R* packages include *ctmm* \cite{calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
 A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.
@@
 <!--a paragraph summarizing possibilities-->
+A comprehensive collection of *R* software package exist to handle the most common tasks and fields of landscape ecology. 
+Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, *R* allows for transparent and reproducible workflows.
+Additionally, this also allows to easily interchange, modify, or adapt methods from other related and unrelated fields.
 
 <!--a paragraph summarizing the survey-->
+The survey revealed that overall the landscape ecology community is well pleased with the capabilities of the *R* programming language for landscape ecology. 
+Furthermore, the survey showed the highly dynamic development of *R* packages by the community itself, showing that avaiable tools are constantly updated.
 
 <!--a paragraph about missing features/future work-->
+Landscape ecology combines and covers many different research topics and methodological approaches and most of them have in common that they heavily rely on spatial data.
+While the *R* programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands in terms of computational time as well as memory requirements the *R* programming language has to face.
+Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the *R* packages related to those fields.
+<!-- MH: I kind of tried to express that it's hard to come up with generel futur works suggestions since its highly sub-field depenend. Not sure that is really clear currently. -->
 

### 2021-01-23 paper/paper.tex
Update Conclusions

@@
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
-The expanding CRAN Task Views (lists of packages related to a certain topic) document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
+The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
 Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely higher.
@@
 The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
-Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable and common programming languages (e.g.~Python) for landscape ecology exist, in this review, we focus on software implemented in the \emph{R} programming language.
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review, we focus on software implemented in the \emph{R} programming language.
 For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 \emph{GitHub} hosts source code under version control, and allows users to install packages with one line of code using the \emph{remotes} \cite{Hester2020} package.
-Additionally, hosting a package on \emph{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing.
+Additionally, hosting a package on \emph{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e.~testing if functions return an expected value).
 
@@
 Thus, most packages can be found on both platforms, and many developers use \emph{GitHub} for regular development and \emph{CRAN} to publish stable releases of the packages.
-Furthermore, communities like \emph{rOpenSci} also provide a peer review process for code quality.
+Furthermore, online communities like \emph{rOpenSci} also provide a peer-review process for code quality.
 However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
@@
 \includegraphics[width=1\linewidth,height=0.3\textheight]{data/Workflow} \caption{Exemplary workflow of spatial data analyses for landscape ecology using the R programming language.
-For all major tasks, a few example R packages are listed.
-.}\label{fig:fig-workflow}
+For all major tasks, some example R packages are listed.}\label{fig:fig-workflow}
 \end{figure}
@@
 Currently, the most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
-A \emph{raster} successor, \emph{terra}, aimed at the simpler interface and improved performance is now being developed \cite{R-terra}; however, it could take several years for this package to be adopted by other developers and users.
+A \emph{raster} successor, \emph{terra}, aimed at the simpler interface and improved performance is being developed \cite{R-terra}, however, it could take several years for this package to be adopted by other developers and users.
 Alternatively, the \emph{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
@@
 However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
+Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019a}.
 
@@
 Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
-Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database to downlad region and country data, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download data from the Landsat 8, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
-Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
+Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
+Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
 Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
@@
 However, most landscape ecology studies should utilize projected CRSs, which use some measurement units (e.g., meters).
-The selection of projected CRS should be based on the property of spatial data that we want to keep intact (e.g., does not distort areas, shapes, distances, or angles) and be appropriate for a given study area.
+The selection of projected CRS should be based on the property of spatial data that needs to be kept intact (e.g., no distortion of areas, shapes, distances, or angles) and be appropriate for a given study area.
 Tools to find an appropriate CRS for a certain region can be found at \url{https://spatialreference.org}, \url{https://epsg.org}, or \url{http://epsg.io}.
@@
 
+\begin{figure}
+\includegraphics[width=1\linewidth,height=0.5\textheight]{data/plot_all} \caption{Comparison of different options to create maps using a) base plot, b) ggplot2, and c) tmap.
+All maps show the total annual precipitation of Switzerland.
+Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders  package.
+The code to create the maps can be found in the appendix.}\label{fig:fig-map}
+\end{figure}
+
 \hypertarget{sec:ecological_analysis}{%
@@
 
-The introduction of the \emph{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019}.
-However, while \emph{FRAGSTATS} is free to download at the developers' homepage, it is not open-source software.
-This restricts its transparency, reproducibility, and extensibility.
-Recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was developed to provide a truly open-source solution to calculate landscape metrics.
-The package allows calculation of the most widely used landscape metrics in a transparent and reproducible workflow within the \emph{R} environment.
+The introduction of the \emph{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \emph{landscapemetrics} package \cite{Hesselbarth2019a} allows to calculate the most widely used landscape metrics in within the \emph{R} environment.
 
@@
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
-All of the methods mentioned above are implemented in the \emph{motif} package \cite{R-motif}.
+All of the spatial signatures methods mentioned above are implemented in the \emph{motif} package \cite{R-motif}.
 
@@
 
-Point pattern analysis is one such method.
-This uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
+Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
 The \emph{spatstat} package \cite{R-spatstat} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
@@
 
-Distance-based methods allow us to detect and correct for spatial autocorrelation in data.
+Distance-based methods allow to detect and correct for spatial autocorrelation in data.
 It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling.
@@
 These analyses often use results of landscape ecological analyses as explanatory variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
-Widely used R packages include \emph{ctmm} \cite{calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
+Widely used \emph{R} packages include \emph{ctmm} \cite{calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
 A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.
@@
 In total, the survey was answered by 103 participants, of which the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
-Other, less frequent answers were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
@@
 
-We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'', and ``nature conservation'' (Fig.
+We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig.
 \ref{fig:fig-survey} A)).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
+Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 
-Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
+Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
 \ref{fig:fig-survey} D)).
@@
 
+A comprehensive collection of \emph{R} software package exist to handle the most common tasks and fields of landscape ecology.
+Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, \emph{R} allows for transparent and reproducible workflows.
+Additionally, this also allows to easily interchange, modify, or adapt methods from other related and unrelated fields.
+
+The survey revealed that overall the landscape ecology community is well pleased with the capabilities of the \emph{R} programming language for landscape ecology.
+Furthermore, the survey showed the highly dynamic development of \emph{R} packages by the community itself, showing that avaiable tools are constantly updated.
+
+Landscape ecology combines and covers many different research topics and methodological approaches and most of them have in common that they heavily rely on spatial data.
+While the \emph{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands in terms of computational time as well as memory requirements the \emph{R} programming language has to face.
+Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \emph{R} packages related to those fields.
+
 \hypertarget{acknowledgments}{%
@@
 
-MHKH was supported by the German Research Association (DFG) Research Training Group 1644 `Scaling Problems in Statistics', grant number 152112243.
+MHKH was supported by the German Research Association (DFG) Research Training Group 1644 `Scaling Problems in Statistics,' grant number 152112243.
 

### 2021-01-25 paper/paper.Rmd
several small updates

@@
 Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
-In this context, landscape ecology predominantly focuses on i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchanges within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) and can guide how to to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
+In this context, landscape ecology predominantly focuses on i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchanges within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) and can guide how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
 
@@
 Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.
-Python) for landscape ecology exist, in this review, we focus on software implemented in the *R* programming language.
+Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
 For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Because in addition to handling spatial data, tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
+<!-- In addition to handling spatial data, tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
 \@ref(fig:fig-workflow)), the *R* programming language can be a very powerful tool.
-Thus, in the first part, we present an overview of existing *R* packages for landscape ecology (Table \@ref(tab:packages)).
+-->
+<!-- JS: I tried to reorganize the previous sentence -->
+The *R* programming language can be a very powerful tool.
+In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig.
+\@ref(fig:fig-workflow)).
+
+Thus, in the first part of this article, we present an overview of existing *R* packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
 In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are presently missing for landscape ecology.
@@
 The guaranteed technical quality standard on *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
-Additionally, the technical quality standard on *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y* \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
+Additionally, the technical quality standard on *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y* <!--\cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}-->.
+<!-- JS: ..the technical quality standard on *CRAN* also... on or of CRAN? not sure here-->
 <!-- MH: Not sure if these references belong here? -->
-CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability of code.
+<!-- JS: I think not. I commented the rerences out, so you can delete if you agree -->
+CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
 Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
@@
 
+<!--JS: for or at many scales? -->
 Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
-A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
+A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
 Since publicly available data is becoming more prominent, so are *R* packages to access them.
@@
 The selection of projected CRS should be based on the property of spatial data that needs to be kept intact (e.g., no distortion of areas, shapes, distances, or angles) and be appropriate for a given study area.
+<!-- JS: added a sentense to EPSG codes here -->
+A common way to refer to different CRS is to used codes developed by the European Petrol Survey Group (EPSG).
 Tools to find an appropriate CRS for a certain region can be found at <https://spatialreference.org>, <https://epsg.org>, or <http://epsg.io>.
@@
 
-Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
+Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R-->
+<!-- JS: consider remove Gladly -->
+Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
 
@@
 Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
-<!--- JS: I wonder if we should also cite js library? However, doing a quick google, I was not able to find a correct citation -->
 
@@
 
-Related to SDMs, there are growing number of *R* packages to analyze data from tracked animals, to study their movement characteristics, space use and interaction with other animals and the environment.
-These analyses often use results of landscape ecological analyses as explanatory variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
+Related to SDMs, there are a growing number of *R* packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
+These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
 Widely used *R* packages include *ctmm* \cite{calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
@@
 <!-- MH: Johannes, do you have more insights into the packages here? See comment LJG in L307 -->
+<!-- JS: No not really, I havent worked landscape genetics for some time -->
 

### 2021-01-25 paper/paper.tex
several small updates

@@
 The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
-Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review, we focus on software implemented in the \emph{R} programming language.
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
 For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-Because in addition to handling spatial data, tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
-\ref{fig:fig-workflow}), the \emph{R} programming language can be a very powerful tool.
-Thus, in the first part, we present an overview of existing \emph{R} packages for landscape ecology (Table \ref{tab:packages}).
+The \emph{R} programming language can be a very powerful tool.
+In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig.
+\ref{fig:fig-workflow}).
+
+Thus, in the first part of this article, we present an overview of existing \emph{R} packages for landscape ecology and closely related fields (Table \ref{tab:packages}).
 In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are presently missing for landscape ecology.
@@
 The guaranteed technical quality standard on \emph{CRAN} requires more initial work for developers compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
-Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y} \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
-CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability of code.
+Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y} .
+CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
 Thus, most packages can be found on both platforms, and many developers use \emph{GitHub} for regular development and \emph{CRAN} to publish stable releases of the packages.
@@
 Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
-A lot of this data is publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
+A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
 Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
@@
 The selection of projected CRS should be based on the property of spatial data that needs to be kept intact (e.g., no distortion of areas, shapes, distances, or angles) and be appropriate for a given study area.
+A common way to refer to different CRS is to used codes developed by the European Petrol Survey Group (EPSG).
 Tools to find an appropriate CRS for a certain region can be found at \url{https://spatialreference.org}, \url{https://epsg.org}, or \url{http://epsg.io}.
@@
 
-Related to SDMs, there are growing number of \emph{R} packages to analyze data from tracked animals, to study their movement characteristics, space use and interaction with other animals and the environment.
-These analyses often use results of landscape ecological analyses as explanatory variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
+Related to SDMs, there are a growing number of \emph{R} packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
+These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
 Widely used \emph{R} packages include \emph{ctmm} \cite{calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
@@
 In total, the survey was answered by 103 participants, of which the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
-Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
+Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
 Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 
-Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
+Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
 \ref{fig:fig-survey} D)).

### 2021-01-26 paper/paper.Rmd
update abstract, ljg affiliation, glossary, and minor typos

@@
 - name: Laura J. Graham
-  address: Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
+  address: 
+    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
+    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
   email: l.graham@bham.ac.uk
@@
 abstract: |
-  max.
-(200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+  Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years.
+In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology.
+In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+Additionally, we outline the results of a survey of *R* users within the landscape ecology community.
+In this we find that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually updating the functionality available.
+Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
 bibliography: bibliography.bib
@@
 
-API & Application programming interface \\
+API & Application programming interface. A set of protocols for sending and retrieving information from a server \\
 cell & Smallest, rectangular unit of raster data model \\
-CRAN & Comprehensive R Archive Network \\
-CRS & Coordinate references systems \\
+CRAN & Comprehensive R Archive Network. A repository for R packages \\
+CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
 GIS & Geographical information system \\
@@
 R package & User-created software extension to R programming language \\
-SDM & Species distribution modeling \\
+SDM & Species distribution modelling \\
 vector & Points, lines, and polygons-based data model \\
@@
 
-While *R* has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+While *R* has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
 However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
@@
 Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package.
-Also, packages specifically designed for SDM exists, including includes the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages.
+Also, packages specifically designed for SDM exist, including the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages.
+Additionally, packages such as *ENMeval* \cite{R-enmeval} provide functionality for model tuning and evaluation.
 <!-- These packages [check!] allow the user to create ensemble models of SDMs based on different underlying model structures such as Maxent, regression trees, and generalized linear or additive models.
@@
 Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
-Further functionality for landscape genetics such as ...
-<!-- LJG: I think a little more info needed here - happy to dig through the packages to add in, but thought I'd leave for now incase anyone knows off top of their heads -->can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.
+Further functionality for landscape genetics such as quantifying and analysing population genetic structure, and hierarchical decomposition analysis can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.
 
@@
 & zoon & Reproducible and remixable species distribution modelling & \cite{Golding2018} \\
+& ENMeval & Model tuning and evaluation & \cite{R-enmeval} \\
 & adehabitatHR, adehabitatHS & Home range and habitat selection modelling & \cite{Calenge2006} \\
@@
 Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey) B).
-Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
+The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
 
@@
 
-Interestingly, more people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
+More people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
 This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-survey) C)).
@@
 <!--a paragraph summarizing possibilities-->
-A comprehensive collection of *R* software package exist to handle the most common tasks and fields of landscape ecology. 
+A comprehensive collection of *R* software packages exist to handle the most common tasks and fields of landscape ecology. 
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, *R* allows for transparent and reproducible workflows.
-Additionally, this also allows to easily interchange, modify, or adapt methods from other related and unrelated fields.
+Additionally, this also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
 
 <!--a paragraph summarizing the survey-->
-The survey revealed that overall the landscape ecology community is well pleased with the capabilities of the *R* programming language for landscape ecology. 
+The survey revealed that overall the landscape ecology community is satisfied with the capabilities of the *R* programming language for landscape ecology. 
 Furthermore, the survey showed the highly dynamic development of *R* packages by the community itself, showing that avaiable tools are constantly updated.

### 2021-01-26 paper/paper.tex
update abstract, ljg affiliation, glossary, and minor typos

@@
         Laura J. Graham \at
-     Geography, Earth \& Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom \\
+     a) Geography, Earth \& Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom b) Biodiversity, Ecology \& Conservation Group, IIASA, Vienna, Austria \\
      \email{\href{mailto:l.graham@bham.ac.uk}{\nolinkurl{l.graham@bham.ac.uk}}}  %  \\
@@
 \begin{abstract}
-max.
-(200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
+Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years.
+In particular, the \emph{R} programming language provides a wealth of community developed tools for landscape ecology.
+In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+Additionally, we outline the results of a survey of \emph{R} users within the landscape ecology community.
+In this we find that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually updating the functionality available.
+Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
 \\
@@
 
-While \emph{R} has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+While \emph{R} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
 However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
@@
 Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
-Also, packages specifically designed for SDM exists, including includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008}, or \emph{zoon} \cite{Golding2018} packages.
+Also, packages specifically designed for SDM exist, including the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008}, or \emph{zoon} \cite{Golding2018} packages.
+Additionally, packages such as \emph{ENMeval} \cite{R-enmeval} provide functionality for model tuning and evaluation.
 
@@
 Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
-Further functionality for landscape genetics such as \ldots{} can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.
+Further functionality for landscape genetics such as quantifying and analysing population genetic structure, and hierarchical decomposition analysis can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.
 
@@
 Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
-Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
+The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
 
-Interestingly, more people use the raster data model (72.8\%) in comparison to the vector data model (27.2\%).
+More people use the raster data model (72.8\%) in comparison to the vector data model (27.2\%).
 This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey} C)).
@@
 
-A comprehensive collection of \emph{R} software package exist to handle the most common tasks and fields of landscape ecology.
+A comprehensive collection of \emph{R} software packages exist to handle the most common tasks and fields of landscape ecology.
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, \emph{R} allows for transparent and reproducible workflows.
-Additionally, this also allows to easily interchange, modify, or adapt methods from other related and unrelated fields.
+Additionally, this also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
 
-The survey revealed that overall the landscape ecology community is well pleased with the capabilities of the \emph{R} programming language for landscape ecology.
+The survey revealed that overall the landscape ecology community is satisfied with the capabilities of the \emph{R} programming language for landscape ecology.
 Furthermore, the survey showed the highly dynamic development of \emph{R} packages by the community itself, showing that avaiable tools are constantly updated.

### 2021-01-26 paper/paper.Rmd
added gdal, proj, epsg and simple features to glossary

@@
 CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
+EPSG & Public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement \\
+GDAL & Open source translator library for raster and vector geospatial data formats \\
 GIS & Geographical information system \\
@@
 patch & Neighboring cells with same value \\
+PROJ & Open source generic coordinate transformation software \\
 raster & Cell grid based data model \\
@@
 SDM & Species distribution modelling \\
+Simple Features & A set of standards that specify a common storage and access model of geographic features \\
 vector & Points, lines, and polygons-based data model \\

### 2021-01-27 paper/paper.Rmd
Adding simple features and adding some reference currently missing

--- 2021-01-26 paper/paper.Rmd
@@
 cell & Smallest, rectangular unit of raster data model \\
-CRAN & Comprehensive R Archive Network. A repository for R packages \\
+CRAN & Comprehensive R Archive Network. A repository for *R* package \\
 CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
+EPSG & Public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement \\
+GDAL & Open source translator library for raster and vector geospatial data formats \\
 GIS & Geographical information system \\
@@
 patch & Neighboring cells with same value \\
+PROJ & Open source generic coordinate transformation software \\
 raster & Cell grid based data model \\
 resolution & Size in meter or degree of one cell \\
-R package & User-created software extension to R programming language \\
+*R* package & User-created software extension to *R* programming language \\
 SDM & Species distribution modelling \\
+Simple Features & A set of standards that specify a common storage and access model of geographic features \\
 vector & Points, lines, and polygons-based data model \\
@@
 
-The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
 Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.
@@
 For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-<!-- In addition to handling spatial data, tasks such as statistical modeling, creation of publication-ready figures, and even preparation of complete reports can be done in the same working environment (Fig.
-\@ref(fig:fig-workflow)), the *R* programming language can be a very powerful tool.
--->
-<!-- JS: I tried to reorganize the previous sentence -->
 The *R* programming language can be a very powerful tool.
@@
 \@ref(fig:fig-workflow)).
+The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
 
@@
 
-The guaranteed technical quality standard on *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
-Additionally, the technical quality standard on *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y* <!--\cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}-->.
-<!-- JS: ..the technical quality standard on *CRAN* also... on or of CRAN? not sure here-->
-<!-- MH: Not sure if these references belong here? -->
-<!-- JS: I think not. I commented the rerences out, so you can delete if you agree -->
+The guaranteed technical quality standard of *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
+Additionally, the technical quality standard of *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y*.
 CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
@@
 ```
-
-<!--JN: the above figure is nice; we should update it when everything else is done-->
 
@@
 The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
-In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands <!--(add reference)-->.
+In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
 The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
@@
 Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019a}.
+*sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 
-<!--JS: for or at many scales? -->
-Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
+Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
 A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
@@
 
-Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps <!-- (add reference) -->.
+Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps \cite{Lovelace2019,Pebesma2019a}.
 This is also referred to as the spatial projection and is often the first barrier in spatial data analysis.
 It is not only required to have all of the used data in the same projection, but also to select a proper CRS.
-This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales <!-- (add reference) -->.
+This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales \cite{Bivand2013,Lovelace2019}.
 Coordinates in spatial data represent one of many coordinate reference systems.
@@
 The selection of projected CRS should be based on the property of spatial data that needs to be kept intact (e.g., no distortion of areas, shapes, distances, or angles) and be appropriate for a given study area.
-<!-- JS: added a sentense to EPSG codes here -->
 A common way to refer to different CRS is to used codes developed by the European Petrol Survey Group (EPSG).
@@
 
-Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R-->
-<!-- JS: consider remove Gladly -->
-Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
+Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
+It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
 
@@
 
+Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
+Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
+Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
 The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
-Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
-Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
-<!-- Advanced plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}. -->
-<!-- JN: I do not know if we should add rasterVis here... -->
-<!-- LJG: out of interest why not? -->
-<!-- MH: I would also be interesed. I sometimes like to use it for specific raster :)  -->
-
 Quick interactive visualization of spatial data can be done with the *mapview* package\cite{R-mapview}.
 Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
-
 A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
@@
 All maps show the total annual precipitation of Switzerland.
-Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders  package.
-The code to create the maps can be found in the appendix."}
+Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders) package.
+The code to create the maps can be found in the Appendix."}
 include_graphics("data/plot_all.pdf")
@@
 Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
+<!-- MH: Do we have a reference here @Jakub? -->
 Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
 <!-- LJG added some examples, but not 100% if they're the distances you're thinking of -->
-<!-- MH: Do we need a reference for L266f.? -->
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
@@
 #### Spatial statistics
-
-<!--JN: I would rather say that geostatistics is a subbranch of spatial statistics (e.g. https://www.geos.ed.ac.uk/~gisteac/gis_book_abridged/files/ch16.pdf). 
-However, we can revert my changes if you want-->
 
@@
 Additionally, packages such as *ENMeval* \cite{R-enmeval} provide functionality for model tuning and evaluation.
-<!-- These packages [check!] allow the user to create ensemble models of SDMs based on different underlying model structures such as Maxent, regression trees, and generalized linear or additive models.
-This is considered the gold standard in contemporary SDM \cite(Araujo2007).-->
 
-Related to SDMs, there are a growing number of *R* packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
+Related to SDMs, there is a growing number of *R* packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
 These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
@@
 A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.
-
-<!-- Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}. However, currently the *maxent* package is not available on *CRAN* anymore. -->
-<!-- LJG: maxent can be applied through biomod2 - I think it's worth mentioning because it's pretty much the most widely used SDM (whether this is a good thing or not is another question)-->
-<!-- LJG: Another thought - it's generally accepted within SDM that ensemble models are superior to using single models.
-I know that biomod and zoon facilitate this.
-Can have a look at the other ones to see if they do.
-Have added a potential sentence as a comment -->
-<!-- MH: I basically have no idea about SDM at all! So please feel free to change this section as you want. Pretty sure Johannes has some inputs about this as well. -->
 
@@
 
-<!-- MH: Johannes, do you have more insights into the packages here? See comment LJG in L307 -->
-<!-- JS: No not really, I havent worked landscape genetics for some time -->
-
 #### Neutral landscape models {#sec:NLM}
@@
 
-\caption{Overview of commonly used R packages for landscape ecology. Packages are sorted by their major application task. Only packages focused on spatial data and landscape ecology are included.}
+\caption{Overview of commonly used R packages for spatial data and landscape ecology. Packages are sorted by their major application task.}
 
@@
 \hline
-\multirow{7}{*}{Creating maps} & mapview & Interactive viewing of spatial data & \cite{R-mapview} \\
+\multirow{7}{*}{Creating maps} & rasterVis & Vizualuzation of raster data & \cite{Lamigueiro2020} \\
+& ggspatial & Spatial extension for ggplot2 & \cite{Dunnington2020} \\
+& cartography & Create cartographic maps & \cite{Giraud2016} \\
 & tmap & Create thematic maps & \cite{Tennekes2018} \\
+& mapview & Interactive viewing of spatial data & \cite{R-mapview} \\
 & leaflet & Create interactive web maps & \cite{R-leaflet} \\
-& cartography & Create cartographic maps & \cite{Giraud2016} \\
-& ggspatial & Spatial extension for ggplot2 & \cite{Dunnington2020} \\
 & rayshader & Create 2D and 3D data visualizations & \cite{Morgen-Wall2020} \\
@@
 ```{r fig-survey, echo = FALSE, fig.align="center", fig.cap = "Results of the online survey about open-source software tools in R for landscape ecology.
-Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology.
+Results include a) which terms describe major reserach topics the best, b) the most important workflow task, c) the most used R packages and d) the overall usefulnes or R for landscape ecology.
 The 'others' category includes all answers with less than five total mentions.", warning=FALSE, out.width="100%"}
@@
   # plot_layout(heights = c(2, 3)) +
-  plot_annotation(tag_levels = "A", tag_suffix = ")")
+  plot_annotation(tag_levels = "a")
 
@@
 
-<!--JN: I think we should answer to some of the comments in the missing packages section: point people to some packages (if possible), mention which of the missing tools are not going to be implemented, and also forward people to methods and tools on how to improve some problems (e.g.
-speed and required RAM for larger data -> use of Rcpp, use external servers, or read data in chunks, etc.)-->
-
-<!-- MH: I am not sure if we should do this explicitly here  That would somehow repeat the section we already have about the different packages.
-We should, however, make sure that in case we know a package that answers one of the points here, is included in the first section of the paper -->
-
 # Conclusions
@@
 <!--a paragraph summarizing possibilities-->
-A comprehensive collection of *R* software packages exist to handle the most common tasks and fields of landscape ecology. 
+A comprehensive collection of *R* software packages exist to handle the most common tasks of landscape ecology. 
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, *R* allows for transparent and reproducible workflows.
@@
 <!--a paragraph summarizing the survey-->
-The survey revealed that overall the landscape ecology community is satisfied with the capabilities of the *R* programming language for landscape ecology. 
-Furthermore, the survey showed the highly dynamic development of *R* packages by the community itself, showing that avaiable tools are constantly updated.
+The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the *R* programming language for landscape ecology. 
+Furthermore, the survey showed the that many members of the landscape ecology community actively develop *R* packages themselves, demonstrating that tools are constantly added and updated.
 
 <!--a paragraph about missing features/future work-->
-Landscape ecology combines and covers many different research topics and methodological approaches and most of them have in common that they heavily rely on spatial data.
-While the *R* programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands in terms of computational time as well as memory requirements the *R* programming language has to face.
+Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
+While the *R* programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the *R* programming language has to face.
 Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the *R* packages related to those fields.
-<!-- MH: I kind of tried to express that it's hard to come up with generel futur works suggestions since its highly sub-field depenend. Not sure that is really clear currently. -->
+But as the development of *R* packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
 
@@
 
+<!-- MH: Make sure to update all sections  -->
+
 We are thankful to all members of the 'Coastal Ecology and Conservation Lab' at University of Michigan for comments on an early draft of the manuscript.
-
-<!-- MH: Make sure to update all sections with information of all authors. Out-of-date currently -->
 
@@
 JN is author of *belg* and *motif* package. JS is author of the *amt* package. LJG is author of the *grainchanger* package.
-<!-- MH: I wasn't sure if we need to list this here, but it kind of felt like a not completely stupid idea. What do you think? -->
-<!-- JS: I don't think its critical, but surely it doesn't do any arm listing this here -->
 
@@
 
+<!-- MH: Add UMich Funding? -->
+
 MHKH was supported by the German Research Association (DFG) Research Training Group 1644 'Scaling Problems in Statistics', grant number 152112243.
-<!-- MH: Add UMich Funding? -->
 

### 2021-01-27 paper/paper.tex
Adding simple features and adding some reference currently missing

--- 2021-01-26 paper/paper.tex
@@
 
-The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
 Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
@@
 \ref{fig:fig-workflow}).
+The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
 
@@
 
-The guaranteed technical quality standard on \emph{CRAN} requires more initial work for developers compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
-Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y} .
+The guaranteed technical quality standard of \emph{CRAN} requires more initial work for developers compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
+Additionally, the technical quality standard of \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y}.
 CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
@@
 The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
-In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands .
+In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
 The \emph{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
@@
 Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019a}.
+\emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 
-Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
+Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
 A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
@@
 
-Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps .
+Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps \cite{Lovelace2019,Pebesma2019a}.
 This is also referred to as the spatial projection and is often the first barrier in spatial data analysis.
 It is not only required to have all of the used data in the same projection, but also to select a proper CRS.
-This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales .
+This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales \cite{Bivand2013,Lovelace2019}.
 Coordinates in spatial data represent one of many coordinate reference systems.
@@
 Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
-Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.
+It is possible to control several GIS software directly from \emph{R} using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.
 
@@
 
+Further plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
+Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
+Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
 The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
-Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
-Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
-
 Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{R-mapview}.
 Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
-
 A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
@@
 All maps show the total annual precipitation of Switzerland.
-Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders  package.
-The code to create the maps can be found in the appendix.}\label{fig:fig-map}
+Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders) package.
+The code to create the maps can be found in the Appendix.}\label{fig:fig-map}
 \end{figure}
@@
 
-Related to SDMs, there are a growing number of \emph{R} packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
+Related to SDMs, there is a growing number of \emph{R} packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
 These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
+Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 \caption{Results of the online survey about open-source software tools in R for landscape ecology.
-Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology.
+Results include a) which terms describe major reserach topics the best, b) the most important workflow task, c) the most used R packages and d) the overall usefulnes or R for landscape ecology.
 The 'others' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
@@
 
-A comprehensive collection of \emph{R} software packages exist to handle the most common tasks and fields of landscape ecology.
+A comprehensive collection of \emph{R} software packages exist to handle the most common tasks of landscape ecology.
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, \emph{R} allows for transparent and reproducible workflows.
@@
 
-The survey revealed that overall the landscape ecology community is satisfied with the capabilities of the \emph{R} programming language for landscape ecology.
-Furthermore, the survey showed the highly dynamic development of \emph{R} packages by the community itself, showing that avaiable tools are constantly updated.
+The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the \emph{R} programming language for landscape ecology.
+Furthermore, the survey showed the that many members of the landscape ecology community actively develop \emph{R} packages themselves, demonstrating that tools are constantly added and updated.
 
-Landscape ecology combines and covers many different research topics and methodological approaches and most of them have in common that they heavily rely on spatial data.
-While the \emph{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands in terms of computational time as well as memory requirements the \emph{R} programming language has to face.
+Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
+While the \emph{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \emph{R} programming language has to face.
 Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \emph{R} packages related to those fields.
+But as the development of \emph{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
 

### 2021-01-27 paper/paper.Rmd
Adding headings to abstract

@@
 title: Open-source tools in R for landscape ecology
-# subtitle: Do you have a subtitle? If so, write it here
-# titlerunning: Open source landscape ecology tools
-# authorrunning: Short form of author list if too long for running head
-# thanks: |
 authors: 
@@
 - scientific software
-- ...
 abstract: |
-  Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years.
+  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years.
 In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology.
-In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+**Recent findings** In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
 Additionally, we outline the results of a survey of *R* users within the landscape ecology community.
-In this we find that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually updating the functionality available.
+**Summary** In this we find that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually updating the functionality available.
 Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.

### 2021-01-27 paper/paper.tex
Adding headings to abstract

@@
 \begin{abstract}
-Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years.
+\textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years.
 In particular, the \emph{R} programming language provides a wealth of community developed tools for landscape ecology.
-In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+\textbf{Recent findings} In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
 Additionally, we outline the results of a survey of \emph{R} users within the landscape ecology community.
-In this we find that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually updating the functionality available.
+\textbf{Summary} In this we find that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually updating the functionality available.
 Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
@@
         scientific software \and
-        \ldots{} \and
     }

### 2021-01-27 paper/paper.Rmd
Final update survey data

@@
   dplyr::arrange(-n_rel)
-
 ```
@@
 
+# remove escape characters
+r_packages <- str_remove_all(r_packages, pattern = "\"")
+
 # create data.frame
@@
   theme(plot.title.position = "plot") +
-  scale_x_continuous(breaks = seq(0, 35, 5), limits = c(0, 35)) +
+  scale_x_continuous(breaks = seq(0, 35, 5), limits = c(0, 36)) +
   scale_fill_manual(values = c("grey50", "black")) +
@@
            "not enough overview to answer this", "?", "i am currently missing no methods",
-           ";", "nothing yet", "i'm thinking", "nil")
+           ";", "nothing yet", "i'm thinking", "nil", "none", 
+           "nothing really but prefer to do programming in python", "i dont miss any")
 

### 2021-01-27 paper/paper.tex
Final update survey data

@@
 To better understand how the landscape ecology community uses \emph{R}, we conducted a short survey using mailing lists and social media to reach the community.
-In total, the survey was answered by 103 participants, of which the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
+In total, the survey was answered by 109 participants, of which the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (14.7\%).
 Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
-Most people use \emph{R} either ``daily'' (54.4\%) or a ``few times a week'' (36.9\%).
-Almost half of all participants described themselves as ``advanced'' users (46.6\%), while 40.8\% described themselves as ``intermediate'' users.
-Related to this, about half of the participants either implemented their own package (21.4\%) or plan to do so in the future (23.3\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
+Most people use \emph{R} either ``daily'' (53.2\%) or a ``few times a week'' (36.7\%).
+Almost half of all participants described themselves as ``advanced'' users (46.8\%), while 40.4\% described themselves as ``intermediate'' users.
+Related to this, about half of the participants either implemented their own package (21.1\%) or plan to do so in the future (22\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
 
@@
 
-More people use the raster data model (72.8\%) in comparison to the vector data model (27.2\%).
+More people use the raster data model (72.5\%) in comparison to the vector data model (27.5\%).
 This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey} C)).
-When asked for the three most used packages, participants of the survey listed 83 packages in total.
+When asked for the three most used packages, participants of the survey listed 93 packages in total.
 The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
 Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey} C).
-Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 33.44\%).
+Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 35.87\%).
 
-Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig.
+Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
 \ref{fig:fig-survey} D)).
@@
 
-Overall, 22.3\% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
+Overall, 23.9\% of the participants reported that currently no packages and functionality are missing for them or they lack the overview to answer the question.
 There were three most common topics across the answers of the participants.
-Firstly, many participants (13.6\%) wished for a better computational performance of \emph{R} in terms of speed and required RAM, especially for larger data.
-Secondly, participants are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (9.7\%).
-Thirdly, many participants (8.7\%) are currently missing advanced and easy-to-apply methods to create high-quality maps or other visualization-related functionality.
+Firstly, many participants (12.8\%) wished for a better computational performance of \emph{R} in terms of speed and required RAM, especially for larger data.
+Secondly, participants are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (9.2\%).
+Thirdly, many participants (8.3\%) are currently missing advanced and easy-to-apply methods to create high-quality maps or other visualization-related functionality.
 

### 2021-01-27 paper/paper.Rmd
Remove Umlaute from author information

@@
 - name: Johannes Signer
-  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany
+  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
   email: jsigner@uni-goettingen.de

### 2021-01-27 paper/paper.tex
Remove Umlaute from author information

@@
         Johannes Signer \at
-     Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany \\
+     Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany \\
      \email{\href{mailto:jsigner@uni-goettingen.de}{\nolinkurl{jsigner@uni-goettingen.de}}}  %  \\

### 2021-01-27 paper/paper.Rmd
Small edits in the abstract.

@@
 In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology.
-**Recent findings** In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+**Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
 Additionally, we outline the results of a survey of *R* users within the landscape ecology community.
-**Summary** In this we find that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually updating the functionality available.
+**Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available.
 Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
@@
 *R* package & User-created software extension to *R* programming language \\
-SDM & Species distribution modelling \\
+SDM & Species distribution modeling \\
 Simple Features & A set of standards that specify a common storage and access model of geographic features \\

### 2021-01-27 paper/paper.tex
Small edits in the abstract.

@@
 In particular, the \emph{R} programming language provides a wealth of community developed tools for landscape ecology.
-\textbf{Recent findings} In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+\textbf{Recent findings} In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
 Additionally, we outline the results of a survey of \emph{R} users within the landscape ecology community.
-\textbf{Summary} In this we find that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually updating the functionality available.
+\textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually further develop the functionality available.
 Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
@@
 Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019a}.
-\emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
+\emph{sf} builds on the idea of ``simple features'', a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 In total, the survey was answered by 109 participants, of which the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (14.7\%).
-Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
+Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 
-Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
+Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
 \ref{fig:fig-survey} D)).

### 2021-01-27 R/code_example.R
Adjusted example panels 1 and 2.

@@ -3,6 +3,7 @@ library(sf)
 library(raster)
 library(rnaturalearth)
 library(tmap)
+library(viridis)
 
 # get precipitation data
 precipitation <- raster::getData("worldclim", lon = 3, lat = 38,
@@ -26,26 +27,32 @@ switzerland <- sf::st_transform(x = switzerland, crs = "epsg:21781")
 precipitation_sum_ch <- raster:::crop(x = precipitation_sum,
                                       y = sf::st_buffer(x = switzerland, dist = 5000))
 
+# reclassify into 5 classes
+precipitation_class_ch <- raster::cut(precipitation_sum_ch, breaks = 5)
+
 # create base plot
 # pdf(file = "R/Figures/base_plot.pdf")
 
-plot(precipitation_sum_ch)
+plot(precipitation_class_ch, col = viridis(5))
+plot(st_geometry(switzerland), add = TRUE, lwd = 4, border = "white")
 plot(st_geometry(switzerland), add = TRUE)
 
 # dev.off()
 
-# reclassify into 5 classes
-precipitation_class_ch <- raster::cut(precipitation_sum_ch, breaks = 5)
 
 # create ggplot
 # MH: Why is this on long/lat? Coordinates seem to be okay?
+# JS: We could keep coords by adding:
+# `coord_sf(datum = st_crs(switzerland))`
+# but I actually like long/lat better, it looks cleaner
 plot_gg <- ggplot(raster::as.data.frame(precipitation_class_ch, xy = TRUE)) +
     geom_raster(aes(x = x, y = y, fill = factor(layer))) +
     geom_sf(data = switzerland, fill = NA, col = "white") +
     scale_fill_viridis_d(name = "Precipitation\nclassified") +
     coord_sf() +
     labs(x = "", y = "") +
-    theme_classic()
+    theme_classic() +
+    theme(legend.position = "bottom")
 
 ggplot2::ggsave(filename = "R/Figures/ggplot2.pdf")
 

### 2021-01-27 paper/paper.Rmd
Checked refs to figs and tables. Added references to Fig. 2 and updated caption.

@@
 changing the displayed values.
-All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function.
+All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
 However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
@@
 Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
-Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
+Also the popular plotting package *ggplot2* \cite{Wickham2016a} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
@@
 Quick interactive visualization of spatial data can be done with the *mapview* package\cite{R-mapview}.
-Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
+Both, *tmap* (Fig. \@ref(fig:fig-map)c) and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
 A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
 
-```{r fig-map, echo = FALSE, out.width="100%", out.height="50%", fig.cap = "Comparison of different options to create maps using a) base plot, b) ggplot2, and c) tmap.
+(ref:capfigmap) Comparison of different options to create maps using a) base plot, b) *ggplot2* package, and c) *tmap* package.
 All maps show the total annual precipitation of Switzerland.
-Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders) package.
-The code to create the maps can be found in the Appendix."}
+Data were downloaded with the *raster* package (precipitation data) and *rnaturalearth* package (country borders) package.
+The code to create the maps can be found in the Appendix.
+```{r fig-map, echo = FALSE, out.width="100%", out.height="50%", fig.cap = "(ref:capfigmap)"}
 include_graphics("data/plot_all.pdf")
@@
 We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10% of participants included "biodiversity", followed by "land use management", "landscape connectivity", and "nature conservation" (Fig.
-\@ref(fig:fig-survey) A)).
+\@ref(fig:fig-survey)a)).
 
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey) B).
+Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey)b).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
@@
 More people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
-This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-survey) C)).
+This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-survey)c).
 When asked for the three most used packages, participants of the survey listed `r length(unique(r_packages))` packages in total.
 The *raster* package was mentioned the most, followed by the *sf* package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey) C).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey)c.
 Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], 2)`%).
@@
 Lastly, when asked how useful *R* is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
-\@ref(fig:fig-survey) D)).
+\@ref(fig:fig-survey)d)).
 

### 2021-01-27 paper/paper.tex
Checked refs to figs and tables. Added references to Fig. 2 and updated caption.

@@
 Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.~changing the displayed values.
-All packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the generic \texttt{plot()} function.
+All packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the generic \texttt{plot()} function (Fig. \ref{fig:fig-map}a).
 However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
@@
 Further plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
-Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
+Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} (Fig. \ref{fig:fig-map}b) has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
@@
 Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{R-mapview}.
-Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
+Both, \emph{tmap} (Fig. \ref{fig:fig-map}c) and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
 A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth,height=0.5\textheight]{data/plot_all} \caption{Comparison of different options to create maps using a) base plot, b) ggplot2, and c) tmap.
+\includegraphics[width=1\linewidth,height=0.5\textheight]{data/plot_all} \caption{Comparison of different options to create maps using a) base plot, b) \emph{ggplot2} package, and c) \emph{tmap} package.
 All maps show the total annual precipitation of Switzerland.
-Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders) package.
+Data were downloaded with the \emph{raster} package (precipitation data) and \emph{rnaturalearth} package (country borders) package.
 The code to create the maps can be found in the Appendix.}\label{fig:fig-map}
@@
 We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'', and ``nature conservation'' (Fig.
-\ref{fig:fig-survey} A)).
+\ref{fig:fig-survey}a)).
 
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
+Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey}b).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 More people use the raster data model (72.5\%) in comparison to the vector data model (27.5\%).
-This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey} C)).
+This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey}c).
 When asked for the three most used packages, participants of the survey listed 93 packages in total.
 The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
-Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey} C).
+Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey}c.
 Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 35.87\%).
@@
 Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
-\ref{fig:fig-survey} D)).
+\ref{fig:fig-survey}d)).
 

### 2021-01-27 R/code_example.R
improves code

@@ -24,19 +24,17 @@ switzerland <- rnaturalearth::ne_countries(scale = 50,
 switzerland <- sf::st_transform(x = switzerland, crs = "epsg:21781")
 
 # crop and mask with buffer
-precipitation_sum_ch <- raster:::crop(x = precipitation_sum,
-                                      y = sf::st_buffer(x = switzerland, dist = 5000))
+precipitation_sum_ch <- raster::crop(x = precipitation_sum,
+                                     y = sf::st_buffer(x = switzerland, dist = 5000))
 
 # reclassify into 5 classes
 precipitation_class_ch <- raster::cut(precipitation_sum_ch, breaks = 5)
 
 # create base plot
 # pdf(file = "R/Figures/base_plot.pdf")
-
 plot(precipitation_class_ch, col = viridis(5))
 plot(st_geometry(switzerland), add = TRUE, lwd = 4, border = "white")
 plot(st_geometry(switzerland), add = TRUE)
-
 # dev.off()
 
 
@@ -53,39 +51,43 @@ plot_gg <- ggplot(raster::as.data.frame(precipitation_class_ch, xy = TRUE)) +
     labs(x = "", y = "") +
     theme_classic() +
     theme(legend.position = "bottom")
-
-ggplot2::ggsave(filename = "R/Figures/ggplot2.pdf")
+# ggplot2::ggsave(filename = "R/Figures/ggplot2.pdf")
 
 # create interactive tmap (take screenshot for paper)
 # tmap_mode("plot")
-tmap_mode("view")
+# tmap_mode("view")
 
 plot_tm = tm_shape(precipitation_class_ch) +
     tm_graticules() +
-    tm_raster(title = "Precipitation\nclassified", style = "cont",
-              palette = "viridis") +
+    tm_raster(title = "Precipitation\nclassified",
+              style = "cat",
+              palette = "viridis",
+              legend.is.portrait = FALSE) +
     tm_shape(switzerland) +
     tm_borders(lwd = 2, col = "white") +
     tm_scale_bar(breaks = c(0, 25, 50),
                  bg.color = "white") +
-    tm_layout(legend.outside = TRUE)
+    tm_layout(legend.outside = TRUE,
+              legend.outside.position = "bottom")
 
-tmap::tmap_save(plot_tm, "R/Figures/tmap.pdf")
[…]

### 2021-01-30 paper/paper.Rmd
Consistenlty named put *R* in italic.

@@
 - statistical programming language
-- R packages
+- "*R* packages"
 - reproducibility
@@
 cell & Smallest, rectangular unit of raster data model \\
-CRAN & Comprehensive R Archive Network. A repository for *R* package \\
+CRAN & Comprehensive R Archive Network. A repository for \textit{R} packages. \\
 CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
@@
 resolution & Size in meter or degree of one cell \\
-*R* package & User-created software extension to *R* programming language \\
+\textit{R} package & User-created software extension to \textit{R} programming language \\
 SDM & Species distribution modeling \\
@@
 
-```{r fig-workflow, echo = FALSE, out.width="100%", out.height="30%", fig.cap = "Exemplary workflow of spatial data analyses for landscape ecology using the R programming language.
-For all major tasks, some example R packages are listed."}
+(ref:fig1cap) Exemplary workflow of spatial data analyses for landscape ecology using the *R* programming language. For all major tasks, some example *R* packages are listed.
+```{r fig-workflow, echo = FALSE, out.width="100%", out.height="30%", fig.cap = "(ref:fig1cap)"}
 include_graphics("data/Workflow.png")
@@
 Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms. 
-R packages *geoR* \cite{R-geoR} and *gstat* \cite{R-gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+*R* packages *geoR* \cite{R-geoR} and *gstat* \cite{R-gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
 Geostatistics also allows for spatial data simulations.
@@
 Widely used *R* packages include *ctmm* \cite{calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
-A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.
+A recent and very comprehensive overview of *R* packages for the analysis of animal movement data is given by \cite{joo2020}.
 
@@
 
-\caption{Overview of commonly used R packages for spatial data and landscape ecology. Packages are sorted by their major application task.}
+\caption{Overview of commonly used \textit{R} packages for spatial data and landscape ecology. Packages are sorted by their major application task.}
 
@@
 \hline
-Task & R package & Describtion & Reference \\
+Task & \textit{R} package & Describtion & Reference \\
 \hline
@@
   labs(x = "Number of package mentioned [%]", y = NULL,
-       title = "List the three R packages related\nto landscape ecology you use the most")
+       title = "List the three *R* packages related\nto landscape ecology you use the most")
 ```
@@
 
-```{r fig-survey, echo = FALSE, fig.align="center", fig.cap = "Results of the online survey about open-source software tools in R for landscape ecology.
-Results include a) which terms describe major reserach topics the best, b) the most important workflow task, c) the most used R packages and d) the overall usefulnes or R for landscape ecology.
-The 'others' category includes all answers with less than five total mentions.", warning=FALSE, out.width="100%"}
+(ref:fig3cap) Results of the online survey about open-source software tools in *R* for landscape ecology.
+Results include a) which terms describe major research topics the best, b) the most important workflow task, c) the most used *R* packages and d) the overall usefulness or *R* for landscape ecology.
+The 'others' category includes all answers with less than five total mentions.
+
+```{r fig-survey, echo = FALSE, fig.align="center", fig.cap = "(ref:fig3cap)", warning=FALSE, out.width="100%"}
 

### 2021-01-30 paper/paper.tex
Consistenlty named put *R* in italic.

@@
         statistical programming language \and
-        R packages \and
+        \emph{R} packages \and
         reproducibility \and
@@
 \begin{figure}
-\includegraphics[width=1\linewidth,height=0.3\textheight]{data/Workflow} \caption{Exemplary workflow of spatial data analyses for landscape ecology using the R programming language.
-For all major tasks, some example R packages are listed.}\label{fig:fig-workflow}
+\includegraphics[width=1\linewidth,height=0.3\textheight]{data/Workflow} \caption{Exemplary workflow of spatial data analyses for landscape ecology using the \emph{R} programming language.
+For all major tasks, some example \emph{R} packages are listed.}\label{fig:fig-workflow}
 \end{figure}
@@
 Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms.
-R packages \emph{geoR} \cite{R-geoR} and \emph{gstat} \cite{R-gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+\emph{R} packages \emph{geoR} \cite{R-geoR} and \emph{gstat} \cite{R-gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
 Geostatistics also allows for spatial data simulations.
@@
 Widely used \emph{R} packages include \emph{ctmm} \cite{calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
-A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.
+A recent and very comprehensive overview of \emph{R} packages for the analysis of animal movement data is given by \cite{joo2020}.
 
@@
 
-\caption{Results of the online survey about open-source software tools in R for landscape ecology.
-Results include a) which terms describe major reserach topics the best, b) the most important workflow task, c) the most used R packages and d) the overall usefulnes or R for landscape ecology.
-The 'others' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
+\caption{Results of the online survey about open-source software tools in \emph{R} for landscape ecology.
+Results include a) which terms describe major research topics the best, b) the most important workflow task, c) the most used \emph{R} packages and d) the overall usefulness or \emph{R} for landscape ecology.
+The `others' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
 \end{figure}

### 2021-02-02 paper/paper.Rmd
Clean references

--- 2021-01-27 paper/paper.Rmd
@@
 In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology.
-**Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+**Recent findings** In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
 Additionally, we outline the results of a survey of *R* users within the landscape ecology community.
-**Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available.
+**Summary** In this we find that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually updating the functionality available.
 Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
@@
 resolution & Size in meter or degree of one cell \\
-*R* package & User-created software extension to *R* programming language \\
-SDM & Species distribution modeling \\
+R package & User-created software extension to R programming language \\
+SDM & Species distribution modelling \\
 Simple Features & A set of standards that specify a common storage and access model of geographic features \\
@@
 
-Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
+Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
 The importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance today \cite{Simova2012,Estes2018}.
@@
 Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
-The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
+The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
 A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.
@@
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
-Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL} and *PROJ* \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL/OGRcontributors2020} and *PROJ* \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
 Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
@@
 Currently, the most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
-A *raster* successor, *terra*, aimed at the simpler interface and improved performance is being developed \cite{R-terra}, however, it could take several years for this package to be adopted by other developers and users.
+A *raster* successor, *terra*, aimed at the simpler interface and improved performance is being developed \cite{Hijmans2021}, however, it could take several years for this package to be adopted by other developers and users.
 Alternatively, the *stars* package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
@@
 However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
-Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019a}.
+Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
 *sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
@@
 Since publicly available data is becoming more prominent, so are *R* packages to access them.
-Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to download region and country data, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download Sentinel-2 optical images.
+Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to download region and country data, the *elevatr* package to access raster elevation data \cite{Hollister2020}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{Ranghetti2020} to download Sentinel-2 optical images.
 Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4.
-Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
+Additionally, the *rgee* package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
 
@@
 
-Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
-It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.
+Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
+It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.
 
@@
 Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
-Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
+Also the popular plotting package *ggplot2* \cite{Wickham2016} has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
 The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
-Quick interactive visualization of spatial data can be done with the *mapview* package\cite{R-mapview}.
-Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
+Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
+Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
 A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
@@
 
-The introduction of the *FRAGSTATS* in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the *landscapemetrics* package \cite{Hesselbarth2019a} allows to calculate the most widely used landscape metrics in within the *R* environment.
+The introduction of the *FRAGSTATS* in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the *landscapemetrics* package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the *R* environment.
 
@@
 The *geodiv* package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
-Additionally, the *belg* package allows calculation of the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
+Additionally, the *belg* package allows calculation of the Boltzmann entropy of a landscape gradient \cite{Nowosad2020}.
 
@@
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
-All of the spatial signatures methods mentioned above are implemented in the *motif* package \cite{R-motif}.
+All of the spatial signatures methods mentioned above are implemented in the *motif* package \cite{Nowosad2021}.
 
@@
 Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
-The *spatstat* package \cite{R-spatstat} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
+The *spatstat* package \cite{Baddeley2005} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
 A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
@@
 The *spdep* package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
-The *rinla* \cite{rue2009approximate} (\url{https://www.r-inla.org}) and *inlabru* \cite{R-inlabru} packages also provide functionality for modeling of spatially structured data.
+The *rinla* \cite{Rue2009} (\url{https://www.r-inla.org}) and *inlabru* \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.
 
 Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms. 
-R packages *geoR* \cite{R-geoR} and *gstat* \cite{R-gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+R packages *geoR* \cite{Diggle2007} and *gstat* \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
 Geostatistics also allows for spatial data simulations.
@@
 Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
-Additionally, textbooks introducing basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2019}.
+Additionally, textbooks introducing basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2018}.
 
-Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs.
+Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2018}, there is also a large number of *R* packages used for SDMs.
 Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package.
 Also, packages specifically designed for SDM exist, including the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages.
-Additionally, packages such as *ENMeval* \cite{R-enmeval} provide functionality for model tuning and evaluation.
+Additionally, packages such as *ENMeval* \cite{Muscarella2014} provide functionality for model tuning and evaluation.
 
 Related to SDMs, there is a growing number of *R* packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
-These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
-Widely used *R* packages include *ctmm* \cite{calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
-A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.
+These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{Signer2015}, behavioral states \cite{Langrock2012}, or habitat selection \cite{Fieberg2020}.
+Widely used *R* packages include *ctmm* \cite{Calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{Michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
+A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{Joo2020}.
 
@@
 Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}.
-Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019a}, and is provided by the recently published *samc* package \cite{Marx2020}.
+Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019}, and is provided by the recently published *samc* package \cite{Marx2020}.
 
@@
 \multirow{8}{*}{Spatial data} & raster & Raster data handling and analysis & \cite{Hijmans2019} \\
-& terra & Raster (and some vector) data handling and analysis & \cite{R-terra} \\
+& terra & Raster (and some vector) data handling and analysis & \cite{Hijmans2021} \\
 & stars & Spatiotemporal raster data handling and analysis & \cite{Pebesma2019} \\
@@
 \multirow{6}{*}{Spatial data download} & rnaturalearth & Download region and country data & \cite{South2017} \\
-& elevatr & Download elevation data & \cite{R-elevatr} \\
-& rgbif & Download biodiversity data & \cite{R-rgbif} \\
+& elevatr & Download elevation data & \cite{Hollister2020} \\
+& rgbif & Download biodiversity data & \cite{Chamberlain2017} \\
 & BIEN & Download plant diversity, function and distribution data & \cite{Maitner2020} \\
 & marmap & Download and manipulate bathymetric data & \cite{Pante2013} \\
-& FedData & Download geospatial data from US federal sources & \cite{R-FedData} \\
+& FedData & Download geospatial data from US federal sources & \cite{Bocinsky2019} \\
 & getlandsat & Download satellite data from Landsat 8 & \cite{Chamberlain2018} \\
 & MODIS & Download satellite data from MODIS & \cite{Mattiuzzi2020} \\
-& sen2r & Download satellite data from Sentinel & \cite{R-sen2r} \\
+& sen2r & Download satellite data from Sentinel & \cite{Ranghetti2020} \\
 \hline
@@
 & tmap & Create thematic maps & \cite{Tennekes2018} \\
-& mapview & Interactive viewing of spatial data & \cite{R-mapview} \\
-& leaflet & Create interactive web maps & \cite{R-leaflet} \\
+& mapview & Interactive viewing of spatial data & \cite{Appelhans2020} \\
+& leaflet & Create interactive web maps & \cite{Cheng2021} \\
 & rayshader & Create 2D and 3D data visualizations & \cite{Morgen-Wall2020} \\
 \hline
-\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & Quantify categorical landscape patterns & \cite{Hesselbarth2019a} \\
-& belg & Calculate the Boltzmann entropy & \cite{R-belgpaper} \\
-& motif & Pattern-based spatial analysis & \cite{R-motif} \\
+\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & Quantify categorical landscape patterns & \cite{Hesselbarth2019} \\
+& belg & Calculate the Boltzmann entropy & \cite{Nowosad2020} \\
+& motif & Pattern-based spatial analysis & \cite{Nowosad2021} \\
 & geodiv & Quantify continuous landscape patterns & \cite{Smith2020} \\
 \hline
-\multirow{6}{*}{\shortstack[l]{Spatial statistics}} & spatstat & Point pattern analysis & \cite{R-spatstat} \\
+\multirow{6}{*}{\shortstack[l]{Spatial statistics}} & spatstat & Point pattern analysis & \cite{Baddeley2005} \\
 & spdep & Quantify and correct for spatial autocorrelation & \cite{Bivand2013} \\
-& rinla & Fitting Bayesian spatio-temporal models using INLA & \cite{rue2009approximate} \\
-& inlabru & Fitting Bayesian spatio-temporal models using INLA & \cite{R-inlabru} \\
-& geoR & Variograms, correlograms and (co-)kriging & \cite{R-geoR} \\
-& gstat & Variograms, correlograms and (co-)kriging & \cite{R-gstat} \\
+& rinla & Fitting Bayesian spatio-temporal models using INLA & \cite{Rue2009} \\
+& inlabru & Fitting Bayesian spatio-temporal models using INLA & \cite{Bachl2019} \\
+& geoR & Variograms, correlograms and (co-)kriging & \cite{Diggle2007} \\
+& gstat & Variograms, correlograms and (co-)kriging & \cite{Pebesma2004} \\
 \hline
@@
 & zoon & Reproducible and remixable species distribution modelling & \cite{Golding2018} \\
-& ENMeval & Model tuning and evaluation & \cite{R-enmeval} \\
+& ENMeval & Model tuning and evaluation & \cite{Muscarella2014} \\
 & adehabitatHR, adehabitatHS & Home range and habitat selection modelling & \cite{Calenge2006} \\
 & amt & Manage and analyze animal movement data &\cite{Signer2019} \\
-& ctmm & Fit continuous time movement models &\cite{calabrese2016} \\
-& moveHMM & Fit hidden Markov models to movement data &\cite{michelot2016} \\
+& ctmm & Fit continuous time movement models &\cite{Calabrese2016} \\
+& moveHMM & Fit hidden Markov models to movement data &\cite{Michelot2016} \\
 \hline
@@
 This article does not contain any studies with human or animal subjects performed by any of the authors.
+
+### Important references
+
+\noindent \textbullet \textbullet \cite{Bivand2020}: Overview of general developments of *R* to analyze spatial data.
+
+\noindent \textbullet \textbullet \cite{Lovelace2019}: Textbook about spatial data handling with applied examples in the *R* programming language.
+
+\noindent \textbullet \cite{Lai2019}: Review of the *R* programming language in ecology.
+
+\noindent \textbullet \cite{Gustafson2019}: Review of quantifying landscape spatial pattern.

### 2021-02-02 paper/paper.tex
Clean references

--- 2021-01-27 paper/paper.tex
@@
 In particular, the \emph{R} programming language provides a wealth of community developed tools for landscape ecology.
-\textbf{Recent findings} In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+\textbf{Recent findings} In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
 Additionally, we outline the results of a survey of \emph{R} users within the landscape ecology community.
-\textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually further develop the functionality available.
+\textbf{Summary} In this we find that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually updating the functionality available.
 Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
@@
 Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely higher.
-The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
+The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
 A recent overview over the progress of \emph{R} to handle spatial data in general can be found in \cite{Bivand2020}.
@@
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
-Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} \cite{GDAL} and \emph{PROJ} \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} \cite{GDAL/OGRcontributors2020} and \emph{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
 Additionally, \emph{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
@@
 Currently, the most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
-A \emph{raster} successor, \emph{terra}, aimed at the simpler interface and improved performance is being developed \cite{R-terra}, however, it could take several years for this package to be adopted by other developers and users.
+A \emph{raster} successor, \emph{terra}, aimed at the simpler interface and improved performance is being developed \cite{Hijmans2021}, however, it could take several years for this package to be adopted by other developers and users.
 Alternatively, the \emph{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
@@
 However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
-Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019a}.
-\emph{sf} builds on the idea of ``simple features'', a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
+Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019}.
+\emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
-Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
+Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \emph{elevatr} package to access raster elevation data \cite{Hollister2020}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{Ranghetti2020} to download Sentinel-2 optical images.
 Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
-Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
+Additionally, the \emph{rgee} package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
 
@@
 
-Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
-It is possible to control several GIS software directly from \emph{R} using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.
+Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
+It is possible to control several GIS software directly from \emph{R} using dedicated packages, such as \emph{rgrass7} \cite{Bivand2021} for GRASS GIS, \emph{RSAGA} \cite{Brenning2018} for SAGA GIS, and \emph{rgee} \cite{Aybar2020} for Google Earth Engine.
 
@@
 Further plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
-Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
+Also the popular plotting package \emph{ggplot2} \cite{Wickham2016} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{Dunnington2020}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
 The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
-Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{R-mapview}.
-Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
+Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{Appelhans2020}.
+Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{Cheng2021}.
 A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
@@
 
-The introduction of the \emph{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \emph{landscapemetrics} package \cite{Hesselbarth2019a} allows to calculate the most widely used landscape metrics in within the \emph{R} environment.
+The introduction of the \emph{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \emph{landscapemetrics} package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the \emph{R} environment.
 
@@
 The \emph{geodiv} package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
-Additionally, the \emph{belg} package allows calculation of the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.
+Additionally, the \emph{belg} package allows calculation of the Boltzmann entropy of a landscape gradient \cite{Nowosad2020}.
 
@@
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
-All of the spatial signatures methods mentioned above are implemented in the \emph{motif} package \cite{R-motif}.
+All of the spatial signatures methods mentioned above are implemented in the \emph{motif} package \cite{Nowosad2021}.
 
@@
 Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
-The \emph{spatstat} package \cite{R-spatstat} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
+The \emph{spatstat} package \cite{Baddeley2005} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
 A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
@@
 The \emph{spdep} package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
-The \emph{rinla} \cite{rue2009approximate} (\url{https://www.r-inla.org}) and \emph{inlabru} \cite{R-inlabru} packages also provide functionality for modeling of spatially structured data.
+The \emph{rinla} \cite{Rue2009} (\url{https://www.r-inla.org}) and \emph{inlabru} \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.
 
 Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms.
-R packages \emph{geoR} \cite{R-geoR} and \emph{gstat} \cite{R-gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+R packages \emph{geoR} \cite{Diggle2007} and \emph{gstat} \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
 Geostatistics also allows for spatial data simulations.
@@
 Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
-Additionally, textbooks introducing basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2019}.
+Additionally, textbooks introducing basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2018}.
 
-Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of \emph{R} packages used for SDMs.
+Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2018}, there is also a large number of \emph{R} packages used for SDMs.
 Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
 Also, packages specifically designed for SDM exist, including the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008}, or \emph{zoon} \cite{Golding2018} packages.
-Additionally, packages such as \emph{ENMeval} \cite{R-enmeval} provide functionality for model tuning and evaluation.
+Additionally, packages such as \emph{ENMeval} \cite{Muscarella2014} provide functionality for model tuning and evaluation.
 
 Related to SDMs, there is a growing number of \emph{R} packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
-These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
-Widely used \emph{R} packages include \emph{ctmm} \cite{calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
-A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.
+These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{Signer2015}, behavioral states \cite{Langrock2012}, or habitat selection \cite{Fieberg2020}.
+Widely used \emph{R} packages include \emph{ctmm} \cite{Calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{Michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
+A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{Joo2020}.
 
@@
 Least-cost paths can be calculated using the \emph{gdistance} package \cite{vanEtten2017}.
-Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019a}, and is provided by the recently published \emph{samc} package \cite{Marx2020}.
+Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019}, and is provided by the recently published \emph{samc} package \cite{Marx2020}.
 
@@
 In total, the survey was answered by 109 participants, of which the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (14.7\%).
-Other, less frequent answers were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
@@
 
-Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
+Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
 \ref{fig:fig-survey} D)).
@@
 
+\hypertarget{important-references}{%
+\subsubsection{Important references}\label{important-references}}
+
+\noindent \textbullet \textbullet \cite{Bivand2020}: Overview of general developments of \emph{R} to analyze spatial data.
+
+\noindent \textbullet \textbullet \cite{Lovelace2019}: Textbook about spatial data handling with applied examples in the \emph{R} programming language.
+
+\noindent \textbullet \cite{Lai2019}: Review of the \emph{R} programming language in ecology.
+
+\noindent \textbullet \cite{Gustafson2019}: Review of quantifying landscape spatial pattern.
+
 \bibliographystyle{spphys}

### 2021-02-02 paper/paper.Rmd

@@
 cell & Smallest, rectangular unit of raster data model \\
-CRAN & Comprehensive R Archive Network. A repository for *R* package \\
+CRAN & Comprehensive \textit{R} Archive Network. A repository for \textit{R} package \\
 CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
@@
 resolution & Size in meter or degree of one cell \\
-R package & User-created software extension to R programming language \\
+\textit{R} package & User-created software extension to \textit{R} programming language \\
 SDM & Species distribution modelling \\
@@
 Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
-Also the popular plotting package *ggplot2* \cite{Wickham2016} has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}.
+Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
@@
 Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
-Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
+Both, *tmap* (Fig. \@ref(fig:fig-map)c) and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
 A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
 
-```{r fig-map, echo = FALSE, out.width="100%", out.height="50%", fig.cap = "Comparison of different options to create maps using a) base plot, b) ggplot2, and c) tmap.
+(ref:capfigmap) Comparison of different options to create maps using a) base plot, b) *ggplot2* package, and c) *tmap* package.
 All maps show the total annual precipitation of Switzerland.
-Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders) package.
-The code to create the maps can be found in the Appendix."}
+Data were downloaded with the *raster* package (precipitation data) and *rnaturalearth* package (country borders) package.
+The code to create the maps can be found in the Appendix.
+```{r fig-map, echo = FALSE, out.width="100%", out.height="50%", fig.cap = "(ref:capfigmap)"}
 include_graphics("data/plot_all.pdf")
@@
 Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms. 
-R packages *geoR* \cite{Diggle2007} and *gstat* \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+*R* packages *geoR* \cite{Diggle2007} and *gstat* \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
 Geostatistics also allows for spatial data simulations.
@@
 Widely used *R* packages include *ctmm* \cite{Calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{Michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
-A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{Joo2020}.
+A recent and very comprehensive overview of *R* packages for the analysis of animal movement data is given by \cite{Joo2020}.
 
@@
 ```
-# Survey of R usage by landscape ecology community
+# Survey of *R* usage by landscape ecology community
 
@@
   labs(x = "Number of package mentioned [%]", y = NULL,
-       title = "List the three R packages related\nto landscape ecology you use the most")
+       title = "List the three*R packages related\nto landscape ecology you use the most")
 ```
@@
 Lastly, when asked how useful *R* is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
-\@ref(fig:fig-survey) D)).
+\@ref(fig:fig-survey)d)).
 
-```{r fig-survey, echo = FALSE, fig.align="center", fig.cap = "Results of the online survey about open-source software tools in R for landscape ecology.
-Results include a) which terms describe major reserach topics the best, b) the most important workflow task, c) the most used R packages and d) the overall usefulnes or R for landscape ecology.
-The 'others' category includes all answers with less than five total mentions.", warning=FALSE, out.width="100%"}
+(ref:fig3cap) Results of the online survey about open-source software tools in *R* for landscape ecology.
+Results include a) which terms describe major research topics the best, b) the most important workflow task, c) the most used *R* packages and d) the overall usefulness or *R* for landscape ecology.
+The 'others' category includes all answers with less than five total mentions.
+
+```{r fig-survey, echo = FALSE, fig.align="center", fig.cap = "(ref:fig3cap)", warning=FALSE, out.width="100%"}
 

### 2021-02-02 paper/paper.tex

@@
 Further plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
-Also the popular plotting package \emph{ggplot2} \cite{Wickham2016} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{Dunnington2020}.
+Also the popular plotting package \emph{ggplot2} \cite{Wickham2016} (Fig.
+\ref{fig:fig-map}b) has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{Dunnington2020}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
@@
 Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{Appelhans2020}.
-Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{Cheng2021}.
+Both, \emph{tmap} (Fig. \ref{fig:fig-map}c) and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{Cheng2021}.
 A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
@@
 Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms.
-R packages \emph{geoR} \cite{Diggle2007} and \emph{gstat} \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+\emph{R} packages \emph{geoR} \cite{Diggle2007} and \emph{gstat} \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
 Geostatistics also allows for spatial data simulations.
@@
 Widely used \emph{R} packages include \emph{ctmm} \cite{Calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{Michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
-A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{Joo2020}.
+A recent and very comprehensive overview of \emph{R} packages for the analysis of animal movement data is given by \cite{Joo2020}.
 
@@
 \hypertarget{survey-of-r-usage-by-landscape-ecology-community}{%
-\section{Survey of R usage by landscape ecology community}\label{survey-of-r-usage-by-landscape-ecology-community}}
+\section{\texorpdfstring{Survey of \emph{R} usage by landscape ecology community}{Survey of R usage by landscape ecology community}}\label{survey-of-r-usage-by-landscape-ecology-community}}
 
@@
 We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig.
-\ref{fig:fig-survey} A)).
+\ref{fig:fig-survey}a)).
 
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
+Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey}b).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
-\ref{fig:fig-survey} D)).
+\ref{fig:fig-survey}d)).
 

### 2021-02-02 paper/paper.Rmd
Resolve last comments within MS

@@
 Most landscape metrics are a represented by a single number depicting specific characteristics of a local landscape.
-Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
-<!-- MH: Do we have a reference here @Jakub? -->
+Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram \cite{Nowosad2021}.
 Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
-<!-- LJG added some examples, but not 100% if they're the distances you're thinking of -->
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
@@
 Geostatistics also allows for spatial data simulations.
-<!-- MH: I am not sure if I understand the last sentence, but is it worth to reference the NLM section here? -->
 
@@
 
-<!--intro-->
 Since its first introduction in 1995, *R* has come a long way from an exclusively statistical programming language to a powerful landscape ecology tool.
@@
 
-<!--a paragraph summarizing possibilities-->
 A comprehensive collection of *R* software packages exist to handle the most common tasks of landscape ecology. 
@@
 
-<!--a paragraph summarizing the survey-->
 The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the *R* programming language for landscape ecology. 
@@
 
-<!--a paragraph about missing features/future work-->
 Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
@@
 
-<!-- MH: Make sure to update all sections  -->
-
-We are thankful to all members of the 'Coastal Ecology and Conservation Lab' at University of Michigan for comments on an early draft of the manuscript.
+We are thankful to all members of the 'Coastal Ecology and Conservation Lab' at the University of Michigan for comments on an early draft of the manuscript.
 
@@
 #### Funding
-
-<!-- MH: Add UMich Funding? -->
 

### 2021-02-02 paper/paper.tex
Resolve last comments within MS

@@
 Most landscape metrics are a represented by a single number depicting specific characteristics of a local landscape.
-Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
+Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram \cite{Nowosad2021}.
 Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g.~Euclidian, Manhattan, Mahlanobis).
@@
 
-We are thankful to all members of the `Coastal Ecology and Conservation Lab' at University of Michigan for comments on an early draft of the manuscript.
+We are thankful to all members of the `Coastal Ecology and Conservation Lab' at the University of Michigan for comments on an early draft of the manuscript.
 

### 2021-02-02 R/code_example.R
Code for appendix

@@ -1,3 +1,16 @@
+
+#---------------------------------------------------#
+#                                                   #
+#   Code to reproduce Fig. 2 in                     #
+#   "Open-source tools in R for landscape ecology"  #
+#                                                   #
+#   Code created by MHKH, JN, JS and LJG            #
+#                                                   #
+#   Contact: mhk.hesselbarth@gmail.com              #
+#                                                   #
+#---------------------------------------------------#
+
+# load libraries
 library(ggplot2)
 library(sf)
 library(raster)
@@ -31,18 +44,16 @@ precipitation_sum_ch <- raster::crop(x = precipitation_sum,
 precipitation_class_ch <- raster::cut(precipitation_sum_ch, breaks = 5)
 
 # create base plot
+
 # pdf(file = "R/Figures/base_plot.pdf")
-plot(precipitation_class_ch, col = viridis(5))
+
+plot(precipitation_class_ch, col = viridis::viridis(5))
 plot(st_geometry(switzerland), add = TRUE, lwd = 4, border = "white")
 plot(st_geometry(switzerland), add = TRUE)
+
 # dev.off()
 
-
 # create ggplot
-# MH: Why is this on long/lat? Coordinates seem to be okay?
-# JS: We could keep coords by adding:
-# `coord_sf(datum = st_crs(switzerland))`
-# but I actually like long/lat better, it looks cleaner
 plot_gg <- ggplot(raster::as.data.frame(precipitation_class_ch, xy = TRUE)) +
     geom_raster(aes(x = x, y = y, fill = factor(layer))) +
     geom_sf(data = switzerland, fill = NA, col = "white") +
@@ -51,13 +62,10 @@ plot_gg <- ggplot(raster::as.data.frame(precipitation_class_ch, xy = TRUE)) +
     labs(x = "", y = "") +
     theme_classic() +
     theme(legend.position = "bottom")
+
 # ggplot2::ggsave(filename = "R/Figures/ggplot2.pdf")
 
-# create interactive tmap (take screenshot for paper)
-# tmap_mode("plot")
-# tmap_mode("view")
-
-plot_tm = tm_shape(precipitation_class_ch) +
+plot_tm <- tm_shape(precipitation_class_ch) +
     tm_graticules() +
     tm_raster(title = "Precipitation\nclassified",
               style = "cat",
[…]

### 2021-02-08 paper/paper.tex
add in LJG funding

@@
 Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019}.
-\emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
+\emph{sf} builds on the idea of ``simple features'', a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 
-We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig.
+We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'', and ``nature conservation'' (Fig.
 \ref{fig:fig-survey}a)).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey}b).
+Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey}b).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 
-Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
+Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
 \ref{fig:fig-survey}d)).
@@
 
-MHKH was supported by the German Research Association (DFG) Research Training Group 1644 `Scaling Problems in Statistics,' grant number 152112243.
+MHKH was supported by the German Research Association (DFG) Research Training Group 1644 `Scaling Problems in Statistics', grant number 152112243.
+
+LJG was supported by the Natural Environment Research Council (NERC) grant \emph{A complex-systems approach to improve understanding of the biodiversity-landscape structure relationship}, grant number NE/T009373/1.
 

### 2021-02-08 paper/paper.Rmd
add in LJG funding
Final polishing

@@
 
-\begin{tabularx}{450pt}{lX}
+\begin{tabularx}{1.15\textwidth}{lX}
 
@@
 (ref:fig1cap) Exemplary workflow of spatial data analyses for landscape ecology using the *R* programming language. For all major tasks, some example *R* packages are listed.
-```{r fig-workflow, echo = FALSE, out.width="100%", out.height="30%", fig.cap = "(ref:fig1cap)"}
-include_graphics("data/Workflow.png")
+```{r fig-workflow, echo = FALSE, out.width="\\linewidth", fig.cap = "(ref:fig1cap)"}
+include_graphics("data/Figure_1.pdf")
 ```
@@
 The code to create the maps can be found in the Appendix.
-```{r fig-map, echo = FALSE, out.width="100%", out.height="50%", fig.cap = "(ref:capfigmap)"}
-include_graphics("data/plot_all.pdf")
+```{r fig-map, echo = FALSE, out.width="\\linewidth", fig.cap = "(ref:capfigmap)"}
+include_graphics("data/Figure_2.pdf")
 ```
+
+\FloatBarrier
 
@@
 
-\begin{tabularx}{450pt}{lrXr}
+\begin{tabularx}{1.35\linewidth}{lrXr}
 
@@
 ```
+
+\FloatBarrier
+
 # Survey of *R* usage by landscape ecology community
@@
 To better understand how the landscape ecology community uses *R*, we conducted a short survey using mailing lists and social media to reach the community. 
+All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).
+
 In total, the survey was answered by `r n_people` participants, of which the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
@@
 The 'others' category includes all answers with less than five total mentions.
-
-```{r fig-survey, echo = FALSE, fig.align="center", fig.cap = "(ref:fig3cap)", warning=FALSE, out.width="100%"}
+```{r fig-survey, echo = FALSE, fig.align="center", fig.cap = "(ref:fig3cap)", warning=FALSE, out.width="\\textwidth"}
 
@@
 
-A comprehensive collection of *R* software packages exist to handle the most common tasks of landscape ecology. 
+A comprehensive collection of *R* software packages exists to handle the most common tasks of landscape ecology. 
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, *R* allows for transparent and reproducible workflows.
@@
 But as the development of *R* packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
+
+\FloatBarrier
 
@@
 
-MHKH was supported by the German Research Association (DFG) Research Training Group 1644 'Scaling Problems in Statistics', grant number 152112243.
+No funding was received by the authors for this research and publication.
+
+####  Availability of data and material (data transparency)
+
+All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).
 

### 2021-02-08 paper/paper.tex
Final polishing

--- 2021-02-02 paper/paper.tex
@@
 \begin{figure}
-\includegraphics[width=1\linewidth,height=0.3\textheight]{data/Workflow} \caption{Exemplary workflow of spatial data analyses for landscape ecology using the \emph{R} programming language.
+\includegraphics[width=\linewidth]{data/Figure_1} \caption{Exemplary workflow of spatial data analyses for landscape ecology using the \emph{R} programming language.
 For all major tasks, some example \emph{R} packages are listed.}\label{fig:fig-workflow}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth,height=0.5\textheight]{data/plot_all} \caption{Comparison of different options to create maps using a) base plot, b) \emph{ggplot2} package, and c) \emph{tmap} package.
+\includegraphics[width=\linewidth]{data/Figure_2} \caption{Comparison of different options to create maps using a) base plot, b) \emph{ggplot2} package, and c) \emph{tmap} package.
 All maps show the total annual precipitation of Switzerland.
@@
 \end{figure}
+
+\FloatBarrier
 
@@
 
+\FloatBarrier
+
 \hypertarget{survey-of-r-usage-by-landscape-ecology-community}{%
@@
 To better understand how the landscape ecology community uses \emph{R}, we conducted a short survey using mailing lists and social media to reach the community.
+All raw data and scripts to analyze the data can be found at \url{https://github.com/r-spatialecology/Hesselbarth_et_al_CLER}.
+
 In total, the survey was answered by 109 participants, of which the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (14.7\%).
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-survey-1}
+{\centering \includegraphics[width=\textwidth]{paper_files/figure-latex/fig-survey-1}
 
@@
 
-A comprehensive collection of \emph{R} software packages exist to handle the most common tasks of landscape ecology.
+A comprehensive collection of \emph{R} software packages exists to handle the most common tasks of landscape ecology.
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, \emph{R} allows for transparent and reproducible workflows.
@@
 But as the development of \emph{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
+
+\FloatBarrier
 
@@
 
-MHKH was supported by the German Research Association (DFG) Research Training Group 1644 `Scaling Problems in Statistics,' grant number 152112243.
+No funding was received by the authors for this research and publication.
+
+\hypertarget{availability-of-data-and-material-data-transparency}{%
+\paragraph{Availability of data and material (data transparency)}\label{availability-of-data-and-material-data-transparency}}
+
+All raw data and scripts to analyze the data can be found at \url{https://github.com/r-spatialecology/Hesselbarth_et_al_CLER}.
 

### 2021-02-08 paper/paper.Rmd
Not sure what is going on...
...
Funding

@@
 
-No funding was received by the authors for this research and publication.
+LJG was supported by the Natural Environment Research Council (NERC) grant *A complex-systems approach to improve understanding of the biodiversity-landscape structure relationship*, grant number NE/T009373/1.
+All other authors did not receive any funding for this research.
 

### 2021-02-08 paper/paper.tex
Not sure what is going on...
...
Funding

@@
 
-No funding was received by the authors for this research and publication.
+LJG was supported by the Natural Environment Research Council (NERC) grant \emph{A complex-systems approach to improve understanding of the biodiversity-landscape structure relationship}, grant number NE/T009373/1.
+All other authors did not receive any funding for this research.
 

### 2021-02-10 paper/paper.Rmd
Clean figures

@@
   labs(x = "Number of package mentioned [%]", y = NULL,
-       title = "List the three*R packages related\nto landscape ecology you use the most")
+       title = "List the three R packages related\nto landscape ecology you use the most")
 ```

### 2021-04-22 Comments_Reviewer.Rmd
Just adding reviewer comments

---
title: "Comments for the author"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section Editor

Thank you for this submission.
I, and the two reviewers, all agree that this has the potential to be a very useful piece for the community and likely will be highly cited.
Both reviewers offer some constructive suggestions -- Reviewer 2's are fairly minor.
Reviewer 1's are more in-depth, but I strongly suggest you implement R1's suggestions.
I concur with the view that you may not need to define landscape ecology given the target journal.
I think adding some of the detail that R1 asks for in terms of descriptions of the packages and recommendations would be a very useful thing to add.
Both reviewers also commented about the survey.
I am less concerned about your addressing those issues, as I do not see the survey as being core to the paper -- although some interpretation of the questions that R1 asks would be worth reflecting on, and a little more detail on the survey reach, as asked by R2 would be good to include.

In my own read of the paper, I wondered if the development of some sort of typology of the available tools (as Rayfield et al. 2011 did for connectivity tools) would be useful.
I also suggest you consider some "boxes" that give point-by-point information.
I would love to see a "Where to start" box for total beginners and then maybe a second one with "Tips for developers".
In addition, can you please better explain the tidyverse to the readers?
It would also be useful if you could just elaborate a little bit in the annotated references for some of the key papers, as has been done for most of the papers published in this journal.

I hope you find these comments helpful.
With a little more revision, I am quite confident that the paper can be accepted in Current Landscape Ecology reports.
Thank you, and I look forward to receiving a revised manuscript!

**Reference:** Rayfield B, MJ Fortin, A Fall. 2011. Connectivity for conservation: a framework to classify network measures. Ecology 92(4): 847-858.

## Reviewer 1

I think the idea of a review of R tools for landscape ecology workflows is a great one, but this manuscript suffers a bit from lack of clarity of purpose.
If it's for landscape ecologists, as the choice of journal suggests, defining landscape ecology seems unnecessary, but a bit more information on R in general and why a landscape ecologist might choose R rather than some other tool would be useful.

I'm reading this manuscript from the perspective of a very experienced user of R for geospatial analyses, and who has spent a lot of time teaching students and post-docs how to do so.
I am familiar with nearly all of the packages mentioned, and even I found the lists of packages with little explanation or information overwhelming.
Why would I use one or another?
Do they have good documentation?
What capabilities might I need from other software?
Have you used all the packages mentioned, or simply compiled lists?

And, mentioned but not elaborated on: How would I combine these into an effective workflow, making my landscape ecology research easier and more reproducible?

1.1 I am a bit impressed but also surprised that you chose to define landscape ecology without using the word "pattern."

"it has no internal support for reading, processing, or visualizing spatial data"
This is misleading - these capabilities are part of well-developed and maintained packages, consonant with the R philosophy and structure.

Because of the lack of clear objective and audience, the manuscript is a mix of overexplanation and complex concepts dropped without adequate explanation.
Some examples of underexplained concepts:

"Packages that deal with such issue include groundhog [42], packrat [43], or renv [44]. For more information about R package development in general, see [40]."
How do they do so?
Why does it matter?

"also integrates into the widely used tidyverse packages"
Which are what, and why should the reader care?

"Finally, there are a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS [69,70,37], terrain analysis methods in SAGA GIS [71], or morphological operations for Google Earth Engine. It is possible to control several GIS software directly from R using dedicated packages, such as rgrass7 [72] for GRASS GIS, RSAGA [73] for SAGA GIS, and rgee [66] for Google Earth Engine."
This should be expanded on.
How can R fit into a workflow using these GIS tools, and why would you want to?

The section on CRS and projections should discuss the recent major changes to PROJ and thus to R packages relying on it.
A lot of the information that appears online is now obsolete.

In some places the formatting is problematic, such as automatic hyphenation of R package names resulting in "random-Forest" instead of "randomForest".

Table 1 is very useful, but does not contain all the packages mentioned in the text.

Survey: do you think the survey results reflect the true state of affairs, or do they reflect lack of training/knowledge/the fragmented nature of the R community?

"However, since landscape ecology constantly develops and improves, the R programming language and its packages need to change and adapt to these changes."
That is indeed a strength of the R community, though the way it is stated here implies that this is not true.
Instead, the problem is that there are so many packages, and constantly more being added.

"Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, R allows for transparent and reproducible workflows."

But the authors do not mention any tools for facililitating reproducibility.
This is a huge part of why R should be more widely used, and should be expanded upon.

Furthermore, the survey showed the that many members of the landscape ecology community actively develop R packages themselves, demonstrating that tools are constantly added and updated.
Yes, agreed.
So why imply not, as called out above?

"While the R programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the R programming language has to face."
Which are?
I know, but the casual reader may not.
How could the R core team improve capabilities?
What efforts are underway?
How do existing packages, notably raster and terra, deal with this?
Are there tools to help parallelize landscape ecology workflows?
This problem statement without any discussion or resolution might turn landscape ecologists away from R as a tool.

There *are* reasons one might choose tools other than R, or alongside R.
This review doesn't really articulate what the pros and cons are.

The reference section gives no journal names.
That can't be right.

I took a look at a 2019 review of R packages for a particular discipline that was cited in this paper (Joo et al, J. Animal Ecology).
Beyond simply listing packages, the authors looked at their documentation, whether they were actively maintained, and suggested how those packages might fit into a workflow.
Their Figure 1 provides an illustration of the workflow, and helps to put the package list in their Table 1 into a sensible context.
Something similar would work well here, and help to make sense of the overwhelming possibilities.

Can you give any recommendations?
There are a LOT of packages dealing with all aspects of landscape ecology, including aspects not covered here.
How can landscape ecologists learn of existing packages and new developments?
How can they choose among similar packages?

How can they put together packages to develop a reproducible workflow, and how can they document and share that?

The script provided in the appendix is both not useful for experienced R users, and not well-enough documented for novices.

The viridis package appears here, but is never mentioned or explained.
ggplot graphics are used, but that is a large and complex field.
While the caption to the associated Figure 2 mentions base vs ggplot graphics, and a subsequent paragraph goes into more detail, the distinction isn't really clear.

I would like to see a really useful review of R tools for landscape ecologists.
There's a lot of potential there, and something that covers both the basic concepts of using R and getting information and the current state of the relevant packages would be of great utility.

## Reviewer 2

The manuscript "Open-source tools in R for landscape ecology" by Maximillian H.K. Hesselbarth, Jakub Nowosad, Johannes Signer and Laura J. Graham (CLER-D-21-00001) is a brief but complete review of the current R packages for landscape ecology studies and their functionality.
The manuscript is well structured as it describes the typical workflow of any landscape ecology study, and details most of the existing R packages to perform these tasks.
I consider this information very useful for those ecologists initiating into the landscape ecology field that would like to analyze landscape data through R.
Finally, the manuscript shows the results of a simple survey to the landscape ecology community that evidence the usage of R for their studies, and exhibit the main points that the community would like to improve for their landscape ecology studies.

I have revised the manuscript and I believe is almost ready for publication.
I have a few comments and suggestions that I believe would clarify some parts of the manuscript, which I detail below.

L16: Can't understand the phrase, please clarify: "and lastly, iv) and can guide how to manage these heterogeneous landscapes".

L42: "today the programming language".
Replace "the" with "this".

L64: Replace "for landscape ecology" with "for landscape ecology studies".

L76: Replace "therefore, we focus on only the former two in this review" with "therefore, in this review we focus on only the former two".

L81: Perhaps you can also mention the package devtools .
Although I see that the package remotes is based in devtools.

L241: "Most landscape metrics are a represented by...".
Remove "a".

L348: "many different generic algorithms to create landscapes can be found across various R packages".
Replace "various" with "a couple" as you mention only two packages in the next sentences.

L355: Please expand in how you reach the "landscape ecology community".
Did you mail colleagues you already knew?
Did you send the survey to scientists of different parts of the world?
For instance, I personally believe that results may differ depending on the age, or on the country or continent in which the scientist works.
I only suggest to expand on how representative is the survey population of the landscape ecology community, and clarify the scope of it.

L357: The link is broken.

L361: Correct the commas, as they are inside the quotation marks.

L375: Correct the commas, as they are inside the quotation marks.

### 2021-04-24 R/code_example.R
Better documentation of R script

@@ -1,4 +1,3 @@
-
 #---------------------------------------------------#
 #                                                   #
 #   Code to reproduce Fig. 2 in                     #
@@ -10,7 +9,9 @@
 #                                                   #
 #---------------------------------------------------#
 
-# load libraries
+# First, we need to load all required packages used in this R script. In case you are
+# missing a package, it can be install from CRAN using e.g., install.packages("raster").
+# By loading the packages, we make sure all functions from them can be used in the script.
 library(ggplot2)
 library(sf)
 library(raster)
@@ -18,46 +19,81 @@ library(rnaturalearth)
 library(tmap)
 library(viridis)
 
-# get precipitation data
-precipitation <- raster::getData("worldclim", lon = 3, lat = 38,
+# This function downloads data and saves it as a RasterLayer. The 'name' argument
+# allows to specify a source to download data from the worldclim database. The
+# 'lon' and 'lat' argument specify the SRTM tile for which the data is downloaded, and
+# the 'res' the resolution of the data in minutes of a degree. The 'var' argument
+# specifies which kind of data is downloaded, in this case precipitation data,
+# and lastly the path on the local disk where the data is saved can be set.
+precipitation <- raster::getData(name = "worldclim", lon = 3, lat = 38,
                                  res = 0.5, var = "prec", path = "R/")
 
-# calculate total precipitation
+# The previous downloaded object is a RasterStack in which each layer corresponds
+# to the precipitation in a month (Jan - Dec). To get the annual precipitation,
+# we can use the calc function which applies basic mathematical function to each
+# layer of the RasterStack. By getting the sum of all layers for each cell, we
+# get the annual precipitation.
 precipitation_sum <- raster::calc(x = precipitation, fun = sum)
 
-# reproject to geographical CRS
+# Next, we need to ensure that all spatial data has the same coordinate reference system.
+# To achieve this we reproject the downloaded data to the CRS epsg:21781.
 precipitation_sum <- raster::projectRaster(precipitation_sum, crs = "+init=epsg:21781")
 
-# get admin boundaries of Switzerland
+# Next, we download administrational boundaries of Switzerland as vector data from
+# the naturalearth database. Again, we need to specify the resolution of the data
+# (called 'scale' here). Additionally, the function allows to select the return type
+# and we select a sf vector object.
[…]

### 2021-04-24 Comments_Reviewer.Rmd
:construction_worker: Starting to address the first reviewer comments

@@
 
-**Reference:** Rayfield B, MJ Fortin, A Fall. 2011. Connectivity for conservation: a framework to classify network measures. Ecology 92(4): 847-858.
+**Reference:** Rayfield B, MJ Fortin, A Fall.
+2011.
+Connectivity for conservation: a framework to classify network measures.
+Ecology 92(4): 847-858.
 
@@
 
-"it has no internal support for reading, processing, or visualizing spatial data"
-This is misleading - these capabilities are part of well-developed and maintained packages, consonant with the R philosophy and structure.
+"it has no internal support for reading, processing, or visualizing spatial data" This is misleading - these capabilities are part of well-developed and maintained packages, consonant with the R philosophy and structure.
+
+**Reply**: We agree with the reviewer that this sentence might be misleading.
+We tried to reword the statement in a way that makes clear base R does not have these capabilities, but many package provide it.
+\vskip 0.5cm
 
@@
 
-"Packages that deal with such issue include groundhog [42], packrat [43], or renv [44]. For more information about R package development in general, see [40]."
-How do they do so?
+"Packages that deal with such issue include groundhog [42], packrat [43], or renv [44]. For more information about R package development in general, see [40]." How do they do so?
 Why does it matter?
 
-"also integrates into the widely used tidyverse packages"
-Which are what, and why should the reader care?
+"also integrates into the widely used tidyverse packages" Which are what, and why should the reader care?
 
 "Finally, there are a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS [69,70,37], terrain analysis methods in SAGA GIS [71], or morphological operations for Google Earth Engine.
-It is possible to control several GIS software directly from R using dedicated packages, such as rgrass7 [72] for GRASS GIS, RSAGA [73] for SAGA GIS, and rgee [66] for Google Earth Engine."
-This should be expanded on.
+It is possible to control several GIS software directly from R using dedicated packages, such as rgrass7 [72] for GRASS GIS, RSAGA [73] for SAGA GIS, and rgee [66] for Google Earth Engine." This should be expanded on.
 How can R fit into a workflow using these GIS tools, and why would you want to?
@@
 
-"However, since landscape ecology constantly develops and improves, the R programming language and its packages need to change and adapt to these changes."
-That is indeed a strength of the R community, though the way it is stated here implies that this is not true.
+**Reply**: We added that we aware that the survey is most likely not representative for the R community and R users due to how we tried to reach the community and also due to the questions asked in the survey.
+Since the survey is not the main idea of the manuscript though, we think that the survery still migth be interesting even with these shortcomings.
+\vskip 0.5cm
+
+"However, since landscape ecology constantly develops and improves, the R programming language and its packages need to change and adapt to these changes." That is indeed a strength of the R community, though the way it is stated here implies that this is not true.
 Instead, the problem is that there are so many packages, and constantly more being added.
 
+**Reply**: We agree that the sentence implied that the R community is not constantly adapting to the new developments of landscape ecology which is certainly not true.
+We reworded the sentence making clearea that indeed the community constantly improves the R programming language.
+However, we do not agree with the reviewer that "[...] the problem is that there are so many packages [..]".
+Even though we acknowledge that the amount of package can be overwhelming, especially for novices users, diversity of packages is a big strength - even for similar specific tasks.
+Possible advantages include e.g., the possibility to compare different implementations of similar/identical task or providing functionality for slightly different data structures.
+\vskip 0.5cm
+
 "Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, R allows for transparent and reproducible workflows."
-
 But the authors do not mention any tools for facililitating reproducibility.
 This is a huge part of why R should be more widely used, and should be expanded upon.
+
+**Reply**: We highly agreed that this one of the main reasons why R should be widely used and we added an explanation how the script-based character of R itself allows to share and reproduce analysis.
+\vskip 0.5cm
 
@@
 
-"While the R programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the R programming language has to face."
-Which are?
+**Reply**: Please see our reply to the previous comment. We reworded the sentence to state clearer that this is indeed one of the strength of the R community. \vskip 0.5cm
+
+"While the R programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the R programming language has to face." Which are?
 I know, but the casual reader may not.
@@
 
+**Reply**: We added some explanation that raster and terra already provide solutions to this issue by processing only chunks that can be stored in memory.
+Additionally, we added some information about parallel processing as another possible solution to these issues.
+However, we believe that discussing underlying issues of R memory management is outside the scope if this paper.
+\vskip 0.5cm
+
 There *are* reasons one might choose tools other than R, or alongside R.
@@
 That can't be right.
+
+**Reply**: We apologise for this.
+The manuscript was written using the rticles packages with the corresponding template for Springer journals which somehow did not include journal names.
+We fixed the citation style including journal names.
+\textcolor{red}{STILL NEED TO DO THIS ACTUALLY!} \vskip 0.5cm
 
@@
 
+**Reply**: We agree with the reviewer that the provided script is not used for experienced R users and targeted to novices.
+Thus, we tried to improve the the documentation of the code. \vskip 0.5cm
+
 The viridis package appears here, but is never mentioned or explained.
@@
 
+**Reply**: We agree that the use of viridis might be confusing without mentioning it in the manuscript and we removed it from the R script.
+We agree that ggplot2 graphics are a very large field.
+This also highlights that it is outside the scope of this paper.
+Nevertheless, ggplot2 is used quite frequently by the R community, so we decided to include it in the example.
+We added a reference to the manuscript pointing towards ggplot2 specific resources. \vskip 0.5cm
+
 I would like to see a really useful review of R tools for landscape ecologists.
 There's a lot of potential there, and something that covers both the basic concepts of using R and getting information and the current state of the relevant packages would be of great utility.
+
+**Reply**: We agree with the reviewer and hope that by trying to address the very helpful and valuable comments by the reviewers, we approach this goal to a higher degree. \vskip 0.5cm
 
@@
 
+**Reply**: Thank you very much for the kind and encouraging words about the manuscript.
+We are happy to hear that the reviewer thinks the manuscript might be useful for ecologists that would like to analyze data through R. \vskip 0.5cm
+
 L16: Can't understand the phrase, please clarify: "and lastly, iv) and can guide how to manage these heterogeneous landscapes".
+
+**Reply**: Since both the section editor and reviewer 1 suggested to remove the definition of landscape ecology due to the target audience of the journal, the introduction was rewritten and shorted including this phrase.
+\vskip 0.5cm
 
@@
 
+**Reply**: We changed the sentence as suggested. \vskip 0.5cm
+
 L64: Replace "for landscape ecology" with "for landscape ecology studies".
 
+**Reply**: We changed the sentence as suggested. \vskip 0.5cm
+
 L76: Replace "therefore, we focus on only the former two in this review" with "therefore, in this review we focus on only the former two".
+
+**Reply**: We changed the sentence as suggested. \vskip 0.5cm
 
@@
 
+**Reply**: We agree that many users are used to devtools to install packages from GitHub.
+However, we decided to only mention remotes in this context since it is the more recent replacement for this task as also stated on the remotes homepage ("[...] This package is a lightweight replacement of the **install\_** functions in devtools.
+Indeed most of the code was copied over from devtools [...]" <https://github.com/r-lib/remotes#readme>) \vskip 0.5cm
+
 L241: "Most landscape metrics are a represented by...".
@@
 
+**Reply**: We changed the sentence as suggested. \vskip 0.5cm
+
 L348: "many different generic algorithms to create landscapes can be found across various R packages".
 Replace "various" with "a couple" as you mention only two packages in the next sentences.
+
+**Reply**: We changed the sentence as suggested. \vskip 0.5cm
 
@@
 
+**Reply**: We strongly agree with the reviewer that this information would be very valuable to judge the representation of the survey.
+However, unfortunately such questions were not included.
+We also acknowledge that the survey is most likely not representative for all skill levels of *R* users because we assume that by using mailing list and social media more advanced users were more likely to take part in the survey.
+We added and clearified those shortcomings of the survey. \vskip 0.5cm
+
 L357: The link is broken.
+
+**Reply**: The GitHub repository is currently set to "private" and will be publicly available upon publication. \vskip 0.5cm
 
@@
 
+**Reply**: We changed the sentence as suggested. \vskip 0.5cm
+
 L375: Correct the commas, as they are inside the quotation marks.
+
+**Reply**: We changed the sentence as suggested. \vskip 0.5cm

### 2021-04-24 paper/paper.Rmd
:construction_worker: Starting to address the first reviewer comments

@@
   \usepackage[modulo]{lineno}
+  \usepackage{xcolor}
 output:
@@
 One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}. 
-Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
+\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
+Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
@@
 Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
-For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
 The *R* programming language can be a very powerful tool.
@@
 There are three major online platforms to host *R* packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
-The last one focuses on tools for the analysis of genomic data; therefore, we focus on only the former two in this review.
+The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two".}
+
 CRAN (the *Comprehensive R Archive Network*) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
@@
 
-While *R* has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
+While \textcolor{blue}{base *R*} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+\textcolor{blue}{However, as discussed previously, one strength and core idea of the *R* programming language are its expandability by pacakges.} 
+Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
@@
 Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}. 
+\textcolor{blue}{Fore more information about **ggplot2** in general see \cite{Wickham2016}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
@@
 
-Most landscape metrics are a represented by a single number depicting specific characteristics of a local landscape.
+Most landscape metrics are represented by a single number depicting specific characteristics of a local landscape.
 Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram \cite{Nowosad2021}.
@@
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across various *R* packages.
+Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across \textcolor{blue}{a couple} *R* packages.
 A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the *NLMR* package \cite{Sciaini2018}.
@@
 To better understand how the landscape ecology community uses *R*, we conducted a short survey using mailing lists and social media to reach the community.
+\textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of *R* users and results are most likely biased towards more advanced users.
+Also, the survey did not include any personal questions, such as age or country of residence, which might have an effect on the representation of the survey.}.
 All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).
@@
 In total, the survey was answered by `r n_people` participants, of which the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
-Other, less frequent answers were "Data scientists", "None of the above", "Government employees" "Master's degree student" and "Bachelor's degree student" (in decreasing order).
+Other, less frequent answers were \textcolor{blue}{"Data scientists", "None of the above", "Government employees", "Master's degree student" and "Bachelor's degree student" (in decreasing order).}
 
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey)b).
+\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey)b).}
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
@@
 The highly dynamic development of *R* packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since landscape ecology constantly develops and improves, the *R* programming language and its packages need to change and adapt to these changes.
+However, since landscape ecology constantly develops and improves, \textcolor{blue}{consequently also the *R* programming language and its packages are under constant change to adapt to these new developments.}
 
 A comprehensive collection of *R* software packages exists to handle the most common tasks of landscape ecology.
-Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, *R* allows for transparent and reproducible workflows.
-Additionally, this also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
+Because it is possible to import, modify, analyze, and visualize spatial data all in the same  \textcolor{blue}{scripted} programming environment, *R* allows for transparent and reproducible workflows.
+\textcolor{blue}{This script-based characteristic of the *R* programming language allows to easily share and reproduce analysis, as demonstrated by the R script in the appendix, which is a big advantage over point-and-click software interfaces.}
+This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
 
@@
 While the *R* programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the *R* programming language has to face.
+\textcolor{blue}{Some packages discussed herein, such as *raster* and *terra*, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
+Furthermore, also parallel processing, and even including the use of high performance clusters, is provided by several *R* packages. 
+While this is outside the scope of this manuscript, further resources about parallel processing in *R* can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.} 
 Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the *R* packages related to those fields.

### 2021-04-24 paper/paper.tex
:construction_worker: Starting to address the first reviewer comments

@@
 \usepackage[modulo]{lineno}
+\usepackage{xcolor}
 
@@
 One successful example of an open-source project is the \emph{R} programming language, and its extensions called packages \cite{RCoreTeam2019}.
-Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
+\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
+Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
@@
 Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
-For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
 The \emph{R} programming language can be a very powerful tool.
@@
 There are three major online platforms to host \emph{R} packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
-The last one focuses on tools for the analysis of genomic data; therefore, we focus on only the former two in this review.
+The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two".}
+
 CRAN (the \emph{Comprehensive R Archive Network}) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
@@
 
-While \emph{R} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
+While \textcolor{blue}{base *R*} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+\textcolor{blue}{However, as discussed previously, one strength and core idea of the *R* programming language are its expandability by pacakges.}
+Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
@@
 \ref{fig:fig-map}b) has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{Dunnington2020}.
+\textbackslash textcolor\{blue\}\{Fore more information about \textbf{ggplot2} in general see \cite{Wickham2016}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
@@
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across various \emph{R} packages.
+Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across \textcolor{blue}{a couple} \emph{R} packages.
 A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \emph{NLMR} package \cite{Sciaini2018}.
@@
 To better understand how the landscape ecology community uses \emph{R}, we conducted a short survey using mailing lists and social media to reach the community.
+\textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of *R* users and results are most likely biased towards more advanced users.
+Also, the survey did not include any personal questions, such as age or country of residence, which might have an effect on the representation of the survey.}.
 All raw data and scripts to analyze the data can be found at \url{https://github.com/r-spatialecology/Hesselbarth_et_al_CLER}.
@@
 In total, the survey was answered by 109 participants, of which the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (14.7\%).
-Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
+Other, less frequent answers were \textcolor{blue}{"Data scientists", "None of the above", "Government employees", "Master's degree student" and "Bachelor's degree student" (in decreasing order).}
 
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey}b).
+\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \\ref{fig:fig-survey}b).}
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 The highly dynamic development of \emph{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since landscape ecology constantly develops and improves, the \emph{R} programming language and its packages need to change and adapt to these changes.
+However, since landscape ecology constantly develops and improves, \textcolor{blue}{consequently also the *R* programming language and its packages are under constant change to adapt to these new developments.}
 
 A comprehensive collection of \emph{R} software packages exists to handle the most common tasks of landscape ecology.
-Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, \emph{R} allows for transparent and reproducible workflows.
-Additionally, this also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
+Because it is possible to import, modify, analyze, and visualize spatial data all in the same \textcolor{blue}{scripted} programming environment, \emph{R} allows for transparent and reproducible workflows.
+\textcolor{blue}{This script-based characteristic of the *R* programming language allows to easily share and reproduce analysis, as demonstrated by the R script in the appendix, which is a big advantage over point-and-click software interfaces.}
+This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
 
@@
 While the \emph{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \emph{R} programming language has to face.
+\textcolor{blue}{Some packages discussed herein, such as *raster* and *terra*, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
+Furthermore, also parallel processing, and even including the use of high performance clusters, is provided by several *R* packages. 
+While this is outside the scope of this manuscript, further resources about parallel processing in *R* can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.}
 Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \emph{R} packages related to those fields.

### 2021-04-25 Comments_Reviewer.Rmd
:construction_worker: Addressing more comments...only a few left

@@
 
-**Reference:** Rayfield B, MJ Fortin, A Fall.
-2011.
-Connectivity for conservation: a framework to classify network measures.
-Ecology 92(4): 847-858.
+**Reference:** Rayfield B, MJ Fortin, A Fall. 2011. Connectivity for conservation: a framework to classify network measures. Ecology 92(4): 847-858.
 
@@
 
+**Reply**: We agree that it was not clear enough how these package could help woth reproducibility, thus we added a more detailed explanation. \vskip 0.5cm
+
 "also integrates into the widely used tidyverse packages" Which are what, and why should the reader care?
+
+**Reply**: Thank you for pointing out that any further explanation is missing. We added a brief explanation of what the tidyverse is and why using it can be beneficial. \vskip 0.5cm
 
@@
 
+**Reply**: We added an explanation how these package allow to incorporate functionality currently only available in GIS software into reproducible and shareable R scripts \vskip 0.5cm
+
 The section on CRS and projections should discuss the recent major changes to PROJ and thus to R packages relying on it.
@@
 
+**Reply**: @Jakub? \vskip 0.5cm
+
 In some places the formatting is problematic, such as automatic hyphenation of R package names resulting in "random-Forest" instead of "randomForest".
 
+**Reply**: We agree that this is problematic and we weren't even aware this is happeing so thanks for pointing it out. We deactivated the automatic hyphenation by LaTeX. \vskip 0.5cm
+
 Table 1 is very useful, but does not contain all the packages mentioned in the text.
+
+**Reply**: Thank you for checking the table so extensively. We made sure all packages should be included now. \textcolor{red}{Check again all packages are there} \vskip 0.5cm
 
@@
 This review doesn't really articulate what the pros and cons are.
+
+**Reply**: We agree with the reviewer that there are many reasons to use other software or programming languages.
+Nevertheless, we think that the scope of this review is not to discuss pro and cons between the different tools, but rather to focus on R.
+At the same time, we believe that the manuscript does give many arguments why R might be a good solution for many tasks.
+\vskip 0.5cm
 

### 2021-04-25 paper/paper.Rmd
:construction_worker: Addressing more comments...only a few left

@@
   \usepackage{xcolor}
+  \usepackage[none]{hyphenat}
 output:
@@
 Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
+\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
 For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
+\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
 For more information about *R* package development in general, see \cite{Wickham2015}.
@@
 Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
+\textcolor{blue}{The *tidyverse* is a collection of *R* packages developed for almost all major tasks of any data science project. 
+Because all *tidyverse* packages follow the same data philosophy structures, one strength of the *tidyverse* is its high consistency of usage across its packages \cite{Wickham2019}.}
 *sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
@@
 It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.
+\textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}
 
@@
 Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}. 
-\textcolor{blue}{Fore more information about **ggplot2** in general see \cite{Wickham2016}.
+\textcolor{blue}{Fore more information about *ggplot2* in general see \cite{Wickham2016}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
@@
 
-\caption{Overview of commonly used \textit{R} packages for spatial data and landscape ecology. Packages are sorted by their major application task.}
+\captionsetup{width=2\linewidth}
+
+\caption{Overview of commonly used \textit{R} packages for spatial data and landscape ecology.
+Packages are sorted by their major application task.
+The table only includes packages specifically desinged for spatial data or landscape ecology.}
 
@@
 & sen2r & Download satellite data from Sentinel & \cite{Ranghetti2020} \\
+\hline
+\multirow{3}{*}{Bridge stuff} & rgrass7 & Interface between GRASS7 and R  & \cite{Bivand2021} \\
+& RSAGA & Interface between SAGA and R & \cite{Brenning2018} \\
+& rgee &  Download data from Google Earth Engine & \cite{Aybar2020} \\
 \hline

### 2021-04-25 paper/paper.tex
:construction_worker: Addressing more comments...only a few left

@@
 \usepackage{xcolor}
+\usepackage[none]{hyphenat}
 
@@
 Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
+\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
 For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 Packages that deal with such issue include \emph{groundhog} \cite{Simonsohn2021}, \emph{packrat} \cite{Ushey2018}, or \emph{renv} \cite{Ushey2020}.
+\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
 For more information about \emph{R} package development in general, see \cite{Wickham2015}.
@@
 Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019}.
+\textcolor{blue}{The *tidyverse* is a collection of *R* packages developed for almost all major tasks of any data science project. 
+Because all *tidyverse* packages follow the same data philosophy structures, one strength of the *tidyverse* is its high consistency of usage across its packages \cite{Wickham2019}.}
 \emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
@@
 It is possible to control several GIS software directly from \emph{R} using dedicated packages, such as \emph{rgrass7} \cite{Bivand2021} for GRASS GIS, \emph{RSAGA} \cite{Brenning2018} for SAGA GIS, and \emph{rgee} \cite{Aybar2020} for Google Earth Engine.
+\textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}
 
@@
 \ref{fig:fig-map}b) has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{Dunnington2020}.
-\textbackslash textcolor\{blue\}\{Fore more information about \textbf{ggplot2} in general see \cite{Wickham2016}.
+\textbackslash textcolor\{blue\}\{Fore more information about \emph{ggplot2} in general see \cite{Wickham2016}.
 Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.

### 2021-04-25 Comments_Reviewer.Rmd
Ups...forgot to knit
Adding synthax info to map section

@@
 **Reply**: We agree with the reviewer that this sentence might be misleading. 
-We tried to reword the statement in a way that makes clear base R does not have these capabilities, but many package provide it.
-\vskip 0.5cm
+We tried to reword the statement in a way that makes clear base R does not have these capabilities, but many package provide it. \vskip 0.5cm
 
@@
 
-**Reply**: Thank you for pointing out that any further explanation is missing. We added a brief explanation of what the tidyverse is and why using it can be beneficial. \vskip 0.5cm
+**Reply**: Thank you for pointing out that any further explanation is missing. 
+We added a brief explanation of what the tidyverse is and why using it can be beneficial. \vskip 0.5cm
 
@@
 
-**Reply**: We added an explanation how these package allow to incorporate functionality currently only available in GIS software into reproducible and shareable R scripts \vskip 0.5cm
+**Reply**: We added an explanation how these package allow to incorporate functionality currently only available in GIS software into reproducible and shareable R scripts. \vskip 0.5cm
 
@@
 
-**Reply**: We agree that this is problematic and we weren't even aware this is happeing so thanks for pointing it out. We deactivated the automatic hyphenation by LaTeX. \vskip 0.5cm
+**Reply**: We agree that this is problematic and we weren't even aware this is happening so thanks for pointing it out. 
+We deactivated the automatic hyphenation by LaTeX. \vskip 0.5cm
 
@@
 
-**Reply**: Thank you for checking the table so extensively. We made sure all packages should be included now. \textcolor{red}{Check again all packages are there} \vskip 0.5cm
+**Reply**: Thank you for checking the table so extensively. 
+We made sure all packages should be included now.
+We also added the information to the table caption that not all packages from the manuscript are included but only packages specifically designed for spatial data and landscape ecology.
+\vskip 0.5cm
 
@@
 **Reply**: We added that we aware that the survey is most likely not representative for the R community and R users due to how we tried to reach the community and also due to the questions asked in the survey.
-Since the survey is not the main idea of the manuscript though, we think that the survery still migth be interesting even with these shortcomings.
-\vskip 0.5cm
+Since the survey is not the main idea of the manuscript though, we think that the survery still migth be interesting even with these shortcomings. \vskip 0.5cm
 
@@
 Even though we acknowledge that the amount of package can be overwhelming, especially for novices users, diversity of packages is a big strength - even for similar specific tasks. 
-Possible advantages include e.g., the possibility to compare different implementations of similar/identical task or providing functionality for slightly different data structures.
-\vskip 0.5cm
+Possible advantages include e.g., the possibility to compare different implementations of similar/identical task or providing functionality for slightly different data structures. \vskip 0.5cm
 
@@
 
-**Reply**: Please see our reply to the previous comment. We reworded the sentence to state clearer that this is indeed one of the strength of the R community. \vskip 0.5cm
+**Reply**: Please see our reply to the previous comment. 
+We reworded the sentence to state clearer that this is indeed one of the strength of the R community. \vskip 0.5cm
 
@@
 Additionally, we added some information about parallel processing as another possible solution to these issues. 
-However, we believe that discussing underlying issues of R memory management is outside the scope if this paper.
-\vskip 0.5cm
+However, we believe that discussing underlying issues of R memory management is outside the scope if this paper. \vskip 0.5cm
 
@@
 Nevertheless, we think that the scope of this review is not to discuss pro and cons between the different tools, but rather to focus on R. 
-At the same time, we believe that the manuscript does give many arguments why R might be a good solution for many tasks.
-\vskip 0.5cm
+At the same time, we believe that the manuscript does give many arguments why R might be a good solution for many tasks. \vskip 0.5cm
 
@@
 
-**Reply**: We apologise for this.
-The manuscript was written using the rticles packages with the corresponding template for Springer journals which somehow did not include journal names.
-We fixed the citation style including journal names.
-\textcolor{red}{STILL NEED TO DO THIS ACTUALLY!} \vskip 0.5cm
+**Reply**: We apologize for this. 
+The manuscript was written using the rticles packages with the corresponding template for Springer journals, which somehow did not include journal names. 
+We fixed the citation style including journal names. \textcolor{red}{STILL NEED TO DO THIS ACTUALLY!} \vskip 0.5cm
 
@@
 Nevertheless, ggplot2 is used quite frequently by the R community, so we decided to include it in the example.
-We added a reference to the manuscript pointing towards ggplot2 specific resources. \vskip 0.5cm
+We added an explanation how base R maps and ggplot2 maps differ in their underlying synthax and capabilities. 
+Also, we added a reference to the manuscript pointing towards ggplot2 specific resources. \vskip 0.5cm
 

### 2021-04-25 paper/paper.Rmd
Adding synthax info to map section

@@
 All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
-However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
+However, the generic functions are focused on quick visual inspection of the data (Fig.
+\@ref(fig:fig-map)a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor multiples maps.}
 
+The popular plotting package *ggplot2* (Fig.
+\@ref(fig:fig-map)b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
+Additionally, there is an extension especially designed to plot spatial data named *ggspatial*, which provides additional map elements \cite{Dunnington2020}.
+For more information about *ggplot2* in general, see \cite{Wickham2016}}.
+The *tmap* package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
+\textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
+Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
+Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
+Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
 Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
-Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}. 
-\textcolor{blue}{Fore more information about *ggplot2* in general see \cite{Wickham2016}.
-Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
-The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
-Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
-Both, *tmap* (Fig. \@ref(fig:fig-map)c) and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
 A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
 
-(ref:capfigmap) Comparison of different options to create maps using a) base plot, b) *ggplot2* package, and c) *tmap* package.
+(ref:capfigmap) Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) *ggplot2* package, and c) *tmap* package.
+All three approaches result in similar-looking outputs, but their syntax and capabilities differ.
 All maps show the total annual precipitation of Switzerland.

### 2021-04-25 paper/paper.tex
Adding synthax info to map section

@@
 All packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the generic \texttt{plot()} function (Fig. \ref{fig:fig-map}a).
-However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
+However, the generic functions are focused on quick visual inspection of the data (Fig.
+\ref{fig:fig-map}a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor multiples maps.}
 
+The popular plotting package \emph{ggplot2} (Fig.
+\ref{fig:fig-map}b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
+Additionally, there is an extension especially designed to plot spatial data named *ggspatial*, which provides additional map elements \cite{Dunnington2020}.
+For more information about *ggplot2* in general, see \cite{Wickham2016}}.
+The \emph{tmap} package provides a provides a coherent plotting system intended for drawing maps (Fig. \ref{fig:fig-map}c) \cite{Tennekes2018}.
+\textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
+Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{Appelhans2020}.
+Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{Cheng2021}.
+Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
 Further plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
-Also the popular plotting package \emph{ggplot2} \cite{Wickham2016} (Fig.
-\ref{fig:fig-map}b) has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{Dunnington2020}.
-\textbackslash textcolor\{blue\}\{Fore more information about \emph{ggplot2} in general see \cite{Wickham2016}.
-Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
-The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
-Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{Appelhans2020}.
-Both, \emph{tmap} (Fig. \ref{fig:fig-map}c) and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{Cheng2021}.
 A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{data/Figure_2} \caption{Comparison of different options to create maps using a) base plot, b) \emph{ggplot2} package, and c) \emph{tmap} package.
+\includegraphics[width=\linewidth]{data/Figure_2} \caption{Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) \emph{ggplot2} package, and c) \emph{tmap} package.
+All three approaches result in similar-looking outputs, but their syntax and capabilities differ.
 All maps show the total annual precipitation of Switzerland.

### 2021-04-25 Comments_Reviewer.Rmd
Updating Fig 1 and some more minor reviewer comments

@@
 
-**Reference:** Rayfield B, MJ Fortin, A Fall. 2011. Connectivity for conservation: a framework to classify network measures. Ecology 92(4): 847-858.
+*Reference:* Rayfield B, MJ Fortin, A Fall. 2011. Connectivity for conservation: a framework to classify network measures. Ecology 92(4): 847-858.
 
@@
 If it's for landscape ecologists, as the choice of journal suggests, defining landscape ecology seems unnecessary, but a bit more information on R in general and why a landscape ecologist might choose R rather than some other tool would be useful.
+
+**Reply**: Thank you for the general encouraging words about the manuscript. 
+We agree that a definition of landscape ecology is not needed given the target audience of the journal. 
+Thus, we shortened the introduction. 
+\textcolor{red}{Maybe move section of powerful R to introduction? But, comprehensive comparison with other software outside of scope} \vskip 0.5cm
 
@@
 Have you used all the packages mentioned, or simply compiled lists?
+
+**Reply**: All authors of the manuscript are experienced R users with many years of experience using the programming language. 
+Additionally, all authors even contributed several packages to the community themselves. 
+Given the different background of all authors, collectively probably most packages were used by at least one of the authors, but probably not all. 
+Nevertheless, we also believe that this is not necessarily important for this review. 
+Furthermore, we agree with the reviewer that more specific recommendation which package to use could be useful, however, we also think that this is very difficult. 
+Reasons for this include, besides others, personal preferences of the users or analysis specific circumstances. 
+Lastly, the review should be as neutral as possible without favoring one package over another because pros and cons of package differ between users and all were developed with a specific idea in mind.
+\textcolor{red}{Refer to new how to get started box} \vskip 0.5cm
 
@@
 
+**Reply**: We agree with the reviewer that Figure 1 of Joo et al.
+places the packages nicely into a context.
+We tried to achieve something similar with our Figure 1 as well, but obviously failed to achieve this goal.
+Thus, we tried to improve Figure 1 in the revised version of the manuscript to highlight better how the packages could be placed in a workflow.
+Also, we added an explanation that only packages were included that were available on CRAN at the time of preparation of the manuscript, which ensures at least that packages are maintained to fulfill all technical requirements by CRAN.
+\vskip 0.5cm
+
 Can you give any recommendations?

### 2021-04-25 paper/paper.Rmd
Updating Fig 1 and some more minor reviewer comments

@@
 For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+
 The *R* programming language can be a very powerful tool.
@@
 The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
-
 Thus, in the first part of this article, we present an overview of existing *R* packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
+\textcolor{blue}{We included only packages that were available on the official *R* package archive network CRAN at the time the manuscript was prepared. 
+This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
 In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are presently missing for landscape ecology.
@@
 
-(ref:fig1cap) Exemplary workflow of spatial data analyses for landscape ecology using the *R* programming language. For all major tasks, some example *R* packages are listed.
+(ref:fig1cap) Exemplary workflow of spatial data analyses for landscape ecology using the *R* programming language.
+For all major tasks, some example *R* packages are listed.
+For more packages related to the specific task, see Table \@ref(tab:packages).
 ```{r fig-workflow, echo = FALSE, out.width="\\linewidth", fig.cap = "(ref:fig1cap)"}

### 2021-04-25 paper/paper.tex
Updating Fig 1 and some more minor reviewer comments

@@
 For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+
 The \emph{R} programming language can be a very powerful tool.
@@
 The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
-
 Thus, in the first part of this article, we present an overview of existing \emph{R} packages for landscape ecology and closely related fields (Table \ref{tab:packages}).
+\textcolor{blue}{We included only packages that were available on the official *R* package archive network CRAN at the time the manuscript was prepared. 
+This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
 In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are presently missing for landscape ecology.
@@
 \includegraphics[width=\linewidth]{data/Figure_1} \caption{Exemplary workflow of spatial data analyses for landscape ecology using the \emph{R} programming language.
-For all major tasks, some example \emph{R} packages are listed.}\label{fig:fig-workflow}
+For all major tasks, some example \emph{R} packages are listed.
+For more packages related to the specific task, see Table \ref{tab:packages}.}\label{fig:fig-workflow}
 \end{figure}

### 2021-04-26 paper/paper.Rmd
Adding Where to Start Box

@@
 \textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}
+
+\begin{tabularx}{1\textwidth}{|X|}
+\hline
+\multicolumn{1}{|c|}{\textbf{Box 1: Where to start}} \\
+
+\\
+
+\textbf{1.} Get familiar with basics of \textit{R} programming language. 
+Good resources can be found at \url{https://cran.r-project.org/manuals.html}. \\
+
+\textbf{2.} Understand the fundamental spatial data types i.e, raster data and vector data.
+This also includes being able to process the data using the corresponding packages such as \textit{raster} or \textit{sf}. 
+These packakage also have excellent documentation at \url{https://rspatial.org/raster} and \url{https://r-spatial.github.io/sf}, respectively. \\
+
+\textbf{3.} Be able to transform coordinate reference systems of data from different sources.
+Depending on the data type, \textit{raster} or \textit{sf} and the corresponding documentations are good starting points.
+\\
+
+\textbf{4.} Do a quick visualization of the data for inspection or a first glance at the data. 
+This can be done with base \textit{R} or if already familiar \textit{ggplot2}. \\
+
+\textbf{5.} Find suitable \textit{R} package for research questions. 
+This can be difficult since research questions, available data and package are diverse. 
+Good starting points are the CRAN Task Views (mainly \url{https://cran.r-project.org/web/views/Spatial.html}) or review papers similar to this one. 
+If several packages exist, try all of them and see which ones suites the question and data the best. \\
+
+\textbf{6.} Create publication-quality figures and maps using e.g. \textit{ggspatial} or \textit{tmap}. 
+Help for both packages can be found on the corresponding homepages \url{https://paleolimbot.github.io/ggspatial/index.html} or \url{https://mtennekes.github.io/tmap}. \\
+
+\textbf{7.} Make reproducible \textit{R} scripts publicly available if possible. \\
+
+\textbf{8.} Further excellent resources with applied help about spatial data in \textit{R} can be found \url{https://keen-swartz-3146c4.netlify.app}, \url{https://rspatial.org}, or \url{https://geocompr.robinlovelace.net}.
+\\
+
+\hline
+\end{tabularx}
 

### 2021-04-26 paper/paper.Rmd
improves glossary

--- 2021-04-25 paper/paper.Rmd
@@
   \usepackage{xcolor}
-  \usepackage[none]{hyphenat}
 output:
@@
 CRAN & Comprehensive \textit{R} Archive Network. A repository for \textit{R} package \\
-CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
+CRS & Coordinate references system. A coordinate-based local, regional or global system used to locate geographical entities \\
 EPSG & Public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement \\
-GDAL & Open source translator library for raster and vector geospatial data formats \\
-GIS & Geographical information system \\
+GDAL & Open source translator library for raster and vector spatial data formats \\
+GIS & Geographic information system \\
 landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
@@
 patch & Neighboring cells with same value \\
-PROJ & Open source generic coordinate transformation software \\
-raster & Cell grid based data model \\
+PROJ & Open-source generic coordinate transformation software \\
+raster & Cell grid based spatial data model \\
 resolution & Size in meter or degree of one cell \\
 \textit{R} package & User-created software extension to \textit{R} programming language \\
-SDM & Species distribution modelling \\
-Simple Features & A set of standards that specify a common storage and access model of geographic features \\
-vector & Points, lines, and polygons-based data model \\
+SDM & Species distribution modeling \\
+Simple Features & A set of standards that specify a common storage and access model of spatial vector data \\
+vector & Points, lines, and polygons-based spatial data model \\
 
@@
 Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
-\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
 For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
-\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
 For more information about *R* package development in general, see \cite{Wickham2015}.
@@
 Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
-\textcolor{blue}{The *tidyverse* is a collection of *R* packages developed for almost all major tasks of any data science project. 
-Because all *tidyverse* packages follow the same data philosophy structures, one strength of the *tidyverse* is its high consistency of usage across its packages \cite{Wickham2019}.}
 *sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
@@
 It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.
-\textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}
 
@@
 All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
-However, the generic functions are focused on quick visual inspection of the data (Fig.
-\@ref(fig:fig-map)a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor multiples maps.}
+However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
 
-The popular plotting package *ggplot2* (Fig.
-\@ref(fig:fig-map)b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
-Additionally, there is an extension especially designed to plot spatial data named *ggspatial*, which provides additional map elements \cite{Dunnington2020}.
-For more information about *ggplot2* in general, see \cite{Wickham2016}}.
-The *tmap* package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
-\textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
+Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
+Also the popular plotting package *ggplot2* \cite{Wickham2016} (Fig. \@ref(fig:fig-map)b) has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}. 
+\textcolor{blue}{Fore more information about **ggplot2** in general see \cite{Wickham2016}.
+Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
+The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
 Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
-Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
-Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
-Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
+Both, *tmap* (Fig. \@ref(fig:fig-map)c) and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
 A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
 
-(ref:capfigmap) Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) *ggplot2* package, and c) *tmap* package.
-All three approaches result in similar-looking outputs, but their syntax and capabilities differ.
+(ref:capfigmap) Comparison of different options to create maps using a) base plot, b) *ggplot2* package, and c) *tmap* package.
 All maps show the total annual precipitation of Switzerland.
@@
 
-\captionsetup{width=2\linewidth}
-
-\caption{Overview of commonly used \textit{R} packages for spatial data and landscape ecology.
-Packages are sorted by their major application task.
-The table only includes packages specifically desinged for spatial data or landscape ecology.}
+\caption{Overview of commonly used \textit{R} packages for spatial data and landscape ecology. Packages are sorted by their major application task.}
 
@@
 & sen2r & Download satellite data from Sentinel & \cite{Ranghetti2020} \\
-\hline
-\multirow{3}{*}{Bridge stuff} & rgrass7 & Interface between GRASS7 and R  & \cite{Bivand2021} \\
-& RSAGA & Interface between SAGA and R & \cite{Brenning2018} \\
-& rgee &  Download data from Google Earth Engine & \cite{Aybar2020} \\
 \hline

### 2021-04-26 Comments_Reviewer.Rmd
Addressing the last final comments

@@
 
+**Reply**: Thank you very much for considering the manuscript for publication. 
+We appreciate the reviewers comments and think they all help to improve the manuscript.
+We agree with reviewer 1 that the definition of landscape ecology could be shortened given the target audience of the journal. 
+Thus, the introduction now puts a stronger focus on advantages of open-source software and especially R. 
+Furthermore, we added explanations for all packages that were missing in the original submission as suggested by reviewer 1.
+However, we think that given recommendations which packages to use is very difficult for several reasons. 
+First, packages have different advantages and disadvantages that differ depending on e.g., the research question, the available data, or the personal preferences of the users.
+Second, we believe that trying out several packages for similar task can actually beneficial and strengthen trust in the method.
+But, by modifying Figure 1 and adding Box 1 we hope to give some more guidelines on how to organize an analysis.
+We agree that the survey is not the core of the paper, nevertheless, the comments by both reviewers are very valuable and improve the conclusions we can draw from it. 
+Thus, we added some explanation about the survey reach and what this means for the results of the survey. \vskip 0.5cm
+
 In my own read of the paper, I wondered if the development of some sort of typology of the available tools (as Rayfield et al. 2011 did for connectivity tools) would be useful.
@@
 
+**Reply**: We agree that the of Rayfield et al.
+2011 is great, but we also want to point out that it was developed for a more specific topic (i.e., connectivity), while this manuscript deals with the whole field of landscape ecology.
+Thus, we believe that a similar typology for all packages included in this manuscript would be too complicated to be of actual help.
+Thank you for this great idea to add boxes. 
+We added a box "Where to start" to give some major steps how to approach landscape ecology in R for novice users. 
+We hope this will also at least partly address the issue raised by reviewer 1 about recommendations.
+We also thought about a second box for mode advanced users and even developers, but decided that the target audience of the paper are most likely novice to intermediate R users and not developers and the topic of developing own packages specifically for the landscape ecology community could maybe be a future manuscript.
+Also, we added a short explanation of the tidyverse and its strength.
+Lastly, we provide more information about the annotated references.
+\vskip 0.5cm
+
 I hope you find these comments helpful.
@@
 
-*Reference:* Rayfield B, MJ Fortin, A Fall. 2011. Connectivity for conservation: a framework to classify network measures. Ecology 92(4): 847-858.
+**Reply**: Again, thank you very much for the appreciated and valuable suggestions for improving the manuscript and we hope that the revised manuscript improved compared to the previous version with help of the reviewers.
+All changes to the manuscript are indicated by the blue color. \vskip 0.5cm
 
@@
 **Reply**: Thank you for the general encouraging words about the manuscript. 
-We agree that a definition of landscape ecology is not needed given the target audience of the journal. 
-Thus, we shortened the introduction. 
-\textcolor{red}{Maybe move section of powerful R to introduction? But, comprehensive comparison with other software outside of scope} \vskip 0.5cm
+We agree that such a long definition of landscape ecology is not needed given the target audience of the journal. 
+Thus, we shortened the introduction related to landscape ecology in general and rearranged other sections of the introduction.
+Most of the introduction now introduces the advantages and strength of open-source software and in particular R.
+While all of the authors are enthusiastic R users, we think that a comprehensive discussion of R vs.
+other open-source software is outside of the scope of this paper and more important not expedient in general since other software might work better for other people (e.g., Python).
+\vskip 0.5cm
 
@@
 Lastly, the review should be as neutral as possible without favoring one package over another because pros and cons of package differ between users and all were developed with a specific idea in mind.
-\textcolor{red}{Refer to new how to get started box} \vskip 0.5cm
+But, we think that our new "Box 1: Where to start" hopefully can assist novice users with their workflow. \vskip 0.5cm
 
@@
 
+**Reply**: We hope that the modified Figure 1, the added Box 1 and some added explanation in the conclusions explain how using R can facilitate a reproducible workflow in landcape ecology.
+\vskip 0.5cm
+
 1.1 I am a bit impressed but also surprised that you chose to define landscape ecology without using the word "pattern."
+
+**Reply**: While there are certainly many ways to define landscape ecology, we agree that pattern is a central term and included it in the introduction. \vskip 0.5cm
 
@@
 Some examples of underexplained concepts:
+
+**Reply**: Thank you for this valuable remark.
+We tried to add additional explanations to concepts which might be underexplained to especially novice users which hopefully increases the value of the manuscript.
+\vskip 0.5cm
 
@@
 We made sure all packages should be included now. 
-We also added the information to the table caption that not all packages from the manuscript are included but only packages specifically designed for spatial data and landscape ecology.
-\vskip 0.5cm
+We also added the information to the table caption that not all packages from the manuscript are included but only packages specifically designed for spatial data and landscape ecology. \vskip 0.5cm
 
@@
 Additionally, we added some information about parallel processing as another possible solution to these issues. 
-However, we believe that discussing underlying issues of R memory management is outside the scope if this paper. \vskip 0.5cm
+However, we believe that discussing underlying issues of R's memory management is outside the scope if this paper. \vskip 0.5cm
 
@@
 
+**Reply**: We are aware that there a many packages listed in this manuscript which might be overwhelming, especially given that some packages cover similar tasks. 
+We hope that especially Box 1, but also the restructured Figure 1 help landscape ecologists to find suitable packages. 
+Furthermore, we think that the structure in Table 1 actually helps to find packages related to specific task. 
+Lastly, we think that trying out similar packages for the same task can actually be beneficial, as also added to Box 1. \vskip 0.5cm
+
 How can they put together packages to develop a reproducible workflow, and how can they document and share that?
+
+**Reply**: We hope that the modified Figure 1, the added Box 1 and some added explanation in the conclusions explain how using R can facilitate a reproducible workflow in landcape ecology.
+\vskip 0.5cm
 
@@
 **Reply**: Thank you very much for the kind and encouraging words about the manuscript.
-We are happy to hear that the reviewer thinks the manuscript might be useful for ecologists that would like to analyze data through R. \vskip 0.5cm
+We are happy to hear that the reviewer thinks the manuscript might be useful for landscape ecologists that would like to analyze data through R. 
+We hope by including the appreciated and valuable comments by two reviewers we even increased the usefulness of the manuscript. \vskip 0.5cm
 

### 2021-04-26 paper/paper.Rmd
Addressing the last final comments

--- 2021-04-26 paper/paper.Rmd
@@
 
-## A short introduction to landscape ecology {#sec:landscape_ecology}
-
-While human activities have altered the landscapes for millennia \cite{Ellis2011,Ellis2015}, in the recent centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
-Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
-Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important for science, conservation, or management \cite{With2019}.
-
 Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
-In this context, landscape ecology predominantly focuses on i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchanges within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) and can guide how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
-
 Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
-The importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance today \cite{Simova2012,Estes2018}.
-Thus, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize spatial data.
-
-## Open-source software and *R* {#sec:open_source}
+Thus, landscape ecology emphasizes \textcolor{blue}{spatial} patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize these \textcolor{blue}{spatial patterns}.
 
@@
 This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.
+For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
 
@@
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
+Thus, the *R* programming language can be a very powerful tool.
+In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig.
+\@ref(fig:fig-workflow)).
+
+The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
 The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
@@
 
-Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.
-Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
 \textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
-For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-
-The *R* programming language can be a very powerful tool.
-In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig.
-\@ref(fig:fig-workflow)).
-The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
 Thus, in the first part of this article, we present an overview of existing *R* packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
@@
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same  \textcolor{blue}{scripted} programming environment, *R* allows for transparent and reproducible workflows.
-\textcolor{blue}{This script-based characteristic of the *R* programming language allows to easily share and reproduce analysis, as demonstrated by the R script in the appendix, which is a big advantage over point-and-click software interfaces.}
+\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the *R* programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the *R* script in the appendix.
+This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore are often only available for certain operating systems.}
 This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
@@
 
-\noindent \textbullet \textbullet \cite{Bivand2020}: Overview of general developments of *R* to analyze spatial data.
+\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of *R* to analyze spatial data demonstrating the capabilities of the programming language to anlyze spatial data.}
 
-\noindent \textbullet \textbullet \cite{Lovelace2019}: Textbook about spatial data handling with applied examples in the *R* programming language.
+\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Textbook about spatial data handling and analyzing covering all common tasks with applied and helpful code examples in the *R* programming language.}
 
-\noindent \textbullet \cite{Lai2019}: Review of the *R* programming language in ecology.
+\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state-of-the-art for quantifying landscape spatial pattern including lessons learned by the author.}
 
-\noindent \textbullet \cite{Gustafson2019}: Review of quantifying landscape spatial pattern.
+\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the *R* programming language is wideley used in ecology including possible reasons for the wide spread use.}

### 2021-04-26 paper/paper.tex
Addressing the last final comments

@@
 
-\hypertarget{sec:landscape_ecology}{%
-\subsection{A short introduction to landscape ecology}\label{sec:landscape_ecology}}
-
-While human activities have altered the landscapes for millennia \cite{Ellis2011,Ellis2015}, in the recent centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
-Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
-Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important for science, conservation, or management \cite{With2019}.
-
 Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
-In this context, landscape ecology predominantly focuses on i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchanges within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) and can guide how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.
-
 Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
-The importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance today \cite{Simova2012,Estes2018}.
-Thus, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize spatial data.
-
-\hypertarget{sec:open_source}{%
-\subsection{\texorpdfstring{Open-source software and \emph{R}}{Open-source software and R}}\label{sec:open_source}}
+Thus, landscape ecology emphasizes \textcolor{blue}{spatial} patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize these \textcolor{blue}{spatial patterns}.
 
@@
 Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
+Thus, the \emph{R} programming language can be a very powerful tool.
+In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig.
+\ref{fig:fig-workflow}).
+
+The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
 The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
@@
 
-Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
 \textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
-For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
-
-The \emph{R} programming language can be a very powerful tool.
-In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig.
-\ref{fig:fig-workflow}).
-The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
 Thus, in the first part of this article, we present an overview of existing \emph{R} packages for landscape ecology and closely related fields (Table \ref{tab:packages}).
@@
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same \textcolor{blue}{scripted} programming environment, \emph{R} allows for transparent and reproducible workflows.
-\textcolor{blue}{This script-based characteristic of the *R* programming language allows to easily share and reproduce analysis, as demonstrated by the R script in the appendix, which is a big advantage over point-and-click software interfaces.}
+\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the *R* programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the *R* script in the appendix.
+This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore are often only available for certain operating systems.}
 This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
@@
 
-\noindent \textbullet \textbullet \cite{Bivand2020}: Overview of general developments of \emph{R} to analyze spatial data.
+\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of *R* to analyze spatial data demonstrating the capabilities of the programming language to anlyze spatial data.}
 
-\noindent \textbullet \textbullet \cite{Lovelace2019}: Textbook about spatial data handling with applied examples in the \emph{R} programming language.
+\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Textbook about spatial data handling and analyzing covering all common tasks with applied and helpful code examples in the *R* programming language.}
 
-\noindent \textbullet \cite{Lai2019}: Review of the \emph{R} programming language in ecology.
+\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state-of-the-art for quantifying landscape spatial pattern including lessons learned by the author.}
 
-\noindent \textbullet \cite{Gustafson2019}: Review of quantifying landscape spatial pattern.
+\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the *R* programming language is wideley used in ecology including possible reasons for the wide spread use.}
 

### 2021-04-26 paper/paper.Rmd
Some layout changes

@@
 abstract: |
-  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years.
-In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology.
-**Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
-Additionally, we outline the results of a survey of *R* users within the landscape ecology community.
-**Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available.
+  \textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years.
+In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology.
+\textbf{Recent findings} In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community.
+\textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \textit{R}, and that as a community they are continually further develop the functionality available.
 Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
@@
 
+\FloatBarrier
+
+\begin{table}
+
+\small
+
 \begin{tabularx}{1.15\textwidth}{lX}
@@
 \end{tabularx}
+\end{table}
 
@@
 
-One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}. 
-\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
+One successful example of an open-source project is the \textit{R} programming language, and its extensions called packages \cite{RCoreTeam2019}. 
+\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the \textit{R} programming language \cite{Wickham2015}}.
 Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
-Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
-Thus, the *R* programming language can be a very powerful tool.
-In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig.
+Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
+Thus, the \textit{R} programming language can be a very powerful tool.
+In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \textit{R} environment (Fig.
 \@ref(fig:fig-workflow)).
 
-The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
-The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
-Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
-The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
-A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.
+The growing body of \textit{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
+The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: \textit{Analysis of Spatial Data} \cite{Bivand2019a} and \textit{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
+Since the task views are maintained manually by just a few people, the actual amount of \textit{R} packages related to spatial data is most likely higher.
+The growing popularity of the \textit{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
+A recent overview over the progress of \textit{R} to handle spatial data in general can be found in \cite{Bivand2020}.
 
-Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the \textit{R} programming language.
 \textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
-Thus, in the first part of this article, we present an overview of existing *R* packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
-\textcolor{blue}{We included only packages that were available on the official *R* package archive network CRAN at the time the manuscript was prepared. 
+Thus, in the first part of this article, we present an overview of existing \textit{R} packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
+\textcolor{blue}{We included only packages that were available on the official \textit{R} package archive network CRAN at the time the manuscript was prepared. 
 This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
-In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are presently missing for landscape ecology.
+In the second part, we present a survey in which we asked the community how they currently use the \textit{R} programming language and to identify topics for which \textit{R} packages are presently missing for landscape ecology.
 
@@
 
-Most *R* packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
-There are three major online platforms to host *R* packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
-The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two".}
+Most \textit{R} packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
+There are three major online platforms to host \textit{R} packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
+The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two.}
 
-CRAN (the *Comprehensive R Archive Network*) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
-*GitHub* hosts source code under version control, and allows users to install packages with one line of code using the *remotes* \cite{Hester2020} package.
-Additionally, hosting a package on *GitHub* provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e.
+CRAN (the \textit{Comprehensive R Archive Network}) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
+\textit{GitHub} hosts source code under version control, and allows users to install packages with one line of code using the \textit{remotes} \cite{Hester2020} package.
+Additionally, hosting a package on \textit{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e.
 testing if functions return an expected value).
 
-The guaranteed technical quality standard of *CRAN* requires more initial work for developers compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
-Additionally, the technical quality standard of *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e., package *x* requires and uses code from package *y*.
+The guaranteed technical quality standard of \textit{CRAN} requires more initial work for developers compared to \textit{GitHub}, while it ensures for users that the package can be installed on their machine.
+Additionally, the technical quality standard of \textit{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \textit{R} packages, i.e., package \textit{x} requires and uses code from package \textit{y}.
 CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
-Thus, most packages can be found on both platforms, and many developers use *GitHub* for regular development and *CRAN* to publish stable releases of the packages.
-Furthermore, online communities like *rOpenSci* also provide a peer-review process for code quality.
+Thus, most packages can be found on both platforms, and many developers use \textit{GitHub} for regular development and \textit{CRAN} to publish stable releases of the packages.
+Furthermore, online communities like \textit{rOpenSci} also provide a peer-review process for code quality.
 However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
-Packages that deal with such issue include *groundhog* \cite{Simonsohn2021}, *packrat* \cite{Ushey2018}, or *renv* \cite{Ushey2020}.
+Packages that deal with such issue include \textit{groundhog} \cite{Simonsohn2021}, \textit{packrat} \cite{Ushey2018}, or \textit{renv} \cite{Ushey2020}.
 \textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
-For more information about *R* package development in general, see \cite{Wickham2015}.
+For more information about \textit{R} package development in general, see \cite{Wickham2015}.
 
-(ref:fig1cap) Exemplary workflow of spatial data analyses for landscape ecology using the *R* programming language.
-For all major tasks, some example *R* packages are listed.
+(ref:fig1cap) Exemplary workflow of spatial data analyses for landscape ecology using the \textit{R} programming language.
+For all major tasks, some example \textit{R} packages are listed.
 For more packages related to the specific task, see Table \@ref(tab:packages).
@@
 
-While \textcolor{blue}{base *R*} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-\textcolor{blue}{However, as discussed previously, one strength and core idea of the *R* programming language are its expandability by pacakges.} 
-Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
-Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
-Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL/OGRcontributors2020} and *PROJ* \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
-Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
+While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by pacakges.} 
+Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
+Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
+Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
 
@@
 The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
-Currently, the most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
-A *raster* successor, *terra*, aimed at the simpler interface and improved performance is being developed \cite{Hijmans2021}, however, it could take several years for this package to be adopted by other developers and users.
-Alternatively, the *stars* package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
-Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as *fasterize* \cite{Ross2020}, *rasterDT* \cite{OBrien2020}, or *exactextractr* \cite{Baston2020}.
+Currently, the most prominent package allowing for raster data representation is \textit{raster} \cite{Hijmans2019}.
+A \textit{raster} successor, \textit{terra}, aimed at the simpler interface and improved performance is being developed \cite{Hijmans2021}, however, it could take several years for this package to be adopted by other developers and users.
+Alternatively, the \textit{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
+Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as \textit{fasterize} \cite{Ross2020}, \textit{rasterDT} \cite{OBrien2020}, or \textit{exactextractr} \cite{Baston2020}.
 
@@
 In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
-The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
-As of 2020, more than 500 *R* packages directly depend or imports *sp*.
-However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
-Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
-\textcolor{blue}{The *tidyverse* is a collection of *R* packages developed for almost all major tasks of any data science project. 
-Because all *tidyverse* packages follow the same data philosophy structures, one strength of the *tidyverse* is its high consistency of usage across its packages \cite{Wickham2019}.}
-*sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
+The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
+As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
+However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
+Besides many advantages and strength of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
+\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data science project. 
+Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
+\textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
-A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
-Since publicly available data is becoming more prominent, so are *R* packages to access them.
-Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to download region and country data, the *elevatr* package to access raster elevation data \cite{Hollister2020}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{Ranghetti2020} to download Sentinel-2 optical images.
-Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4.
-Additionally, the *rgee* package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
+A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \textit{R} session.
+Since publicly available data is becoming more prominent, so are \textit{R} packages to access them.
+Packages include \textit{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \textit{elevatr} package to access raster elevation data \cite{Hollister2020}, the \textit{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the \textit{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \textit{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \textit{FedData} package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the \textit{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \textit{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \textit{sen2r} \cite{Ranghetti2020} to download Sentinel-2 optical images.
+Also, the `getData()` function from the \textit{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
+Additionally, the \textit{rgee} package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
 
@@
 Additionally, they offer many other operations, such as merging or joining spatial data, extracting values from one dataset into another, raster resolution changes, or vector data simplifications.
-A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the *grainchanger* package \cite{Graham2019}.
-Furthermore, *landscapetools* is a collection of various utility functions for the raster data model \cite{Sciaini2018}.
+A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the \textit{grainchanger} package \cite{Graham2019}.
+Furthermore, \textit{landscapetools} is a collection of various utility functions for the raster data model \cite{Sciaini2018}.
 
 Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
-It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.
-\textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}
+It is possible to control several GIS software directly from \textit{R} using dedicated packages, such as \textit{rgrass7} \cite{Bivand2021} for GRASS GIS, \textit{RSAGA} \cite{Brenning2018} for SAGA GIS, and \textit{rgee} \cite{Aybar2020} for Google Earth Engine.
+\textcolor{blue}{By being able to access these GIS software directly from within \textit{R}, these packages can allow to create easily reproducible and sharable \textit{R} scripts, even if functionality is currently only available in the corresponding GIS software.}
+
+\begin{table}
+
+\small
 
 \begin{tabularx}{1\textwidth}{|X|}
+
 \hline
@@
 Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
-*R* allows two major types of maps.
+\textit{R} allows two major types of maps.
 Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.
@@
 
-The popular plotting package *ggplot2* (Fig.
+The popular plotting package \textit{ggplot2} (Fig.
 \@ref(fig:fig-map)b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
-Additionally, there is an extension especially designed to plot spatial data named *ggspatial*, which provides additional map elements \cite{Dunnington2020}.
-For more information about *ggplot2* in general, see \cite{Wickham2016}}.
-The *tmap* package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
+Additionally, there is an extension especially designed to plot spatial data named \textit{ggspatial}, which provides additional map elements \cite{Dunnington2020}.
+For more information about \textit{ggplot2} in general, see \cite{Wickham2016}}.
+The \textit{tmap} package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
 \textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
-Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
-Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
-Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
-Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
-A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
+Quick interactive visualization of spatial data can be done with the \textit{mapview} package\cite{Appelhans2020}.
+Both, \textit{tmap} and \textit{mapview} build upon the \textit{leaflet} package and leaflet javascript library \cite{Cheng2021}.
+Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \textit{cartography} package \cite{Giraud2016}.
+Further plotting methods for raster objects can be found in the \textit{rasterVis} package \cite{Lamigueiro2020}.
+A slightly different approach to visualizing spatial data is adapted by the \textit{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
 
-(ref:capfigmap) Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) *ggplot2* package, and c) *tmap* package.
+(ref:capfigmap) Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) \textit{ggplot2} package, and c) \textit{tmap} package.
 All three approaches result in similar-looking outputs, but their syntax and capabilities differ.
 All maps show the total annual precipitation of Switzerland.
-Data were downloaded with the *raster* package (precipitation data) and *rnaturalearth* package (country borders) package.
+Data were downloaded with the \textit{raster} package (precipitation data) and \textit{rnaturalearth} package (country borders) package.
 The code to create the maps can be found in the Appendix.
@@
 
-The introduction of the *FRAGSTATS* in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the *landscapemetrics* package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the *R* environment.
+The introduction of the \textit{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \textit{landscapemetrics} package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the \textit{R} environment.
 
 More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
-The *geodiv* package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
-Additionally, the *belg* package allows calculation of the Boltzmann entropy of a landscape gradient \cite{Nowosad2020}.
+The \textit{geodiv} package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
+Additionally, the \textit{belg} package allows calculation of the Boltzmann entropy of a landscape gradient \cite{Nowosad2020}.
 
@@
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
-All of the spatial signatures methods mentioned above are implemented in the *motif* package \cite{Nowosad2021}.
+All of the spatial signatures methods mentioned above are implemented in the \textit{motif} package \cite{Nowosad2021}.
 
@@
 Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
-The *spatstat* package \cite{Baddeley2005} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
+The \textit{spatstat} package \cite{Baddeley2005} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
 A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
@@
 It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling. 
-The *spdep* package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
-The *rinla* \cite{Rue2009} (\url{https://www.r-inla.org}) and *inlabru* \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.
+The \textit{spdep} package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
+The \textit{rinla} \cite{Rue2009} (\url{https://www.r-inla.org}) and \textit{inlabru} \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.
 
 Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms. 
-*R* packages *geoR* \cite{Diggle2007} and *gstat* \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+\textit{R} packages \textit{geoR} \cite{Diggle2007} and \textit{gstat} \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
 Geostatistics also allows for spatial data simulations.
@@
 Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
-Additionally, textbooks introducing basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2018}.
+Additionally, textbooks introducing basic concepts of SDM in \textit{R} exists \cite{Guisan2017,Fletcher2018}.
 
-Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2018}, there is also a large number of *R* packages used for SDMs.
-Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package.
-Also, packages specifically designed for SDM exist, including the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages.
-Additionally, packages such as *ENMeval* \cite{Muscarella2014} provide functionality for model tuning and evaluation.
+Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2018}, there is also a large number of \textit{R} packages used for SDMs.
+Popular approaches and packages include generalized linear models using, e.g., the \textit{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \textit{mgcv} \cite{Wood2017} or \textit{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \textit{rpart} \cite{Therneau2019}, \textit{randomForest} \cite{Liaw2002} or \textit{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \textit{ade4} \cite{Dray2007} or \textit{vegan} \cite{Oksanen2019} package.
+Also, packages specifically designed for SDM exist, including the \textit{dismo} \cite{Hijmans2017}, \textit{sdm} \cite{Naimi2016}, \textit{ecospat} \cite{Broennimann2020}, \textit{biomod2} \cite{Thuiller2020}, \textit{PresenceAbsence} \cite{Freeman2008}, or \textit{zoon} \cite{Golding2018} packages.
+Additionally, packages such as \textit{ENMeval} \cite{Muscarella2014} provide functionality for model tuning and evaluation.
 
-Related to SDMs, there is a growing number of *R* packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
+Related to SDMs, there is a growing number of \textit{R} packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
 These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{Signer2015}, behavioral states \cite{Langrock2012}, or habitat selection \cite{Fieberg2020}.
-Widely used *R* packages include *ctmm* \cite{Calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{Michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
-A recent and very comprehensive overview of *R* packages for the analysis of animal movement data is given by \cite{Joo2020}.
+Widely used \textit{R} packages include \textit{ctmm} \cite{Calabrese2016} and \textit{adehabitatHR} \cite{Calenge2006} for home-range estimation, \textit{moveHMM} \cite{Michelot2016} for the classification of behavioral states, and \textit{amt} \cite{Signer2019} for habitat selection.
+A recent and very comprehensive overview of \textit{R} packages for the analysis of animal movement data is given by \cite{Joo2020}.
 
@@
 At the patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
-Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}).
-Furthermore, the *lconnect* package \cite{Mestre2019} and *Makurhini* package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
+Such measures are provided within the \textit{landscapemetrics} package (see \ref{sec:landscape_metrics}).
+Furthermore, the \textit{lconnect} package \cite{Mestre2019} and \textit{Makurhini} package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
 Another way to describe connectivity is based on graph theory with the advantage that functional connectivity can also be included \cite{Kindlmann2008}.
 In graph theory \cite{Laita2011}, landscapes are described by nodes (i.e., habitat patches) connected by and functional connections called links (or edges) \cite{Laita2011}.
-The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
-More generally, the *igraph* package \cite{Csardi2006} provides functionality for graph theoretic analyses.
+The \textit{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
+More generally, the \textit{igraph} package \cite{Csardi2006} provides functionality for graph theoretic analyses.
 Resistance surfaces and least-cost paths are other tools to model functional connectivity which include attributes of the matrix.
 The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
-Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}.
-Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019}, and is provided by the recently published *samc* package \cite{Marx2020}.
+Least-cost paths can be calculated using the \textit{gdistance} package \cite{vanEtten2017}.
+Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019}, and is provided by the recently published \textit{samc} package \cite{Marx2020}.
 
@@
 Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
-Further functionality for landscape genetics such as quantifying and analysing population genetic structure, and hierarchical decomposition analysis can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.
+Further functionality for landscape genetics such as quantifying and analysing population genetic structure, and hierarchical decomposition analysis can be found in the \textit{graphs4lg} \cite{Savary2020}, \textit{PopGenReport} \cite{Adamack2014,Gruber2015}, \textit{HierDpart} packages \cite{Qin2019}, or \textit{GeNetIt} \cite{Murphy2010}.
 
@@
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across \textcolor{blue}{a couple} *R* packages.
-A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the *NLMR* package \cite{Sciaini2018}.
-Furthermore, the *RandomFields* package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.
+Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across \textcolor{blue}{a couple} \textit{R} packages.
+A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \textit{NLMR} package \cite{Sciaini2018}.
+Furthermore, the \textit{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.
 
@@
 
-\captionsetup{width=2\linewidth}
+\scriptsize
+
+\captionsetup{font=scriptsize}
 
@@
 \hline
-\multirow{5}{*}{Connectivity} & lconnect & Calculate landscape connectivity metrics & \cite{Mestre2019} \\
+\multirow{5}{*}{Connectivity} & \textit{lconnect} & Calculate landscape connectivity metrics & \cite{Mestre2019} \\
 & Makurhini & Calculate fragmentation and landscape connectivity indices & \cite{Godinez-Gomez2020} \\
@@
 
-# Survey of *R* usage by landscape ecology community
+# Survey of \textit{R} usage by landscape ecology community
 
@@
 
-To better understand how the landscape ecology community uses *R*, we conducted a short survey using mailing lists and social media to reach the community.
-\textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of *R* users and results are most likely biased towards more advanced users.
+To better understand how the landscape ecology community uses \textit{R}, we conducted a short survey using mailing lists and social media to reach the community.
+\textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
 Also, the survey did not include any personal questions, such as age or country of residence, which might have an effect on the representation of the survey.}.
@@
 
-Most people use *R* either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%).
+Most people use \textit{R} either "daily" (`r usage_freq[[1, 2]]`%) or a "few times a week" (`r usage_freq[[2, 2]]`%).
 Almost half of all participants described themselves as "advanced" users (`r expertise[[1, 2]]`%), while `r expertise[[2, 2]]`% described themselves as "intermediate" users.
-Related to this, about half of the participants either implemented their own package (`r own_package[[3, 2]]`%) or plan to do so in the future (`r own_package[[2, 2]]`%) and most of these packages are hosted on *GitHub* and/or *CRAN*.
+Related to this, about half of the participants either implemented their own package (`r own_package[[3, 2]]`%) or plan to do so in the future (`r own_package[[2, 2]]`%) and most of these packages are hosted on \textit{GitHub} and/or \textit{CRAN}.
 
@@
 More people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
-This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-survey)c).
+This was also represented in the most used \textit{R} packages (Fig. \@ref(fig:fig-survey)c).
 When asked for the three most used packages, participants of the survey listed `r length(unique(r_packages))` packages in total.
-The *raster* package was mentioned the most, followed by the *sf* package.
+The \textit{raster} package was mentioned the most, followed by the \textit{sf} package.
 Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey)c.
-Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], 2)`%).
+Nevertheless, the large availability and usage of different \textit{R} packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], 2)`%).
 
@@
 
-Lastly, when asked how useful *R* is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only very few participants evaluated *R* as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
+Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only very few participants evaluated \textit{R} as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
 \@ref(fig:fig-survey)d)).
 
-(ref:fig3cap) Results of the online survey about open-source software tools in *R* for landscape ecology.
-Results include a) which terms describe major research topics the best, b) the most important workflow task, c) the most used *R* packages and d) the overall usefulness or *R* for landscape ecology.
+(ref:fig3cap) Results of the online survey about open-source software tools in \textit{R} for landscape ecology.
+Results include a) which terms describe major research topics the best, b) the most important workflow task, c) the most used \textit{R} packages and d) the overall usefulness or \textit{R} for landscape ecology.
 The 'others' category includes all answers with less than five total mentions.
@@
 
-The survey also included a section in which participants could list methods and tools currently missing in *R* and answers to this question were very diverse.
+The survey also included a section in which participants could list methods and tools currently missing in \textit{R} and answers to this question were very diverse.
 
@@
 There were three most common topics across the answers of the participants.
-Firstly, many participants (`r perc_performance`%) wished for a better computational performance of *R* in terms of speed and required RAM, especially for larger data.
+Firstly, many participants (`r perc_performance`%) wished for a better computational performance of \textit{R} in terms of speed and required RAM, especially for larger data.
 Secondly, participants are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (`r perc_characterisation`%).
@@
 
-Since its first introduction in 1995, *R* has come a long way from an exclusively statistical programming language to a powerful landscape ecology tool.
-Today, many *R* packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed at spatial data handling and analysis.
-The highly dynamic development of *R* packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since landscape ecology constantly develops and improves, \textcolor{blue}{consequently also the *R* programming language and its packages are under constant change to adapt to these new developments.}
+Since its first introduction in 1995, \textit{R} has come a long way from an exclusively statistical programming language to a powerful landscape ecology tool.
+Today, many \textit{R} packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed at spatial data handling and analysis.
+The highly dynamic development of \textit{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
+However, since landscape ecology constantly develops and improves, \textcolor{blue}{consequently also the \textit{R} programming language and its packages are under constant change to adapt to these new developments.}
 
-A comprehensive collection of *R* software packages exists to handle the most common tasks of landscape ecology.
-Because it is possible to import, modify, analyze, and visualize spatial data all in the same  \textcolor{blue}{scripted} programming environment, *R* allows for transparent and reproducible workflows.
-\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the *R* programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the *R* script in the appendix.
+A comprehensive collection of \textit{R} software packages exists to handle the most common tasks of landscape ecology.
+Because it is possible to import, modify, analyze, and visualize spatial data all in the same  \textcolor{blue}{scripted} programming environment, \textit{R} allows for transparent and reproducible workflows.
+\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the \textit{R} programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
 This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore are often only available for certain operating systems.}
@@
 
-The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the *R* programming language for landscape ecology. 
-Furthermore, the survey showed the that many members of the landscape ecology community actively develop *R* packages themselves, demonstrating that tools are constantly added and updated.
+The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the \textit{R} programming language for landscape ecology. 
+Furthermore, the survey showed the that many members of the landscape ecology community actively develop \textit{R} packages themselves, demonstrating that tools are constantly added and updated.
 
 Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
-While the *R* programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the *R* programming language has to face.
-\textcolor{blue}{Some packages discussed herein, such as *raster* and *terra*, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
-Furthermore, also parallel processing, and even including the use of high performance clusters, is provided by several *R* packages. 
-While this is outside the scope of this manuscript, further resources about parallel processing in *R* can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.} 
-Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the *R* packages related to those fields.
-But as the development of *R* packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
+While the \textit{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \textit{R} programming language has to face.
+\textcolor{blue}{Some packages discussed herein, such as \textit{raster} and \textit{terra}, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
+Furthermore, also parallel processing, and even including the use of high performance clusters, is provided by several \textit{R} packages. 
+While this is outside the scope of this manuscript, further resources about parallel processing in \textit{R} can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.} 
+Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \textit{R} packages related to those fields.
+But as the development of \textit{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
 
@@
 
-MHKH and JN are authors of the *landscapemetrics* and *landscapetools* package.
-JN is author of *belg* and *motif* package. JS is author of the *amt* package. LJG is author of the *grainchanger* package.
+MHKH and JN are authors of the \textit{landscapemetrics} and \textit{landscapetools} package.
+MHKH is author of the \textit{NLMR} package.
+JN is author of \textit{belg} and \textit{motif} package. JS is author of the \textit{amt} package. LJG is author of the \textit{grainchanger} package.
 
@@
 
-LJG was supported by the Natural Environment Research Council (NERC) grant *A complex-systems approach to improve understanding of the biodiversity-landscape structure relationship*, grant number NE/T009373/1.
+LJG was supported by the Natural Environment Research Council (NERC) grant \textit{A complex-systems approach to improve understanding of the biodiversity-landscape structure relationship}, grant number NE/T009373/1.
 All other authors did not receive any funding for this research.
@@
 
-\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of *R* to analyze spatial data demonstrating the capabilities of the programming language to anlyze spatial data.}
+\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of \textit{R} to analyze spatial data demonstrating the capabilities of the programming language to analyze spatial data.}
 
-\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Textbook about spatial data handling and analyzing covering all common tasks with applied and helpful code examples in the *R* programming language.}
+\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Comprehensive textbook about spatial data handling and analyzing including many useful applied code examples.}
 
-\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state-of-the-art for quantifying landscape spatial pattern including lessons learned by the author.}
+\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state of the art for quantifying landscape spatial pattern including lessons learned by the author.}
 
-\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the *R* programming language is wideley used in ecology including possible reasons for the wide spread use.}
+\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the \textit{R} programming language is widely used in ecology including possible reasons for the wide spread use.}

### 2021-04-26 paper/paper.tex
Some layout changes

@@
 \textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years.
-In particular, the \emph{R} programming language provides a wealth of community developed tools for landscape ecology.
+In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology.
 \textbf{Recent findings} In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
-Additionally, we outline the results of a survey of \emph{R} users within the landscape ecology community.
-\textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually further develop the functionality available.
+Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community.
+\textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \textit{R}, and that as a community they are continually further develop the functionality available.
 Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
@@
 \section{Introduction}\label{sec:intro}}
+
+\FloatBarrier
+
+\begin{table}
+
+\small
 
@@
 
-One successful example of an open-source project is the \emph{R} programming language, and its extensions called packages \cite{RCoreTeam2019}.
-\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
+One successful example of an open-source project is the \textit{R} programming language, and its extensions called packages \cite{RCoreTeam2019}.
+\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the \textit{R} programming language \cite{Wickham2015}}.
 Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
-Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
-Thus, the \emph{R} programming language can be a very powerful tool.
-In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig.
+Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
+Thus, the \textit{R} programming language can be a very powerful tool.
+In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \textit{R} environment (Fig.
 \ref{fig:fig-workflow}).
 
-The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
-The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
-Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely higher.
-The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
-A recent overview over the progress of \emph{R} to handle spatial data in general can be found in \cite{Bivand2020}.
+The growing body of \textit{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
+The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: \textit{Analysis of Spatial Data} \cite{Bivand2019a} and \textit{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
+Since the task views are maintained manually by just a few people, the actual amount of \textit{R} packages related to spatial data is most likely higher.
+The growing popularity of the \textit{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
+A recent overview over the progress of \textit{R} to handle spatial data in general can be found in \cite{Bivand2020}.
 
-Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the \textit{R} programming language.
 \textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
-Thus, in the first part of this article, we present an overview of existing \emph{R} packages for landscape ecology and closely related fields (Table \ref{tab:packages}).
-\textcolor{blue}{We included only packages that were available on the official *R* package archive network CRAN at the time the manuscript was prepared. 
+Thus, in the first part of this article, we present an overview of existing \textit{R} packages for landscape ecology and closely related fields (Table \ref{tab:packages}).
+\textcolor{blue}{We included only packages that were available on the official \textit{R} package archive network CRAN at the time the manuscript was prepared. 
 This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
-In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are presently missing for landscape ecology.
+In the second part, we present a survey in which we asked the community how they currently use the \textit{R} programming language and to identify topics for which \textit{R} packages are presently missing for landscape ecology.
 
@@
 
-Most \emph{R} packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
-There are three major online platforms to host \emph{R} packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
-The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two".}
+Most \textit{R} packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
+There are three major online platforms to host \textit{R} packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
+The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two.}
 
-CRAN (the \emph{Comprehensive R Archive Network}) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
-\emph{GitHub} hosts source code under version control, and allows users to install packages with one line of code using the \emph{remotes} \cite{Hester2020} package.
-Additionally, hosting a package on \emph{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e.~testing if functions return an expected value).
+CRAN (the \textit{Comprehensive R Archive Network}) provides large visibility to the community, ease of installation, and a technical quality standard, including checks for common problems on all major operating systems \cite{Wickham2015}.
+\textit{GitHub} hosts source code under version control, and allows users to install packages with one line of code using the \textit{remotes} \cite{Hester2020} package.
+Additionally, hosting a package on \textit{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e.~testing if functions return an expected value).
 
-The guaranteed technical quality standard of \emph{CRAN} requires more initial work for developers compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
-Additionally, the technical quality standard of \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y}.
+The guaranteed technical quality standard of \textit{CRAN} requires more initial work for developers compared to \textit{GitHub}, while it ensures for users that the package can be installed on their machine.
+Additionally, the technical quality standard of \textit{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \textit{R} packages, i.e., package \textit{x} requires and uses code from package \textit{y}.
 CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
-Thus, most packages can be found on both platforms, and many developers use \emph{GitHub} for regular development and \emph{CRAN} to publish stable releases of the packages.
-Furthermore, online communities like \emph{rOpenSci} also provide a peer-review process for code quality.
+Thus, most packages can be found on both platforms, and many developers use \textit{GitHub} for regular development and \textit{CRAN} to publish stable releases of the packages.
+Furthermore, online communities like \textit{rOpenSci} also provide a peer-review process for code quality.
 However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
-Packages that deal with such issue include \emph{groundhog} \cite{Simonsohn2021}, \emph{packrat} \cite{Ushey2018}, or \emph{renv} \cite{Ushey2020}.
+Packages that deal with such issue include \textit{groundhog} \cite{Simonsohn2021}, \textit{packrat} \cite{Ushey2018}, or \textit{renv} \cite{Ushey2020}.
 \textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
-For more information about \emph{R} package development in general, see \cite{Wickham2015}.
+For more information about \textit{R} package development in general, see \cite{Wickham2015}.
 
 \begin{figure}
-\includegraphics[width=\linewidth]{data/Figure_1} \caption{Exemplary workflow of spatial data analyses for landscape ecology using the \emph{R} programming language.
-For all major tasks, some example \emph{R} packages are listed.
+\includegraphics[width=\linewidth]{data/Figure_1} \caption{Exemplary workflow of spatial data analyses for landscape ecology using the \textit{R} programming language.
+For all major tasks, some example \textit{R} packages are listed.
 For more packages related to the specific task, see Table \ref{tab:packages}.}\label{fig:fig-workflow}
@@
 
-While \textcolor{blue}{base *R*} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-\textcolor{blue}{However, as discussed previously, one strength and core idea of the *R* programming language are its expandability by pacakges.}
-Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
-Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
-Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} \cite{GDAL/OGRcontributors2020} and \emph{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
-Additionally, \emph{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
+While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
+\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by pacakges.}
+Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
+Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
+Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
 
@@
 The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
-Currently, the most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
-A \emph{raster} successor, \emph{terra}, aimed at the simpler interface and improved performance is being developed \cite{Hijmans2021}, however, it could take several years for this package to be adopted by other developers and users.
-Alternatively, the \emph{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
-Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as \emph{fasterize} \cite{Ross2020}, \emph{rasterDT} \cite{OBrien2020}, or \emph{exactextractr} \cite{Baston2020}.
+Currently, the most prominent package allowing for raster data representation is \textit{raster} \cite{Hijmans2019}.
+A \textit{raster} successor, \textit{terra}, aimed at the simpler interface and improved performance is being developed \cite{Hijmans2021}, however, it could take several years for this package to be adopted by other developers and users.
+Alternatively, the \textit{stars} package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
+Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as \textit{fasterize} \cite{Ross2020}, \textit{rasterDT} \cite{OBrien2020}, or \textit{exactextractr} \cite{Baston2020}.
 
@@
 In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
-The \emph{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
-As of 2020, more than 500 \emph{R} packages directly depend or imports \emph{sp}.
-However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
-Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019}.
-\textcolor{blue}{The *tidyverse* is a collection of *R* packages developed for almost all major tasks of any data science project. 
-Because all *tidyverse* packages follow the same data philosophy structures, one strength of the *tidyverse* is its high consistency of usage across its packages \cite{Wickham2019}.}
-\emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
+The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
+As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
+However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
+Besides many advantages and strength of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
+\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data science project. 
+Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
+\textit{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
-A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
-Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
-Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \emph{elevatr} package to access raster elevation data \cite{Hollister2020}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
-Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{Ranghetti2020} to download Sentinel-2 optical images.
-Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
-Additionally, the \emph{rgee} package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
+A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \textit{R} session.
+Since publicly available data is becoming more prominent, so are \textit{R} packages to access them.
+Packages include \textit{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \textit{elevatr} package to access raster elevation data \cite{Hollister2020}, the \textit{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the \textit{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \textit{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \textit{FedData} package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
+Furthermore, the \textit{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \textit{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \textit{sen2r} \cite{Ranghetti2020} to download Sentinel-2 optical images.
+Also, the \texttt{getData()} function from the \textit{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
+Additionally, the \textit{rgee} package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
 
@@
 Additionally, they offer many other operations, such as merging or joining spatial data, extracting values from one dataset into another, raster resolution changes, or vector data simplifications.
-A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the \emph{grainchanger} package \cite{Graham2019}.
-Furthermore, \emph{landscapetools} is a collection of various utility functions for the raster data model \cite{Sciaini2018}.
+A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the \textit{grainchanger} package \cite{Graham2019}.
+Furthermore, \textit{landscapetools} is a collection of various utility functions for the raster data model \cite{Sciaini2018}.
 
 Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
-It is possible to control several GIS software directly from \emph{R} using dedicated packages, such as \emph{rgrass7} \cite{Bivand2021} for GRASS GIS, \emph{RSAGA} \cite{Brenning2018} for SAGA GIS, and \emph{rgee} \cite{Aybar2020} for Google Earth Engine.
-\textcolor{blue}{By being able to access these GIS software directly from within *R*, these packages can allow to create easily reproducibale and sharable *R* scripts, even if functionality is currently only available in the corresponding GIS software.}
+It is possible to control several GIS software directly from \textit{R} using dedicated packages, such as \textit{rgrass7} \cite{Bivand2021} for GRASS GIS, \textit{RSAGA} \cite{Brenning2018} for SAGA GIS, and \textit{rgee} \cite{Aybar2020} for Google Earth Engine.
+\textcolor{blue}{By being able to access these GIS software directly from within \textit{R}, these packages can allow to create easily reproducible and sharable \textit{R} scripts, even if functionality is currently only available in the corresponding GIS software.}
+
+\begin{table}
+
+\small
 
 \begin{tabularx}{1\textwidth}{|X|}
+
 \hline
@@
 Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
-\emph{R} allows two major types of maps.
+\textit{R} allows two major types of maps.
 Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.~changing the displayed values.
@@
 
-The popular plotting package \emph{ggplot2} (Fig.
+The popular plotting package \textit{ggplot2} (Fig.
 \ref{fig:fig-map}b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
-Additionally, there is an extension especially designed to plot spatial data named *ggspatial*, which provides additional map elements \cite{Dunnington2020}.
-For more information about *ggplot2* in general, see \cite{Wickham2016}}.
-The \emph{tmap} package provides a provides a coherent plotting system intended for drawing maps (Fig. \ref{fig:fig-map}c) \cite{Tennekes2018}.
+Additionally, there is an extension especially designed to plot spatial data named \textit{ggspatial}, which provides additional map elements \cite{Dunnington2020}.
+For more information about \textit{ggplot2} in general, see \cite{Wickham2016}}.
+The \textit{tmap} package provides a provides a coherent plotting system intended for drawing maps (Fig. \ref{fig:fig-map}c) \cite{Tennekes2018}.
 \textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
-Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{Appelhans2020}.
-Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{Cheng2021}.
-Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
-Further plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
-A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
+Quick interactive visualization of spatial data can be done with the \textit{mapview} package\cite{Appelhans2020}.
+Both, \textit{tmap} and \textit{mapview} build upon the \textit{leaflet} package and leaflet javascript library \cite{Cheng2021}.
+Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \textit{cartography} package \cite{Giraud2016}.
+Further plotting methods for raster objects can be found in the \textit{rasterVis} package \cite{Lamigueiro2020}.
+A slightly different approach to visualizing spatial data is adapted by the \textit{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.
 
 \begin{figure}
-\includegraphics[width=\linewidth]{data/Figure_2} \caption{Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) \emph{ggplot2} package, and c) \emph{tmap} package.
+\includegraphics[width=\linewidth]{data/Figure_2} \caption{Comparison of different options to create maps of the total annual precipitation in Switzerland using a) base plot, b) \textit{ggplot2} package, and c) \textit{tmap} package.
 All three approaches result in similar-looking outputs, but their syntax and capabilities differ.
 All maps show the total annual precipitation of Switzerland.
-Data were downloaded with the \emph{raster} package (precipitation data) and \emph{rnaturalearth} package (country borders) package.
+Data were downloaded with the \textit{raster} package (precipitation data) and \textit{rnaturalearth} package (country borders) package.
 The code to create the maps can be found in the Appendix.}\label{fig:fig-map}
@@
 
-The introduction of the \emph{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \emph{landscapemetrics} package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the \emph{R} environment.
+The introduction of the \textit{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \textit{landscapemetrics} package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the \textit{R} environment.
 
 More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
-The \emph{geodiv} package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
-Additionally, the \emph{belg} package allows calculation of the Boltzmann entropy of a landscape gradient \cite{Nowosad2020}.
+The \textit{geodiv} package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
+Additionally, the \textit{belg} package allows calculation of the Boltzmann entropy of a landscape gradient \cite{Nowosad2020}.
 
@@
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
-All of the spatial signatures methods mentioned above are implemented in the \emph{motif} package \cite{Nowosad2021}.
+All of the spatial signatures methods mentioned above are implemented in the \textit{motif} package \cite{Nowosad2021}.
 
@@
 Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
-The \emph{spatstat} package \cite{Baddeley2005} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
+The \textit{spatstat} package \cite{Baddeley2005} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics.
 A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.
@@
 It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling.
-The \emph{spdep} package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
-The \emph{rinla} \cite{Rue2009} (\url{https://www.r-inla.org}) and \emph{inlabru} \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.
+The \textit{spdep} package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
+The \textit{rinla} \cite{Rue2009} (\url{https://www.r-inla.org}) and \textit{inlabru} \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.
 
 Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms.
-\emph{R} packages \emph{geoR} \cite{Diggle2007} and \emph{gstat} \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
+\textit{R} packages \textit{geoR} \cite{Diggle2007} and \textit{gstat} \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
 Geostatistics also allows for spatial data simulations.
@@
 Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
-Additionally, textbooks introducing basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2018}.
+Additionally, textbooks introducing basic concepts of SDM in \textit{R} exists \cite{Guisan2017,Fletcher2018}.
 
-Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2018}, there is also a large number of \emph{R} packages used for SDMs.
-Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
-Also, packages specifically designed for SDM exist, including the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008}, or \emph{zoon} \cite{Golding2018} packages.
-Additionally, packages such as \emph{ENMeval} \cite{Muscarella2014} provide functionality for model tuning and evaluation.
+Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2018}, there is also a large number of \textit{R} packages used for SDMs.
+Popular approaches and packages include generalized linear models using, e.g., the \textit{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \textit{mgcv} \cite{Wood2017} or \textit{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \textit{rpart} \cite{Therneau2019}, \textit{randomForest} \cite{Liaw2002} or \textit{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \textit{ade4} \cite{Dray2007} or \textit{vegan} \cite{Oksanen2019} package.
+Also, packages specifically designed for SDM exist, including the \textit{dismo} \cite{Hijmans2017}, \textit{sdm} \cite{Naimi2016}, \textit{ecospat} \cite{Broennimann2020}, \textit{biomod2} \cite{Thuiller2020}, \textit{PresenceAbsence} \cite{Freeman2008}, or \textit{zoon} \cite{Golding2018} packages.
+Additionally, packages such as \textit{ENMeval} \cite{Muscarella2014} provide functionality for model tuning and evaluation.
 
-Related to SDMs, there is a growing number of \emph{R} packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
+Related to SDMs, there is a growing number of \textit{R} packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
 These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{Signer2015}, behavioral states \cite{Langrock2012}, or habitat selection \cite{Fieberg2020}.
-Widely used \emph{R} packages include \emph{ctmm} \cite{Calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{Michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
-A recent and very comprehensive overview of \emph{R} packages for the analysis of animal movement data is given by \cite{Joo2020}.
+Widely used \textit{R} packages include \textit{ctmm} \cite{Calabrese2016} and \textit{adehabitatHR} \cite{Calenge2006} for home-range estimation, \textit{moveHMM} \cite{Michelot2016} for the classification of behavioral states, and \textit{amt} \cite{Signer2019} for habitat selection.
+A recent and very comprehensive overview of \textit{R} packages for the analysis of animal movement data is given by \cite{Joo2020}.
 
@@
 At the patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
-Such measures are provided within the \emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
-Furthermore, the \emph{lconnect} package \cite{Mestre2019} and \emph{Makurhini} package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
+Such measures are provided within the \textit{landscapemetrics} package (see \ref{sec:landscape_metrics}).
+Furthermore, the \textit{lconnect} package \cite{Mestre2019} and \textit{Makurhini} package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
 Another way to describe connectivity is based on graph theory with the advantage that functional connectivity can also be included \cite{Kindlmann2008}.
 In graph theory \cite{Laita2011}, landscapes are described by nodes (i.e., habitat patches) connected by and functional connections called links (or edges) \cite{Laita2011}.
-The \emph{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
-More generally, the \emph{igraph} package \cite{Csardi2006} provides functionality for graph theoretic analyses.
+The \textit{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
+More generally, the \textit{igraph} package \cite{Csardi2006} provides functionality for graph theoretic analyses.
 Resistance surfaces and least-cost paths are other tools to model functional connectivity which include attributes of the matrix.
 The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
-Least-cost paths can be calculated using the \emph{gdistance} package \cite{vanEtten2017}.
-Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019}, and is provided by the recently published \emph{samc} package \cite{Marx2020}.
+Least-cost paths can be calculated using the \textit{gdistance} package \cite{vanEtten2017}.
+Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019}, and is provided by the recently published \textit{samc} package \cite{Marx2020}.
 
@@
 Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
-Further functionality for landscape genetics such as quantifying and analysing population genetic structure, and hierarchical decomposition analysis can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.
+Further functionality for landscape genetics such as quantifying and analysing population genetic structure, and hierarchical decomposition analysis can be found in the \textit{graphs4lg} \cite{Savary2020}, \textit{PopGenReport} \cite{Adamack2014,Gruber2015}, \textit{HierDpart} packages \cite{Qin2019}, or \textit{GeNetIt} \cite{Murphy2010}.
 
@@
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across \textcolor{blue}{a couple} \emph{R} packages.
-A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \emph{NLMR} package \cite{Sciaini2018}.
-Furthermore, the \emph{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.
+Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across \textcolor{blue}{a couple} \textit{R} packages.
+A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \textit{NLMR} package \cite{Sciaini2018}.
+Furthermore, the \textit{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.
 
@@
 
-\hypertarget{survey-of-r-usage-by-landscape-ecology-community}{%
-\section{\texorpdfstring{Survey of \emph{R} usage by landscape ecology community}{Survey of R usage by landscape ecology community}}\label{survey-of-r-usage-by-landscape-ecology-community}}
+\hypertarget{survey-of-usage-by-landscape-ecology-community}{%
+\section{\texorpdfstring{Survey of \textit{R} usage by landscape ecology community}{Survey of  usage by landscape ecology community}}\label{survey-of-usage-by-landscape-ecology-community}}
 
-To better understand how the landscape ecology community uses \emph{R}, we conducted a short survey using mailing lists and social media to reach the community.
-\textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of *R* users and results are most likely biased towards more advanced users.
+To better understand how the landscape ecology community uses \textit{R}, we conducted a short survey using mailing lists and social media to reach the community.
+\textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
 Also, the survey did not include any personal questions, such as age or country of residence, which might have an effect on the representation of the survey.}.
@@
 
-Most people use \emph{R} either ``daily'' (53.2\%) or a ``few times a week'' (36.7\%).
+Most people use \textit{R} either ``daily'' (53.2\%) or a ``few times a week'' (36.7\%).
 Almost half of all participants described themselves as ``advanced'' users (46.8\%), while 40.4\% described themselves as ``intermediate'' users.
-Related to this, about half of the participants either implemented their own package (21.1\%) or plan to do so in the future (22\%) and most of these packages are hosted on \emph{GitHub} and/or \emph{CRAN}.
+Related to this, about half of the participants either implemented their own package (21.1\%) or plan to do so in the future (22\%) and most of these packages are hosted on \textit{GitHub} and/or \textit{CRAN}.
 
@@
 More people use the raster data model (72.5\%) in comparison to the vector data model (27.5\%).
-This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey}c).
+This was also represented in the most used \textit{R} packages (Fig. \ref{fig:fig-survey}c).
 When asked for the three most used packages, participants of the survey listed 93 packages in total.
-The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
+The \textit{raster} package was mentioned the most, followed by the \textit{sf} package.
 Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey}c.
-Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 35.87\%).
+Nevertheless, the large availability and usage of different \textit{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 35.87\%).
 
-Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
+Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \textit{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
 \ref{fig:fig-survey}d)).
@@
 
-\caption{Results of the online survey about open-source software tools in \emph{R} for landscape ecology.
-Results include a) which terms describe major research topics the best, b) the most important workflow task, c) the most used \emph{R} packages and d) the overall usefulness or \emph{R} for landscape ecology.
+\caption{Results of the online survey about open-source software tools in \textit{R} for landscape ecology.
+Results include a) which terms describe major research topics the best, b) the most important workflow task, c) the most used \textit{R} packages and d) the overall usefulness or \textit{R} for landscape ecology.
 The `others' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
@@
 
-The survey also included a section in which participants could list methods and tools currently missing in \emph{R} and answers to this question were very diverse.
+The survey also included a section in which participants could list methods and tools currently missing in \textit{R} and answers to this question were very diverse.
 
@@
 There were three most common topics across the answers of the participants.
-Firstly, many participants (12.8\%) wished for a better computational performance of \emph{R} in terms of speed and required RAM, especially for larger data.
+Firstly, many participants (12.8\%) wished for a better computational performance of \textit{R} in terms of speed and required RAM, especially for larger data.
 Secondly, participants are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (9.2\%).
@@
 
-Since its first introduction in 1995, \emph{R} has come a long way from an exclusively statistical programming language to a powerful landscape ecology tool.
-Today, many \emph{R} packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed at spatial data handling and analysis.
-The highly dynamic development of \emph{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since landscape ecology constantly develops and improves, \textcolor{blue}{consequently also the *R* programming language and its packages are under constant change to adapt to these new developments.}
+Since its first introduction in 1995, \textit{R} has come a long way from an exclusively statistical programming language to a powerful landscape ecology tool.
+Today, many \textit{R} packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed at spatial data handling and analysis.
+The highly dynamic development of \textit{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
+However, since landscape ecology constantly develops and improves, \textcolor{blue}{consequently also the \textit{R} programming language and its packages are under constant change to adapt to these new developments.}
 
-A comprehensive collection of \emph{R} software packages exists to handle the most common tasks of landscape ecology.
-Because it is possible to import, modify, analyze, and visualize spatial data all in the same \textcolor{blue}{scripted} programming environment, \emph{R} allows for transparent and reproducible workflows.
-\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the *R* programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the *R* script in the appendix.
+A comprehensive collection of \textit{R} software packages exists to handle the most common tasks of landscape ecology.
+Because it is possible to import, modify, analyze, and visualize spatial data all in the same \textcolor{blue}{scripted} programming environment, \textit{R} allows for transparent and reproducible workflows.
+\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the \textit{R} programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
 This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore are often only available for certain operating systems.}
@@
 
-The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the \emph{R} programming language for landscape ecology.
-Furthermore, the survey showed the that many members of the landscape ecology community actively develop \emph{R} packages themselves, demonstrating that tools are constantly added and updated.
+The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the \textit{R} programming language for landscape ecology.
+Furthermore, the survey showed the that many members of the landscape ecology community actively develop \textit{R} packages themselves, demonstrating that tools are constantly added and updated.
 
 Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
-While the \emph{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \emph{R} programming language has to face.
-\textcolor{blue}{Some packages discussed herein, such as *raster* and *terra*, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
-Furthermore, also parallel processing, and even including the use of high performance clusters, is provided by several *R* packages. 
-While this is outside the scope of this manuscript, further resources about parallel processing in *R* can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.}
-Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \emph{R} packages related to those fields.
-But as the development of \emph{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
+While the \textit{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \textit{R} programming language has to face.
+\textcolor{blue}{Some packages discussed herein, such as \textit{raster} and \textit{terra}, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
+Furthermore, also parallel processing, and even including the use of high performance clusters, is provided by several \textit{R} packages. 
+While this is outside the scope of this manuscript, further resources about parallel processing in \textit{R} can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.}
+Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \textit{R} packages related to those fields.
+But as the development of \textit{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
 
@@
 
-MHKH and JN are authors of the \emph{landscapemetrics} and \emph{landscapetools} package.
-JN is author of \emph{belg} and \emph{motif} package. JS is author of the \emph{amt} package. LJG is author of the \emph{grainchanger} package.
+MHKH and JN are authors of the \textit{landscapemetrics} and \textit{landscapetools} package.
+MHKH is author of the \textit{NLMR} package.
+JN is author of \textit{belg} and \textit{motif} package. JS is author of the \textit{amt} package. LJG is author of the \textit{grainchanger} package.
 
@@
 
-LJG was supported by the Natural Environment Research Council (NERC) grant \emph{A complex-systems approach to improve understanding of the biodiversity-landscape structure relationship}, grant number NE/T009373/1.
+LJG was supported by the Natural Environment Research Council (NERC) grant \textit{A complex-systems approach to improve understanding of the biodiversity-landscape structure relationship}, grant number NE/T009373/1.
 All other authors did not receive any funding for this research.
@@
 
-\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of *R* to analyze spatial data demonstrating the capabilities of the programming language to anlyze spatial data.}
+\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of \textit{R} to analyze spatial data demonstrating the capabilities of the programming language to analyze spatial data.}
 
-\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Textbook about spatial data handling and analyzing covering all common tasks with applied and helpful code examples in the *R* programming language.}
+\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Comprehensive textbook about spatial data handling and analyzing including many useful applied code examples.}
 
-\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state-of-the-art for quantifying landscape spatial pattern including lessons learned by the author.}
+\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state of the art for quantifying landscape spatial pattern including lessons learned by the author.}
 
-\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the *R* programming language is wideley used in ecology including possible reasons for the wide spread use.}
+\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the \textit{R} programming language is widely used in ecology including possible reasons for the wide spread use.}
 

### 2021-04-26 paper/paper.Rmd
Adding CRS section

@@
   \usepackage{placeins}
+  \usepackage{float}
   \usepackage[modulo]{lineno}
@@
 
-\begin{table}
+\begin{table}[H]
 
@@
 All packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations, allowing unification of spatial projections when the used data have different CRS.
+
+\textcolor{blue}{Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
+However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and started to use a new representation called WKT2-2019 for CRS storing.
+Thus, spatial objects saved with degraded Proj4 string representation to \textit{R} file formats could be incorrectly read, and special attention should be given to check their correctness.
+Proj4 string representation is also not recommended to set or transform CRS and instead the previously introduced EPSG codes should be used.
+A more detailed explanation of the recent CRS changes can be found in \cite{Bivand2020}.}
 

### 2021-04-26 paper/paper.tex
Adding CRS section

@@
 \usepackage{placeins}
+\usepackage{float}
 \usepackage[modulo]{lineno}
@@
 All packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations, allowing unification of spatial projections when the used data have different CRS.
+
+\textcolor{blue}{Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
+However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and started to use a new representation called WKT2-2019 for CRS storing.
+Thus, spatial objects saved with degraded Proj4 string representation to \textit{R} file formats could be incorrectly read, and special attention should be given to check their correctness.
+Proj4 string representation is also not recommended to set or transform CRS and instead the previously introduced EPSG codes should be used.
+A more detailed explanation of the recent CRS changes can be found in \cite{Bivand2020}.}
 

### 2021-04-27 paper/paper.Rmd

@@
 In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology.
-\textbf{Recent findings} In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis.
+\textbf{Recent findings} In this paper, we examine existing packages for downloading, processing and visualisation of spatial data, as well as, those specifically developed for ecological analysis.
 Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community.
@@
 
-API & Application programming interface. A set of protocols for sending and retrieving information from a server \\
+API & Application programming interface. A set of protocols for sending and retrieving information, for example, from a server \\
 cell & Smallest, rectangular unit of raster data model \\
@@
 patch & Neighboring cells with same value \\
-PROJ & Open-source generic coordinate transformation software \\
+PROJ & Open-source generic coordinate transformation library \\
 raster & Cell grid based spatial data model \\
@@
 
-Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
-Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
+Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
+Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is essential and ecological processes can vary spatially \cite{With2019}.
 Thus, landscape ecology emphasizes \textcolor{blue}{spatial} patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize these \textcolor{blue}{spatial patterns}.
@@
 Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
-Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
+Additionally, open-source software facilitates (computational) reproducibility and allows a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
 Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
 Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
-This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.
+This democratizes scientific research as free-of-cost software removes a gatekeeper for researchers without access to proprietary software.
 For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 \textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the \textit{R} programming language \cite{Wickham2015}}.
-Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
-Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
+Firstly released in 1995 \cite{Smith2016}, \textcolor{blue}{this} programming language is among the most popular programming languages today, especially in ecology \cite{Lai2019}.
+Originally introduced as a statistical programming language, a large number of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
 Thus, the \textit{R} programming language can be a very powerful tool. 
@@
 The growing popularity of the \textit{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
-A recent overview over the progress of \textit{R} to handle spatial data in general can be found in \cite{Bivand2020}.
+A recent overview over the progress of \textit{R} to handle spatial data can be found in \cite{Bivand2020}.
 
-Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the \textit{R} programming language.
-\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review, we focus on software implemented in the \textit{R} programming language.
+\textcolor{blue}{We acknowledge that there are reasons and circumstances also to use other software tools and programming languages, however, a comprehensive discussion of pros and cons is outside the scope of this manuscript.}
 Thus, in the first part of this article, we present an overview of existing \textit{R} packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
@@
 \textit{GitHub} hosts source code under version control, and allows users to install packages with one line of code using the \textit{remotes} \cite{Hester2020} package.
-Additionally, hosting a package on \textit{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e.
-testing if functions return an expected value).
+Additionally, hosting a package on \textit{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e., testing if functions return an expected value).
 
-The guaranteed technical quality standard of \textit{CRAN} requires more initial work for developers compared to \textit{GitHub}, while it ensures for users that the package can be installed on their machine.
+The guaranteed technical quality standard of \textit{CRAN} requires more initial work for developers compared to \textit{GitHub}. 
+At the same time, it ensures for users that the package can be installed on their machine.
 Additionally, the technical quality standard of \textit{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \textit{R} packages, i.e., package \textit{x} requires and uses code from package \textit{y}.
-CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
-Thus, most packages can be found on both platforms, and many developers use \textit{GitHub} for regular development and \textit{CRAN} to publish stable releases of the packages.
+CRAN also provides archived versions of outdated or orphaned packages, thus ensuring long-term availability and reproducibility of code.
+That being said, most R packages can be found on both platforms, and many developers use \textit{GitHub} for regular development and \textit{CRAN} to publish stable releases of the packages.
 Furthermore, online communities like \textit{rOpenSci} also provide a peer-review process for code quality.
-However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
+It is important to mention that while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backward compatibility is not always ensured.
 Packages that deal with such issue include \textit{groundhog} \cite{Simonsohn2021}, \textit{packrat} \cite{Ushey2018}, or \textit{renv} \cite{Ushey2020}.
-\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
+\textcolor{blue}{These packages facilitate reproducibility to a high degree by preserving the project environment, including specific package versions used for the analysis.}
 For more information about \textit{R} package development in general, see \cite{Wickham2015}.
@@
 While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by pacakges.} 
+\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.} 
 Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
-Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and for coordinates transformation.
 Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
@@
 The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
-In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
+In many cases, this data model allows a more realistic representation of landscape features, however, with the higher computational cost \cite{Lovelace2019}.
 The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
@@
 However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
-Besides many advantages and strength of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
+Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
 \textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data science project. 
 Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
-\textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
+\textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 
-Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps \cite{Lovelace2019,Pebesma2019a}.
-This is also referred to as the spatial projection and is often the first barrier in spatial data analysis.
+Coordinate reference systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps \cite{Lovelace2019,Pebesma2019a}.
+This is also referred to as spatial projection and is often the first barrier in spatial data analysis.
 It is not only required to have all of the used data in the same projection, but also to select a proper CRS.
@@
 \textcolor{blue}{Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
-However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and started to use a new representation called WKT2-2019 for CRS storing.
+However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and switched to use a new representation called WKT2-2019 for CRS storing.
 Thus, spatial objects saved with degraded Proj4 string representation to \textit{R} file formats could be incorrectly read, and special attention should be given to check their correctness.
@@
 In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
-Related to that, masking certain areas of the study region using spatial filters (e.g. water bodies, urban areas) can be required.
+Related to that, masking certain areas of the study region using spatial filters (e.g., water bodies, urban areas) can be required.
 Packages from Section \@ref(sec:spatrep) also allow for these operations.
@@
 
-Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
+Finally, there are several tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
 It is possible to control several GIS software directly from \textit{R} using dedicated packages, such as \textit{rgrass7} \cite{Bivand2021} for GRASS GIS, \textit{RSAGA} \cite{Brenning2018} for SAGA GIS, and \textit{rgee} \cite{Aybar2020} for Google Earth Engine.
-\textcolor{blue}{By being able to access these GIS software directly from within \textit{R}, these packages can allow to create easily reproducible and sharable \textit{R} scripts, even if functionality is currently only available in the corresponding GIS software.}
+\textcolor{blue}{Direct access to these GIS software within \textit{R} allows to create reproducible and sharable \textit{R} scripts, even if functionality is currently not available in \textit{R}.}
 
@@
 
-\textbf{2.} Understand the fundamental spatial data types i.e, raster data and vector data.
+\textbf{2.} Understand the fundamental spatial data types, i.e., raster data and vector data.
 This also includes being able to process the data using the corresponding packages such as \textit{raster} or \textit{sf}. 
-These packakage also have excellent documentation at \url{https://rspatial.org/raster} and \url{https://r-spatial.github.io/sf}, respectively. \\
+These packages also have excellent documentation at \url{https://rspatial.org/raster} and \url{https://r-spatial.github.io/sf}, respectively. \\
 
@@
 \textbf{4.} Do a quick visualization of the data for inspection or a first glance at the data. 
-This can be done with base \textit{R} or if already familiar \textit{ggplot2}. \\
+This can be done with base \textit{R} or if already familiar - \textit{ggplot2}. \\
 
 \textbf{5.} Find suitable \textit{R} package for research questions. 
-This can be difficult since research questions, available data and package are diverse. 
+This can be difficult since research questions, available data and packages are diverse. 
 Good starting points are the CRAN Task Views (mainly \url{https://cran.r-project.org/web/views/Spatial.html}) or review papers similar to this one. 
@@
 
-\textbf{6.} Create publication-quality figures and maps using e.g. \textit{ggspatial} or \textit{tmap}. 
+\textbf{6.} Create publication-quality figures and maps using, e.g., \textit{ggspatial} or \textit{tmap}. 
 Help for both packages can be found on the corresponding homepages \url{https://paleolimbot.github.io/ggspatial/index.html} or \url{https://mtennekes.github.io/tmap}. \\
@@
 
-\textbf{8.} Further excellent resources with applied help about spatial data in \textit{R} can be found \url{https://keen-swartz-3146c4.netlify.app}, \url{https://rspatial.org}, or \url{https://geocompr.robinlovelace.net}.
+\textbf{8.} Further excellent resources with applied help about spatial data in \textit{R} can be found \url{https://geocompr.robinlovelace.net}, \url{https://rspatial.org}, or \url{https://r-spatial.org/book/}.
 \\
@@
 \textit{R} allows two major types of maps.
-Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.
-changing the displayed values.
+Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user can modify the map by, e.g., changing the displayed values.
 All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
 However, the generic functions are focused on quick visual inspection of the data (Fig.
-\@ref(fig:fig-map)a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor multiples maps.}
+\@ref(fig:fig-map)a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor small multiples maps.}
 
@@
 Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram \cite{Nowosad2021}.
-Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
+Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g., Euclidian, Jensen-Shannon, Jaccard).
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
@@
 \textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
-Also, the survey did not include any personal questions, such as age or country of residence, which might have an effect on the representation of the survey.}.
+Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey.}.
 All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey)b).}
+\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options} (Fig. \@ref(fig:fig-survey)b).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
@@
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same  \textcolor{blue}{scripted} programming environment, \textit{R} allows for transparent and reproducible workflows.
-\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the \textit{R} programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
-This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore are often only available for certain operating systems.}
+\textcolor{blue}{Scripts including all analysis steps and parameter settings can easily be shared, thus the script-based characteristic of the \textit{R} programming language allows to easily distribution and reproduction of analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
+This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore, are often only available for certain operating systems.}
 This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
@@
 The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the \textit{R} programming language for landscape ecology. 
-Furthermore, the survey showed the that many members of the landscape ecology community actively develop \textit{R} packages themselves, demonstrating that tools are constantly added and updated.
+Furthermore, the survey showed that many members of the landscape ecology community actively develop \textit{R} packages themselves, demonstrating that tools are constantly added and updated.
 
-Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
+Landscape ecology combines many different research topics and methodological approaches, and most of them heavily rely on spatial data.
 While the \textit{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \textit{R} programming language has to face.
 \textcolor{blue}{Some packages discussed herein, such as \textit{raster} and \textit{terra}, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
-Furthermore, also parallel processing, and even including the use of high performance clusters, is provided by several \textit{R} packages. 
+Furthermore, also parallel processing, including the use of high performance clusters, is provided by several \textit{R} packages. 
 While this is outside the scope of this manuscript, further resources about parallel processing in \textit{R} can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.} 
 Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \textit{R} packages related to those fields.
-But as the development of \textit{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
+Nevertheless, as the development of \textit{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
 
@@
 LJG was supported by the Natural Environment Research Council (NERC) grant \textit{A complex-systems approach to improve understanding of the biodiversity-landscape structure relationship}, grant number NE/T009373/1.
-All other authors did not receive any funding for this research.
+Other authors did not receive any funding for this research.
 

### 2021-04-27 paper/paper.tex
I think data analysis is good for tidyverse

@@
 Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
-Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
+Additionally, open-source software facilitates (computational) reproducibility and allows a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
 Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
 Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
-This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.
+This democratizes scientific research as free-of-cost software removes a gatekeeper for researchers without access to proprietary software.
 For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
@@
 \textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the \textit{R} programming language \cite{Wickham2015}}.
-Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
-Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
+Firstly released in 1995 \cite{Smith2016}, \textcolor{blue}{this} programming language is among the most popular programming languages today, especially in ecology \cite{Lai2019}.
+Originally introduced as a statistical programming language, a large number of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
 Thus, the \textit{R} programming language can be a very powerful tool.
@@
 The growing popularity of the \textit{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
-A recent overview over the progress of \textit{R} to handle spatial data in general can be found in \cite{Bivand2020}.
+A recent overview over the progress of \textit{R} to handle spatial data can be found in \cite{Bivand2020}.
 
-Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the \textit{R} programming language.
-\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
+Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review, we focus on software implemented in the \textit{R} programming language.
+\textcolor{blue}{We acknowledge that there are reasons and circumstances also to use other software tools and programming languages, however, a comprehensive discussion of pros and cons is outside the scope of this manuscript.}
 Thus, in the first part of this article, we present an overview of existing \textit{R} packages for landscape ecology and closely related fields (Table \ref{tab:packages}).
@@
 \textit{GitHub} hosts source code under version control, and allows users to install packages with one line of code using the \textit{remotes} \cite{Hester2020} package.
-Additionally, hosting a package on \textit{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e.~testing if functions return an expected value).
+Additionally, hosting a package on \textit{GitHub} provides many useful features to collaborate and communicate between developers and users \cite{Wickham2015}, or integrated unit testing (i.e., testing if functions return an expected value).
 
-The guaranteed technical quality standard of \textit{CRAN} requires more initial work for developers compared to \textit{GitHub}, while it ensures for users that the package can be installed on their machine.
+The guaranteed technical quality standard of \textit{CRAN} requires more initial work for developers compared to \textit{GitHub}.
+At the same time, it ensures for users that the package can be installed on their machine.
 Additionally, the technical quality standard of \textit{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \textit{R} packages, i.e., package \textit{x} requires and uses code from package \textit{y}.
-CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
-Thus, most packages can be found on both platforms, and many developers use \textit{GitHub} for regular development and \textit{CRAN} to publish stable releases of the packages.
+CRAN also provides archived versions of outdated or orphaned packages, thus ensuring long-term availability and reproducibility of code.
+That being said, most R packages can be found on both platforms, and many developers use \textit{GitHub} for regular development and \textit{CRAN} to publish stable releases of the packages.
 Furthermore, online communities like \textit{rOpenSci} also provide a peer-review process for code quality.
-However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
+It is important to mention that while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backward compatibility is not always ensured.
 Packages that deal with such issue include \textit{groundhog} \cite{Simonsohn2021}, \textit{packrat} \cite{Ushey2018}, or \textit{renv} \cite{Ushey2020}.
-\textcolor{blue}{These packages facilitate reproduciability to a high degree by preserving the project enviroment, including specific package versions, used for an analysis.}
+\textcolor{blue}{These packages facilitate reproducibility to a high degree by preserving the project environment, including specific package versions used for the analysis.}
 For more information about \textit{R} package development in general, see \cite{Wickham2015}.
@@
 While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by pacakges.}
+\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.}
 Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
-Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
+Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and for coordinates transformation.
 Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.
@@
 However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
-Besides many advantages and strength of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
-\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data science project. 
+Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
+\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
 Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
-\textit{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.
+\textit{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 In this case, the pre-processing of spatial data should include vector clipping or raster cropping.
-Related to that, masking certain areas of the study region using spatial filters (e.g.~water bodies, urban areas) can be required.
+Related to that, masking certain areas of the study region using spatial filters (e.g., water bodies, urban areas) can be required.
 Packages from Section \ref{sec:spatrep} also allow for these operations.
@@
 
-Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
+Finally, there are several tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
 It is possible to control several GIS software directly from \textit{R} using dedicated packages, such as \textit{rgrass7} \cite{Bivand2021} for GRASS GIS, \textit{RSAGA} \cite{Brenning2018} for SAGA GIS, and \textit{rgee} \cite{Aybar2020} for Google Earth Engine.
-\textcolor{blue}{By being able to access these GIS software directly from within \textit{R}, these packages can allow to create easily reproducible and sharable \textit{R} scripts, even if functionality is currently only available in the corresponding GIS software.}
+\textcolor{blue}{Direct access to these GIS software within \textit{R} allows to create reproducible and sharable \textit{R} scripts, even if functionality is currently not available in \textit{R}.}
 
@@
 \textit{R} allows two major types of maps.
-Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.~changing the displayed values.
+Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user can modify the map by, e.g., changing the displayed values.
 All packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the generic \texttt{plot()} function (Fig. \ref{fig:fig-map}a).
 However, the generic functions are focused on quick visual inspection of the data (Fig.
-\ref{fig:fig-map}a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor multiples maps.}
+\ref{fig:fig-map}a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor small multiples maps.}
 
@@
 Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram \cite{Nowosad2021}.
-Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g.~Euclidian, Manhattan, Mahlanobis).
+Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g., Euclidian, Jensen-Shannon, Jaccard).
 This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
@@
 \textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
-Also, the survey did not include any personal questions, such as age or country of residence, which might have an effect on the representation of the survey.}.
+Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey.}.
 All raw data and scripts to analyze the data can be found at \url{https://github.com/r-spatialecology/Hesselbarth_et_al_CLER}.
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \\ref{fig:fig-survey}b).}
+\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options} (Fig. \ref{fig:fig-survey}b).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 Because it is possible to import, modify, analyze, and visualize spatial data all in the same \textcolor{blue}{scripted} programming environment, \textit{R} allows for transparent and reproducible workflows.
-\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the \textit{R} programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
-This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore are often only available for certain operating systems.}
+\textcolor{blue}{Scripts including all analysis steps and parameter settings can easily be shared, thus the script-based characteristic of the \textit{R} programming language allows to easily distribution and reproduction of analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
+This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore, are often only available for certain operating systems.}
 This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
@@
 The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the \textit{R} programming language for landscape ecology.
-Furthermore, the survey showed the that many members of the landscape ecology community actively develop \textit{R} packages themselves, demonstrating that tools are constantly added and updated.
+Furthermore, the survey showed that many members of the landscape ecology community actively develop \textit{R} packages themselves, demonstrating that tools are constantly added and updated.
 
-Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
+Landscape ecology combines many different research topics and methodological approaches, and most of them heavily rely on spatial data.
 While the \textit{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \textit{R} programming language has to face.
 \textcolor{blue}{Some packages discussed herein, such as \textit{raster} and \textit{terra}, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
-Furthermore, also parallel processing, and even including the use of high performance clusters, is provided by several \textit{R} packages. 
+Furthermore, also parallel processing, including the use of high performance clusters, is provided by several \textit{R} packages. 
 While this is outside the scope of this manuscript, further resources about parallel processing in \textit{R} can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.}
 Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \textit{R} packages related to those fields.
-But as the development of \textit{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
+Nevertheless, as the development of \textit{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
 

### 2021-04-28 Comments_Reviewer.Rmd
Some layout changes
Adding CRS section
Several fixes in the reply to the comments.

@@
 However, we think that given recommendations which packages to use is very difficult for several reasons. 
-First, packages have different advantages and disadvantages that differ depending on e.g., the research question, the available data, or the personal preferences of the users.
+First, packages have different advantages and disadvantages that differ depending on e.g., the research question, the available data, or the personal preferences and experiences of the users.
 Second, we believe that trying out several packages for similar task can actually beneficial and strengthen trust in the method.
@@
 
-**Reply**: We agree that the of Rayfield et al.
-2011 is great, but we also want to point out that it was developed for a more specific topic (i.e., connectivity), while this manuscript deals with the whole field of landscape ecology.
+**Reply**: We agree that the approach of Rayfield et al.
+2011 is great, but we also want to point out that it was developed for a more specific domain (i.e., connectivity), while this manuscript deals with the whole field of landscape ecology.
 Thus, we believe that a similar typology for all packages included in this manuscript would be too complicated to be of actual help.
@@
 
-**Reply**: We hope that the modified Figure 1, the added Box 1 and some added explanation in the conclusions explain how using R can facilitate a reproducible workflow in landcape ecology.
+**Reply**: We hope that the modified Figure 1, the added Box 1 and some added explanation in the conclusions explain how using R can facilitate a reproducible workflow in landscape ecology.
 \vskip 0.5cm
@@
 
-**Reply**: We agree that it was not clear enough how these package could help woth reproducibility, thus we added a more detailed explanation. \vskip 0.5cm
+**Reply**: We agree that it was not clear enough how these package could help with reproducibility, thus we added a more detailed explanation. \vskip 0.5cm
 
@@
 
-**Reply**: @Jakub? \vskip 0.5cm
+**Reply**: We added an explanation of the recent changes to PROJ and make users aware to pay attention if old Proj4 strings are used. 
+Furthermore, we advise to use ESPG codes for CRS. \vskip 0.5cm
 
@@
 **Reply**: We added that we aware that the survey is most likely not representative for the R community and R users due to how we tried to reach the community and also due to the questions asked in the survey.
-Since the survey is not the main idea of the manuscript though, we think that the survery still migth be interesting even with these shortcomings. \vskip 0.5cm
+Since the survey is not the main idea of the manuscript though, we think that the survey still might be interesting even with these shortcomings. \vskip 0.5cm
 
@@
 **Reply**: We agree that the sentence implied that the R community is not constantly adapting to the new developments of landscape ecology which is certainly not true. 
-We reworded the sentence making clearea that indeed the community constantly improves the R programming language. 
+We reworded the sentence making clearer that indeed the community constantly improves the R programming language. 
 However, we do not agree with the reviewer that "[...] the problem is that there are so many packages [..]".
@@
 "Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, R allows for transparent and reproducible workflows."
-But the authors do not mention any tools for facililitating reproducibility.
+But the authors do not mention any tools for facilitating reproducibility.
 This is a huge part of why R should be more widely used, and should be expanded upon.
@@
 We also acknowledge that the survey is most likely not representative for all skill levels of *R* users because we assume that by using mailing list and social media more advanced users were more likely to take part in the survey.
-We added and clearified those shortcomings of the survey. \vskip 0.5cm
+We added and clarified those shortcomings of the survey. \vskip 0.5cm
 

### 2021-04-28 R/code_example.Rmd
Adding code example as Rmd

---
title: "Code example"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Code to reproduce Fig. 2 in "Open-source tools in R for landscape ecology"

Code created by MHKH, JN, JS and LJG

Contact: mhk.hesselbarth@gmail.com

### Setup

First, we need to load all required packages used in this R script. 
In case you are missing a package, it can be install from CRAN using e.g.,
install.packages("raster"). 
By loading the packages, we make sure all functions from them can be used in the script.

```{r load_libs, message = FALSE}
library(ggplot2)
library(ggspatial)
library(raster)
library(rnaturalearth)
library(sf)
library(tmap)
```

### Download data

This function downloads data and saves it as a RasterLayer. 
The 'name' argument allows to specify a source to download data from the worldclim database. 
The 'lon' and 'lat' argument specify the SRTM tile for which the data is downloaded, and the 'res' the resolution of the data in minutes of a degree. 
The 'var' argument specifies which kind of data is downloaded, in this case precipitation data, and lastly the path on the local disk where the data is saved can be set.

```{r download_data_dummy, eval = FALSE}
precipitation <- raster::getData(name = "worldclim", lon = 3, lat = 38,
                                 res = 0.5, var = "prec", path = "R/")
```

```{r load, eval = TRUE, include = FALSE}
library(stringr)

files <- list.files("wc0.5", pattern = "*.bil", full.names = TRUE)

files <- stringr::str_sort(files, numeric = TRUE)

precipitation <- lapply(files, raster::raster)

precipitation <- raster::stack(precipitation)

raster::crs(precipitation) <- "+proj=longlat +datum=WGS84 +no_defs"

```

### Pre-process data

The previous downloaded object is a RasterStack in which each layer corresponds to the precipitation in a month (Jan - Dec). 
To get the annual precipitation, we can use the calc function which applies basic mathematical function to each layer of the RasterStack. 
By getting the sum of all layers for each cell, we get the annual precipitation.

```{r sum_precipitaiton}
precipitation_sum <- raster::calc(x = precipitation, fun = sum)
```

Next, we need to ensure that all spatial data has the same coordinate reference system. 
To achieve this we reproject the downloaded data to the CRS epsg:21781. 
This returns a warning due to the Proj4 definition, which can be ignored at this point.

```{r reproject}
precipitation_sum <- raster::projectRaster(precipitation_sum, crs = "EPSG:21781")
```

Next, we download administration boundaries of Switzerland as vector data from the naturalearth database. Again, we need to specify the resolution of the data (called 'scale' here). 
Additionally, the function allows to select the return type and we select a sf vector object.

```{r boundaries_ch}
switzerland <- rnaturalearth::ne_countries(scale = 50,
                                           country = "switzerland",
                                           returnclass = "sf")
```

To plot the precipitation data together with the administration boundaries, we also need to reproject the boundaries to the same coordinate reference system epsg:21781.

```{r reproject_ch}
switzerland <- sf::st_transform(x = switzerland, crs = "EPSG:21781")
```

Because the precipitation data includes not just Switzerland but whole Europe, we want to crop the raster data by including only cells that are within the the administration boundaries of Switzerland. 
However, we add a buffer of 5000m around the boundaries.

```{r crop}
precipitation_sum_ch <- raster::crop(x = precipitation_sum,
                                     y = sf::st_buffer(x = switzerland, dist = 5000))
```

The precipitation data is continuous, however, we want to plot discrete classes. 
To classify the continuous values into 5 classes, we can use the cut function and simply specify the number of breaks.

```{r cut}
precipitation_class_ch <- raster::cut(precipitation_sum_ch, breaks = 5)
```

We are going to use the same colors for all maps set by hex color codes.
There are many homepages to find such codes e.g., <https://htmlcolorcodes.com>. 
We need one color for each class i.e. 5 different colors.

```{r col}
colors <- c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF")
```

### Create maps

To create a map in base R we first plot the raster with precipitation classes, setting the colors of the 5 classes to the previously set hex codes. 
Next, we are going to add the vector data to the plot by plotting the sf object and setting the 'add' argument to TRUE which just adds them to the already present map. 
For better visualization, we add the vector data twice using white and black colors.

```{r base_plot, eval = FALSE}
# pdf(file = "R/Figures/base_plot.pdf")

plot(precipitation_class_ch, col = colors)
plot(st_geometry(switzerland), add = TRUE, lwd = 4, border = "white")
plot(st_geometry(switzerland), add = TRUE)

# dev.off()
```

We can create the same map as a ggplot2 map. 
Therefore, we first specify which data to use for the plot and then add subsequently the different so-called geoms. 
This allows to first plot the raster and then add the vector data. 
Lastly, we can modify some parts of the map as the axis labels or the general appearance (theme). 
For more information about ggplot2 in general, please see <https://ggplot2-book.org>

```{r ggplot, eval = FALSE}
plot_gg <- ggplot(raster::as.data.frame(precipitation_class_ch, xy = TRUE)) +
    geom_raster(aes(x = x, y = y, fill = factor(layer))) +
    geom_sf(data = switzerland, fill = NA, col = "white") +
    scale_fill_manual(name = "Precipitation\nclassified", values = colors) +
    coord_sf() +
    annotation_scale(location = "br") +
    annotation_north_arrow(location = "tr", which_north = "true") +
    labs(x = "", y = "") +
    theme_classic() +
    theme(legend.position = "bottom")

# ggplot2::ggsave(filename = "R/Figures/ggplot2.pdf")
```

Creating the map with tmap is comparable to ggplot2. 
First, we need to specify the data to use and subsequently we can add elements to the map. 
This includes the raster data, the vector data, a scale bar, a legend (besides other options). 
For more information about tmap in general, please see <https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html>

```{r tmap, eval = FALSE}
plot_tm <- tm_shape(precipitation_class_ch) +
    tm_graticules() +
    tm_raster(title = "Precipitation\nclassified",
              style = "cat",
              palette = colors,
              legend.is.portrait = FALSE) +
    tm_shape(switzerland) +
    tm_borders(lwd = 2, col = "white") +
    tm_scale_bar(breaks = c(0, 25, 50),
                 bg.color = "white") +
    tm_compass(position = c("right", "top"),
               bg.color = "white") +
    tm_layout(legend.outside = TRUE,
              legend.outside.position = "bottom")

# tmap::tmap_save(plot_tm, "R/Figures/tmap.pdf")
```

### 2021-04-28 R/code_example.R

@@ -12,12 +12,11 @@
 # First, we need to load all required packages used in this R script. In case you are
 # missing a package, it can be install from CRAN using e.g., install.packages("raster").
 # By loading the packages, we make sure all functions from them can be used in the script.
-library(sf)
-library(raster)
-library(viridis)
-library(rnaturalearth)
 library(ggplot2)
 library(ggspatial)
+library(raster)
+library(rnaturalearth)
+library(sf)
 library(tmap)
 
 # This function downloads data and saves it as a RasterLayer. The 'name' argument

### 2021-04-28 paper/paper.Rmd
I think data analysis is good for tidyverse
Small typo

@@
 Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
-\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data science project. 
+\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
 Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
@@
 Packages are sorted by their major application task.
-The table only includes packages specifically desinged for spatial data or landscape ecology.}
+The table only includes packages specifically designed for spatial data or landscape ecology.}
 

### 2021-04-29 paper/paper.tex
Small typo
A few small changes.

@@
 Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
-\textit{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.
+\textit{sf} builds on the idea of ``simple features'', a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.
 
@@
 \textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
-Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey.}.
+Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey}.
 All raw data and scripts to analyze the data can be found at \url{https://github.com/r-spatialecology/Hesselbarth_et_al_CLER}.
@@
 
-We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig.
-\ref{fig:fig-survey}a)).
+We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'', and ``nature conservation'' (Fig.
+\ref{fig:fig-survey}a).
 
@@
 
-Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \textit{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
-\ref{fig:fig-survey}d)).
+Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \textit{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
+\ref{fig:fig-survey}d).
 

### 2021-04-29 Comments_Reviewer.Rmd
Some last typos

@@
 Furthermore, we added explanations for all packages that were missing in the original submission as suggested by reviewer 1.
-However, we think that given recommendations which packages to use is very difficult for several reasons. 
+However, we think that recommendations which packages to use are very difficult for several reasons. 
 First, packages have different advantages and disadvantages that differ depending on e.g., the research question, the available data, or the personal preferences and experiences of the users.
-Second, we believe that trying out several packages for similar task can actually beneficial and strengthen trust in the method.
+Second, we believe that trying out several packages for similar tasks can actually be beneficial and strengthen trust in the method.
 But, by modifying Figure 1 and adding Box 1 we hope to give some more guidelines on how to organize an analysis.
@@
 Thus, we believe that a similar typology for all packages included in this manuscript would be too complicated to be of actual help.
-Thank you for this great idea to add boxes. 
+Thank you for the great idea to add boxes. 
 We added a box "Where to start" to give some major steps how to approach landscape ecology in R for novice users. 
 We hope this will also at least partly address the issue raised by reviewer 1 about recommendations.
-We also thought about a second box for mode advanced users and even developers, but decided that the target audience of the paper are most likely novice to intermediate R users and not developers and the topic of developing own packages specifically for the landscape ecology community could maybe be a future manuscript.
+Further, we thought about a second box for more advanced users and/or developers, but decided that the target audience of the manuscript are most likely novice to intermediate R users and not developers.
+The topic of developing own packages specifically for the landscape ecology community could be an interesting future manuscript.
 Also, we added a short explanation of the tidyverse and its strength.
@@
 **Reply**: Again, thank you very much for the appreciated and valuable suggestions for improving the manuscript and we hope that the revised manuscript improved compared to the previous version with help of the reviewers.
-All changes to the manuscript are indicated by the blue color. \vskip 0.5cm
+All changes to the manuscript are indicated by the blue color in the track change version. \vskip 0.5cm
 
@@
 While all of the authors are enthusiastic R users, we think that a comprehensive discussion of R vs.
-other open-source software is outside of the scope of this paper and more important not expedient in general since other software might work better for other people (e.g., Python).
+other open-source software is outside of the scope of this paper and more importantly not expedient in general since other software might work better for other people (e.g., Python).
 \vskip 0.5cm
@@
 Reasons for this include, besides others, personal preferences of the users or analysis specific circumstances. 
-Lastly, the review should be as neutral as possible without favoring one package over another because pros and cons of package differ between users and all were developed with a specific idea in mind.
+Lastly, we believe that the review should be as neutral as possible without favoring one package over another because pros and cons of packages differ between users and all were developed with a specific idea in mind.
 But, we think that our new "Box 1: Where to start" hopefully can assist novice users with their workflow. \vskip 0.5cm
@@
 
-**Reply**: Thank you for this valuable remark.
-We tried to add additional explanations to concepts which might be underexplained to especially novice users which hopefully increases the value of the manuscript.
-\vskip 0.5cm
+**Reply**: Thank you for this valuable remark. We tried to add additional explanations to concepts which might be underexplained especially for novice users. 
+We hope that this increases the value of the manuscript. \vskip 0.5cm
 
@@
 
-**Reply**: We added an explanation how these package allow to incorporate functionality currently only available in GIS software into reproducible and shareable R scripts. \vskip 0.5cm
+**Reply**: We added an explanation how these packages allow to incorporate functionality currently only available in GIS software into reproducible and shareable R scripts. \vskip 0.5cm
 
@@
 
-**Reply**: We agree that this is problematic and we weren't even aware this is happening so thanks for pointing it out. 
+**Reply**: We agree that this is problematic and we were not aware this is happening so thanks for pointing it out. 
 We deactivated the automatic hyphenation by LaTeX. \vskip 0.5cm
@@
 **Reply**: Thank you for checking the table so extensively. 
-We made sure all packages should be included now. 
-We also added the information to the table caption that not all packages from the manuscript are included but only packages specifically designed for spatial data and landscape ecology. \vskip 0.5cm
+We made sure all packages related to spatial data and landscape ecology should be included now. 
+Thus, we also added the information to the table caption that not all packages from the manuscript are included but only packages specifically designed for spatial data and landscape ecology.
+\vskip 0.5cm
 
@@
 
-**Reply**: We added that we aware that the survey is most likely not representative for the R community and R users due to how we tried to reach the community and also due to the questions asked in the survey.
-Since the survey is not the main idea of the manuscript though, we think that the survey still might be interesting even with these shortcomings. \vskip 0.5cm
+**Reply**: We added that we aware that the survey is most likely not representative for the R community due to how we tried to reach the community and also due to the questions asked in the survey. 
+Since the survey is not the main idea of the manuscript though, we think that the survey still might be interesting - even with these shortcomings. \vskip 0.5cm
 
@@
 Even though we acknowledge that the amount of package can be overwhelming, especially for novices users, diversity of packages is a big strength - even for similar specific tasks. 
-Possible advantages include e.g., the possibility to compare different implementations of similar/identical task or providing functionality for slightly different data structures. \vskip 0.5cm
+Possible advantages include e.g., the possibility to compare different implementations of similar/identical task or providing functionality for slightly different data structures and user preferences.
+\vskip 0.5cm
 
@@
 
-**Reply**: We highly agreed that this one of the main reasons why R should be widely used and we added an explanation how the script-based character of R itself allows to share and reproduce analysis.
+**Reply**: We highly agreed that this one of the main reasons why R should be widely used and we added an explanation how the script-based characteristic of R itself allows to share and reproduce analysis.
 \vskip 0.5cm
@@
 
-**Reply**: Please see our reply to the previous comment. 
-We reworded the sentence to state clearer that this is indeed one of the strength of the R community. \vskip 0.5cm
+**Reply**: We reworded the sentence to state clearer that this is indeed one of the strength of the R community (Please see also our reply to the previous comment regarding this). \vskip 0.5cm
 
@@
 **Reply**: We apologize for this.
-The manuscript was written using the rticles packages with the corresponding template for Springer journals, which somehow did not include journal names. 
-We fixed the citation style including journal names. \textcolor{red}{STILL NEED TO DO THIS ACTUALLY!} \vskip 0.5cm
+There was an issue with our bibliography file, which we fixed and the citation style includes journal names now. \vskip 0.5cm
 
@@
 
-**Reply**: We agree with the reviewer that the provided script is not used for experienced R users and targeted to novices.
+**Reply**: We agree with the reviewer that the provided script is not useful for experienced R users and targeted to novices users.
 Thus, we tried to improve the the documentation of the code. \vskip 0.5cm
@@
 However, unfortunately such questions were not included.
-We also acknowledge that the survey is most likely not representative for all skill levels of *R* users because we assume that by using mailing list and social media more advanced users were more likely to take part in the survey.
+We also acknowledge that the survey is most likely not representative for all skill levels of R users because we assume that by using mailing list and social media more advanced users were more likely to take part in the survey.
 We added and clarified those shortcomings of the survey. \vskip 0.5cm

### 2021-04-29 paper/paper.Rmd
A few small changes.
Some last typos

@@
   chunk_output_type: console
-  markdown:
-    wrap: sentence
 ---
@@
 \textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.} 
-Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
+Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, Table \@ref(tab:packages)).
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
@@
 \textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
-Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
+Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
 \textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.
@@
 
-Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
+Nowadays, spatial data at various scales is available from many online-accessible sources.
 A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \textit{R} session.
@@
 \textcolor{blue}{Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
-However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and switched to use a new representation called WKT2-2019 for CRS storing.
+However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and switched to use a new representation called WKT2-2019 for storing the CRS.
 Thus, spatial objects saved with degraded Proj4 string representation to \textit{R} file formats could be incorrectly read, and special attention should be given to check their correctness.
@@
 \textbf{4.} Do a quick visualization of the data for inspection or a first glance at the data. 
-This can be done with base \textit{R} or if already familiar - \textit{ggplot2}. \\
+This can be done with base \textit{R} or -- if already familiar -- \textit{ggplot2}. \\
 
@@
 \hline
-\multirow{3}{*}{Bridge stuff} & rgrass7 & Interface between GRASS7 and R  & \cite{Bivand2021} \\
+\multirow{3}{*}{Interfacing to GIS} & rgrass7 & Interface between GRASS7 and R  & \cite{Bivand2021} \\
 & RSAGA & Interface between SAGA and R & \cite{Brenning2018} \\
@@
 \hline
-\multirow{5}{*}{Connectivity} & \textit{lconnect} & Calculate landscape connectivity metrics & \cite{Mestre2019} \\
+\multirow{5}{*}{Connectivity} & lconnect & Calculate landscape connectivity metrics & \cite{Mestre2019} \\
 & Makurhini & Calculate fragmentation and landscape connectivity indices & \cite{Godinez-Gomez2020} \\
@@
 \textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
-Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey.}.
+Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey}.
 All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).
@@
 We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10% of participants included "biodiversity", followed by "land use management", "landscape connectivity", and "nature conservation" (Fig.
-\@ref(fig:fig-survey)a)).
+\@ref(fig:fig-survey)a).
 
@@
 Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only very few participants evaluated \textit{R} as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig.
-\@ref(fig:fig-survey)d)).
+\@ref(fig:fig-survey)d).
 

### 2021-04-29 paper/paper.tex
Some last typos

--- 2021-04-27 paper/paper.tex
@@
 \textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.}
-Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
+Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, Table \ref{tab:packages}).
 Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
@@
 \textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
-Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
+Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
 \textit{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.
@@
 We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig.
-\ref{fig:fig-survey}a)).
+\ref{fig:fig-survey}a).
 
@@
 Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \textit{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
-\ref{fig:fig-survey}d)).
+\ref{fig:fig-survey}d).
 

### 2021-05-03 Comments_Reviewer_R2.Rmd
Making changes to manuscript addressing R2

---
title: "Comments for the author R2"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Comments for the Author

Thank you for your rapid revision and for the thoughtful way in which you addressed the comments. 
I am happy with what you have done to respond to the reviewers and I think it will be a valuable contribution to the journal. 
I have only a few minor edits - these are just minor wording/syntax issues and should not take long to fix.
Please see the attached marked-up PDF. 
It might also be nice to acknowledge the two anonymous reviewers in your acknowledgments section.

**Reply**:

### 2021-05-03 paper/paper.Rmd
Making changes to manuscript addressing R2

@@
 In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology.
-\textbf{Recent findings} In this paper, we examine existing packages for downloading, processing and visualisation of spatial data, as well as, those specifically developed for ecological analysis.
+\textbf{Recent findings} In this paper, we examine existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for spatial ecological analysis.
 Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community.
@@
 base_size <- 7.5
+
+digits <- 1
 ```
@@
 Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is essential and ecological processes can vary spatially \cite{With2019}.
-Thus, landscape ecology emphasizes \textcolor{blue}{spatial} patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize these \textcolor{blue}{spatial patterns}.
+Thus, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize these spatial patterns.
 
 Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
-Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
+Open-source software includes all software that is released under licenses that allow for the use, modification and distribution of the software \cite{St.Laurent2008}.
 Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
@@
 Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
-This democratizes scientific research as free-of-cost software removes a gatekeeper for researchers without access to proprietary software.
-For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+This democratizes scientific research.
+For more general overviews of open-source software for landscape ecology studies see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
 
 One successful example of an open-source project is the \textit{R} programming language, and its extensions called packages \cite{RCoreTeam2019}. 
-\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the \textit{R} programming language \cite{Wickham2015}}.
-Firstly released in 1995 \cite{Smith2016}, \textcolor{blue}{this} programming language is among the most popular programming languages today, especially in ecology \cite{Lai2019}.
+Packages allow to easily share code and make functionality and documentation available to users and by that extend the capability of the \textit{R} programming language \cite{Wickham2015}.
+First released in 1995 \cite{Smith2016}, this programming language is among the most popular programming languages today, especially in ecology \cite{Lai2019}.
 Originally introduced as a statistical programming language, a large number of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
@@
 Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review, we focus on software implemented in the \textit{R} programming language.
-\textcolor{blue}{We acknowledge that there are reasons and circumstances also to use other software tools and programming languages, however, a comprehensive discussion of pros and cons is outside the scope of this manuscript.}
+We acknowledge that there are reasons and circumstances to use other software tools and programming languages, however, a comprehensive discussion of pros and cons is outside the scope of this manuscript.
 Thus, in the first part of this article, we present an overview of existing \textit{R} packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
-\textcolor{blue}{We included only packages that were available on the official \textit{R} package archive network CRAN at the time the manuscript was prepared.  
-This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
+We included only packages that were available on the official \textit{R} package archive network CRAN at the time the manuscript was prepared.  
+This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.
 In the second part, we present a survey in which we asked the community how they currently use the \textit{R} programming language and to identify topics for which \textit{R} packages are presently missing for landscape ecology.
@@
 There are three major online platforms to host \textit{R} packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
-The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two.}
+The last one focuses on tools for the analysis of genomic data; therefore, in this review we focus on only the former two.
 
@@
 Packages that deal with such issue include \textit{groundhog} \cite{Simonsohn2021}, \textit{packrat} \cite{Ushey2018}, or \textit{renv} \cite{Ushey2020}.
-\textcolor{blue}{These packages facilitate reproducibility to a high degree by preserving the project environment, including specific package versions used for the analysis.}
+These packages facilitate reproducibility to a high degree by preserving the project environment, including specific package versions used for the analysis.
 For more information about \textit{R} package development in general, see \cite{Wickham2015}.
@@
 
-While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.} 
+While base \textit{R} has several built-in data structures, including vectors, matrices, data frames, and lists, yet it has no internal support for reading, processing, or visualizing spatial data. 
+However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.
 Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, Table \@ref(tab:packages)).
@@
 The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
-In many cases, this data model allows a more realistic representation of landscape features, however, with the higher computational cost \cite{Lovelace2019}.
+In many cases, this data model allows a more realistic representation of landscape features, however, with higher computational cost \cite{Lovelace2019}.
 The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
-As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
+As of 2020, more than 500 \textit{R} packages directly depend on or import \textit{sp}.
 However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
 Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
-\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
-Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
+The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
+Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.
 \textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.
@@
 
-\textcolor{blue}{Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
+Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
 However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and switched to use a new representation called WKT2-2019 for storing the CRS.
@@
 Proj4 string representation is also not recommended to set or transform CRS and instead the previously introduced EPSG codes should be used.
-A more detailed explanation of the recent CRS changes can be found in \cite{Bivand2020}.}
+A more detailed explanation of the recent CRS changes can be found in \cite{Bivand2020}.
 
@@
 It is possible to control several GIS software directly from \textit{R} using dedicated packages, such as \textit{rgrass7} \cite{Bivand2021} for GRASS GIS, \textit{RSAGA} \cite{Brenning2018} for SAGA GIS, and \textit{rgee} \cite{Aybar2020} for Google Earth Engine.
-\textcolor{blue}{Direct access to these GIS software within \textit{R} allows to create reproducible and sharable \textit{R} scripts, even if functionality is currently not available in \textit{R}.}
+Direct access to these GIS software within \textit{R} allows to create reproducible and sharable \textit{R} scripts, even if functionality is currently not available in \textit{R}.
 
@@
 However, the generic functions are focused on quick visual inspection of the data (Fig.
-\@ref(fig:fig-map)a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor small multiples maps.}
+\@ref(fig:fig-map)a), rather than the creating complete maps as they do not directly support additional map elements (e.g., scale bar, north arrow) nor small multiples maps.
 
 The popular plotting package \textit{ggplot2} (Fig.
-\@ref(fig:fig-map)b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
+\@ref(fig:fig-map)b) can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
 Additionally, there is an extension especially designed to plot spatial data named \textit{ggspatial}, which provides additional map elements \cite{Dunnington2020}.
-For more information about \textit{ggplot2} in general, see \cite{Wickham2016}}.
+For more information about \textit{ggplot2} in general, see \cite{Wickham2016}.
 The \textit{tmap} package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
-\textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
+This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.
 Quick interactive visualization of spatial data can be done with the \textit{mapview} package\cite{Appelhans2020}.
@@
 
-The introduction of the \textit{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \textit{landscapemetrics} package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the \textit{R} environment.
+The introduction of \textit{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \textit{landscapemetrics} package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics within the \textit{R} environment.
 
@@
 
-Distance-based methods allow to detect and correct for spatial autocorrelation in data.
+Distance-based methods allow for the detection and correction of spatial autocorrelation in data.
 It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling. 
@@
 Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
-Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
-Additionally, textbooks introducing basic concepts of SDM in \textit{R} exists \cite{Guisan2017,Fletcher2018}.
+Originated in the 1970s, SDM has experienced numerous methodological advancement, and a large body of literature exists today \cite{Zimmermann2010,Norberg2019}.
+Additionally, textbooks introducing basic concepts of SDM in \textit{R} exist \cite{Guisan2017,Fletcher2018}.
 
@@
 Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
-Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across \textcolor{blue}{a couple} \textit{R} packages.
+Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across several \textit{R} packages.
 A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \textit{NLMR} package \cite{Sciaini2018}.
@@
 ```{r participation, include = FALSE}
-digits <- 1
-
 # number of people that took part
@@
 To better understand how the landscape ecology community uses \textit{R}, we conducted a short survey using mailing lists and social media to reach the community. 
-\textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
-Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey}.
+Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
+Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey.
 All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).
@@
 In total, the survey was answered by `r n_people` participants, of which the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
-Other, less frequent answers were \textcolor{blue}{"Data scientists", "None of the above", "Government employees", "Master's degree student" and "Bachelor's degree student" (in decreasing order).}
+Other, less frequent answers were "Data scientists", "None of the above", "Government employees", "Master's degree student" and "Bachelor's degree student" (in decreasing order).
 
@@
   dplyr::summarise(n_rel = sum(n_rel)) %>%
-  dplyr::mutate(label = paste0(round(n_rel, digits = 1), "%"),
+  dplyr::mutate(label = paste0(round(n_rel, digits = digits), "%"),
                 col = dplyr::case_when(tasks_reclass == "others" ~ 0,
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options} (Fig. \@ref(fig:fig-survey)b).
+Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey)b).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").
@@
 Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey)c.
-Nevertheless, the large availability and usage of different \textit{R} packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], 2)`%).
+Nevertheless, the large availability and usage of different \textit{R} packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], digits)`%).
 
@@
   dplyr::arrange(-usefulness) %>%
-  dplyr::mutate(label = paste0(round(n_rel, digits = 1), "%"),
+  dplyr::mutate(label = paste0(round(n_rel, digits = digits), "%"),
                 usefulness_class = classes,
@@
   sum() %>%
-  round(digits = 2)
+  round(digits = digits)
 
@@
 There were three most common topics across the answers of the participants.
-Firstly, many participants (`r perc_performance`%) wished for a better computational performance of \textit{R} in terms of speed and required RAM, especially for larger data.
-Secondly, participants are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (`r perc_characterisation`%).
-Thirdly, many participants (`r perc_viz`%) are currently missing advanced and easy-to-apply methods to create high-quality maps or other visualization-related functionality.
+Firstly, many participants (`r round(perc_performance, digits)`%) wished for a better computational performance of \textit{R} in terms of speed and required RAM, especially for larger data sets.
+Secondly, participants are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (`r round(perc_characterisation, digits)`%).
+Thirdly, many participants (`r round(perc_viz, digits)`%) are currently missing advanced and easy-to-apply methods to create high-quality maps or other visualization-related functionality.
 
@@
 The highly dynamic development of \textit{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since landscape ecology constantly develops and improves, \textcolor{blue}{consequently also the \textit{R} programming language and its packages are under constant change to adapt to these new developments.}
+However, since landscape ecology constantly develops and improves, consequently also the \textit{R} programming language and its packages are under constant change to adapt to these new developments.
 
 A comprehensive collection of \textit{R} software packages exists to handle the most common tasks of landscape ecology.
-Because it is possible to import, modify, analyze, and visualize spatial data all in the same  \textcolor{blue}{scripted} programming environment, \textit{R} allows for transparent and reproducible workflows.
-\textcolor{blue}{Scripts including all analysis steps and parameter settings can easily be shared, thus the script-based characteristic of the \textit{R} programming language allows to easily distribution and reproduction of analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
-This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore, are often only available for certain operating systems.}
+Because it is possible to import, modify, analyze, and visualize spatial data all in the same scripted programming environment, \textit{R} allows for transparent and reproducible workflows.
+Scripts including all analysis steps and parameter settings can easily be shared, thus the script-based characteristic of the \textit{R} programming language allows to easily distribution and reproduction of analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
+This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore, are often only available for certain operating systems.
 This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
@@
 While the \textit{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \textit{R} programming language has to face.
-\textcolor{blue}{Some packages discussed herein, such as \textit{raster} and \textit{terra}, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
-Furthermore, also parallel processing, including the use of high performance clusters, is provided by several \textit{R} packages. 
-While this is outside the scope of this manuscript, further resources about parallel processing in \textit{R} can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.} 
-Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \textit{R} packages related to those fields.
+Some packages discussed herein, such as \textit{raster} and \textit{terra}, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
+Furthermore, parallel processing, including the use of high performance clusters, is also provided by several \textit{R} packages. 
+While this is outside the scope of this manuscript, further resources about parallel processing in \textit{R} can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.
+Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so will the need for \textit{R} packages related to those fields.
 Nevertheless, as the development of \textit{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
@@
 We are thankful to all members of the 'Coastal Ecology and Conservation Lab' at the University of Michigan for comments on an early draft of the manuscript. 
+We would like to thank two anonymous reviewers for their valuable comments on earlier drafts of the manuscript.
 
@@
 
-\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of \textit{R} to analyze spatial data demonstrating the capabilities of the programming language to analyze spatial data.}
+\noindent \textbullet \textbullet \cite{Bivand2020}: Recent and comprehensive overview of general developments of \textit{R} to analyze spatial data demonstrating the capabilities of the programming language to analyze spatial data.
 
-\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Comprehensive textbook about spatial data handling and analyzing including many useful applied code examples.}
+\noindent \textbullet \textbullet \cite{Lovelace2019}: Comprehensive textbook about spatial data handling and analyzing including many useful applied code examples.
 
-\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state of the art for quantifying landscape spatial pattern including lessons learned by the author.}
+\noindent \textbullet \cite{Gustafson2019}: This review presents an overview of the developments and the state of the art for quantifying landscape spatial pattern including lessons learned by the author.
 
-\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the \textit{R} programming language is widely used in ecology including possible reasons for the wide spread use.}
+\noindent \textbullet \cite{Lai2019}: This review demonstrates that the \textit{R} programming language is widely used in ecology including possible reasons for the wide spread use.

### 2021-05-03 paper/paper.tex
Making changes to manuscript addressing R2

@@
 Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
-This democratizes scientific research as free-of-cost software removes a gatekeeper for researchers without access to proprietary software.
-For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
+This democratizes scientific research.
+For more general overviews of open-source software for landscape ecology studies see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
 
 One successful example of an open-source project is the \textit{R} programming language, and its extensions called packages \cite{RCoreTeam2019}.
-\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the \textit{R} programming language \cite{Wickham2015}}.
-Firstly released in 1995 \cite{Smith2016}, \textcolor{blue}{this} programming language is among the most popular programming languages today, especially in ecology \cite{Lai2019}.
+Packages allow to easily share code and make functionality and documentation available to users and by that extend the capability of the \textit{R} programming language \cite{Wickham2015}.
+First released in 1995 \cite{Smith2016}, this programming language is among the most popular programming languages today, especially in ecology \cite{Lai2019}.
 Originally introduced as a statistical programming language, a large number of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
@@
 Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review, we focus on software implemented in the \textit{R} programming language.
-\textcolor{blue}{We acknowledge that there are reasons and circumstances also to use other software tools and programming languages, however, a comprehensive discussion of pros and cons is outside the scope of this manuscript.}
+We acknowledge that there are reasons and circumstances to use other software tools and programming languages, however, a comprehensive discussion of pros and cons is outside the scope of this manuscript.
 Thus, in the first part of this article, we present an overview of existing \textit{R} packages for landscape ecology and closely related fields (Table \ref{tab:packages}).
-\textcolor{blue}{We included only packages that were available on the official \textit{R} package archive network CRAN at the time the manuscript was prepared.  
-This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
+We included only packages that were available on the official \textit{R} package archive network CRAN at the time the manuscript was prepared.\\
+This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.
 In the second part, we present a survey in which we asked the community how they currently use the \textit{R} programming language and to identify topics for which \textit{R} packages are presently missing for landscape ecology.
@@
 
-While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
-\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.}
+While base \textit{R} has several built-in data structures, including vectors, matrices, data frames, and lists, yet it has no internal support for reading, processing, or visualizing spatial data.
+However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.
 Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, Table \ref{tab:packages}).
@@
 The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
-In many cases, this data model allows a more realistic representation of landscape features, however, with the higher computational cost \cite{Lovelace2019}.
+In many cases, this data model allows a more realistic representation of landscape features, however, with higher computational cost \cite{Lovelace2019}.
 The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
-As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
+As of 2020, more than 500 \textit{R} packages directly depend on or import \textit{sp}.
 However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
 Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
-\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
-Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
+The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project.
+Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.
 \textit{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.
@@
 However, the generic functions are focused on quick visual inspection of the data (Fig.
-\ref{fig:fig-map}a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor small multiples maps.}
+\ref{fig:fig-map}a), rather than the creating complete maps as they do not directly support additional map elements (e.g., scale bar, north arrow) nor small multiples maps.
 
 The popular plotting package \textit{ggplot2} (Fig.
-\ref{fig:fig-map}b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
+\ref{fig:fig-map}b) can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
 Additionally, there is an extension especially designed to plot spatial data named \textit{ggspatial}, which provides additional map elements \cite{Dunnington2020}.
-For more information about \textit{ggplot2} in general, see \cite{Wickham2016}}.
+For more information about \textit{ggplot2} in general, see \cite{Wickham2016}.
 The \textit{tmap} package provides a provides a coherent plotting system intended for drawing maps (Fig. \ref{fig:fig-map}c) \cite{Tennekes2018}.
-\textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
+This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.
 Quick interactive visualization of spatial data can be done with the \textit{mapview} package\cite{Appelhans2020}.
@@
 
-Distance-based methods allow to detect and correct for spatial autocorrelation in data.
+Distance-based methods allow for the detection and correction of spatial autocorrelation in data.
 It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling.
@@
 To better understand how the landscape ecology community uses \textit{R}, we conducted a short survey using mailing lists and social media to reach the community.
-\textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
-Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey}.
+Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
+Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey.
 All raw data and scripts to analyze the data can be found at \url{https://github.com/r-spatialecology/Hesselbarth_et_al_CLER}.
@@
 In total, the survey was answered by 109 participants, of which the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (14.7\%).
-Other, less frequent answers were \textcolor{blue}{"Data scientists", "None of the above", "Government employees", "Master's degree student" and "Bachelor's degree student" (in decreasing order).}
+Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees,'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).
 
@@
 Next, we were interested in the most important tasks to the workflow of the participants.
-\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options} (Fig. \ref{fig:fig-survey}b).
+Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey}b).
 The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').
@@
 Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey}c.
-Nevertheless, the large availability and usage of different \textit{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 35.87\%).
+Nevertheless, the large availability and usage of different \textit{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 35.9\%).
 
-Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \textit{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig.
+Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.8\%) and only very few participants evaluated \textit{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.2\%; Fig.
 \ref{fig:fig-survey}d).
@@
 There were three most common topics across the answers of the participants.
-Firstly, many participants (12.8\%) wished for a better computational performance of \textit{R} in terms of speed and required RAM, especially for larger data.
+Firstly, many participants (12.8\%) wished for a better computational performance of \textit{R} in terms of speed and required RAM, especially for larger data sets.
 Secondly, participants are missing specific approaches to quantify landscape characteristics (such as surface metrics), or are wishing for an improvement of currently available approaches to quantify landscape characteristics (9.2\%).
@@
 The highly dynamic development of \textit{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
-However, since landscape ecology constantly develops and improves, \textcolor{blue}{consequently also the \textit{R} programming language and its packages are under constant change to adapt to these new developments.}
+However, since landscape ecology constantly develops and improves, consequently also the \textit{R} programming language and its packages are under constant change to adapt to these new developments.
 
 A comprehensive collection of \textit{R} software packages exists to handle the most common tasks of landscape ecology.
-Because it is possible to import, modify, analyze, and visualize spatial data all in the same \textcolor{blue}{scripted} programming environment, \textit{R} allows for transparent and reproducible workflows.
-\textcolor{blue}{Scripts including all analysis steps and parameter settings can easily be shared, thus the script-based characteristic of the \textit{R} programming language allows to easily distribution and reproduction of analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
-This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore, are often only available for certain operating systems.}
+Because it is possible to import, modify, analyze, and visualize spatial data all in the same scripted programming environment, \textit{R} allows for transparent and reproducible workflows.
+Scripts including all analysis steps and parameter settings can easily be shared, thus the script-based characteristic of the \textit{R} programming language allows to easily distribution and reproduction of analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the \textit{R} script in the appendix.
+This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore, are often only available for certain operating systems.
 This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.
@@
 While the \textit{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \textit{R} programming language has to face.
-\textcolor{blue}{Some packages discussed herein, such as \textit{raster} and \textit{terra}, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
-Furthermore, also parallel processing, including the use of high performance clusters, is provided by several \textit{R} packages. 
-While this is outside the scope of this manuscript, further resources about parallel processing in \textit{R} can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.}
-Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \textit{R} packages related to those fields.
+Some packages discussed herein, such as \textit{raster} and \textit{terra}, already provide solutions by processing data in chunks small enough to be stored in RAM memory.
+Furthermore, parallel processing, including the use of high performance clusters, is also provided by several \textit{R} packages.
+While this is outside the scope of this manuscript, further resources about parallel processing in \textit{R} can be found here \cite{Bengtsson2020,Schubert2019,McCallum2012}.
+Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so will the need for \textit{R} packages related to those fields.
 Nevertheless, as the development of \textit{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
@@
 We are thankful to all members of the `Coastal Ecology and Conservation Lab' at the University of Michigan for comments on an early draft of the manuscript.
+We would like to thank two anonymous reviewers for their valuable comments on earlier drafts of the manuscript.
 
@@
 
-\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of \textit{R} to analyze spatial data demonstrating the capabilities of the programming language to analyze spatial data.}
+\noindent \textbullet \textbullet \cite{Bivand2020}: Recent and comprehensive overview of general developments of \textit{R} to analyze spatial data demonstrating the capabilities of the programming language to analyze spatial data.
 
-\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Comprehensive textbook about spatial data handling and analyzing including many useful applied code examples.}
+\noindent \textbullet \textbullet \cite{Lovelace2019}: Comprehensive textbook about spatial data handling and analyzing including many useful applied code examples.
 
-\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state of the art for quantifying landscape spatial pattern including lessons learned by the author.}
+\noindent \textbullet \cite{Gustafson2019}: This review presents an overview of the developments and the state of the art for quantifying landscape spatial pattern including lessons learned by the author.
 
-\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the \textit{R} programming language is widely used in ecology including possible reasons for the wide spread use.}
+\noindent \textbullet \cite{Lai2019}: This review demonstrates that the \textit{R} programming language is widely used in ecology including possible reasons for the wide spread use.
 

### 2021-05-03 Comments_Reviewer_R2.Rmd
Reply comments R2

@@
 I am happy with what you have done to respond to the reviewers and I think it will be a valuable contribution to the journal.
+
+**Reply**: Thank you very much for the positive feedback. 
+We are happy to hear that our changes to the manuscript improved it and addressed the appreciated comments by the reviewers. \vskip 0.5cm
+
 I have only a few minor edits - these are just minor wording/syntax issues and should not take long to fix.
 Please see the attached marked-up PDF.
+
+**Reply**: Thank you for the careful read of the manuscript and spotting these wording/syntax issues. 
+We used the attached PDF to improve the manuscript and implemented all comments and rewording suggestions.
+All changes are indicated by the blue color in the manuscript version that includes track changes.
+\vskip 0.5cm
+
 It might also be nice to acknowledge the two anonymous reviewers in your acknowledgments section.
 
-**Reply**:
+**Reply**: Thank you for this reminder. We agree that this is a nice acknowledgments and we added a corresponding sentence. \vskip 0.5cm

### 2021-05-06 01_Manuscript/Manuscript.tex
:truck: :bathtub: Clean and rename files

--- 2021-05-03 paper/paper.tex
@@
 
-{\centering \includegraphics[width=\textwidth]{paper_files/figure-latex/fig-survey-1}
+{\centering \includegraphics[width=\textwidth]{Manuscript_files/figure-latex/fig-survey-1}
 
@@
 Results include a) which terms describe major research topics the best, b) the most important workflow task, c) the most used \textit{R} packages and d) the overall usefulness or \textit{R} for landscape ecology.
-The `others' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
+The ``others'' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
 \end{figure}

### 2021-05-06 03_R_Scripts/code_example.Rmd
:truck: :bathtub: Clean and rename files

--- 2021-04-28 R/code_example.Rmd
@@
 precipitation <- raster::getData(name = "worldclim", lon = 3, lat = 38,
-                                 res = 0.5, var = "prec", path = "R/")
+                                 res = 0.5, var = "prec", path = "03_R_Scripts/")
 ```
@@
 ```{r base_plot, eval = FALSE}
-# pdf(file = "R/Figures/base_plot.pdf")
+# pdf(file = "03_R_Scripts/Figures/base_plot.pdf")
 
@@
 
-# ggplot2::ggsave(filename = "R/Figures/ggplot2.pdf")
+# ggplot2::ggsave(filename = "03_R_Scripts/Figures/ggplot2.pdf")
 ```
@@
 
-# tmap::tmap_save(plot_tm, "R/Figures/tmap.pdf")
+# tmap::tmap_save(plot_tm, "03_R_Scripts/Figures/tmap.pdf")
 ```

### 2021-05-06 03_R_Scripts/get_survery_data.R

@@ -14,4 +14,4 @@ names(answers) <-  new_names
 
 results <- list(answers = answers, old_names = old_names, new_names = new_names)
 
-saveRDS(object = results, file = "paper/data/results.rds")
+saveRDS(object = results, file = "01_Manuscript/data/results.rds")

### 2021-05-06 R/code_example.R

@@ -1,124 +1,0 @@
-#---------------------------------------------------#
-#                                                   #
-#   Code to reproduce Fig. 2 in                     #
-#   "Open-source tools in R for landscape ecology"  #
-#                                                   #
-#   Code created by MHKH, JN, JS and LJG            #
-#                                                   #
-#   Contact: mhk.hesselbarth@gmail.com              #
-#                                                   #
-#---------------------------------------------------#
-
-# First, we need to load all required packages used in this R script. In case you are
-# missing a package, it can be install from CRAN using e.g., install.packages("raster").
-# By loading the packages, we make sure all functions from them can be used in the script.
-library(ggplot2)
-library(ggspatial)
-library(raster)
-library(rnaturalearth)
-library(sf)
-library(tmap)
-
-# This function downloads data and saves it as a RasterLayer. The 'name' argument
-# allows to specify a source to download data from the worldclim database. The
-# 'lon' and 'lat' argument specify the SRTM tile for which the data is downloaded, and
-# the 'res' the resolution of the data in minutes of a degree. The 'var' argument
-# specifies which kind of data is downloaded, in this case precipitation data,
-# and lastly the path on the local disk where the data is saved can be set.
-precipitation <- raster::getData(name = "worldclim", lon = 3, lat = 38,
-                                 res = 0.5, var = "prec", path = "R/")
-
-# The previous downloaded object is a RasterStack in which each layer corresponds
-# to the precipitation in a month (Jan - Dec). To get the annual precipitation,
-# we can use the calc function which applies basic mathematical function to each
-# layer of the RasterStack. By getting the sum of all layers for each cell, we
-# get the annual precipitation.
-precipitation_sum <- raster::calc(x = precipitation, fun = sum)
-
-# Next, we need to ensure that all spatial data has the same coordinate reference system.
-# To achieve this we reproject the downloaded data to the CRS epsg:21781.
-precipitation_sum <- raster::projectRaster(precipitation_sum, crs = "EPSG:21781")
-
-# Next, we download administration boundaries of Switzerland as vector data from
-# the naturalearth database. Again, we need to specify the resolution of the data
-# (called 'scale' here). Additionally, the function allows to select the return type
-# and we select a sf vector object.
[…]
